---
title: "Task 1: Arctic Storminess Climatology"
author: Francesco Lucano
format: html
engine: knitr
editor: visual
warning: false
error: false
toc: true
---

# Setup

A self-contained and reproducible computational environment was established as the foundation for the analysis.

-   **Project Setup**: The workflow was situated within an **RStudio Project** to ensure all file paths were relative and the project was portable. The R **`reticulate`** package was used to create and manage a dedicated Python virtual environment (`"arctic-env"`), ensuring all Python package dependencies are isolated and reproducible.

```{r installation, eval=FALSE}
# --- ONE-TIME INSTALLATION ---

# -- R Packages --
# Install all required R packages from CRAN
install.packages(c(
    "reticulate", "tidyverse", "terra", "sf", "rnaturalearth", 
    "devtools", "future", "furrr", "glue", "progressr"
))

# Install the high-resolution map data package from GitHub
# devtools::install_github("ropensci/rnaturalearthhires")


# -- Python Environment and Packages --
# Create the virtual environment if it doesn't already exist
if (!reticulate::virtualenv_exists(envname = "arctic-env")) {
    reticulate::virtualenv_create(envname = "arctic-env")
}

# Install all required Python packages into the new environment
reticulate::py_install(
    packages = c(
        "numpy", "xarray", "netcdf4", "h5netcdf", 
        "cdsapi", "scipy", "tqdm", "dask", "matplotlib", "bottleneck"
    ),
    envname = "arctic-env"
)
```

```{r setup-environment}
# Load R Libraries
library(reticulate)
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(glue)
library(future)
library(furrr)
library(progressr)
library(tictoc)
library(ncdf4)
library(glue)
#library(discordrpc)   # Disable this if you run the code on your pc

# Configure Python Environment
use_virtualenv("arctic-env", required = TRUE)

# Define the absolute path to the data storage folder
DATA_DIRECTORY <- "E:/ERA5_data"
PROCESSED_DATA_DIR <- "E:/ERA5_data/processed"

# Setting for more parallel downloads
#plan(multisession, workers = 20)
```

# Data download

```{python single-download}
import cdsapi
import os
import xarray as xr
from scipy import signal
import numpy as np

# --- 1. CONFIGURATION ---
# Access the DATA_DIRECTORY variable from your R configuration chunk
data_dir = r.DATA_DIRECTORY

# Create the directory on your hard disk if it doesn't exist
if not os.path.exists(data_dir):
    os.makedirs(data_dir)
    print(f"Created directory: {data_dir}")

# Let's start with just two years to test the monthly download loop.
# For the final run, you will change this to range(1958, 2025).
YEARS_TO_DOWNLOAD = range(1958, 2024) 

AREA_PAN_ARCTIC = [90, -180, 55, 180] # North, West, South, East

# --- 2. NESTED LOOP AND DOWNLOAD ---
c = cdsapi.Client()
print("Starting batch download process (one file per month)...")

# Outer loop for each year
for year in YEARS_TO_DOWNLOAD:
    # Inner loop for each month (from 1 to 12)
    for month in range(1, 13):
        # Format the month with a leading zero (e.g., 1 -> "01", 10 -> "10")
        month_str = f"{month:02d}"
        
        # Create a specific filename for each YEAR and MONTH
        final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-{month_str}.nc")
        
        # Check if this specific monthly file already exists
        if not os.path.exists(final_path):
            print(f"--- Downloading data for: {year}-{month_str} ---")
            
            request = {
                "product_type": "reanalysis",
                "variable": ["u_component_of_wind", "v_component_of_wind"],
                "year": str(year),
                "month": month_str, # Dynamically set the month
                "day": [
                    "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
                    "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", 
                    "31"
                ],
                "time": [
                    '00:00', '06:00', '12:00', '18:00'
                ],
                "pressure_level": "850",
                "format": "netcdf",
                "area": AREA_PAN_ARCTIC
            }
            
            # Use a try-except block to handle potential API errors for a single month
            try:
                c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
                print(f"Successfully downloaded to {final_path}")
            except Exception as e:
                print(f"!!! FAILED to download data for {year}-{month_str}. Error: {e}")

        else:
            print(f"--- File for {year}-{month_str} already exists. Skipping. ---")

print("\nFull data download process complete for the specified years.")
```

```{python yearly-download}
import cdsapi
import os

# --- 1. CONFIGURATION ---
data_dir = r.DATA_DIRECTORY
if not os.path.exists(data_dir):
    os.makedirs(data_dir)
    print(f"Created directory: {data_dir}")

YEARS_TO_DOWNLOAD = range(1958, 2024) 
AREA_PAN_ARCTIC = [90, -180, 55, 180]

# --- 2. YEARLY LOOP AND DOWNLOAD ---
c = cdsapi.Client()
print("Starting batch download process (one file per year)...")

# Outer loop for each year (The inner month loop has been removed)
for year in YEARS_TO_DOWNLOAD:
    
    # 1. Adjust the filename to be yearly instead of monthly
    final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}.nc")
    
    # Check if this specific yearly file already exists
    if not os.path.exists(final_path):
        print(f"--- Downloading data for year: {year} ---")
        
        request = {
            "product_type": "reanalysis",
            "variable": ["u_component_of_wind", "v_component_of_wind"],
            "year": str(year),
            # 2. Modify the 'month' parameter to request all 12 months
            "month": [
                "01", "02", "03", "04", "05", "06",
                "07", "08", "09", "10", "11", "12"
            ],
            "day": [
                "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
                "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
                "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", 
                "31"
            ],
            "time": ['00:00', '06:00', '12:00', '18:00'],
            "pressure_level": "850",
            "format": "netcdf",
            "area": AREA_PAN_ARCTIC
        }
        
        # Use a try-except block to handle potential API errors for the year
        try:
            c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
            print(f"Successfully downloaded to {final_path}")
        except Exception as e:
            print(f"!!! FAILED to download data for {year}. Error: {e}")

    else:
        print(f"--- File for {year} already exists. Skipping. ---")

print("\nFull data download process complete for the specified years.")
```

```{r}
YEARS_TO_DOWNLOAD <- 1958:2024
AREA_OF_INTEREST <- c(90, -180, 55, 180)
```

```{python parallel-download}
# === Complete Python Parallel Download Script ===

import cdsapi
import os
import concurrent.futures
from tqdm.auto import tqdm

# --- 1. CONFIGURATION ---
# Get the configuration variables from your R 'configuration' chunk
data_dir = r.DATA_DIRECTORY
years_to_download = r.YEARS_TO_DOWNLOAD
area_of_interest = r.AREA_OF_INTEREST

# Set the number of parallel workers. 2 is a safe number for the CDS API.
MAX_WORKERS = 5

# --- 2. DEFINE THE DOWNLOAD FUNCTION FOR A SINGLE JOB ---
def download_era5_month(job_info):
    year, month = job_info # Unpack the year and month for this job
    
    # Ensure year is a clean integer to prevent '.0' in filenames/requests
    year_int = int(year)
    month_str = f"{month:02d}"
    
    final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year_int}-{month_str}.nc")

    # Check if the file already exists
    if os.path.exists(final_path):
        return f"Skipped (already exists): {year_int}-{month_str}"

    # Each worker needs its own API client instance
    c = cdsapi.Client(quiet=True, progress=False)
    
    request = {rrr
        'product_type': 'reanalysis', 'variable': ['u_component_of_wind', 'v_component_of_wind'],
        'year': str(year_int), 
        'month': month_str,
        'day': [
            '01','02','03','04','05','06','07','08','09','10','11','12','13',
            '14','15','16','17','18','19','20','21','22','23','24','25','26',
            '27','28','29','30','31'
        ],
        'time': ['00:00', '06:00', '12:00', '18:00'], # Using 6-hourly data as requested
        'pressure_level': '850', 
        'format': 'netcdf', 
        'area': area_of_interest
    }
    
    try:
        c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
        return f"Success: {year_int}-{month_str}"
    except Exception as e:
        return f"!!! FAILED: {year_int}-{month_str}. Error: {e}"

# --- 3. CREATE THE LIST OF JOBS ---
# A list of tuples, e.g., [(1958, 1), (1958, 2), ...]
jobs = []
for year in years_to_download:
    for month in range(1, 13):
        jobs.append((year, month))

# --- 4. EXECUTE THE DOWNLOADS IN PARALLEL WITH A PROGRESS BAR ---
print(f"Starting parallel download for {len(jobs)} files using {MAX_WORKERS} workers...")

with concurrent.futures.ThreadPoolExecutor(max_workers=MAX_WORKERS) as executor:
    # 'tqdm' wraps the executor to create a live progress bar.
    results = list(tqdm(executor.map(download_era5_month, jobs), total=len(jobs)))

print("\n--- Download Summary ---")
for res in results:
    if "Success" in res or "FAILED" in res:
         print(res)

print("\nAll download tasks have been processed.")
```

# 2000-2001 workflow test

## Resolution standardisation

```{python monthly-file-combination}
import xarray as xr
import os
import glob

# Access the R variable for the data directory
data_directory = r.DATA_DIRECTORY
processed_dir = r.PROCESSED_DATA_DIR
years_to_process = [2000, 2001]

# Define the output path for the combined raw file at the beginning
output_path_raw = os.path.join(processed_dir, "ERA5_2000-2001_raw.nc")

# --- NEW: Check if the file already exists using Python syntax ---
if not os.path.exists(output_path_raw):
    print("Combined raw file not found. Starting combination process...\n")
    
    # Find all relevant monthly files for the specified years
    all_files = []
    for year in years_to_process:
        pattern = os.path.join(data_directory, f"era5_850hpa_panarctic_{year}-*.nc")
        all_files.extend(glob.glob(pattern))

    all_files.sort()

    if not all_files:
        raise FileNotFoundError("No NetCDF files found for the specified years.")

    print(f"Found {len(all_files)} monthly files for the period {min(years_to_process)}-{max(years_to_process)}.")

    # Open all files as a single xarray dataset
    ds_raw_combined = xr.open_mfdataset(
        all_files, 
        combine='by_coords', 
        parallel=True, 
        engine='h5netcdf'
    )

    # Save the combined dataset to a single intermediate file
    print(f"Saving combined high-resolution data to:\n{output_path_raw}")
    ds_raw_combined.to_netcdf(output_path_raw)

    print("\nCombining raw files complete.")

else:
    # If the file exists, print a message and skip the process
    print(f"Combined raw file already exists. Skipping combination.\nPath: {output_path_raw}")
```

```{r standardization}
# --- UPDATED STANDARDIZATION CHUNK WITH POLE COVERAGE ---
#raw_combined_path <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_raw.nc")
#output_path_correct <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_std_res_vars.nc")

raw_combined_path <- file.path(DATA_DIRECTORY, "era5_850hpa_panarctic_1958-01.nc")
output_path_correct <- file.path(PROCESSED_DATA_DIR, "jan_58_std")

if (!file.exists(output_path_correct)) {

  cat("Standardized file not found. Starting full resampling process...\n")
  
  era5_raw <- terra::rast(raw_combined_path)
  cat("--- Properties of RAW High-Resolution Data ---\n")
  print(era5_raw)

  # --- FIX: Create template with explicit nrows to ensure pole coverage ---
  # We need 36 rows (not 35) to reach from ~55°N to past 90°N with 1° resolution
  template_raster <- terra::rast(
    nrows = 35,           # Explicitly set to 36 rows to cover the pole
    ncols = 360,          # 360 columns for 1° longitude resolution
    xmin = -180.125,
    xmax = 179.875,
    ymin = 55.875,
    ymax = 90.875,        # Extend to 90.875 to ensure pole coverage
    crs = "EPSG:4326"
  )
  
  cat("\nTemplate grid properties:\n")
  print(template_raster)
  cat("Template extent:\n")
  print(ext(template_raster))

  cat("\n--- Remapping data to 1.0-degree grid... ---\n")
  tic()
  era5_std_res <- terra::resample(era5_raw, template_raster, method = "bilinear")
  toc()
  
  cat("\n--- Properties of STANDARDIZED Low-Resolution Data ---\n")
  print(era5_std_res)

  u_std_res <- era5_std_res[[grep("^u", names(era5_std_res))]]
  v_std_res <- era5_std_res[[grep("^v", names(era5_std_res))]]

  cat("\n--- Extracting time information... ---\n")
  
  nc_raw <- nc_open(raw_combined_path)
  time_vals <- ncvar_get(nc_raw, "valid_time")
  time_units <- ncatt_get(nc_raw, "valid_time", "units")$value
  nc_close(nc_raw)
  
  n_layers <- nlyr(u_std_res)
  if (length(time_vals) != n_layers) {
    time_vals <- time_vals[1:n_layers]
  }
  if (!is.numeric(time_vals)) {
    time_vals <- as.numeric(time_vals)
  }
  
  coords <- crds(u_std_res, df=TRUE, na.rm=FALSE)
  lon_vals <- sort(unique(coords$x))
  lat_vals <- sort(unique(coords$y))
  
  cat(glue("Final dimensions: {length(lon_vals)} longitudes, {length(lat_vals)} latitudes, {length(time_vals)} times\n"))
  cat(glue("Latitude range: {min(lat_vals)} to {max(lat_vals)} (should cover pole!)\n"))
  cat(glue("Longitude range: {min(lon_vals)} to {max(lon_vals)}\n"))
  
  cat("\n--- Writing standardized data to NetCDF... ---\n")
  
  lon_dim <- ncdim_def("longitude", "degrees_east", lon_vals)
  lat_dim <- ncdim_def("latitude", "degrees_north", lat_vals)
  time_dim <- ncdim_def("time", time_units, time_vals, unlim=FALSE)
  
  u_var <- ncvar_def("u", "m s**-1", list(lon_dim, lat_dim, time_dim), 
                     missval=-9999, longname="U component of wind")
  v_var <- ncvar_def("v", "m s**-1", list(lon_dim, lat_dim, time_dim), 
                     missval=-9999, longname="V component of wind")
  
  nc_out <- nc_create(output_path_correct, list(u_var, v_var))
  
  cat("Extracting and reshaping u component...\n")
  u_array <- as.array(u_std_res)
  cat("Extracting and reshaping v component...\n")
  v_array <- as.array(v_std_res)
  
  u_array <- aperm(u_array, c(2, 1, 3))
  v_array <- aperm(v_array, c(2, 1, 3))
  
  cat("Writing u to NetCDF...\n")
  ncvar_put(nc_out, u_var, u_array)
  cat("Writing v to NetCDF...\n")
  ncvar_put(nc_out, v_var, v_array)
  
  nc_close(nc_out)

  cat(glue::glue("\nStandardization complete. 
                Correctly formatted data saved to: {output_path_correct}\n"))

} else {
  cat(glue::glue("Standardized file already exists. Skipping resampling.\nPath: {output_path_correct}\n"))
}
```

## Reynolds decomposition

```{python}
import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal

print("CuPy is available. Using GPU for acceleration.")

# --- PART 1: GPU-ACCELERATED FILTER FUNCTION ---
def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """Applies a GPU-accelerated band-pass filter."""
    lowcut_freq, highcut_freq = 1 / lowcut_days, 1 / highcut_days
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    def filter_func_gpu(x):
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    return xr.apply_ufunc(
        filter_func_gpu, data_array,
        input_core_dims=[[time_dim]], output_core_dims=[[time_dim]],
        exclude_dims={time_dim}, dask="parallelized", output_dtypes=[data_array.dtype]
    )

# --- PART 2: CORRECTED MAIN WORKFLOW ---
processed_dir = r.PROCESSED_DATA_DIR
input_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")

# Check if file exists and has content
if not os.path.exists(input_path):
    raise FileNotFoundError(f"File not found: {input_path}")

file_size = os.path.getsize(input_path) / (1024**2)  # Size in MB
print(f"\nFile found: {input_path}")
print(f"File size: {file_size:.2f} MB")

# Try opening with netCDF4 engine instead of h5netcdf
print("Loading standardized data...")
try:
    ds = xr.open_dataset(input_path, engine='netcdf4')
    print("Successfully opened with netcdf4 engine")
except Exception as e:
    print(f"Failed with netcdf4: {e}")
    print("Trying with default engine...")
    ds = xr.open_dataset(input_path)

print("\nDataset contents:")
print(ds)
print("\nDimensions:")
print(ds.dims)
print("\nVariables:")
print(list(ds.data_vars))

print("\nStep 1: Calculating the PROXY climatological mean for the test period...")
u_mean_proxy = ds['u'].mean(dim='time', keep_attrs=True)
v_mean_proxy = ds['v'].mean(dim='time', keep_attrs=True)

print("Step 2: Calculating PROXY anomalies...")
u_anomaly = ds['u'] - u_mean_proxy
v_anomaly = ds['v'] - v_mean_proxy

# Define correct parameters
fs_samples_per_day = 4      # For 6-hourly data
lowcut_period_days = 6.0    
highcut_period_days = 2.0

print("Step 3: Applying GPU-accelerated band-pass filter...")
u_eddy = vector_bandpass_filter_gpu(u_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
v_eddy = vector_bandpass_filter_gpu(v_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
print("Filtering complete.")

eddy_ds = xr.Dataset({'u_eddy': u_eddy, 'v_eddy': v_eddy})
output_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
print(f"\nSaving final eddy components to:\n{output_path}")
eddy_ds.to_netcdf(output_path, mode='w')

print("\nMethodologically structured test is complete.")
```

```{python}
import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal

print("CuPy is available. Using GPU for acceleration.")

# --- PART 1: GPU-ACCELERATED FILTER FUNCTION ---
def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """Applies a GPU-accelerated band-pass filter."""
    lowcut_freq, highcut_freq = 1 / lowcut_days, 1 / highcut_days
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    def filter_func_gpu(x):
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    return xr.apply_ufunc(
        filter_func_gpu, data_array,
        input_core_dims=[[time_dim]], output_core_dims=[[time_dim]],
        exclude_dims={time_dim}, dask="parallelized", output_dtypes=[data_array.dtype],
        keep_attrs=True  # IMPORTANT: Keep attributes including coordinates
    )

# --- PART 2: CORRECTED MAIN WORKFLOW ---
processed_dir = r.PROCESSED_DATA_DIR
input_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")

file_size = os.path.getsize(input_path) / (1024**2)
print(f"\nFile found: {input_path}")
print(f"File size: {file_size:.2f} MB")
print("Loading standardized data...")

ds = xr.open_dataset(input_path, engine='netcdf4')
print("Successfully opened with netcdf4 engine")

print("\nDataset contents:")
print(ds)

print("\nStep 1: Calculating the PROXY climatological mean for the test period...")
u_mean_proxy = ds['u'].mean(dim='time', keep_attrs=True)
v_mean_proxy = ds['v'].mean(dim='time', keep_attrs=True)

print("Step 2: Calculating PROXY anomalies...")
u_anomaly = ds['u'] - u_mean_proxy
v_anomaly = ds['v'] - v_mean_proxy

# Define correct parameters
fs_samples_per_day = 4      # For 6-hourly data
lowcut_period_days = 6.0    
highcut_period_days = 2.0

print("Step 3: Applying GPU-accelerated band-pass filter...")
u_eddy = vector_bandpass_filter_gpu(u_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
v_eddy = vector_bandpass_filter_gpu(v_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)

# CRITICAL: Ensure time coordinates are preserved
u_eddy['time'] = ds['time']
v_eddy['time'] = ds['time']

print("Filtering complete.")

# Create dataset with proper coordinates
eddy_ds = xr.Dataset(
    {
        'u_eddy': u_eddy, 
        'v_eddy': v_eddy
    },
    coords={
        'time': ds['time'],
        'latitude': ds['latitude'],
        'longitude': ds['longitude']
    }
)

output_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
print(f"\nSaving final eddy components to:\n{output_path}")
eddy_ds.to_netcdf(output_path, mode='w')

print("\nEddy components dataset:")
print(eddy_ds)
print("\nMethodologically structured test is complete.")
```

## Indices calculation

```{python calculation}
import xarray as xr
import os

print("--- Starting Final EKE and Storminess Index Calculation ---")

# --- PART 1: LOAD REQUIRED DATASETS ---
processed_dir = r.PROCESSED_DATA_DIR

eddy_components_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
ds_eddy = xr.open_dataset(eddy_components_path, engine='netcdf4')

print("Eddy components dataset:")
print(ds_eddy)
print(f"\nTime coordinate type: {type(ds_eddy['time'].values[0])}")

std_res_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")
ds_std = xr.open_dataset(std_res_path, engine='netcdf4')

# --- PART 2: CALCULATE KINETIC ENERGY INDICES ---
print("\nStep 1: Calculating instantaneous Eddy Kinetic Energy (EKE)...")
eke_inst = 0.5 * (ds_eddy['u_eddy']**2 + ds_eddy['v_eddy']**2)
eke_inst.name = 'eke'

print("Step 2: Calculating Mean Kinetic Energy (MKE)...")
u_mean = ds_std['u'].mean(dim='time')
v_mean = ds_std['v'].mean(dim='time')
mke = 0.5 * (u_mean**2 + v_mean**2)
mke.name = 'mke'

print("Step 3: Calculating the monthly storminess index (variance of EKE)...")
# Use resample instead of groupby for proper time-series handling
storminess_index_timeseries = eke_inst.resample(time='ME').var(dim='time')
storminess_index_timeseries.name = 'monthly_eke_variance'

print(f"\nStorminess index shape: {storminess_index_timeseries.shape}")
print(f"Time dimension: {storminess_index_timeseries.time.values}")

# --- PART 3: SAVE FINAL DATA PRODUCTS ---
final_ds = xr.Dataset({
    'storminess_index': storminess_index_timeseries,
    'mke': mke
})

output_path = os.path.join(processed_dir, "ERA5_2000-2001_storminess_metrics.nc")
print(f"\nSaving final storminess metrics to:\n{output_path}")
final_ds.to_netcdf(output_path, mode='w')

print("\n--- Final Dataset Contents ---")
print(final_ds)
print("\nTask complete. Ready for statistical analysis and visualization.")
```

## Map visualisation

```{r}
metrics_path <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_storminess_metrics.nc")

nc <- nc_open(metrics_path)
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
storminess <- ncvar_get(nc, "storminess_index")
nc_close(nc)

# Use original latitudes - the slight overshoot past 90° is fine for visualization
cat("Latitude range:", range(lat), "\n")

# Reorder and average
storminess_reordered <- aperm(storminess, c(3, 2, 1))
mean_storminess <- apply(storminess_reordered, c(1, 2), mean, na.rm = TRUE)

# Add wraparound for longitude
lon_extended <- c(lon, 180)
mean_storminess_extended <- cbind(mean_storminess, mean_storminess[, 1])

# Create raster with original extent (including the 90.375°N)
storminess_rast <- rast(
  mean_storminess_extended,
  extent = c(min(lon_extended), max(lon_extended), min(lat), max(lat)),
  crs = "EPSG:4326"
)
names(storminess_rast) <- "mean_eke_variance"

cat("\nOriginal raster:\n")
print(storminess_rast)

# Project (warnings are expected and harmless)
arctic_crs <- "+proj=laea +lat_0=90 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m"

cat("\nProjecting (warnings about invalid latitude are expected and can be ignored)...\n")
suppressWarnings({
  storminess_rast_proj <- project(storminess_rast, arctic_crs, method = "bilinear")
})

cat("Projected raster:\n")
print(storminess_rast_proj)

storminess_df_proj <- as.data.frame(storminess_rast_proj, xy = TRUE, na.rm = TRUE)

world_coastlines <- ne_coastline(scale = "medium", returnclass = "sf")
world_coastlines_proj <- st_transform(world_coastlines, crs = arctic_crs)

# Create plot
p_arctic <- ggplot() +
  geom_raster(
    data = storminess_df_proj, 
    aes(x = x, y = y, fill = mean_eke_variance),
    interpolate = FALSE
  ) +
  geom_sf(data = world_coastlines_proj, color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = expression("Mean EKE Variance\n(m"^4*"/s"^4*")"),
    option = "plasma",
    na.value = "transparent"
  ) +
  coord_sf(
    xlim = c(-4500000, 4500000),
    ylim = c(-4500000, 4500000),
    expand = FALSE,
    crs = arctic_crs,
    datum = arctic_crs
  ) +
  labs(
    title = "Mean Pan-Arctic Storminess (2000-2001 Test)",
    subtitle = "Average Monthly Variance of 850 hPa Eddy Kinetic Energy",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "gray90", color = NA),
    panel.grid = element_line(color = "gray70", linetype = "dashed", linewidth = 0.3),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_arctic)

ggsave(
  filename = file.path(PROCESSED_DATA_DIR, "storminess_arctic_final.png"),
  plot = p_arctic,
  width = 10,
  height = 10,
  dpi = 300
)

cat("\n Arctic plot saved! (Small white dot at pole is normal for polar projections)\n")
```

# Full range study

```{r}
# ============================================================================
# Step 1-3: Data Preparation (R)
# ============================================================================

# Processing parameters
FULL_YEARS <- 1958:2024

# Create directories
dir.create(file.path(PROCESSED_DATA_DIR, "temp_monthly_std"), showWarnings = FALSE, recursive = TRUE)

# Create log file
log_file <- file.path(PROCESSED_DATA_DIR, paste0("processing_log_continuous_", format(Sys.Date(), "%Y%m%d"), ".txt"))

log_message <- function(msg) {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  message_text <- paste0("[", timestamp, "] ", msg)
  cat(message_text, "\n")
  cat(message_text, "\n", file = log_file, append = TRUE)
}

log_message("=== STARTING CONTINUOUS PROCESSING (1958-2024) ===")
log_message("Processing entire period as single continuous time series")

# ============================================================================
# STEP 1: STANDARDIZE INDIVIDUAL MONTHLY FILES
# ============================================================================
log_message("\n=== STEP 1: STANDARDIZING INDIVIDUAL MONTHLY FILES TO 1° GRID ===")

# Create template once
template_raster <- terra::rast(
  nrows = 36, ncols = 360,
  xmin = -180.125, xmax = 179.875,
  ymin = 54.875, ymax = 90.875,
  crs = "EPSG:4326"
)

log_message("Created 1° template grid")

# Function to standardize a single month
standardize_single_month <- function(year, month) {
  month_str <- sprintf("%02d", month)
  
  input_file <- file.path(DATA_DIRECTORY, glue("era5_850hpa_panarctic_{year}-{month_str}.nc"))
  output_file <- file.path(PROCESSED_DATA_DIR, "temp_monthly_std", glue("std_{year}_{month_str}.nc"))
  
  if (file.exists(output_file)) {
    return(glue("SKIP: {year}-{month_str}"))
  }
  
  if (!file.exists(input_file)) {
    return(glue("MISSING: {year}-{month_str}"))
  }
  
  tryCatch({
    era5_month <- terra::rast(input_file)
    era5_month_std <- terra::resample(era5_month, template_raster, method = "bilinear")
    
    u_std <- era5_month_std[[grep("^u", names(era5_month_std))]]
    v_std <- era5_month_std[[grep("^v", names(era5_month_std))]]
    
    nc_raw <- nc_open(input_file)
    time_vals <- as.numeric(ncvar_get(nc_raw, "valid_time"))
    time_units <- ncatt_get(nc_raw, "valid_time", "units")$value
    nc_close(nc_raw)
    
    coords <- crds(u_std, df=TRUE, na.rm=FALSE)
    lon_vals <- sort(unique(coords$x))
    lat_vals <- sort(unique(coords$y))
    
    lon_dim <- ncdim_def("longitude", "degrees_east", lon_vals)
    lat_dim <- ncdim_def("latitude", "degrees_north", lat_vals)
    time_dim <- ncdim_def("time", time_units, time_vals, unlim=FALSE)
    
    u_var <- ncvar_def("u", "m s**-1", list(lon_dim, lat_dim, time_dim), missval=-9999)
    v_var <- ncvar_def("v", "m s**-1", list(lon_dim, lat_dim, time_dim), missval=-9999)
    
    nc_out <- nc_create(output_file, list(u_var, v_var))
    ncvar_put(nc_out, u_var, aperm(as.array(u_std), c(2, 1, 3)))
    ncvar_put(nc_out, v_var, aperm(as.array(v_std), c(2, 1, 3)))
    nc_close(nc_out)
    
    rm(era5_month, era5_month_std, u_std, v_std)
    gc()
    
    return(glue("SUCCESS: {year}-{month_str}"))
    
  }, error = function(e) {
    return(glue("ERROR: {year}-{month_str} - {e$message}"))
  })
}

# Process all months
total_months <- length(FULL_YEARS) * 12
processed_count <- 0

for (year in FULL_YEARS) {
  for (month in 1:12) {
    result <- standardize_single_month(year, month)
    processed_count <- processed_count + 1
    
    if (processed_count %% 10 == 0) {
      log_message(glue("Progress: {processed_count}/{total_months} months - Latest: {result}"))
    }
  }
}

log_message(glue("Step 1 complete: {total_months} monthly files standardized"))

# ============================================================================
# STEP 2: COMBINE ALL STANDARDIZED MONTHS INTO ONE CONTINUOUS FILE
# ============================================================================
log_message("\n=== STEP 2: COMBINING ALL MONTHS INTO SINGLE CONTINUOUS FILE ===")
log_message("This creates one file covering entire 1958-2024 period")

py_run_string("
import xarray as xr
import os
import glob

def combine_all_continuous(processed_dir, years_list):
    output_path = os.path.join(processed_dir, 'ERA5_1958-2024_std_res_vars_continuous.nc')
    
    if os.path.exists(output_path):
        return 'SKIP: Continuous file already exists'
    
    # Find ALL standardized monthly files
    monthly_files = []
    for year in years_list:
        for month in range(1, 13):
            month_str = f'{month:02d}'
            file_path = os.path.join(processed_dir, 'temp_monthly_std', f'std_{year}_{month_str}.nc')
            if os.path.exists(file_path):
                monthly_files.append(file_path)
    
    monthly_files.sort()
    
    if not monthly_files:
        return 'ERROR: No standardized files found'
    
    print(f'Combining {len(monthly_files)} monthly files into continuous dataset...')
    
    try:
        # Combine into single continuous file
        ds = xr.open_mfdataset(monthly_files, combine='by_coords', parallel=True, engine='netcdf4')
        
        print(f'Total timesteps: {len(ds.time)}')
        print(f'Time range: {ds.time.values[0]} to {ds.time.values[-1]}')
        print(f'Writing continuous file (this may take a while)...')
        
        ds.to_netcdf(output_path)
        ds.close()
        
        return f'SUCCESS: Combined {len(monthly_files)} months into continuous file'
    except Exception as e:
        return f'ERROR: {str(e)}'
")

result <- py$combine_all_continuous(PROCESSED_DATA_DIR, as.list(FULL_YEARS))
log_message(result)

log_message("Step 2 complete: Single continuous file created")

# ============================================================================
# STEP 3: CALCULATE FULL PERIOD CLIMATOLOGY (1958-2024)
# ============================================================================
log_message("\n=== STEP 3: CALCULATING FULL PERIOD CLIMATOLOGY ===")

py_run_string("
import xarray as xr
import os

def calculate_climatology_continuous(processed_dir):
    climatology_path = os.path.join(processed_dir, 'ERA5_climatology_1958-2024.nc')
    
    if os.path.exists(climatology_path):
        return 'SKIP: Climatology already exists'
    
    continuous_file = os.path.join(processed_dir, 'ERA5_1958-2024_std_res_vars_continuous.nc')
    
    if not os.path.exists(continuous_file):
        return 'ERROR: Continuous file not found'
    
    try:
        print('Loading continuous dataset...')
        ds = xr.open_dataset(continuous_file, engine='netcdf4')
        
        print(f'Calculating climatology from {len(ds.time)} timesteps...')
        
        u_clim = ds['u'].mean(dim='time', keep_attrs=True)
        v_clim = ds['v'].mean(dim='time', keep_attrs=True)
        
        clim_ds = xr.Dataset({'u_climatology': u_clim, 'v_climatology': v_clim})
        clim_ds.to_netcdf(climatology_path)
        
        ds.close()
        
        return 'SUCCESS: Climatology calculated from full continuous period'
    except Exception as e:
        return f'ERROR: {str(e)}'
")

result <- py$calculate_climatology_continuous(PROCESSED_DATA_DIR)
log_message(result)

log_message("Step 3 complete: Climatology calculated")
```

```{python}
# ============================================================================
# STEP 4 & 5: REYNOLDS DECOMPOSITION AND STORMINESS CALCULATION (Python)
# ============================================================================
# This script performs bandpass filtering and calculates the storminess index
# Processing is done in GPU-friendly chunks with overlap to maintain continuity
# ============================================================================

import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal
import numpy as np
from datetime import datetime

# Define paths (adjust if needed for your setup)
PROCESSED_DATA_DIR = "E:/ERA5_data/processed"

# Create log function
def log_message(msg):
    """Log messages with timestamp"""
    timestamp = datetime.now().strftime("%Y-%m-%d %H:%M:%S")
    print(f"[{timestamp}] {msg}")

# ============================================================================
# FUNCTION: GPU-ACCELERATED BANDPASS FILTER
# ============================================================================
def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """
    Apply 2-6 day bandpass filter using GPU acceleration.
    
    This implements the Butterworth bandpass filter following Yau & Chang (2020):
    - Filters out slow-moving features (>6 days) and high-frequency noise (<2 days)
    - Retains synoptic-scale eddies (storms) in the 2-6 day period band
    - Uses GPU (CuPy) for massive speedup on large datasets
    
    Parameters:
    -----------
    data_array : xarray.DataArray
        Input data to filter (e.g., u wind anomalies)
    time_dim : str
        Name of time dimension (usually 'time')
    fs : float
        Sampling frequency (4 = four times per day for 6-hourly data)
    lowcut_days : float
        Lower period cutoff in days (6.0 days)
    highcut_days : float
        Upper period cutoff in days (2.0 days)
    order : int
        Butterworth filter order (5 is standard)
    
    Returns:
    --------
    xarray.DataArray : Filtered data (eddy components)
    """
    
    # Convert period (days) to frequency (cycles per timestep)
    lowcut_freq = 1 / lowcut_days   # Low frequency bound (long period)
    highcut_freq = 1 / highcut_days  # High frequency bound (short period)
    
    # Design Butterworth bandpass filter coefficients
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    
    def filter_func_gpu(x):
        """
        Inner function that does the actual GPU filtering.
        
        Process:
        1. Transfer data to GPU (CuPy array)
        2. Detrend to remove linear trends (important for filtering)
        3. Apply forward-backward filter (filtfilt) for zero phase shift
        4. Transfer result back to CPU (NumPy array)
        """
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    
    # Apply filter along time dimension using xarray's apply_ufunc
    return xr.apply_ufunc(
        filter_func_gpu, 
        data_array,
        input_core_dims=[[time_dim]],
        output_core_dims=[[time_dim]],
        exclude_dims={time_dim},
        dask='parallelized',
        output_dtypes=[data_array.dtype],
        keep_attrs=True
    )


# ============================================================================
# STEP 4: REYNOLDS DECOMPOSITION WITH CONTINUOUS FILTERING
# ============================================================================
log_message("=== STEP 4: REYNOLDS DECOMPOSITION ===")
log_message("Applying 2-6 day bandpass filter to isolate synoptic eddies")
log_message("Processing in memory-efficient chunks with overlap")

# Define file paths
continuous_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_1958-2024_std_res_vars_continuous.nc')
climatology_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_climatology_1958-2024.nc')
eddy_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_1958-2024_eddy_components_continuous.nc')

# Check if already processed
if os.path.exists(eddy_path):
    log_message("Checking existing eddy components file...")
    try:
        ds_test = xr.open_dataset(eddy_path, engine='netcdf4')
        if 'u_eddy' in ds_test and 'v_eddy' in ds_test:
            log_message(">> Eddy components already exist and are complete - skipping Step 4")
            ds_test.close()
        else:
            log_message(">> Eddy components file is incomplete - will reprocess")
            ds_test.close()
            os.remove(eddy_path)
    except:
        log_message(">> Eddy components file is corrupted - will reprocess")
        if os.path.exists(eddy_path):
            os.remove(eddy_path)

if not os.path.exists(eddy_path):
    # ------------------------------------------------------------------------
    # 4.1: LOAD DATA
    # ------------------------------------------------------------------------
    log_message("Loading continuous dataset (9.67 GB)...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    log_message(f"  Loaded: {len(ds.time)} timesteps from {ds.time.values[0]} to {ds.time.values[-1]}")
    
    log_message("Loading climatology...")
    ds_clim = xr.open_dataset(climatology_path, engine='netcdf4')
    
    # ------------------------------------------------------------------------
    # 4.2: CALCULATE ANOMALIES
    # ------------------------------------------------------------------------
    log_message("Calculating anomalies from full-period climatology...")
    u_anomaly = ds['u'] - ds_clim['u_climatology']
    v_anomaly = ds['v'] - ds_clim['v_climatology']
    
    # ------------------------------------------------------------------------
    # 4.3: SETUP CHUNKED PROCESSING
    # ------------------------------------------------------------------------
    chunk_size = 10000  # ~2.5 years per chunk
    overlap = 500       # ~125 days overlap
    
    total_timesteps = len(ds.time)
    n_chunks = int(np.ceil(total_timesteps / chunk_size))
    
    log_message(f"Processing strategy:")
    log_message(f"  - {n_chunks} temporal chunks of {chunk_size} timesteps each")
    log_message(f"  - {overlap} timestep overlap between chunks")
    log_message(f"  - Overlap trimmed after filtering for seamless results")
    
    u_eddy_list = []
    v_eddy_list = []
    
    # ------------------------------------------------------------------------
    # 4.4: PROCESS EACH CHUNK
    # ------------------------------------------------------------------------
    for i in range(n_chunks):
        start_idx = max(0, i * chunk_size - overlap)
        end_idx = min(total_timesteps, (i + 1) * chunk_size + overlap)
        
        log_message(f"\n--- Chunk {i+1}/{n_chunks} ---")
        log_message(f"  Processing timesteps {start_idx} to {end_idx} ({end_idx - start_idx} total)")
        
        # Extract temporal chunk
        u_chunk = u_anomaly.isel(time=slice(start_idx, end_idx))
        v_chunk = v_anomaly.isel(time=slice(start_idx, end_idx))
        
        # Apply bandpass filter on GPU
        log_message(f"  Applying 2-6 day bandpass filter on GPU...")
        u_eddy_chunk = vector_bandpass_filter_gpu(u_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        v_eddy_chunk = vector_bandpass_filter_gpu(v_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        
        # Trim overlap
        if i == 0:
            trim_end = -overlap if i < n_chunks - 1 else None
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(None, trim_end))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(None, trim_end))
        elif i == n_chunks - 1:
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, None))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, None))
        else:
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, -overlap))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, -overlap))
        
        u_eddy_list.append(u_eddy_trimmed)
        v_eddy_list.append(v_eddy_trimmed)
        
        log_message(f"  >> Chunk filtered: {len(u_eddy_trimmed.time)} timesteps retained after trimming")
    
    # ------------------------------------------------------------------------
    # 4.5: COMBINE ALL CHUNKS
    # ------------------------------------------------------------------------
    log_message("\nCombining all filtered chunks into continuous time series...")
    u_eddy = xr.concat(u_eddy_list, dim='time')
    v_eddy = xr.concat(v_eddy_list, dim='time')
    
    log_message(f"Final check: {len(u_eddy.time)} timesteps (expected: {total_timesteps})")
    if len(u_eddy.time) != total_timesteps:
        log_message(f"WARNING: Timestep mismatch!")
    else:
        log_message(">> All timesteps accounted for")
    
    # Restore original time coordinates
    u_eddy['time'] = ds['time']
    v_eddy['time'] = ds['time']
    
    # ------------------------------------------------------------------------
    # 4.6: SAVE EDDY COMPONENTS (WITHOUT CHUNKING)
    # ------------------------------------------------------------------------
    log_message("Saving filtered eddy components to file...")
    
    eddy_ds = xr.Dataset(
        {'u_eddy': u_eddy, 'v_eddy': v_eddy},
        coords={'time': ds['time'], 'latitude': ds['latitude'], 'longitude': ds['longitude']}
    )
    
    # CRITICAL FIX: Save without chunking specification to avoid errors
    # Just use compression
    encoding = {
        'u_eddy': {'zlib': True, 'complevel': 4},
        'v_eddy': {'zlib': True, 'complevel': 4}
    }
    
    # Force computation before saving (loads data into memory)
    log_message("  Computing data arrays (loading into RAM)...")
    eddy_ds = eddy_ds.compute()
    
    log_message("  Writing to disk...")
    eddy_ds.to_netcdf(eddy_path, encoding=encoding, format='NETCDF4')
    
    # Get file size
    file_size_gb = os.path.getsize(eddy_path) / (1024**3)
    log_message(f"  >> File saved: {file_size_gb:.2f} GB")
    
    # Clean up
    ds.close()
    ds_clim.close()
    eddy_ds.close()
    del u_anomaly, v_anomaly, u_eddy, v_eddy, u_eddy_list, v_eddy_list, eddy_ds
    
    log_message(">> Step 4 complete: Eddy components saved successfully")


# ============================================================================
# STEP 5 COMPLETE KINETIC ENERGY BUDGET
# ============================================================================
log_message("\n=== STEP 5: CALCULATING COMPLETE KINETIC ENERGY BUDGET ===")

# Define output paths
tke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_TKE_monthly.nc')
mean_eke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_mean_EKE_monthly.nc')
storm_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_1958-2024_storminess_timeseries.nc')
mke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_MKE.nc')

# ------------------------------------------------------------------------
# 5.1: CALCULATE MKE 
# ------------------------------------------------------------------------
log_message("Calculating Mean Kinetic Energy (MKE)...")
if os.path.exists(mke_path):
    log_message(">> MKE already exists - skipping")
else:
    ds_clim = xr.open_dataset(climatology_path, engine='netcdf4')
    mke = 0.5 * (ds_clim['u_climatology']**2 + ds_clim['v_climatology']**2)
    mke.name = 'mke'
    mke.to_netcdf(mke_path)
    ds_clim.close()
    log_message(">> MKE calculated and saved")

# ------------------------------------------------------------------------
# 5.2: CALCULATE TKE (from original continuous file)
# ------------------------------------------------------------------------
log_message("\nCalculating Total Kinetic Energy (TKE)...")
if os.path.exists(tke_path):
    log_message(">> TKE already exists - skipping")
else:
    log_message("  Loading continuous wind data...")
    ds_continuous = xr.open_dataset(continuous_path, engine='netcdf4')
    
    log_message("  Formula: TKE = 0.5 x (u^2 + v^2)")
    log_message("  Calculating instantaneous TKE...")
    tke_instant = 0.5 * (ds_continuous['u']**2 + ds_continuous['v']**2)
    tke_instant.name = 'tke_instantaneous'
    
    log_message("  Calculating monthly mean TKE...")
    tke_monthly = tke_instant.resample(time='ME').mean(dim='time')
    tke_monthly.name = 'monthly_mean_tke'
    
    log_message("  Saving TKE time series...")
    tke_ds = xr.Dataset({'monthly_mean_tke': tke_monthly})
    tke_ds.to_netcdf(tke_path)
    
    n_months = len(tke_monthly.time)
    log_message(f">> TKE calculated: {n_months} months")
    log_message(f"   TKE range: {float(tke_monthly.min()):.2f} to {float(tke_monthly.max()):.2f} m^2/s^2")
    
    ds_continuous.close()
    del tke_instant, tke_monthly, tke_ds

# ------------------------------------------------------------------------
# 5.3: CALCULATE MEAN EKE (from eddy components)
# ------------------------------------------------------------------------
log_message("\nCalculating Mean Eddy Kinetic Energy (Mean EKE)...")
if os.path.exists(mean_eke_path):
    log_message(">> Mean EKE already exists - skipping")
else:
    log_message("  Loading eddy components...")
    ds_eddy = xr.open_dataset(eddy_path, engine='netcdf4')
    
    if 'u_eddy' not in ds_eddy or 'v_eddy' not in ds_eddy:
        log_message("ERROR: Eddy components file is incomplete!")
    else:
        log_message("  Formula: Mean EKE = monthly mean of [0.5 x (u_eddy^2 + v_eddy^2)]")
        log_message("  Calculating instantaneous EKE...")
        eke_instant = 0.5 * (ds_eddy['u_eddy']**2 + ds_eddy['v_eddy']**2)
        eke_instant.name = 'eke_instantaneous'
        
        log_message("  Calculating monthly mean EKE...")
        mean_eke_monthly = eke_instant.resample(time='ME').mean(dim='time')
        mean_eke_monthly.name = 'monthly_mean_eke'
        
        log_message("  Saving Mean EKE time series...")
        mean_eke_ds = xr.Dataset({'monthly_mean_eke': mean_eke_monthly})
        mean_eke_ds.to_netcdf(mean_eke_path)
        
        n_months = len(mean_eke_monthly.time)
        log_message(f">> Mean EKE calculated: {n_months} months")
        log_message(f"   Mean EKE range: {float(mean_eke_monthly.min()):.2f} to {float(mean_eke_monthly.max()):.2f} m^2/s^2")
        
        ds_eddy.close()
        del eke_instant, mean_eke_monthly, mean_eke_ds

# ------------------------------------------------------------------------
# 5.4: CALCULATE EKE VARIANCE (Storminess Index)
# ------------------------------------------------------------------------
log_message("\nCalculating EKE Variance (Storminess Index)...")
if os.path.exists(storm_path):
    log_message(">> Storminess index already exists - skipping")
else:
    # [Your existing code for EKE variance]
    pass

# ------------------------------------------------------------------------
# 5.5: VERIFY ENERGY BUDGET CONSISTENCY
# ------------------------------------------------------------------------
log_message("\n=== VERIFYING KINETIC ENERGY BUDGET ===")
log_message("Theoretical relationship: <TKE> = MKE + <EKE>")

try:
    ds_tke = xr.open_dataset(tke_path)
    ds_mke = xr.open_dataset(mke_path)
    ds_mean_eke = xr.open_dataset(mean_eke_path)
    
    # Calculate spatial mean for one sample month
    sample_month = 100  # Pick a month to verify
    
    tke_sample = float(ds_tke['monthly_mean_tke'].isel(time=sample_month).mean())
    mke_sample = float(ds_mke['mke'].mean())
    eke_sample = float(ds_mean_eke['monthly_mean_eke'].isel(time=sample_month).mean())
    
    log_message(f"\nSample month {sample_month} (spatial averages):")
    log_message(f"  TKE = {tke_sample:.4f} m^2/s^2")
    log_message(f"  MKE = {mke_sample:.4f} m^2/s^2")
    log_message(f"  Mean EKE = {eke_sample:.4f} m^2/s^2")
    log_message(f"  MKE + Mean EKE = {mke_sample + eke_sample:.4f} m^2/s^2")
    log_message(f"  Difference: {abs(tke_sample - (mke_sample + eke_sample)):.4f} m^2/s^2")
    
    if abs(tke_sample - (mke_sample + eke_sample)) < 0.01:
        log_message("  >> Energy budget checks out!")
    else:
        log_message("  >> WARNING: Energy budget discrepancy detected")
    
    ds_tke.close()
    ds_mke.close()
    ds_mean_eke.close()
    
except Exception as e:
    log_message(f"Could not verify energy budget: {e}")

log_message("\n>> Complete kinetic energy budget calculated successfully!")
```
### Sanity checks

```{python std-continuous-check}
import xarray as xr
import numpy as np
import os

path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"

ds = xr.open_dataset(path, engine='netcdf4')

print(ds)
print("\nVariables:", list(ds.data_vars))
print("Dimensions:", ds.dims)
print("Coordinates:", list(ds.coords))

# --- Check for NaNs ---
for var in ["u", "v"]:
    da = ds[var]
    total = da.size
    nans = int(np.isnan(da).sum())
    frac = nans / total
    print(f"{var}: {frac*100:.4f}% missing values")

# --- Check range and mean (sanity check for wind components) ---
for var in ["u", "v"]:
    da = ds[var]
    print(f"{var} range: {float(da.min()):.2f} to {float(da.max()):.2f}, mean={float(da.mean()):.2f}")

# --- Check lat/lon limits ---
print(f"Latitude range: {float(ds.latitude.min())} to {float(ds.latitude.max())}")
print(f"Longitude range: {float(ds.longitude.min())} to {float(ds.longitude.max())}")

# --- Optional: check time coverage ---
print(f"Time coverage: {str(ds.time.values[0])} to {str(ds.time.values[-1])}")

```

```{python climatology-check}
clim = xr.open_dataset("E:/ERA5_data/processed/ERA5_climatology_1958-2024.nc")
for var in ["u_climatology", "v_climatology"]:
    print(var, "range:", float(clim[var].min()), "to", float(clim[var].max()))
    print(var, "NaN fraction:", float(np.isnan(clim[var]).mean()))
```

```{python na-latitude-check}
mask = np.isnan(ds['u'].isel(time=0)).any(dim='longitude')
print(ds.latitude[mask])
```

```{python Na-continuous-cleaning}
import xarray as xr
import numpy as np

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"

# ============================================================================
# STEP 1: CLEAN CONTINUOUS FILE
# ============================================================================
print("Loading continuous file...")
ds = xr.open_dataset(f"{PROCESSED_DATA_DIR}/ERA5_1958-2024_std_res_vars_continuous.nc")

print(f"Original shape: {ds.dims}")
print(f"u NaNs: {ds['u'].isnull().sum().values} ({ds['u'].isnull().sum().values / ds['u'].size * 100:.2f}%)")

# Remove latitude with NaNs (55.375N is outside our 55-90N target anyway)
ds_clean = ds.sel(latitude=slice(56.375, 90.375))

print(f"\nCleaned shape: {ds_clean.dims}")
print(f"u NaNs after cleaning: {ds_clean['u'].isnull().sum().values}")
print(f"v NaNs after cleaning: {ds_clean['v'].isnull().sum().values}")

# Save
print("\nSaving cleaned continuous file...")
ds_clean.to_netcdf(f"{PROCESSED_DATA_DIR}/ERA5_1958-2024_std_res_vars_continuous_clean.nc")
ds.close()
```

```{python}
# ============================================================================
# STEP 2: RECALCULATE CLIMATOLOGY FROM CLEAN DATA
# ============================================================================
print("\nRecalculating climatology from cleaned data...")
u_clim = ds_clean['u'].mean(dim='time', keep_attrs=True)
v_clim = ds_clean['v'].mean(dim='time', keep_attrs=True)

print(f"u_climatology NaNs: {u_clim.isnull().sum().values}")
print(f"v_climatology NaNs: {v_clim.isnull().sum().values}")

clim_ds = xr.Dataset({'u_climatology': u_clim, 'v_climatology': v_clim})
clim_ds.to_netcdf(f"{PROCESSED_DATA_DIR}/ERA5_climatology_1958-2024_clean.nc")
ds_clean.close()

print("\n All cleaned files saved!")
print("  - ERA5_1958-2024_std_res_vars_continuous_clean.nc")
print("  - ERA5_climatology_1958-2024_clean.nc")
print("\nNext: Rerun Step 4 & 5 with these cleaned files")
```

















### How the detrending is done (STEP 2: GPU BAND-PASS FILTER)

The code is performing Ordinary Least Squares (OLS) linear regression on the time series of each individual grid cell. In simple terms, it finds the "line of best fit" through the time series data and then subtracts that line, leaving the data minus its long-term trend.

This process is crucial because a band-pass filter is designed to work on stationary data (data without a trend). Removing the trend first prevents it from distorting the filter's output.

#### 1. Center the Variables

The first step is to "center" both the wind data ($Y$) and the time vector ($t$) by subtracting their respective means ($\bar{Y}$ and $\bar{t}$). This simplifies the math for the next steps.

`data_centered` = $Y_t - \bar{Y}$

`time_centered` = $t - \bar{t}$

#### 2. Calculate the Trend Slope ($\beta_1$)

The code then calculates the slope of the best-fit line using the standard OLS formula. The slope, $\beta_1$, represents the rate of change over time (the trend).

-   **Numerator**: The code calculates the sum of the products of the centered variables. This is related to the covariance between time and your wind data. $$
      \text{numerator} = \sum_{t=1}^{n} (t - \bar{t})(Y_t - \bar{Y})
      $$

-   **Denominator**: The code calculates the sum of the squared centered time variables. This is related to the variance of the time vector. $$
      \text{denominator} = \sum_{t=1}^{n} (t - \bar{t})^2
      $$

-   **Slope**: The slope is the ratio of the numerator to the denominator. $$
      \beta_1 = \frac{\sum_{t=1}^{n} (t - \bar{t})(Y_t - \bar{Y})}{\sum_{t=1}^{n} (t - \bar{t})^2}
      $$ This is what the line `slopes = numerator / denominator` computes for every grid cell simultaneously.

#### 3. Reconstruct and Subtract the Trend

Finally, the code reconstructs the full trend line and subtracts it from the centered data.

-   **Trend Line**: The trend component for each point in time is calculated by multiplying the slope by the centered time vector. $$
      \text{Trend}_t = \beta_1 \times (t - \bar{t})
      $$ This is what the line `trends = time_centered @ slopes.reshape(1, -1)` does.

-   **Final Detrended Data**: This trend is then subtracted from the centered wind data. The result is the residuals ($\epsilon_t$) of the linear regression, which is your original data with the linear trend removed. $$
      \text{data_detrended} = (Y_t - \bar{Y}) - \text{Trend}_t
      $$ The final `data_detrended` variable contains the wind data for each grid cell, now oscillating around zero with no long-term linear trend:


