---
title: "Task 1: Arctic Storminess Climatology"
author: Francesco Lucano
format: html
engine: knitr
warning: false
error: false
toc: true
---

# Setup

A self-contained and reproducible computational environment was established as the foundation for the analysis.

-   **Project Setup**: The workflow was situated within an **RStudio Project** to ensure all file paths were relative and the project was portable. The R **`reticulate`** package was used to create and manage a dedicated Python virtual environment (`"arctic-env"`), ensuring all Python package dependencies are isolated and reproducible.

```{r installation, eval=FALSE}
# --- ONE-TIME INSTALLATION ---

# -- R Packages --
# Install all required R packages from CRAN
install.packages(c(
    "reticulate", "tidyverse", "terra", "sf", "rnaturalearth", 
    "devtools", "future", "furrr", "glue", "progressr"
))

# Install the high-resolution map data package from GitHub
# devtools::install_github("ropensci/rnaturalearthhires")


# -- Python Environment and Packages --
# Create the virtual environment if it doesn't already exist
if (!reticulate::virtualenv_exists(envname = "arctic-env")) {
    reticulate::virtualenv_create(envname = "arctic-env")
}

# Install all required Python packages into the new environment
reticulate::py_install(
    packages = c(
        "numpy", "xarray", "netcdf4", "h5netcdf", 
        "cdsapi", "scipy", "tqdm", "dask",
    ),
    envname = "arctic-env"
)
```

```{r setup-environment}
# Load R Libraries
library(reticulate)
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(glue)
library(future)
library(furrr)
library(progressr)

# Configure Python Environment
use_virtualenv("arctic-env", required = TRUE)


# Define the absolute path to the data storage folder
DATA_DIRECTORY <- "E:/ERA5_data"

# Setting for more parallel downloads
#plan(multisession, workers = 20)
```


# EKE, TKE and MKE 850hPa January 2022 test

**Data Acquisition**: The source data was programmatically downloaded from the Copernicus Climate Data Store (CDS) using the Python **`cdsapi`** library.

-   The specific dataset requested was **`reanalysis-era5-pressure-levels`**.

-   The variables `u_component_of_wind` and `v_component_of_wind` were downloaded at the **850 hPa** pressure level in NetCDF format.

-   The downloaded data was loaded into a Python session as an **`xarray.Dataset`** object, which is ideal for handling labeled, multi-dimensional climate data.

```{python API-request}
import cdsapi
import os
import xarray as xr
from scipy import signal
import numpy as np

# Define the final path where we want the file to be
final_path = os.path.join('data', 'era5_850hpa_2022_01_svalbard.nc')

# First, check if the file already exists at the final destination
if not os.path.exists(final_path):
    print(f"File not found at {final_path}. Starting download from CDS...")
    c = cdsapi.Client()

    dataset = "reanalysis-era5-pressure-levels"
    request = {
        "product_type": "reanalysis",
        "variable": [
            "u_component_of_wind", "v_component_of_wind"
        ],
        "year": "2022",
        "month": "01",
        "day": [
            "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
            "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
            "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31"
        ],
        "time": [
            "00:00", "01:00", "02:00", "03:00", "04:00", "05:00", "06:00",
            "07:00", "08:00", "09:00", "10:00", "11:00", "12:00", "13:00",
            "14:00", "15:00", "16:00", "17:00", "18:00", "19:00", "20:00",
            "21:00", "22:00", "23:00"
        ],
        "pressure_level": "850",
        "format": "netcdf",
        "area": [81, 10, 76, 35]
    }
    
    # Retrieve the data AND download it to the final path in one step by adding .download(target=final_path)
    c.retrieve(dataset, request).download(target=final_path)
    
    print(f"Data successfully downloaded to {final_path}")
else:
    print(f"File already exists at {final_path}. Skipping download.")
```

## Step 1: Reynolds decomposition via Band-Pass filtering

To isolate the transient eddy components (u′ and v′), a digital band-pass filter was applied.

-   **Function Definition**: A reusable Python function, `robust_bandpass_filter`, was created to encapsulate the filtering logic.

-   **Filter Design**: A 5th-order **Butterworth filter** was designed using the `scipy.signal.butter` function. The filter was configured with a passband corresponding to periods of **2 to 6 days**, consistent with the timescale of synoptic-scale storms. The sampling frequency (`fs`) was set to 24 to reflect the hourly resolution of the input data.

-   **Filter Application**: To ensure numerical stability, the time series at each grid cell was first **detrended** using `scipy.signal.detrend`. The Butterworth filter was then applied using `scipy.signal.filtfilt`, which performs a **zero-phase filtering** operation to prevent any temporal shift in the output data. This entire process was applied to each grid cell's time series via a Python loop for maximum stability. The output of this step consists of two new `xarray.DataArray` variables, `u_eddy` and `v_eddy`.

```{python filtering}
# --- PART 1: Define the New, Robust Filter Function ---
def bandpass_filter(ds, u_var, v_var, time_dim, fs, lowcut_days, highcut_days, order=5):
    """
    Applies a robust, detrending band-pass filter to u and v components of an xarray Dataset.
    Uses a simple loop for stability and returns the Dataset with eddy components.
    """
    print("Designing the filter...")
    # Design the filter once
    lowcut_freq = 1 / lowcut_days
    highcut_freq = 1 / highcut_days
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)

    # Get data as numpy arrays
    u_data = ds[u_var].values.astype('float64')
    v_data = ds[v_var].values.astype('float64')
    
    # Create empty arrays for results
    u_eddy_data = np.empty_like(u_data)
    v_eddy_data = np.empty_like(v_data)

    print("Applying filter cell by cell. This may take a moment...")
    # Loop through each spatial grid cell
    for lat_idx in range(u_data.shape[1]):
        for lon_idx in range(u_data.shape[2]):
            # Detrend and filter the time series for this cell
            u_filtered = signal.filtfilt(b, a, signal.detrend(u_data[:, lat_idx, lon_idx]))
            v_filtered = signal.filtfilt(b, a, signal.detrend(v_data[:, lat_idx, lon_idx]))
            
            # Store the result
            u_eddy_data[:, lat_idx, lon_idx] = u_filtered
            v_eddy_data[:, lat_idx, lon_idx] = v_filtered
    
    print("Filtering complete.")

    # Add the results back to the original dataset
    ds['u_eddy'] = ((time_dim, 'latitude', 'longitude'), u_eddy_data)
    ds['v_eddy'] = ((time_dim, 'latitude', 'longitude'), v_eddy_data)
    
    return ds
```

```{python}
# --- PART 2: Load Data and Apply the Function ---
# Load the dataset
ds_real = xr.open_dataset('data/era5_850hpa_2022_01_svalbard.nc').squeeze()

# Define parameters
fs_samples_per_day = 24
lowcut_period_days = 6.0
highcut_period_days = 2.0

# Call the self-contained function
ds_real_filtered = bandpass_filter(
    ds=ds_real,
    u_var='u',
    v_var='v',
    time_dim='valid_time',
    fs=fs_samples_per_day,
    lowcut_days=lowcut_period_days,
    highcut_days=highcut_period_days
)

# --- PART 3: Sanity Check ---
print("\nSanity Check of filtered 'u_eddy' values:")
print(f"  Min: {ds_real_filtered['u_eddy'].min().values:.2f}")
print(f"  Max: {ds_real_filtered['u_eddy'].max().values:.2f}")
print(f"  Mean: {ds_real_filtered['u_eddy'].mean().values:.2f}")
```

**\## 1. The `bandpass_filter` Function**

First, we defined a Python function.

```         
def bandpass_filter(data_array, fs, lowcut_days, highcut_days, order=5):     # ... code inside ...     return filtered_data 
```

**Why a function?** We need to perform the exact same filtering operation on both the `u` and `v` wind components. Encapsulating the logic in a function makes our code **reusable, readable, and less prone to errors** from copy-pasting.

**\## 2. The Function's Inputs (Parameters)**

Let's look at the function's "control knobs":

-   `data_array`: This is the input data we want to filter (e.g., the `xarray` DataArray containing our `u` wind component).

-   `fs`: This stands for **sampling frequency**. Think of it as frames per second in a video. It tells the filter how many data points we have per time unit. Since our filter is defined in **days** (a 2-6 **day** band) and our data is **hourly**, we have **24 samples per day**. So, `fs = 24`.

-   `lowcut_days` and `highcut_days`: These are the edges of our "band," directly from your proposal. We set them to `6.0` and `2.0`. This tells the filter we want to *keep* the signals with a period *between* 2 and 6 days.

-   `order=5`: This controls the "steepness" of the filter. A higher order creates a sharper cutoff between the signals we keep and those we reject. An order of 5 is a robust, standard choice that provides a clean separation.

**\## 3. Inside the Function: The "How"**

This is where the scientific computation happens.

A. Designing the Filter

```         
from scipy import signal  # ... b, a = signal.butter(order, [low, high], btype='band') 
```

Digital filters are defined by a set of mathematical coefficients. This line uses the `scipy.signal.butter` function to design a **Butterworth filter**, which is a standard type known for being very stable and not distorting the signal in the "passband" (the 2-6 day range we want to keep). The function takes our parameters and calculates the necessary coefficients, which it returns as `b` and `a`.

B. Applying the Filter

```         
lambda x: signal.filtfilt(b, a, signal.detrend(x))
```

This is the core operation. The `signal.filtfilt` function takes the filter coefficients (`b`, `a`) and applies them to the data (`x`). The reason we use `filtfilt` ("filter-filter") is crucial:

-   It filters the data once in the forward direction.

-   Then, it filters the resulting signal again in the backward direction.

This two-way pass **perfectly corrects for any time lag (phase shift)** that a normal filter would introduce. This ensures that the peaks and troughs in our filtered storm signal are perfectly aligned in time with the original data. The data is also detrended with `signal.detrend()` prior to filtering to remove any linear trends that could cause filter instability.

C. Applying Across the Entire grid

```         
filtered_data = xr.apply_ufunc(...) 
```

This is the most powerful part of the code. The `filtfilt` function is designed to work on a simple, 1D time series. But our data is a 3D grid (time, latitude, longitude).

`xr.apply_ufunc` is `xarray`'s magic wand for this problem. It intelligently applies the 1D `filtfilt` function to every single time series at each individual latitude/longitude grid point. It handles all the looping for us in a highly optimized way. The `input_core_dims=[['time']]` argument is our instruction to `xarray`, telling it that `filtfilt` operates along the `time` dimension.

**\## 4. Putting It All Together**

Finally, we call our function:

```         
u_eddy = bandpass_filter(ds_real['u'], fs_samples_per_day, ...) v_eddy = bandpass_filter(ds_real['v'], fs_samples_per_day, ...) 
```

This executes the whole process. It feeds the raw `u` and `v` wind data into our function. The function then designs the filter and applies it across the entire grid, returning the perfectly isolated eddy components (u' and v'). These are now ready for the final EKE calculation.

## Step 2: Calculate EKE, MKE and TKE

The three key kinetic energy indices were calculated in Python.

-   **Eddy Kinetic Energy (EKE)**: The instantaneous EKE was calculated at each grid cell for every time step from the eddy components. The result was stored as a new 3D `DataArray` named `eke`.

-   **Mean Kinetic Energy (MKE)**: The time-averaged wind components, u and v, were calculated by applying the `.mean(dim='valid_time')` method to the original `u` and `v` DataArrays. MKE was then calculated and resulted in a 2D DataArray.

-   **Total Kinetic Energy (TKE)**: The time-averaged EKE, was first calculated. The mean TKE was then computed as the sum of the two time-averaged components:.

```{python eke-calculation}
# 'ds_real' is the xarray Dataset from the previous chunk, containing u_eddy and v_eddy

# 1. Calculate EKE using the eddy components
# EKE = 0.5 * (u'^2 + v'^2)
eke = 0.5 * (np.square(ds_real['u_eddy']) + np.square(ds_real['v_eddy']))

# 2. Add EKE as a new data variable to the dataset
ds_real['eke'] = eke

# 3. Print the dataset to confirm the new 'eke' variable is present
print(ds_real)

# 4. Calculate the time-averaged wind components (u_bar, v_bar)
# We take the mean of the original wind fields across the 'valid_time' dimension.
u_bar = ds_real['u'].mean(dim='valid_time')
v_bar = ds_real['v'].mean(dim='valid_time')

# 5. Calculate Mean Kinetic Energy (MKE) using the formula
mke = 0.5 * (np.square(u_bar) + np.square(v_bar))

# 6. Calculate the time-average of our previously calculated EKE
eke_mean = ds_real['eke'].mean(dim='valid_time')

# 7. Calculate the time-average of Total Kinetic Energy (TKE)
tke = mke + eke_mean

# 8. Combine these into a final summary dataset for easy access
ds_summary = xr.Dataset({
    'mke': mke, 
    'eke_mean': eke_mean, 
    'tke': tke
})

# Rename the 'eke_mean' to just 'eke' for simplicity, as it's the mean EKE
ds_summary = ds_summary.rename({'eke_mean': 'eke'})

print("Summary of Monthly-Mean Kinetic Energy Components (m²/s²):")
print(ds_summary)
```

## Step 3: Calculate EKE variance

The primary storminess metric was calculated as the final analytical step in Python.

-   The **variance of EKE** was computed by applying the `.var(dim='valid_time')` method to the 3D `eke` DataArray. This operation collapses the time dimension, resulting in a final 2D `xarray.DataArray` (`eke_variance`) which serves as the "fundamental metric for overall storm track activity."

```{python eke-variance}
# 'ds_real' now contains our 'eke' variable

# Calculate the variance across the time dimension ('valid_time')
# This leaves us with a 2D grid of (latitude, longitude)
eke_variance = ds_real['eke'].var(dim='valid_time')

# Print the final 2D DataArray
print("Final 2D EKE Variance grid:")
print(eke_variance)
```

## Step 3: Conversion and remapping

The final 2D data products were transferred from Python to R for remapping and visualization.

-   **Data Transfer**: The `eke_variance`, `mke`, and `tke` `xarray` objects were accessed in R using **`reticulate`**'s `py$` accessor.

-   **Remapping to Standard Grid**: As required by your proposal for robust trend analysis, each data grid was remapped to a common **1.0° x 1.0° grid**. This was achieved by:

    -   First converting the Python objects into high-resolution **`terra::SpatRaster`** objects in R.

    -   Creating a blank 1.0° target grid (`template_raster`).

    -   Using the **`terra::resample()`** function with `bilinear` interpolation to project the high-resolution data onto the target grid.

```{r}
# --- PART 1: Define the Integrated Function ---
# This single function handles both conversion from Python and remapping.
process_and_remap_py <- function(py_xarray_object, layer_name, template_raster) {
  
  # Step A: Convert the Python object to a high-resolution R SpatRaster
  high_res_raster <- terra::rast(
    t(py_xarray_object$values),
    ext = terra::ext(
      min(py_xarray_object$longitude$values),
      max(py_xarray_object$longitude$values),
      min(py_xarray_object$latitude$values),
      max(py_xarray_object$latitude$values)
    ),
    crs = "EPSG:4326"
  )
  
  # Step B: Remap the high-resolution raster to the template grid
  resampled_raster <- terra::resample(high_res_raster, template_raster, method = "bilinear")
  
  # Set the final layer name
  names(resampled_raster) <- layer_name
  
  # Return the final, remapped raster
  return(resampled_raster)
}

# --- PART 2: Prepare the Target Grid (Blueprint) ---
# We create the 1.0-degree blueprint once. All data will be remapped to this grid.
template_raster <- terra::rast(
  resolution = 1.0, 
  extent = terra::ext(
    min(py$eke_variance$longitude$values),
    max(py$eke_variance$longitude$values),
    min(py$eke_variance$latitude$values),
    max(py$eke_variance$latitude$values)
  ),
  crs = "EPSG:4326"
)

# --- PART 3: Apply the New Function to All Variables ---
# Now we just call our powerful new function for each variable.
eke_variance_resampled <- process_and_remap_py(py$eke_variance, "eke_variance", template_raster)
mke_resampled <- process_and_remap_py(py$ds_summary$mke, "mke", template_raster)
tke_resampled <- process_and_remap_py(py$ds_summary$tke, "tke", template_raster)

# Let's print one of the final products to verify.
# The resolution should now be 1 x 1.
print("--- Final Resampled EKE Raster (1.0-degree resolution) ---")
print(eke_variance_resampled)
```

## Step 4: Visualize the final maps

The final, remapped `SpatRaster` objects were converted to `data.frame` objects.

Publication-quality maps were generated using the **`ggplot2`** package, with `geom_raster` to display the energy data.

A geographical outline of Svalbard was acquired using the **`rnaturalearth`** package and overlaid using **`geom_sf`** to provide essential context.

```{r visualization-EKE}
# Devtools to install the high-resolution map data package from its source
# devtools::install_github("ropensci/rnaturalearthhires")

# Acquire all Norwegian state-level polygons using the function: ne_states().
norway_territories <- ne_states(country = "Norway", returnclass = "sf")

# Now, filter that collection to get only the feature named "Svalbard"
svalbard_sf <- norway_territories |>
  filter(name == "Svalbard")

# Get the bounding box of the Svalbard polygon
svalbard_bbox <- st_bbox(svalbard_sf)

# Convert the raster to a data frame for plotting with ggplot2
eke_variance_resampled_df <- as.data.frame(eke_variance_resampled, xy = TRUE)

# Create the final map with the new contour layer
ggplot() +
  # First, draw the EKE raster data
  geom_raster(data = eke_variance_resampled_df, aes(x = x, y = y, fill = eke_variance)) +
  
  # Next, draw the Svalbard outline on top
  geom_sf(data = svalbard_sf, fill = NA, color = "white", linewidth = 0.5) +
  
  # Apply color scale and labels
  scale_fill_viridis_c(name = "EKE Variance\n(m²/s²)²)") +
  labs(
    title = "Storm Track Activity over Svalbard",
    subtitle = "January 2022 | Variance of 2-6 day band-pass filtered EKE at 850 hPa",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # IMPORTANT: Set the plot's coordinate system and limits
  # We add a small buffer (e.g., +/- 1 degree) for a nice margin
  coord_sf(
    xlim = c(svalbard_bbox["xmin"], svalbard_bbox["xmax"]),
    ylim = c(svalbard_bbox["ymin"] + 1, svalbard_bbox["ymax"]),
    expand = FALSE
  ) +
  theme_minimal()
```

```{r visualization-MKE}
mke_resampled_df <- as.data.frame(mke_resampled, xy = TRUE)

ggplot() +
  geom_raster(data = mke_resampled_df, aes(x = x, y = y, fill = mke)) +
  
  geom_sf(data = svalbard_sf, fill = NA, color = "white", linewidth = 0.5) +
  
  scale_fill_viridis_c(name = "MKE\n(m²/s²)") +
  labs(
    title = "Mean Kinetic Energy (MKE) over Svalbard",
    subtitle = "January 2022 | Energy of the time-averaged flow at 850 hPa",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  coord_sf(
    xlim = c(svalbard_bbox["xmin"], svalbard_bbox["xmax"]),
    ylim = c(svalbard_bbox["ymin"] + 1, svalbard_bbox["ymax"]),
    expand = FALSE
  ) +
  theme_minimal()
```

```{r visualization-TKE}
# Convert the raster to a data frame for plotting with ggplot2
tke_resampled_df <- as.data.frame(tke_resampled, xy = TRUE)

# Create the final map with the new contour layer
ggplot() +
  # First, draw the EKE raster data
  geom_raster(data = tke_resampled_df, aes(x = x, y = y, fill = tke)) +
  
  # Next, draw the Svalbard outline on top
  geom_sf(data = svalbard_sf, fill = NA, color = "white", linewidth = 0.5) +
  
  # Apply color scale and labels
  scale_fill_viridis_c(name = "TzKE\n(m²/s²)") +
  labs(
    title = "Total Kinetic Energy (TKE) over Svalbard",
    subtitle = "January 2022 | Total mean kinetic energy (MKE + EKE) at 850 hPa",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # IMPORTANT: Set the plot's coordinate system and limits
  # We add a small buffer (e.g., +/- 1 degree) for a nice margin
  coord_sf(
    xlim = c(svalbard_bbox["xmin"], svalbard_bbox["xmax"]),
    ylim = c(svalbard_bbox["ymin"] + 1, svalbard_bbox["ymax"]),
    expand = FALSE
  ) +
  theme_minimal()
```


# 1979-present EKE, TKE and MKE 850hPa

```{python single-download}
import cdsapi
import os
import xarray as xr
from scipy import signal
import numpy as np

# --- 1. CONFIGURATION ---
# Access the DATA_DIRECTORY variable from your R configuration chunk
data_dir = r.DATA_DIRECTORY

# Create the directory on your hard disk if it doesn't exist
if not os.path.exists(data_dir):
    os.makedirs(data_dir)
    print(f"Created directory: {data_dir}")

# Let's start with just two years to test the monthly download loop.
# For the final run, you will change this to range(1958, 2025).
YEARS_TO_DOWNLOAD = range(1958, 2024) 

AREA_PAN_ARCTIC = [90, -180, 55, 180] # North, West, South, East

# --- 2. NESTED LOOP AND DOWNLOAD ---
c = cdsapi.Client()
print("Starting batch download process (one file per month)...")

# Outer loop for each year
for year in YEARS_TO_DOWNLOAD:
    # Inner loop for each month (from 1 to 12)
    for month in range(1, 13):
        # Format the month with a leading zero (e.g., 1 -> "01", 10 -> "10")
        month_str = f"{month:02d}"
        
        # Create a specific filename for each YEAR and MONTH
        final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-{month_str}.nc")
        
        # Check if this specific monthly file already exists
        if not os.path.exists(final_path):
            print(f"--- Downloading data for: {year}-{month_str} ---")
            
            request = {
                "product_type": "reanalysis",
                "variable": ["u_component_of_wind", "v_component_of_wind"],
                "year": str(year),
                "month": month_str, # Dynamically set the month
                "day": [
                    "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
                    "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", 
                    "31"
                ],
                "time": [
                    "00:00", "01:00", "02:00", "03:00", "04:00", "05:00", "06:00",
                    "07:00", "08:00", "09:00", "10:00", "11:00", "12:00", "13:00",
                    "14:00", "15:00", "16:00", "17:00", "18:00", "19:00", "20:00",
                    "21:00", "22:00", "23:00"
                ],
                "pressure_level": "850",
                "format": "netcdf",
                "area": AREA_PAN_ARCTIC
            }
            
            # Use a try-except block to handle potential API errors for a single month
            try:
                c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
                print(f"Successfully downloaded to {final_path}")
            except Exception as e:
                print(f"!!! FAILED to download data for {year}-{month_str}. Error: {e}")

        else:
            print(f"--- File for {year}-{month_str} already exists. Skipping. ---")

print("\nFull data download process complete for the specified years.")
```

```{r}
YEARS_TO_DOWNLOAD <- 1958:2024
AREA_OF_INTEREST <- c(90, -180, 55, 180)
```

```{python parallel-download}
import cdsapi
import os
import concurrent.futures
from tqdm.auto import tqdm

# --- 1. CONFIGURATION ---
# Get the configuration variables from your R 'configuration' chunk
data_dir = r.DATA_DIRECTORY
years_to_download = r.YEARS_TO_DOWNLOAD
area_of_interest = r.AREA_OF_INTEREST

# Set the number of parallel download
MAX_WORKERS = 20

# --- 2. DEFINE THE DOWNLOAD FUNCTION FOR A SINGLE JOB ---
def download_era5_month(job_info):
    year, month = job_info # Unpack the year and month for this job
    month_str = f"{month:02d}"
    final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-{month_str}.nc")

    if os.path.exists(final_path):
        return f"Skipped (already exists): {year}-{month_str}"

    # Each worker needs its own API client instance
    c = cdsapi.Client(quiet=True, progress=False)
    
    request = {
        'product_type': 'reanalysis', 'variable': ['u_component_of_wind', 'v_component_of_wind'],
        'year': str(int(year)), 'month': month_str,
        'day': ['01','02','03','04','05','06','07','08','09','10','11','12','13','14','15','16','17','18','19','20','21','22','23','24','25','26','27','28','29','30','31'],
        'time': ['00:00','01:00','02:00','03:00','04:00','05:00','06:00','07:00','08:00','09:00','10:00','11:00','12:00','13:00','14:00','15:00','16:00','17:00','18:00','19:00','20:00','21:00','22:00','23:00'],
        'pressure_level': '850', 'format': 'netcdf', 'area': area_of_interest
    }
    
    try:
        c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
        return f"Success: {year}-{month_str}"
    except Exception as e:
        return f"!!! FAILED: {year}-{month_str}. Error: {e}"

# --- 3. CREATE THE LIST OF JOBS ---
jobs = []
for year in years_to_download:
    for month in range(1, 13):
        jobs.append((year, month))

# --- 4. EXECUTE THE DOWNLOADS IN PARALLEL WITH A PROGRESS BAR ---
print(f"Starting parallel download for {len(jobs)} files using {MAX_WORKERS} workers...")

with concurrent.futures.ThreadPoolExecutor(max_workers=MAX_WORKERS) as executor:
    # 'tqdm' wraps the executor to create a live progress bar.
    results = list(tqdm(executor.map(download_era5_month, jobs), total=len(jobs)))

print("\nAll download tasks have been processed.")
```

## Analytical time series

```{python cuda-optimized}
import xarray as xr
from scipy import signal
import numpy as np
import cupy as cp
import cupyx.scipy.signal as cp_signal
import glob
import os
from tqdm.auto import tqdm

# ============================================================================
# CONFIGURATION
# ============================================================================
raw_data_dir = r.DATA_DIRECTORY
output_dir = os.path.join(raw_data_dir, "processed_correct")
if not os.path.exists(output_dir):
    os.makedirs(output_dir)

YEARS_TO_PROCESS = range(1958, 2025)
cp.cuda.set_allocator(cp.cuda.MemoryPool().malloc)

# Band-pass filter parameters (2-6 day synoptic scale)
FS = 24.0
LOWCUT_DAYS = 6.0  # Low-frequency cutoff
HIGHCUT_DAYS = 2.0  # High-frequency cutoff
ORDER = 5

# Pre-compute filter coefficients
FILTER_B, FILTER_A = signal.butter(
    ORDER, 
    [1/LOWCUT_DAYS, 1/HIGHCUT_DAYS], 
    btype='band', 
    fs=FS
)
FILTER_B_GPU = cp.asarray(FILTER_B)
FILTER_A_GPU = cp.asarray(FILTER_A)

# ============================================================================
# STEP 0: RESOLUTION STANDARDIZATION
# ============================================================================
def standardize_resolution():
    """
    Remap all raw ERA5 data to standard 1.0° x 1.0° grid.
    """
    standardized_dir = os.path.join(raw_data_dir, "standardized_1deg")
    if not os.path.exists(standardized_dir):
        os.makedirs(standardized_dir)
    
    print("\n" + "="*60)
    print("RESOLUTION STANDARDIZATION TO 1.0° x 1.0°")
    print("="*60)
    
    raw_files = sorted(glob.glob(os.path.join(raw_data_dir, "era5_850hpa_panarctic_*.nc")))
    
    for file in tqdm(raw_files, desc="Standardizing resolution"):
        filename = os.path.basename(file)
        output_file = os.path.join(standardized_dir, filename)
        
        if os.path.exists(output_file):
            continue
        
        # Load and squeeze
        ds = xr.open_dataset(file).squeeze()  # Remove singleton dims
        
        # Create 1.0° grid
        lat_1deg = np.arange(ds.latitude.min(), ds.latitude.max() + 1.0, 1.0)
        lon_1deg = np.arange(ds.longitude.min(), ds.longitude.max() + 1.0, 1.0)
        
        # Regrid
        ds_regrid = ds.interp(latitude=lat_1deg, longitude=lon_1deg, method='linear')
        
        # Ensure clean dimensions
        ds_regrid = ds_regrid.squeeze()
        
        # Save
        ds_regrid.to_netcdf(output_file)
        ds.close()
        ds_regrid.close()
    
    print(f"Standardized files saved to: {standardized_dir}")
    return standardized_dir

# ============================================================================
# STEP 1: CALCULATE CLIMATOLOGY
# ============================================================================
def calculate_climatology(data_dir):
    """
    Calculate long-term mean u and v for Reynolds decomposition.
    """
    climatology_file = os.path.join(output_dir, "climatology_1958-2024.nc")
    
    if os.path.exists(climatology_file):
        print("Climatology already exists.")
        return xr.open_dataset(climatology_file)
    
    print("\n" + "="*60)
    print("CALCULATING CLIMATOLOGY (1958-2024)")
    print("="*60)
    
    all_files = sorted(glob.glob(os.path.join(data_dir, "era5_850hpa_panarctic_*.nc")))
    
    print(f"Loading {len(all_files)} files...")
    ds_all = xr.open_mfdataset(all_files, combine='by_coords', parallel=True)
    
    print("Computing time-mean...")
    u_clim = ds_all['u'].mean(dim='valid_time').compute()
    v_clim = ds_all['v'].mean(dim='valid_time').compute()
    
    ds_clim = xr.Dataset({
        'u_climatology': u_clim,
        'v_climatology': v_clim
    })
    ds_clim.to_netcdf(climatology_file)
    
    ds_all.close()
    return ds_clim

# ============================================================================
# STEP 2: GPU BAND-PASS FILTER
# ============================================================================
def bandpass_filter_gpu(data_3d):
    """
    Apply 2-6 day band-pass filter on GPU.
    Handles data with extra singleton dimensions.
    """
    # Squeeze out any singleton dimensions
    data_np = np.squeeze(data_3d)
    
    # Verify we now have 3D data
    if data_np.ndim != 3:
        raise ValueError(f"Expected 3D data after squeeze, got {data_np.ndim}D: {data_np.shape}")
    
    data_gpu = cp.asarray(data_np, dtype=cp.float64)
    ntime, nlat, nlon = data_gpu.shape
    
    data_reshaped = data_gpu.reshape(ntime, nlat * nlon)
    
    # Detrend
    mean_vals = cp.mean(data_reshaped, axis=0, keepdims=True)
    data_centered = data_reshaped - mean_vals
    
    time_indices = cp.arange(ntime, dtype=cp.float64).reshape(-1, 1)
    time_centered = time_indices - cp.mean(time_indices)
    
    numerator = cp.sum(time_centered * data_centered, axis=0)
    denominator = cp.sum(time_centered ** 2)
    slopes = numerator / denominator
    trends = time_centered @ slopes.reshape(1, -1)
    data_detrended = data_centered - trends
    
    # Apply filter
    data_transposed = data_detrended.T
    filtered_transposed = cp_signal.filtfilt(
        FILTER_B_GPU, FILTER_A_GPU, data_transposed, axis=1,
        padlen=min(3 * max(len(FILTER_B_GPU), len(FILTER_A_GPU)), ntime - 1)
    )
    
    filtered_3d = filtered_transposed.T.reshape(ntime, nlat, nlon)
    result = cp.asnumpy(filtered_3d)
    
    del data_gpu, data_reshaped, data_centered, data_detrended
    del data_transposed, filtered_transposed, filtered_3d
    cp.get_default_memory_pool().free_all_blocks()
    
    return result

# ============================================================================
# STEP 3: PROCESS YEAR WITH CORRECT REYNOLDS DECOMPOSITION
# ============================================================================
def process_year_correct(year, data_dir, climatology):
    """
    CORRECT IMPLEMENTATION:
    1. Load data
    2. Subtract climatology (Reynolds decomposition) → get u', v'
    3. Apply band-pass filter to u', v' → get u'_filtered, v'_filtered
    4. Calculate EKE, MKE, TKE
    """
    print(f"\n{'='*60}")
    print(f"PROCESSING YEAR {year}")
    print(f"{'='*60}")
    
    year_file = os.path.join(output_dir, f"storminess_{year}.nc")
    
    if os.path.exists(year_file):
        print(f"Year {year} already processed.")
        return f"[SKIP] {year}"
    
    try:
        # Load year with overlap
        year_files = sorted(glob.glob(
            os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-*.nc")
        ))
        
        if len(year_files) == 0:
            return f"[WARN] No files for {year}"
        
        # Add overlap
        extended_files = []
        prev_dec = os.path.join(data_dir, f"era5_850hpa_panarctic_{year-1}-12.nc")
        if os.path.exists(prev_dec):
            extended_files.append(prev_dec)
        
        extended_files.extend(year_files)
        
        next_jan = os.path.join(data_dir, f"era5_850hpa_panarctic_{year+1}-01.nc")
        if os.path.exists(next_jan):
            extended_files.append(next_jan)
        
        ds = ds = xr.open_mfdataset(extended_files, combine='by_coords').squeeze(drop=True).load()
        
        print(f"   Data shape: {ds['u'].shape}")
        
        # CRITICAL STEP: Reynolds decomposition FIRST
        print(f"   Step 1: Reynolds decomposition (u' = u - ū)...")
        u_clim = climatology['u_climatology']
        v_clim = climatology['v_climatology']
        
        # Calculate eddy components (deviation from climatology)
        u_prime = ds['u'] - u_clim
        v_prime = ds['v'] - v_clim
        
        # THEN apply band-pass filter to the eddy components
        print(f"   Step 2: Band-pass filtering u', v'...")
        u_prime_filtered = bandpass_filter_gpu(u_prime.values.astype(np.float64))
        v_prime_filtered = bandpass_filter_gpu(v_prime.values.astype(np.float64))
        
        # Add to dataset
        ds['u_eddy'] = (ds['u'].dims, u_prime_filtered)
        ds['v_eddy'] = (ds['v'].dims, v_prime_filtered)
        
        # Calculate instantaneous EKE
        ds['eke'] = 0.5 * (ds['u_eddy']**2 + ds['v_eddy']**2)
        
        # Calculate instantaneous TKE (for proper averaging)
        ds['tke'] = 0.5 * (ds['u']**2 + ds['v']**2)
        
        # Extract target year
        ds_year = ds.sel(valid_time=slice(f"{year}-01-01", f"{year}-12-31"))
        
        print(f"   Step 3: Calculating monthly summaries...")
        
        # MKE from climatology (same for all months)
        mke = 0.5 * (u_clim**2 + v_clim**2)
        
        monthly_summaries = []
        
        for month in range(1, 13):
            ds_month = ds_year.sel(valid_time=ds_year.valid_time.dt.month == month)
            
            if len(ds_month.valid_time) == 0:
                continue
            
            # Time-averaged quantities for this month
            eke_mean = ds_month['eke'].mean('valid_time')
            eke_variance = ds_month['eke'].var('valid_time')
            tke_mean = ds_month['tke'].mean('valid_time')
            
            ds_summary = xr.Dataset({
                'eke_mean': eke_mean,
                'eke_variance': eke_variance,  # PRIMARY STORMINESS INDEX
                'mke': mke,
                'tke_mean': tke_mean
            })
            
            ds_summary = ds_summary.assign_coords({'month': month})
            monthly_summaries.append(ds_summary)
        
        # Combine months
        ds_year_summary = xr.concat(monthly_summaries, dim='month')
        ds_year_summary['year'] = year
        
        # Save
        ds_year_summary.to_netcdf(year_file)
        
        ds.close()
        print(f"   Saved: {year_file}")
        
        return f"[DONE] {year}"
        
    except Exception as e:
        cp.get_default_memory_pool().free_all_blocks()
        return f"[ERROR] {year}: {str(e)}"

# ============================================================================
# MAIN EXECUTION
# ============================================================================
print("="*60)
print("SCIENTIFICALLY CORRECT ARCTIC STORMINESS ANALYSIS")
print("="*60)
print(f"GPU: {cp.cuda.runtime.getDeviceProperties(0)['name'].decode()}")
print(f"GPU Memory: {cp.cuda.runtime.getDeviceProperties(0)['totalGlobalMem'] / 1e9:.1f} GB")
print("="*60)

# Step 0: Resolution standardization
print("\nSTEP 0: Resolution standardization...")
standardized_dir = standardize_resolution()

# Step 1: Calculate climatology
print("\nSTEP 1: Calculate climatology...")
climatology = calculate_climatology(standardized_dir)

# Step 2: Process all years
print("\nSTEP 2: Process all years...")
results = []

for year in tqdm(YEARS_TO_PROCESS, desc="Years"):
    result = process_year_correct(year, standardized_dir, climatology)
    results.append(result)

# Summary
print("\n" + "="*60)
print("SUMMARY")
print("="*60)
print(f"Completed: {sum('[DONE]' in r for r in results)}")
print(f"Skipped: {sum('[SKIP]' in r for r in results)}")
print(f"Errors: {sum('[ERROR]' in r for r in results)}")
print("="*60)
print(f"\nOutputs in: {output_dir}")
print("Ready for trend analysis!")
```


