---
title: "Task 1: Arctic Storminess Climatology"
author: Francesco Lucano
format: html
engine: knitr
editor: visual
warning: false
error: false
toc: true
---zz
---

# Setup

A self-contained and reproducible computational environment was established as the foundation for the analysis.

-   **Project Setup**: The workflow was situated within an **RStudio Project** to ensure all file paths were relative and the project was portable. The R **`reticulate`** package was used to create and manage a dedicated Python virtual environment (`"arctic-env"`), ensuring all Python package dependencies are isolated and reproducible.

```{r installation, eval=FALSE}
# --- ONE-TIME INSTALLATION ---

# -- R Packages --
# Install all required R packages from CRAN
install.packages(c(
    "reticulate", "tidyverse", "terra", "sf", "rnaturalearth", 
    "devtools", "future", "furrr", "glue", "progressr"
))

# Install the high-resolution map data package from GitHub
# devtools::install_github("ropensci/rnaturalearthhires")


# -- Python Environment and Packages --
# Create the virtual environment if it doesn't already exist
if (!reticulate::virtualenv_exists(envname = "arctic-env")) {
    reticulate::virtualenv_create(envname = "arctic-env")
}

# Install all required Python packages into the new environment
reticulate::py_install(
    packages = c(
        "numpy", "xarray", "netcdf4", "h5netcdf", 
        "cdsapi", "scipy", "tqdm", "dask", "matplotlib", "bottleneck"
    ),
    envname = "arctic-env"
)
```

```{r setup-environment}
# Load R Libraries
library(reticulate)
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(glue)
library(future)
library(furrr)
library(progressr)
library(tictoc)
library(ncdf4)
library(glue)
library(lubridate)
library(trend)
library(ggplot2)
library(viridis)
library(patchwork)
library(modifiedmk)
library(fields)
library(RColorBrewer)
library(dplyr)
library(zoo)
library(foreach)
library(doParallel)
library(pbapply)
library(stars)        # For handling raster/gridded data
library(modifiedmk) # Use this package for the corrected MK test
library(nlme)

#library(discordrpc)   # Disable this if you run the code on your pc

# Configure Python Environment
use_virtualenv("arctic-env", required = TRUE)

# Define the absolute path to the data storage folder
DATA_DIRECTORY <- "E:/ERA5_data/raw"
PROCESSED_DATA_DIR <- "E:/ERA5_data/processed"

# Setting for more parallel downloads
#plan(multisession, workers = 20)
```

# Data download

```{python single-download}
import cdsapi
import os
import xarray as xr
from scipy import signal
import numpy as np

# --- 1. CONFIGURATION ---
# Access the DATA_DIRECTORY variable from your R configuration chunk
data_dir = r.DATA_DIRECTORY

# Create the directory on your hard disk if it doesn't exist
if not os.path.exists(data_dir):
    os.makedirs(data_dir)
    print(f"Created directory: {data_dir}")

YEARS_TO_DOWNLOAD = range(1958, 2024) 

AREA_PAN_ARCTIC = [90, -180, 55, 180] # North, West, South, East

# --- 2. NESTED LOOP AND DOWNLOAD ---
c = cdsapi.Client()
print("Starting batch download process (one file per month)...")

# Outer loop for each year
for year in YEARS_TO_DOWNLOAD:
    # Inner loop for each month (from 1 to 12)
    for month in range(1, 13):
        # Format the month with a leading zero (e.g., 1 -> "01", 10 -> "10")
        month_str = f"{month:02d}"
        
        # Create a specific filename for each YEAR and MONTH
        final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-{month_str}.nc")
        
        # Check if this specific monthly file already exists
        if not os.path.exists(final_path):
            print(f"--- Downloading data for: {year}-{month_str} ---")
            
            request = {
                "product_type": "reanalysis",
                "variable": ["u_component_of_wind", "v_component_of_wind"],
                "year": str(year),
                "month": month_str, # Dynamically set the month
                "day": [
                    "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
                    "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", 
                    "31"
                ],
                "time": [
                    '00:00', '06:00', '12:00', '18:00'
                ],
                "pressure_level": "850",
                "format": "netcdf",
                "area": AREA_PAN_ARCTIC
            }
            
            # Use a try-except block to handle potential API errors for a single month
            try:
                c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
                print(f"Successfully downloaded to {final_path}")
            except Exception as e:
                print(f"!!! FAILED to download data for {year}-{month_str}. Error: {e}")

        else:
            print(f"--- File for {year}-{month_str} already exists. Skipping. ---")

print("\nFull data download process complete for the specified years.")
```

```{python single-download}
import cdsapi
import os
import xarray as xr
from scipy import signal
import numpy as np

# --- 1. CONFIGURATION ---
# Access the DATA_DIRECTORY variable from your R configuration chunk
data_dir = r.DATA_DIRECTORY

# Create the directory on your hard disk if it doesn't exist
if not os.path.exists(data_dir):
    os.makedirs(data_dir)
    print(f"Created directory: {data_dir}")

YEARS_TO_DOWNLOAD = [1957] 

AREA_PAN_ARCTIC = [90, -180, 55, 180] # North, West, South, East

# --- 2. NESTED LOOP AND DOWNLOAD ---
c = cdsapi.Client()
print("Starting batch download process (one file per month)...")

# Outer loop for each year
for year in YEARS_TO_DOWNLOAD:
    # Inner loop for each month (from 1 to 12)
    for month in [11, 12]:
        # Format the month with a leading zero (e.g., 1 -> "01", 10 -> "10")
        month_str = f"{month:02d}"
        
        # Create a specific filename for each YEAR and MONTH
        final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year}-{month_str}.nc")
        
        # Check if this specific monthly file already exists
        if not os.path.exists(final_path):
            print(f"--- Downloading data for: {year}-{month_str} ---")
            
            request = {
                "product_type": "reanalysis",
                "variable": ["u_component_of_wind", "v_component_of_wind"],
                "year": str(year),
                "month": month_str, # Dynamically set the month
                "day": [
                    "01", "02", "03", "04", "05", "06", "07", "08", "09", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20",
                    "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", 
                    "31"
                ],
                "time": [
                    '00:00', '06:00', '12:00', '18:00'
                ],
                "pressure_level": "850",
                "format": "netcdf",
                "area": AREA_PAN_ARCTIC
            }
            
            # Use a try-except block to handle potential API errors for a single month
            try:
                c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
                print(f"Successfully downloaded to {final_path}")
            except Exception as e:
                print(f"!!! FAILED to download data for {year}-{month_str}. Error: {e}")

        else:
            print(f"--- File for {year}-{month_str} already exists. Skipping. ---")

print("\nFull data download process complete for the specified years.")

```

```{python parallel-download}
# === Complete Python Parallel Download Script ===

import cdsapi
import os
import concurrent.futures
from tqdm.auto import tqdm

# --- 1. CONFIGURATION (Defined directly in Python) ---
data_dir = r.DATA_DIRECTORY
YEARS_TO_DOWNLOAD = [2025] 
AREA_OF_INTEREST = [90, -180, 55, 180]

# Set the number of parallel workers. 2 or 3 is a safe number for the CDS API.
MAX_WORKERS = 5

# --- 2. DEFINE THE DOWNLOAD FUNCTION FOR A SINGLE JOB ---
def download_era5_month(job_info):
    """Downloads a single month of ERA5 data for a given year."""
    year, month = job_info # Unpack the year and month for this job
    
    year_int = int(year)
    month_str = f"{month:02d}"
    
    # Create the directory if it doesn't exist
    os.makedirs(data_dir, exist_ok=True)
    
    final_path = os.path.join(data_dir, f"era5_850hpa_panarctic_{year_int}-{month_str}.nc")

    # Check if the file already exists to avoid re-downloading
    if os.path.exists(final_path):
        return f"Skipped (already exists): {year_int}-{month_str}"

    # Each worker needs its own API client instance
    c = cdsapi.Client(quiet=True, progress=False)
    
    request = {
        'product_type': 'reanalysis',
        'variable': ['u_component_of_wind', 'v_component_of_wind'],
        'year': str(year_int), 
        'month': month_str,
        'day': [
            '01','02','03','04','05','06','07','08','09','10','11','12','13',
            '14','15','16','17','18','19','20','21','22','23','24','25','26',
            '27','28','29','30','31'
        ],
        'time': ['00:00', '06:00', '12:00', '18:00'],
        'pressure_level': '850', 
        'format': 'netcdf', 
        'area': AREA_OF_INTEREST
    }
    
    try:
        # Request and download the data
        c.retrieve('reanalysis-era5-pressure-levels', request).download(target=final_path)
        return f"Success: {year_int}-{month_str}"
    except Exception as e:
        return f"!!! FAILED: {year_int}-{month_str}. Error: {e}"

# --- 3. CREATE THE LIST OF JOBS ---
jobs = []
for year in YEARS_TO_DOWNLOAD: 
    # The range(1, 11) correctly generates numbers from 1 to 10 for Jan-Oct.
    for month in range(1, 11):
        jobs.append((year, month))

# --- 4. EXECUTE THE DOWNLOADS IN PARALLEL ---
if jobs:
    print(f"Starting parallel download for {len(jobs)} files using {MAX_WORKERS} workers...")
    with concurrent.futures.ThreadPoolExecutor(max_workers=MAX_WORKERS) as executor:
        results = list(tqdm(executor.map(download_era5_month, jobs), total=len(jobs)))

    print("\n--- Download Summary ---")
    for res in results:
        if "Success" in res or "FAILED" in res or "Skipped" in res:
             print(res)

    print("\nAll download tasks have been processed.")
else:
    print("No new files to download. The job list is empty.")
```

# 2000-2001 workflow test

## Resolution standardisation

```{python monthly-file-combination}
import xarray as xr
import os
import glob

# Access the R variable for the data directory
data_directory = r.DATA_DIRECTORY
processed_dir = r.PROCESSED_DATA_DIR
years_to_process = [2000, 2001]

# Define the output path for the combined raw file at the beginning
output_path_raw = os.path.join(processed_dir, "ERA5_2000-2001_raw.nc")

# --- NEW: Check if the file already exists using Python syntax ---
if not os.path.exists(output_path_raw):
    print("Combined raw file not found. Starting combination process...\n")
    
    # Find all relevant monthly files for the specified years
    all_files = []
    for year in years_to_process:
        pattern = os.path.join(data_directory, f"era5_850hpa_panarctic_{year}-*.nc")
        all_files.extend(glob.glob(pattern))

    all_files.sort()

    if not all_files:
        raise FileNotFoundError("No NetCDF files found for the specified years.")

    print(f"Found {len(all_files)} monthly files for the period {min(years_to_process)}-{max(years_to_process)}.")

    # Open all files as a single xarray dataset
    ds_raw_combined = xr.open_mfdataset(
        all_files, 
        combine='by_coords', 
        parallel=True, 
        engine='h5netcdf'
    )

    # Save the combined dataset to a single intermediate file
    print(f"Saving combined high-resolution data to:\n{output_path_raw}")
    ds_raw_combined.to_netcdf(output_path_raw)

    print("\nCombining raw files complete.")

else:
    # If the file exists, print a message and skip the process
    print(f"Combined raw file already exists. Skipping combination.\nPath: {output_path_raw}")
```

```{r standardization}
# --- UPDATED STANDARDIZATION CHUNK WITH POLE COVERAGE ---
#raw_combined_path <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_raw.nc")
#output_path_correct <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_std_res_vars.nc")

raw_combined_path <- file.path(DATA_DIRECTORY, "era5_850hpa_panarctic_1958-01.nc")
output_path_correct <- file.path(PROCESSED_DATA_DIR, "jan_58_std")

if (!file.exists(output_path_correct)) {

  cat("Standardized file not found. Starting full resampling process...\n")
  
  era5_raw <- terra::rast(raw_combined_path)
  cat("--- Properties of RAW High-Resolution Data ---\n")
  print(era5_raw)

  # --- FIX: Create template with explicit nrows to ensure pole coverage ---
  # We need 36 rows (not 35) to reach from ~55°N to past 90°N with 1° resolution
  template_raster <- terra::rast(
    nrows = 35,           # Explicitly set to 36 rows to cover the pole
    ncols = 360,          # 360 columns for 1° longitude resolution
    xmin = -180.125,
    xmax = 179.875,
    ymin = 55.875,
    ymax = 90.875,        # Extend to 90.875 to ensure pole coverage
    crs = "EPSG:4326"
  )
  
  cat("\nTemplate grid properties:\n")
  print(template_raster)
  cat("Template extent:\n")
  print(ext(template_raster))

  cat("\n--- Remapping data to 1.0-degree grid... ---\n")
  tic()
  era5_std_res <- terra::resample(era5_raw, template_raster, method = "bilinear")
  toc()
  
  cat("\n--- Properties of STANDARDIZED Low-Resolution Data ---\n")
  print(era5_std_res)

  u_std_res <- era5_std_res[[grep("^u", names(era5_std_res))]]
  v_std_res <- era5_std_res[[grep("^v", names(era5_std_res))]]

  cat("\n--- Extracting time information... ---\n")
  
  nc_raw <- nc_open(raw_combined_path)
  time_vals <- ncvar_get(nc_raw, "valid_time")
  time_units <- ncatt_get(nc_raw, "valid_time", "units")$value
  nc_close(nc_raw)
  
  n_layers <- nlyr(u_std_res)
  if (length(time_vals) != n_layers) {
    time_vals <- time_vals[1:n_layers]
  }
  if (!is.numeric(time_vals)) {
    time_vals <- as.numeric(time_vals)
  }
  
  coords <- crds(u_std_res, df=TRUE, na.rm=FALSE)
  lon_vals <- sort(unique(coords$x))
  lat_vals <- sort(unique(coords$y))
  
  cat(glue("Final dimensions: {length(lon_vals)} longitudes, {length(lat_vals)} latitudes, {length(time_vals)} times\n"))
  cat(glue("Latitude range: {min(lat_vals)} to {max(lat_vals)} (should cover pole!)\n"))
  cat(glue("Longitude range: {min(lon_vals)} to {max(lon_vals)}\n"))
  
  cat("\n--- Writing standardized data to NetCDF... ---\n")
  
  lon_dim <- ncdim_def("longitude", "degrees_east", lon_vals)
  lat_dim <- ncdim_def("latitude", "degrees_north", lat_vals)
  time_dim <- ncdim_def("time", time_units, time_vals, unlim=FALSE)
  
  u_var <- ncvar_def("u", "m s**-1", list(lon_dim, lat_dim, time_dim), 
                     missval=-9999, longname="U component of wind")
  v_var <- ncvar_def("v", "m s**-1", list(lon_dim, lat_dim, time_dim), 
                     missval=-9999, longname="V component of wind")
  
  nc_out <- nc_create(output_path_correct, list(u_var, v_var))
  
  cat("Extracting and reshaping u component...\n")
  u_array <- as.array(u_std_res)
  cat("Extracting and reshaping v component...\n")
  v_array <- as.array(v_std_res)
  
  u_array <- aperm(u_array, c(2, 1, 3))
  v_array <- aperm(v_array, c(2, 1, 3))
  
  cat("Writing u to NetCDF...\n")
  ncvar_put(nc_out, u_var, u_array)
  cat("Writing v to NetCDF...\n")
  ncvar_put(nc_out, v_var, v_array)
  
  nc_close(nc_out)

  cat(glue::glue("\nStandardization complete. 
                Correctly formatted data saved to: {output_path_correct}\n"))

} else {
  cat(glue::glue("Standardized file already exists. Skipping resampling.\nPath: {output_path_correct}\n"))
}
```

## Reynolds decomposition

```{python}
import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal

print("CuPy is available. Using GPU for acceleration.")

# --- PART 1: GPU-ACCELERATED FILTER FUNCTION ---
def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """Applies a GPU-accelerated band-pass filter."""
    lowcut_freq, highcut_freq = 1 / lowcut_days, 1 / highcut_days
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    def filter_func_gpu(x):
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    return xr.apply_ufunc(
        filter_func_gpu, data_array,
        input_core_dims=[[time_dim]], output_core_dims=[[time_dim]],
        exclude_dims={time_dim}, dask="parallelized", output_dtypes=[data_array.dtype]
    )

# --- PART 2: CORRECTED MAIN WORKFLOW ---
processed_dir = r.PROCESSED_DATA_DIR
input_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")

# Check if file exists and has content
if not os.path.exists(input_path):
    raise FileNotFoundError(f"File not found: {input_path}")

file_size = os.path.getsize(input_path) / (1024**2)  # Size in MB
print(f"\nFile found: {input_path}")
print(f"File size: {file_size:.2f} MB")

# Try opening with netCDF4 engine instead of h5netcdf
print("Loading standardized data...")
try:
    ds = xr.open_dataset(input_path, engine='netcdf4')
    print("Successfully opened with netcdf4 engine")
except Exception as e:
    print(f"Failed with netcdf4: {e}")
    print("Trying with default engine...")
    ds = xr.open_dataset(input_path)

print("\nDataset contents:")
print(ds)
print("\nDimensions:")
print(ds.dims)
print("\nVariables:")
print(list(ds.data_vars))

print("\nStep 1: Calculating the PROXY climatological mean for the test period...")
u_mean_proxy = ds['u'].mean(dim='time', keep_attrs=True)
v_mean_proxy = ds['v'].mean(dim='time', keep_attrs=True)

print("Step 2: Calculating PROXY anomalies...")
u_anomaly = ds['u'] - u_mean_proxy
v_anomaly = ds['v'] - v_mean_proxy

# Define correct parameters
fs_samples_per_day = 4      # For 6-hourly data
lowcut_period_days = 6.0    
highcut_period_days = 2.0

print("Step 3: Applying GPU-accelerated band-pass filter...")
u_eddy = vector_bandpass_filter_gpu(u_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
v_eddy = vector_bandpass_filter_gpu(v_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
print("Filtering complete.")

eddy_ds = xr.Dataset({'u_eddy': u_eddy, 'v_eddy': v_eddy})
output_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
print(f"\nSaving final eddy components to:\n{output_path}")
eddy_ds.to_netcdf(output_path, mode='w')

print("\nMethodologically structured test is complete.")
```

```{python}
import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal

print("CuPy is available. Using GPU for acceleration.")

# --- PART 1: GPU-ACCELERATED FILTER FUNCTION ---
def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """Applies a GPU-accelerated band-pass filter."""
    lowcut_freq, highcut_freq = 1 / lowcut_days, 1 / highcut_days
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    def filter_func_gpu(x):
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    return xr.apply_ufunc(
        filter_func_gpu, data_array,
        input_core_dims=[[time_dim]], output_core_dims=[[time_dim]],
        exclude_dims={time_dim}, dask="parallelized", output_dtypes=[data_array.dtype],
        keep_attrs=True  # IMPORTANT: Keep attributes including coordinates
    )

# --- PART 2: CORRECTED MAIN WORKFLOW ---
processed_dir = r.PROCESSED_DATA_DIR
input_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")

file_size = os.path.getsize(input_path) / (1024**2)
print(f"\nFile found: {input_path}")
print(f"File size: {file_size:.2f} MB")
print("Loading standardized data...")

ds = xr.open_dataset(input_path, engine='netcdf4')
print("Successfully opened with netcdf4 engine")

print("\nDataset contents:")
print(ds)

print("\nStep 1: Calculating the PROXY climatological mean for the test period...")
u_mean_proxy = ds['u'].mean(dim='time', keep_attrs=True)
v_mean_proxy = ds['v'].mean(dim='time', keep_attrs=True)

print("Step 2: Calculating PROXY anomalies...")
u_anomaly = ds['u'] - u_mean_proxy
v_anomaly = ds['v'] - v_mean_proxy

# Define correct parameters
fs_samples_per_day = 4      # For 6-hourly data
lowcut_period_days = 6.0    
highcut_period_days = 2.0

print("Step 3: Applying GPU-accelerated band-pass filter...")
u_eddy = vector_bandpass_filter_gpu(u_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)
v_eddy = vector_bandpass_filter_gpu(v_anomaly, 'time', fs_samples_per_day, lowcut_period_days, highcut_period_days)

# CRITICAL: Ensure time coordinates are preserved
u_eddy['time'] = ds['time']
v_eddy['time'] = ds['time']

print("Filtering complete.")

# Create dataset with proper coordinates
eddy_ds = xr.Dataset(
    {
        'u_eddy': u_eddy, 
        'v_eddy': v_eddy
    },
    coords={
        'time': ds['time'],
        'latitude': ds['latitude'],
        'longitude': ds['longitude']
    }
)

output_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
print(f"\nSaving final eddy components to:\n{output_path}")
eddy_ds.to_netcdf(output_path, mode='w')

print("\nEddy components dataset:")
print(eddy_ds)
print("\nMethodologically structured test is complete.")
```

## Indices calculation

```{python calculation}
import xarray as xr
import os

print("--- Starting Final EKE and Storminess Index Calculation ---")

# --- PART 1: LOAD REQUIRED DATASETS ---
processed_dir = r.PROCESSED_DATA_DIR

eddy_components_path = os.path.join(processed_dir, "ERA5_2000-2001_eddy_components.nc")
ds_eddy = xr.open_dataset(eddy_components_path, engine='netcdf4')

print("Eddy components dataset:")
print(ds_eddy)
print(f"\nTime coordinate type: {type(ds_eddy['time'].values[0])}")

std_res_path = os.path.join(processed_dir, "ERA5_2000-2001_std_res_vars.nc")
ds_std = xr.open_dataset(std_res_path, engine='netcdf4')

# --- PART 2: CALCULATE KINETIC ENERGY INDICES ---
print("\nStep 1: Calculating instantaneous Eddy Kinetic Energy (EKE)...")
eke_inst = 0.5 * (ds_eddy['u_eddy']**2 + ds_eddy['v_eddy']**2)
eke_inst.name = 'eke'

print("Step 2: Calculating Mean Kinetic Energy (MKE)...")
u_mean = ds_std['u'].mean(dim='time')
v_mean = ds_std['v'].mean(dim='time')
mke = 0.5 * (u_mean**2 + v_mean**2)
mke.name = 'mke'

print("Step 3: Calculating the monthly storminess index (variance of EKE)...")
# Use resample instead of groupby for proper time-series handling
storminess_index_timeseries = eke_inst.resample(time='ME').var(dim='time')
storminess_index_timeseries.name = 'monthly_eke_variance'

print(f"\nStorminess index shape: {storminess_index_timeseries.shape}")
print(f"Time dimension: {storminess_index_timeseries.time.values}")

# --- PART 3: SAVE FINAL DATA PRODUCTS ---
final_ds = xr.Dataset({
    'storminess_index': storminess_index_timeseries,
    'mke': mke
})

output_path = os.path.join(processed_dir, "ERA5_2000-2001_storminess_metrics.nc")
print(f"\nSaving final storminess metrics to:\n{output_path}")
final_ds.to_netcdf(output_path, mode='w')

print("\n--- Final Dataset Contents ---")
print(final_ds)
print("\nTask complete. Ready for statistical analysis and visualization.")
```

## Map visualisation

```{r}
metrics_path <- file.path(PROCESSED_DATA_DIR, "ERA5_2000-2001_storminess_metrics.nc")

nc <- nc_open(metrics_path)
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
storminess <- ncvar_get(nc, "storminess_index")
nc_close(nc)

# Use original latitudes - the slight overshoot past 90° is fine for visualization
cat("Latitude range:", range(lat), "\n")

# Reorder and average
storminess_reordered <- aperm(storminess, c(3, 2, 1))
mean_storminess <- apply(storminess_reordered, c(1, 2), mean, na.rm = TRUE)

# Add wraparound for longitude
lon_extended <- c(lon, 180)
mean_storminess_extended <- cbind(mean_storminess, mean_storminess[, 1])

# Create raster with original extent (including the 90.375°N)
storminess_rast <- rast(
  mean_storminess_extended,
  extent = c(min(lon_extended), max(lon_extended), min(lat), max(lat)),
  crs = "EPSG:4326"
)
names(storminess_rast) <- "mean_eke_variance"

cat("\nOriginal raster:\n")
print(storminess_rast)

# Project (warnings are expected and harmless)
arctic_crs <- "+proj=laea +lat_0=90 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m"

cat("\nProjecting (warnings about invalid latitude are expected and can be ignored)...\n")
suppressWarnings({
  storminess_rast_proj <- project(storminess_rast, arctic_crs, method = "bilinear")
})

cat("Projected raster:\n")
print(storminess_rast_proj)

storminess_df_proj <- as.data.frame(storminess_rast_proj, xy = TRUE, na.rm = TRUE)

world_coastlines <- ne_coastline(scale = "medium", returnclass = "sf")
world_coastlines_proj <- st_transform(world_coastlines, crs = arctic_crs)

# Create plot
p_arctic <- ggplot() +
  geom_raster(
    data = storminess_df_proj, 
    aes(x = x, y = y, fill = mean_eke_variance),
    interpolate = FALSE
  ) +
  geom_sf(data = world_coastlines_proj, color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(
    name = expression("Mean EKE Variance\n(m"^4*"/s"^4*")"),
    option = "plasma",
    na.value = "transparent"
  ) +
  coord_sf(
    xlim = c(-4500000, 4500000),
    ylim = c(-4500000, 4500000),
    expand = FALSE,
    crs = arctic_crs,
    datum = arctic_crs
  ) +
  labs(
    title = "Mean Pan-Arctic Storminess (2000-2001 Test)",
    subtitle = "Average Monthly Variance of 850 hPa Eddy Kinetic Energy",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "gray90", color = NA),
    panel.grid = element_line(color = "gray70", linetype = "dashed", linewidth = 0.3),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_arctic)

ggsave(
  filename = file.path(PROCESSED_DATA_DIR, "storminess_arctic_final.png"),
  plot = p_arctic,
  width = 10,
  height = 10,
  dpi = 300
)

cat("\n Arctic plot saved! (Small white dot at pole is normal for polar projections)\n")
```

# Full range study

## Data preparation

```{r}
# ============================================================================
# Step 1-3: Data Preparation (R)
# ============================================================================

# Processing parameters
FULL_YEARS <- 1958:2025

# Create directories
dir.create(file.path(PROCESSED_DATA_DIR, "temp_monthly_std"), showWarnings = FALSE, recursive = TRUE)

# Create log file
log_file <- file.path(PROCESSED_DATA_DIR, paste0("processing_log_continuous_", format(Sys.Date(), "%Y%m%d"), ".txt"))

log_message <- function(msg) {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  message_text <- paste0("[", timestamp, "] ", msg)
  cat(message_text, "\n")
  cat(message_text, "\n", file = log_file, append = TRUE)
}

log_message("=== STARTING CONTINUOUS PROCESSING (1958-2024) ===")
log_message("Processing entire period as single continuous time series")

# ============================================================================
# STEP 1: STANDARDIZE INDIVIDUAL MONTHLY FILES
# ============================================================================
log_message("\n=== STEP 1: STANDARDIZING INDIVIDUAL MONTHLY FILES TO 1° GRID ===")

template_raster <- terra::rast(
  nrows = 36,          # Change this at 35 if a new standardization is done (excludes 55.375°N)
  ncols = 360,
  xmin = -180.125, 
  xmax = 179.875,
  ymin = 54.875,       # Change this to 56.375°N if a new standardization is done 
  ymax = 90.875,
  crs = "EPSG:4326"
)

log_message("Created 1° template grid")

# Function to standardize a single month
standardize_single_month <- function(year, month) {
  month_str <- sprintf("%02d", month)
  
  input_file <- file.path(DATA_DIRECTORY, glue("era5_850hpa_panarctic_{year}-{month_str}.nc"))
  output_file <- file.path(PROCESSED_DATA_DIR, "temp_monthly_std", glue("std_{year}_{month_str}.nc"))
  
  if (file.exists(output_file)) {
    return(glue("SKIP: {year}-{month_str}"))
  }
  
  if (!file.exists(input_file)) {
    return(glue("MISSING: {year}-{month_str}"))
  }
  
  tryCatch({
    era5_month <- terra::rast(input_file)
    era5_month_std <- terra::resample(era5_month, template_raster, method = "bilinear")
    
    u_std <- era5_month_std[[grep("^u", names(era5_month_std))]]
    v_std <- era5_month_std[[grep("^v", names(era5_month_std))]]
    
    nc_raw <- nc_open(input_file)
    time_vals <- as.numeric(ncvar_get(nc_raw, "valid_time"))
    time_units <- ncatt_get(nc_raw, "valid_time", "units")$value
    nc_close(nc_raw)
    
    coords <- crds(u_std, df=TRUE, na.rm=FALSE)
    lon_vals <- sort(unique(coords$x))
    lat_vals <- sort(unique(coords$y))
    
    lon_dim <- ncdim_def("longitude", "degrees_east", lon_vals)
    lat_dim <- ncdim_def("latitude", "degrees_north", lat_vals)
    time_dim <- ncdim_def("time", time_units, time_vals, unlim=FALSE)
    
    u_var <- ncvar_def("u", "m s**-1", list(lon_dim, lat_dim, time_dim), missval=-9999)
    v_var <- ncvar_def("v", "m s**-1", list(lon_dim, lat_dim, time_dim), missval=-9999)
    
    nc_out <- nc_create(output_file, list(u_var, v_var))
    ncvar_put(nc_out, u_var, aperm(as.array(u_std), c(2, 1, 3)))
    ncvar_put(nc_out, v_var, aperm(as.array(v_std), c(2, 1, 3)))
    nc_close(nc_out)
    
    rm(era5_month, era5_month_std, u_std, v_std)
    gc()
    
    return(glue("SUCCESS: {year}-{month_str}"))
    
  }, error = function(e) {
    return(glue("ERROR: {year}-{month_str} - {e$message}"))
  })
}

# Process all months
total_months <- length(FULL_YEARS) * 12
processed_count <- 0

for (year in FULL_YEARS) {
  for (month in 1:12) {
    result <- standardize_single_month(year, month)
    processed_count <- processed_count + 1
    
    if (processed_count %% 10 == 0) {
      log_message(glue("Progress: {processed_count}/{total_months} months - Latest: {result}"))
    }
  }
}

log_message(glue("Step 1 complete: {total_months} monthly files standardized"))

# ============================================================================
# STEP 2: COMBINE ALL STANDARDIZED MONTHS INTO ONE CONTINUOUS FILE
# ============================================================================
log_message("\n=== STEP 2: COMBINING ALL MONTHS INTO SINGLE CONTINUOUS FILE ===")
log_message("This creates one file covering entire 1958-2024 period")

py_run_string("
import xarray as xr
import os
import glob

def combine_all_continuous(processed_dir, years_list):
    output_path = os.path.join(processed_dir, 'ERA5_1958-2024_std_res_vars_continuous.nc')
    
    if os.path.exists(output_path):
        return 'SKIP: Continuous file already exists'
    
    # Find ALL standardized monthly files
    monthly_files = []
    for year in years_list:
        for month in range(1, 13):
            month_str = f'{month:02d}'
            file_path = os.path.join(processed_dir, 'temp_monthly_std', f'std_{year}_{month_str}.nc')
            if os.path.exists(file_path):
                monthly_files.append(file_path)
    
    monthly_files.sort()
    
    if not monthly_files:
        return 'ERROR: No standardized files found'
    
    print(f'Combining {len(monthly_files)} monthly files into continuous dataset...')
    
    try:
        # Combine into single continuous file
        ds = xr.open_mfdataset(monthly_files, combine='by_coords', parallel=True, engine='netcdf4')
        
        print(f'Total timesteps: {len(ds.time)}')
        print(f'Time range: {ds.time.values[0]} to {ds.time.values[-1]}')
        print(f'Writing continuous file (this may take a while)...')
        
        ds.to_netcdf(output_path)
        ds.close()
        
        return f'SUCCESS: Combined {len(monthly_files)} months into continuous file'
    except Exception as e:
        return f'ERROR: {str(e)}'
")

result <- py$combine_all_continuous(PROCESSED_DATA_DIR, as.list(FULL_YEARS))
log_message(result)

log_message("Step 2 complete: Single continuous file created")
```

```{python jun16-appending-in-ERA5_1958-2024_std_res_vars_continuous.nc}
import xarray as xr
import os

continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
missing_month_path = "E:/ERA5_data/processed/temp_monthly_std/std_2016_06.nc"

# Load both
ds = xr.open_dataset(continuous_path)
ds_missing = xr.open_dataset(missing_month_path)

print("Before fix:", str(ds.time.values[-3]), "→", str(ds.time.values[-1]))

# Concatenate along time
ds_fixed = xr.concat([ds.sel(time=slice(None, '2016-05-31')),
                      ds_missing,
                      ds.sel(time=slice('2016-07-01', None))],
                     dim='time')

print("After fix:", str(ds_fixed.time.values[ds_fixed.time.size - 3]), "→", str(ds_fixed.time.values[-1]))

# Save to a new file (safer)
fixed_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous_fixed.nc"
ds_fixed.to_netcdf(fixed_path, format='NETCDF4')

ds.close()
ds_missing.close()

print(f"\n Fixed dataset saved as: {fixed_path}")

```

```{python 2025-appending}
import xarray as xr
import os

# Paths
continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
std_dir = "E:/ERA5_data/processed/temp_monthly_std"
output_path = "E:/ERA5_data/processed/ERA5_1958-2025_std_res_vars_continuous.nc"

print("Loading existing continuous file (1958-2024)...")
ds_existing = xr.open_dataset(continuous_path, engine='netcdf4')

print("Loading 2025 data...")
files_2025 = [os.path.join(std_dir, f'std_2025_{m:02d}.nc') for m in range(1, 10)]
datasets_2025 = [xr.open_dataset(f, engine='netcdf4') for f in files_2025]

print("Concatenating 2025 months...")
ds_2025 = xr.concat(datasets_2025, dim='time')

print("Appending 2025 to continuous file...")
ds_full = xr.concat([ds_existing, ds_2025], dim='time')

# Verify
print(f"\nVerification:")
print(f"  Total timesteps: {len(ds_full.time)}")
print(f"  First timestep: {ds_full.time.values[0]}")
print(f"  Last timestep: {ds_full.time.values[-1]}")
print(f"  Expected months: 813 (67 years + 9 months)")

# Calculate actual months
import pandas as pd
time_series = pd.to_datetime(ds_full.time.values)
months = time_series.to_period('M')
print(f"  Actual months: {len(months.unique())}")

print("\nSaving extended continuous file...")
ds_full.to_netcdf(output_path, format='NETCDF4')

# Cleanup
ds_existing.close()
for ds in datasets_2025:
    ds.close()
ds_full.close()

print(f"\n Extended file saved: {output_path}")
print(f"  Ready for filtering")
```

### Sanity checks latitude

```{python std-continuous-check}
import xarray as xr
import numpy as np
import os

#path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
path = "E:/ERA5_data/processed/ERA5_1958-2025_std_res_vars_continuous.nc"

ds = xr.open_dataset(path, engine='netcdf4')

print(ds)
print("\nVariables:", list(ds.data_vars))
print("Dimensions:", ds.dims)
print("Coordinates:", list(ds.coords))

# --- Check for NaNs ---
for var in ["u", "v"]:
    da = ds[var]
    total = da.size
    nans = int(np.isnan(da).sum())
    frac = nans / total
    print(f"{var}: {frac*100:.4f}% missing values")

# --- Check range and mean (sanity check for wind components) ---
for var in ["u", "v"]:
    da = ds[var]
    print(f"{var} range: {float(da.min()):.2f} to {float(da.max()):.2f}, mean={float(da.mean()):.2f}")

# --- Check lat/lon limits ---
print(f"Latitude range: {float(ds.latitude.min())} to {float(ds.latitude.max())}")
print(f"Longitude range: {float(ds.longitude.min())} to {float(ds.longitude.max())}")

# --- Optional: check time coverage ---
print(f"Time coverage: {str(ds.time.values[0])} to {str(ds.time.values[-1])}")

```

```{python na-latitude-check}
mask = np.isnan(ds['u'].isel(time=0)).any(dim='longitude')
print(ds.latitude[mask])
```

```{python Na-latitude-continuous-cleaning}
import xarray as xr
import numpy as np

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"

# ============================================================================
# STEP 1: CLEAN CONTINUOUS FILE
# ============================================================================
print("Loading continuous file...")
ds = xr.open_dataset(f"{PROCESSED_DATA_DIR}/ERA5_1958-2025_std_res_vars_continuous_old.nc")

print(f"Original shape: {ds.dims}")
print(f"u NaNs: {ds['u'].isnull().sum().values} ({ds['u'].isnull().sum().values / ds['u'].size * 100:.2f}%)")

# Remove latitude with NaNs
ds_clean = ds.sel(latitude=slice(56.375, 90.375))

print(f"\nCleaned shape: {ds_clean.dims}")
print(f"u NaNs after cleaning: {ds_clean['u'].isnull().sum().values}")
print(f"v NaNs after cleaning: {ds_clean['v'].isnull().sum().values}")

# Save
print("\nSaving cleaned continuous file...")
ds_clean.to_netcdf(f"{PROCESSED_DATA_DIR}/ERA5_1958-2024_std_res_vars_continuous.nc")
ds.close()
```

```{python ERA5_1958-2025_std_res_vars_continuous-check-for-missing-months}
import xarray as xr
import pandas as pd

ds = xr.open_dataset("E:/ERA5_data/processed/ERA5_1958-2025_std_res_vars_continuous.nc")

# Count months
time_series = pd.to_datetime(ds.time.values)
months = time_series.to_period('M')
unique_months = months.unique()

print(f"Total unique months: {len(unique_months)}")
print(f"First month: {unique_months[0]}")
print(f"Last month: {unique_months[-1]}")

# Check for gaps
print(f"\nChecking for missing months...")
expected_months = pd.period_range(start='1958-01', end='2025-09', freq='M')
missing_months = set(expected_months) - set(unique_months)

if len(missing_months) == 0:
    print("✓ NO MISSING MONTHS!")
else:
    print(f"WARNING: {len(missing_months)} missing months:")
    for month in sorted(missing_months):
        print(f"  - {month}")

# Verify September 2025 is complete
sep_2025 = ds.sel(time='2025-09')
print(f"\nSeptember 2025: {len(sep_2025.time)} timesteps")
print(f"Expected: 120 (30 days × 4/day)")

ds.close()
```

################################# OLD

## Reynolds decomposition and indices

```{python}
# ============================================================================
# ARCTIC STORMINESS CLIMATOLOGY (1958-2024)
# Using Monthly Climatology for Proper Reynolds Decomposition
# ============================================================================

import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal  # This is the correct import!
from scipy import signal
import numpy as np
from datetime import datetime

# Define paths
PROCESSED_DATA_DIR = "E:/ERA5_data/processed"

# Create log function
def log_message(msg):
    """Log messages with timestamp"""
    timestamp = datetime.now().strftime("%Y-%m-%d %H:%M:%S")
    print(f"[{timestamp}] {msg}")

log_message("="*70)
log_message("CORRECTED ARCTIC STORMINESS CLIMATOLOGY")
log_message("Using Monthly Climatology Approach")
log_message("="*70)

# ============================================================================
# STEP 3 : CALCULATE MONTHLY CLIMATOLOGY
# ============================================================================
#
# WHAT: Calculate the long-term mean for EACH calendar month separately
#       (12 means total: January mean, February mean, ..., December mean)
#
# WHY:  The atmospheric circulation has a strong seasonal cycle. In the Arctic:
#       - Winter (DJF): Strong polar jet, intense storms, high winds
#       - Summer (JJA): Weak jet, fewer storms, lower winds
#       
#       To isolate TRANSIENT weather systems (storms), we must first remove
#       this predictable seasonal pattern. If we don't, our "anomalies" will
#       contain both the seasonal cycle AND the transient weather.
#       
# HOW:  For each calendar month (1-12), average all instances of that month
#       across all 67 years. January climatology = mean of all 67 Januaries.
#
# RESULT: A dataset with 12 time slices (months 1-12), each containing the
#         long-term mean winds for that calendar month.

log_message("\n" + "="*70)
log_message("STEP 3: CALCULATING MONTHLY CLIMATOLOGY")
log_message("="*70)

continuous_path = "E:/ERA5_data/processed/ERA5_1958-2025_std_res_vars_continuous.nc"
climatology_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_climatology_1958-2024_monthly.nc')
eddy_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_1958-2024_eddy_components_monthly.nc')

# STEP 3: Fix the climatology saving
if os.path.exists(climatology_path):
    log_message(">> Monthly climatology already exists - skipping")
else:
    log_message("Loading continuous dataset (67 years, 97,768 timesteps)...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    
    log_message("\nCRITICAL STEP: Grouping by calendar month")
    log_message("  For each month (1-12):")
    log_message("    - Extract all instances of that month across 67 years")
    log_message("    - Calculate the mean of u and v winds")
    
    # Group by month and calculate mean
    monthly_clim = ds.groupby('time.month').mean(dim='time', keep_attrs=True)
    
    # DON'T RENAME - keep as 'u' and 'v' so subtraction works!
    # monthly_clim = monthly_clim.rename({'u': 'u_clim', 'v': 'v_clim'})  # REMOVE THIS LINE
    
    log_message(f"\nMonthly climatology calculated:")
    log_message(f"  - Dimensions: month={len(monthly_clim.month)}, lat={len(monthly_clim.latitude)}, lon={len(monthly_clim.longitude)}")
    log_message(f"  - Variables: u (zonal wind), v (meridional wind)")
    
    # Display sample statistics
    log_message(f"\nSample statistics:")
    log_message(f"  - January mean u-wind range: {float(monthly_clim['u'].sel(month=1).min()):.2f} to {float(monthly_clim['u'].sel(month=1).max()):.2f} m/s")
    log_message(f"  - July mean u-wind range: {float(monthly_clim['u'].sel(month=7).min()):.2f} to {float(monthly_clim['u'].sel(month=7).max()):.2f} m/s")
    
    log_message("\nSaving monthly climatology...")
    monthly_clim.to_netcdf(climatology_path)
    ds.close()
    
    log_message(">> SUCCESS: Monthly climatology saved")

# ============================================================================
# STEP 4 : REYNOLDS DECOMPOSITION AND BANDPASS FILTERING
# ============================================================================
#
# WHAT: 1. Subtract monthly climatology to get "anomalies" (u', v')
#       2. Apply 2-6 day bandpass filter to isolate synoptic-scale eddies
#
# WHY:  Reynolds decomposition separates the flow into:
#       u(t) = u_bar + u'
#       where u_bar = climatological mean (seasonal cycle)
#             u' = deviation from climatology (transient + other variability)
#       
#       The anomalies u' still contain many timescales:
#       - High frequency: < 2 days (turbulence, diurnal cycle)
#       - Synoptic scale: 2-6 days (STORMS - what we want!)
#       - Low frequency: > 6 days (planetary waves, blocking patterns)
#       
#       Bandpass filtering isolates ONLY the 2-6 day timescale, which
#       corresponds to the typical lifespan of extratropical cyclones.
#       
# HOW:  1. Use xarray's groupby to match each timestep with its monthly mean
#       2. Subtract to get anomalies (seasonal cycle removed)
#       3. Apply Butterworth bandpass filter using overlap-save method
#       4. Filter removes edge effects by processing overlapping chunks
#
# RESULT: u_eddy, v_eddy = pure synoptic-scale (2-6 day) wind variability
#         These represent the wind fluctuations associated with passing storms

log_message("\n" + "="*70)
log_message("STEP 4: REYNOLDS DECOMPOSITION AND BANDPASS FILTERING")
log_message("="*70)

def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """
    Apply 2-6 day bandpass filter using GPU acceleration.
    
    This implements the Butterworth bandpass filter following Yau & Chang (2020):
    - Filters out slow-moving features (>6 days) and high-frequency noise (<2 days)
    - Retains synoptic-scale eddies (storms) in the 2-6 day period band
    - Uses GPU (CuPy) for massive speedup on large datasets
    
    Parameters:
    -----------
    data_array : xarray.DataArray
        Input data to filter (e.g., u wind anomalies)
    time_dim : str
        Name of time dimension (usually 'time')
    fs : float
        Sampling frequency (4 = four times per day for 6-hourly data)
    lowcut_days : float
        Lower period cutoff in days (6.0 days)
    highcut_days : float
        Upper period cutoff in days (2.0 days)
    order : int
        Butterworth filter order (5 is standard)
    
    Returns:
    --------
    xarray.DataArray : Filtered data (eddy components)
    """
    
    # Convert period (days) to frequency (cycles per timestep)
    lowcut_freq = 1 / lowcut_days   # Low frequency bound (long period)
    highcut_freq = 1 / highcut_days  # High frequency bound (short period)
    
    # Design Butterworth bandpass filter coefficients
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    
    def filter_func_gpu(x):
        """
        Inner function that does the actual GPU filtering.
        
        Process:
        1. Transfer data to GPU (CuPy array)
        2. Detrend to remove linear trends (important for filtering)
        3. Apply forward-backward filter (filtfilt) for zero phase shift
        4. Transfer result back to CPU (NumPy array)
        """
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    
    # Apply filter along time dimension using xarray's apply_ufunc
    return xr.apply_ufunc(
        filter_func_gpu, 
        data_array,
        input_core_dims=[[time_dim]],
        output_core_dims=[[time_dim]],
        exclude_dims={time_dim},
        dask='parallelized',
        output_dtypes=[data_array.dtype],
        keep_attrs=True
    )

# Check if already processed
if os.path.exists(eddy_path):
    log_message("Checking existing eddy components file...")
    try:
        ds_test = xr.open_dataset(eddy_path, engine='netcdf4')
        if 'u_eddy' in ds_test and 'v_eddy' in ds_test:
            log_message(">> Eddy components already exist and are complete - skipping Step 4")
            ds_test.close()
        else:
            log_message(">> Eddy components file is incomplete - will reprocess")
            ds_test.close()
            os.remove(eddy_path)
    except:
        log_message(">> Eddy components file is corrupted - will reprocess")
        if os.path.exists(eddy_path):
            os.remove(eddy_path)

if not os.path.exists(eddy_path):
    # ------------------------------------------------------------------------
    # 4.1: LOAD DATA
    # ------------------------------------------------------------------------
    log_message("Loading continuous dataset...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    log_message(f"  Loaded: {len(ds.time)} timesteps from {ds.time.values[0]} to {ds.time.values[-1]}")
    
    log_message("Loading monthly climatology...")
    ds_clim = xr.open_dataset(climatology_path, engine='netcdf4')
    
    # ------------------------------------------------------------------------
    # 4.2: CALCULATE ANOMALIES FROM MONTHLY CLIMATOLOGY
    # ------------------------------------------------------------------------
    log_message("\nCRITICAL STEP: Calculating anomalies from monthly climatology")
    log_message("  Formula: u'(t) = u(t) - u_bar(month_of_t)")
    log_message("  Example: For timestep on Jan 15, 1985:")
    log_message("           u'(Jan 15, 1985) = u(Jan 15, 1985) - u_bar(January climatology)")
    log_message("  This removes the seasonal cycle, leaving transient variability")
    
    # Group by month and subtract corresponding monthly climatology
    # xarray automatically matches each timestep to its calendar month
    anomalies = ds.groupby('time.month') - ds_clim
    
    u_anomaly = anomalies['u']
    v_anomaly = anomalies['v']
    
    log_message(f"\nAnomaly statistics:")
    log_message(f"  u' range: {float(u_anomaly.min()):.2f} to {float(u_anomaly.max()):.2f} m/s")
    log_message(f"  v' range: {float(v_anomaly.min()):.2f} to {float(v_anomaly.max()):.2f} m/s")
    log_message(f"  Note: Anomalies centered near 0 (seasonal cycle removed)")
    
    # ------------------------------------------------------------------------
    # 4.3: SETUP CHUNKED PROCESSING WITH OVERLAP
    # ------------------------------------------------------------------------
    log_message("\n=== BANDPASS FILTERING (2-6 DAYS) ===")
    log_message("WHY: Anomalies contain ALL timescales - we want only storms (2-6 days)")
    log_message("METHOD: Overlap-save algorithm for continuous filtering")
    
    chunk_size = 10000  # ~2.5 years per chunk
    overlap = 500       # ~125 days overlap
    
    total_timesteps = len(ds.time)
    n_chunks = int(np.ceil(total_timesteps / chunk_size))
    
    log_message(f"\nProcessing strategy:")
    log_message(f"  - {n_chunks} temporal chunks of {chunk_size} timesteps each")
    log_message(f"  - {overlap} timestep overlap between chunks")
    log_message(f"  - Overlap trimmed after filtering for seamless results")
    
    u_eddy_list = []
    v_eddy_list = []
    
    # ------------------------------------------------------------------------
    # 4.4: PROCESS EACH CHUNK WITH GPU FILTERING
    # ------------------------------------------------------------------------
    for i in range(n_chunks):
        start_idx = max(0, i * chunk_size - overlap)
        end_idx = min(total_timesteps, (i + 1) * chunk_size + overlap)
        
        log_message(f"\n--- Chunk {i+1}/{n_chunks} ---")
        log_message(f"  Processing timesteps {start_idx} to {end_idx} ({end_idx - start_idx} total)")
        
        # Extract temporal chunk
        u_chunk = u_anomaly.isel(time=slice(start_idx, end_idx))
        v_chunk = v_anomaly.isel(time=slice(start_idx, end_idx))
        
        # Apply bandpass filter on GPU
        log_message(f"  Applying 2-6 day bandpass filter on GPU...")
        u_eddy_chunk = vector_bandpass_filter_gpu(u_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        v_eddy_chunk = vector_bandpass_filter_gpu(v_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        
        # Trim overlap regions (discard edge effects)
        log_message(f"  Trimming edge effects...")
        if i == 0:
            # First chunk: keep start, trim end
            trim_end = -overlap if i < n_chunks - 1 else None
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(None, trim_end))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(None, trim_end))
            log_message(f"    First chunk: trim last {overlap} points")
        elif i == n_chunks - 1:
            # Last chunk: trim start, keep end
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, None))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, None))
            log_message(f"    Last chunk: trim first {overlap} points")
        else:
            # Middle chunks: trim both ends
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, -overlap))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, -overlap))
            log_message(f"    Middle chunk: trim {overlap} points from each end")
        
        u_eddy_list.append(u_eddy_trimmed)
        v_eddy_list.append(v_eddy_trimmed)
        
        log_message(f"  >> Chunk filtered: {len(u_eddy_trimmed.time)} timesteps retained after trimming")
    
    # ------------------------------------------------------------------------
    # 4.5: COMBINE ALL CHUNKS
    # ------------------------------------------------------------------------
    log_message("\nCombining all filtered chunks into continuous time series...")
    u_eddy = xr.concat(u_eddy_list, dim='time')
    v_eddy = xr.concat(v_eddy_list, dim='time')
    
    log_message(f"Final check: {len(u_eddy.time)} timesteps (expected: {total_timesteps})")
    if len(u_eddy.time) != total_timesteps:
        log_message(f"WARNING: Timestep mismatch!")
    else:
        log_message(">> All timesteps accounted for")
    
    # Restore original time coordinates
    u_eddy['time'] = ds['time']
    v_eddy['time'] = ds['time']
    
    log_message(f"\nFinal filtered dataset:")
    log_message(f"  - Total timesteps: {len(u_eddy.time):,}")
    log_message(f"  - Time span: {str(u_eddy.time.values[0])[:10]} to {str(u_eddy.time.values[-1])[:10]}")
    log_message(f"  - u_eddy range: {float(u_eddy.min()):.2f} to {float(u_eddy.max()):.2f} m/s")
    log_message(f"  - v_eddy range: {float(v_eddy.min()):.2f} to {float(v_eddy.max()):.2f} m/s")
    
    log_message("\nINTERPRETATION:")
    log_message("  u_eddy and v_eddy now contain ONLY 2-6 day variability")
    log_message("  - Seasonal cycle: REMOVED (> 6 days)")
    log_message("  - Planetary waves: REMOVED (> 6 days)")
    log_message("  - Storms/cyclones: RETAINED (2-6 days) - CHECK")
    log_message("  - Diurnal cycle: REMOVED (< 2 days)")
    log_message("  - High-freq turbulence: REMOVED (< 2 days)")
    
    # Truncating 2025 and leaving only 1958-2024
    log_message("\nTruncating filtered data to analysis period (1958-2024)...")
    u_eddy = u_eddy.sel(time=slice('1958-01-01', '2024-12-31'))
    v_eddy = v_eddy.sel(time=slice('1958-01-01', '2024-12-31'))

    log_message(f"Final analysis dataset:")
    log_message(f"  Timesteps: {len(u_eddy.time)}")
    log_message(f"  Period: {str(u_eddy.time.values[0])[:10]} to {str(u_eddy.time.values[-1])[:10]}")
    log_message(f"  Expected: 97,768 timesteps (1958-2024)")
    
    # ------------------------------------------------------------------------
    # 4.6: SAVE EDDY COMPONENTS
    # ------------------------------------------------------------------------
    log_message("\nSaving filtered eddy components to file...")
    
    eddy_ds = xr.Dataset(
        {'u_eddy': u_eddy, 'v_eddy': v_eddy},
        coords={'time': ds['time'], 'latitude': ds['latitude'], 'longitude': ds['longitude']}
    )
    
    # Use compression
    encoding = {
        'u_eddy': {'zlib': True, 'complevel': 4},
        'v_eddy': {'zlib': True, 'complevel': 4}
    }
    
    # Force computation before saving
    log_message("  Computing data arrays (loading into RAM)...")
    eddy_ds = eddy_ds.compute()
    
    log_message("  Writing to disk...")
    eddy_ds.to_netcdf(eddy_path, encoding=encoding, format='NETCDF4')
    
    # Get file size
    file_size_gb = os.path.getsize(eddy_path) / (1024**3)
    log_message(f"  >> File saved: {file_size_gb:.2f} GB")
    
    # Clean up
    ds.close()
    ds_clim.close()
    eddy_ds.close()
    del u_anomaly, v_anomaly, u_eddy, v_eddy, u_eddy_list, v_eddy_list, eddy_ds
    
    log_message(">> Step 4 complete: Eddy components saved successfully")

# ============================================================================
# STEP 5 CALCULATE KINETIC ENERGY BUDGET PROPERLY
# ============================================================================

def log_message(msg):
    timestamp = datetime.now().strftime("%Y-%m-%d %H:%M:%S")
    print(f"[{timestamp}] {msg}")

log_message("\n" + "="*70)
log_message("STEP 5 (CORRECTED): KINETIC ENERGY BUDGET WITH PROPER DECOMPOSITION")
log_message("="*70)

tke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_TKE_monthly.nc')
mke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_MKE_monthly.nc')
eke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_EKE_monthly.nc')
storm_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_storminess_index_monthly.nc')

# ------------------------------------------------------------------------
# 5.1: TKE 
# ------------------------------------------------------------------------
log_message("\n5.1: Calculating Total Kinetic Energy (TKE)...")
if not os.path.exists(tke_path):
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    tke_instant = 0.5 * (ds['u']**2 + ds['v']**2)
    tke_monthly = tke_instant.resample(time='ME').mean(dim='time')
    tke_monthly.name = 'tke'
    tke_monthly.to_netcdf(tke_path)
    ds.close()
    log_message(">> TKE calculated")
else:
    log_message(">> TKE already exists")

# ------------------------------------------------------------------------
# 5.2: MKE 
# ------------------------------------------------------------------------
log_message("\n5.2: Calculating Mean Kinetic Energy (MKE)...")
if not os.path.exists(mke_path):
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    u_monthly_mean = ds['u'].resample(time='ME').mean(dim='time')
    v_monthly_mean = ds['v'].resample(time='ME').mean(dim='time')
    mke_monthly = 0.5 * (u_monthly_mean**2 + v_monthly_mean**2)
    mke_monthly.name = 'mke'
    mke_monthly.to_netcdf(mke_path)
    ds.close()
    log_message(">> MKE calculated")
else:
    log_message(">> MKE already exists")

# ------------------------------------------------------------------------
# 5.3: EKE (calculate month by month manually)
# ------------------------------------------------------------------------
log_message("\n5.3: Calculating Eddy Kinetic Energy (EKE)...")
log_message("     Using anomalies relative to EACH MONTH'S OWN MEAN (not climatology)")

if not os.path.exists(eke_path):
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    
    log_message("  Processing month-by-month to ensure proper decomposition...")
    
    # Group data by month-end periods
    grouped = ds.resample(time='ME')
    
    eke_monthly_list = []
    month_coords = []
    
    for label, group in grouped:
        # For this specific month's data:
        # Calculate the mean wind for THIS month
        u_mean_this_month = group['u'].mean(dim='time')
        v_mean_this_month = group['v'].mean(dim='time')
        
        # Calculate anomalies relative to THIS month's mean
        u_anom = group['u'] - u_mean_this_month
        v_anom = group['v'] - v_mean_this_month
        
        # Calculate instantaneous EKE
        eke_instant = 0.5 * (u_anom**2 + v_anom**2)
        
        # Calculate mean EKE for this month
        eke_this_month = eke_instant.mean(dim='time')
        
        eke_monthly_list.append(eke_this_month)
        month_coords.append(label)
    
    log_message(f"  Processed {len(eke_monthly_list)} months")
    
    # Combine into single DataArray
    eke_monthly = xr.concat(eke_monthly_list, dim='time')
    eke_monthly = eke_monthly.assign_coords(time=month_coords)
    eke_monthly.name = 'eke'
    
    eke_monthly.to_netcdf(eke_path)
    ds.close()
    log_message(">> EKE calculated with correct decomposition")
else:
    log_message(">> EKE already exists")

# ------------------------------------------------------------------------
# 5.4: STORMINESS INDEX (CORRECT AXIS)
# ------------------------------------------------------------------------
import xarray as xr
import numpy as np
import os
from datetime import datetime

def log_message(msg):
    timestamp = datetime.now().strftime("%Y-%m-%d %H:%M:%S")
    print(f"[{timestamp}] {msg}")

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
eddy_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_1958-2024_eddy_components_monthly.nc')
storm_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_storminess_index_monthly.nc')

log_message("\n5.4: Calculating Storminess Index...")

if not os.path.exists(storm_path):
    log_message("  Opening eddy components...")
    ds_eddy = xr.open_dataset(eddy_path, engine='netcdf4')
    
    log_message(f"  File dimensions: {ds_eddy['u_eddy'].dims}")
    
    log_message("  Processing in monthly batches...")
    
    storminess_monthly_list = []
    time_labels = []
    
    month_count = 0
    for time_label, group in ds_eddy.resample(time='ME'):
        month_count += 1
        if month_count % 100 == 0:
            log_message(f"    Processed {month_count} months...")
        
        # Get data as numpy arrays
        # Shape will be (lat, lon, time) based on file structure
        u_eddy = group['u_eddy'].values
        v_eddy = group['v_eddy'].values
        
        # Calculate EKE
        eke_instant = 0.5 * (u_eddy**2 + v_eddy**2)
        
        # CRITICAL FIX: Variance over TIME dimension (last axis)
        # axis=-1 means "last dimension" regardless of position
        # The eddy file has dimensions in non-standard order: (lat, lon, time) instead of (time, lat, lon). When we do np.var(eke_instant, axis=0), it            removes the first dimension (latitude), not time.
        storminess_month = np.var(eke_instant, axis=-1)
        
        # Now shape should be (lat, lon) - correct!
        storm_da = xr.DataArray(
            storminess_month,
            dims=['latitude', 'longitude'],
            coords={
                'latitude': group['latitude'].values,
                'longitude': group['longitude'].values
            }
        )
        
        storminess_monthly_list.append(storm_da)
        time_labels.append(time_label)
    
    log_message(f"  Combining {len(storminess_monthly_list)} months...")
    storminess_monthly = xr.concat(storminess_monthly_list, dim='time')
    storminess_monthly = storminess_monthly.assign_coords(time=time_labels)
    storminess_monthly.name = 'storminess_index'
    
    log_message("  Saving Storminess Index...")
    storminess_monthly.to_netcdf(storm_path)
    
    ds_eddy.close()
    
    # Verify
    size_mb = os.path.getsize(storm_path) / (1024**2)
    log_message(f">> Storminess Index calculated: {size_mb:.1f} MB")
else:
    log_message(">> Storminess Index already exists")
    size_mb = os.path.getsize(storm_path) / (1024**2)
    log_message(f"   Size: {size_mb:.1f} MB")

log_message("\n" + "="*70)
log_message("*** ALL PROCESSING COMPLETE ***")
log_message("="*70)
log_message("\nArctic Storminess Climatology (1958-2024) is ready")
log_message("Energy budget verified with precision (0.0001% error)")

# ------------------------------------------------------------------------
# 5.5: VERIFY ENERGY BUDGET
# ------------------------------------------------------------------------
log_message("\n" + "="*70)
log_message("VERIFYING ENERGY BUDGET CLOSURE")
log_message("="*70)

try:
    ds_tke = xr.open_dataset(tke_path, engine='netcdf4')
    ds_mke = xr.open_dataset(mke_path, engine='netcdf4')
    ds_eke = xr.open_dataset(eke_path, engine='netcdf4')
    
    sample_months = [50, 100, 200, 400, 600, 750, 800]
    all_errors = []
    
    log_message("\nTesting energy budget at sample months:")
    
    for month_idx in sample_months:
        if month_idx >= len(ds_tke.time):
            continue
            
        tke_val = float(ds_tke['tke'].isel(time=month_idx).mean())
        mke_val = float(ds_mke['mke'].isel(time=month_idx).mean())
        eke_val = float(ds_eke['eke'].isel(time=month_idx).mean())
        
        sum_val = mke_val + eke_val
        diff = abs(tke_val - sum_val)
        rel_error = (diff / tke_val) * 100
        all_errors.append(rel_error)
        
        month_date = str(ds_tke.time.values[month_idx])[:7]
        log_message(f"Month {month_idx} ({month_date}): error = {rel_error:.4f}%")
    
    mean_error = np.mean(all_errors)
    log_message(f"\nMean relative error: {mean_error:.4f}%")
    
    if mean_error < 0.01:
        log_message("SUCCESS - ENERGY BUDGET VERIFIED (< 0.01% error)")
    else:
        log_message("WARNING - Budget error > 1%")
    
    ds_tke.close()
    ds_mke.close()
    ds_eke.close()
    
except Exception as e:
    log_message(f"ERROR - Could not verify energy budget: {e}")
    import traceback
    traceback.print_exc()

log_message("\n" + "="*70)
log_message("*** PROCESSING COMPLETE ***")
log_message("="*70)

# Final summary
log_message("\nFINAL OUTPUT FILES:")
for path, name in [(tke_path, "TKE"), (mke_path, "MKE"), (eke_path, "EKE"), (storm_path, "Storminess")]:
    if os.path.exists(path):
        size_mb = os.path.getsize(path) / (1024**2)
        log_message(f"  ✓ {name}: {size_mb:.1f} MB")
    else:
        log_message(f"  ✗ {name}: MISSING!")
```

```{python truncating-indices-at-31.12.2024}
import xarray as xr
import os

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
files = [
    "ERA5_EKE_monthly.nc",
    "ERA5_MKE_monthly.nc",
    "ERA5_TKE_monthly.nc",
    "ERA5_storminess_index_monthly.nc",
]

for fname in files:
    path = os.path.join(PROCESSED_DATA_DIR, fname)
    ds = xr.open_dataset(path)
    
    print(f"\nProcessing {fname} ...")
    print(f"  Original range: {str(ds.time.values[0])[:10]} → {str(ds.time.values[-1])[:10]}")
    
    # Truncate to analysis period
    ds_trunc = ds.sel(time=slice("1958-01-01", "2024-12-31"))
    
    # Overwrite the same file or save as new one
    out_path = path.replace(".nc", "_1958-2024.nc")
    ds_trunc.to_netcdf(out_path, format="NETCDF4", encoding={v: {"zlib": True, "complevel": 4} for v in ds_trunc.data_vars})
    
    print(f"  → Truncated range: {str(ds_trunc.time.values[0])[:10]} → {str(ds_trunc.time.values[-1])[:10]}")
    print(f"  Saved to: {out_path}")
    
    ds.close()
    ds_trunc.close()

```

```{python truncation-check}
import xarray as xr
for fname in files:
    path = fname.replace(".nc", "_1958-2024.nc")
    ds = xr.open_dataset(os.path.join(PROCESSED_DATA_DIR, path))
    print(f"{fname}: {len(ds.time)} months, last = {str(ds.time.values[-1])[:10]}")
    ds.close()

```

### Sanity checks budget

```{python budget-verification}
log_message("\n" + "="*70)
log_message("DETAILED CHECK: Testing edge effects near end of dataset")
log_message("="*70)

# Test the last 10 months specifically
edge_months = list(range(794, 804))  # Last 10 months
log_message(f"\nChecking final months {edge_months[0]} to {edge_months[-1]}:")

edge_errors = []
for month_idx in edge_months:
    tke_val = float(ds_tke['tke'].isel(time=month_idx).mean())
    mke_val = float(ds_mke['mke'].isel(time=month_idx).mean())
    eke_val = float(ds_eke['eke'].isel(time=month_idx).mean())
    
    sum_val = mke_val + eke_val
    diff = abs(tke_val - sum_val)
    rel_error = (diff / tke_val) * 100
    edge_errors.append(rel_error)
    
    month_date = str(ds_tke.time.values[month_idx])[:7]
    status = "OK" if rel_error < 0.01 else "ISSUE"
    log_message(f"  Month {month_idx} ({month_date}): error = {rel_error:.4f}% [{status}]")

# Test middle months for comparison
log_message(f"\nChecking middle months for comparison:")
middle_months = [300, 350, 400, 450, 500]
middle_errors = []

for month_idx in middle_months:
    tke_val = float(ds_tke['tke'].isel(time=month_idx).mean())
    mke_val = float(ds_mke['mke'].isel(time=month_idx).mean())
    eke_val = float(ds_eke['eke'].isel(time=month_idx).mean())
    
    sum_val = mke_val + eke_val
    diff = abs(tke_val - sum_val)
    rel_error = (diff / tke_val) * 100
    middle_errors.append(rel_error)
    
    month_date = str(ds_tke.time.values[month_idx])[:7]
    log_message(f"  Month {month_idx} ({month_date}): error = {rel_error:.4f}%")

log_message(f"\nEdge months mean error: {np.mean(edge_errors):.4f}%")
log_message(f"Middle months mean error: {np.mean(middle_errors):.4f}%")

if np.mean(edge_errors) > 1.0:
    log_message("\nWARNING: Edge effects detected near end of dataset!")
    log_message("This is likely due to the final chunk in bandpass filtering.")
    log_message("Recommendation: Exclude final 10-15 months from analysis, or")
    log_message("              reprocess with extended data beyond 2024 if available.")
else:
    log_message("\nNo systematic edge effects detected.")
```

```{python month-na-verification}
import xarray as xr
import numpy as np

# Check the filtered eddy file at problem month
eddy_path = "E:/ERA5_data/processed/ERA5_1958-2024_eddy_components_monthly.nc"
ds_eddy = xr.open_dataset(eddy_path, engine='netcdf4')

# July 2020 - find timesteps
july_2020 = ds_eddy.sel(time='2020-07')
print(f"July 2020 timesteps: {len(july_2020.time)}")
print(f"First: {july_2020.time.values[0]}")
print(f"Last: {july_2020.time.values[-1]}")

# Check for NaNs or anomalies
print(f"\nU_eddy NaNs: {july_2020['u_eddy'].isnull().sum().values}")
print(f"V_eddy NaNs: {july_2020['v_eddy'].isnull().sum().values}")

print(f"\nU_eddy range: {float(july_2020['u_eddy'].min())} to {float(july_2020['u_eddy'].max())}")
print(f"V_eddy range: {float(july_2020['v_eddy'].min())} to {float(july_2020['v_eddy'].max())}")

# Check for infinite values
print(f"\nU_eddy infinite: {np.isinf(july_2020['u_eddy'].values).sum()}")
print(f"V_eddy infinite: {np.isinf(july_2020['v_eddy'].values).sum()}")

ds_eddy.close()
```

```{python full-dataset-month-timesteps-verification}
import xarray as xr
import numpy as np
import pandas as pd

continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
ds = xr.open_dataset(continuous_path, engine='netcdf4')

# Better way: manually count timesteps per month
print("Checking timestep distribution by month...")

# Convert to pandas for easier grouping
time_series = pd.to_datetime(ds.time.values)
year_month = time_series.to_period('M')

# Count timesteps per month
counts = year_month.value_counts().sort_index()

print(f"\nTotal months: {len(counts)}")
print(f"Timesteps per month range: {counts.min()} to {counts.max()}")
print(f"Expected: 120-124 timesteps per month (4 per day)")

# Find anomalous months
anomalous = counts[counts < 120]
print(f"\nMonths with < 120 timesteps: {len(anomalous)}")

if len(anomalous) > 0:
    print("\nProblematic months:")
    for month, count in anomalous.items():
        print(f"  {month}: {count} timesteps")

# Check July 2020 specifically
july_2020_count = counts[counts.index == '2020-07']
print(f"\nJuly 2020 timestep count: {july_2020_count.values[0] if len(july_2020_count) > 0 else 'NOT FOUND'}")

# Also check December 2024
dec_2024_count = counts[counts.index == '2024-12']
print(f"December 2024 timestep count: {dec_2024_count.values[0] if len(dec_2024_count) > 0 else 'NOT FOUND'}")

ds.close()
```

```{python time-allignement-and-manual-budget-verification}
import xarray as xr
import numpy as np

# Load all three datasets
continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
tke_path = "E:/ERA5_data/processed/ERA5_TKE_monthly.nc"
mke_path = "E:/ERA5_data/processed/ERA5_MKE_monthly.nc"
eke_path = "E:/ERA5_data/processed/ERA5_EKE_monthly.nc"

ds_cont = xr.open_dataset(continuous_path, engine='netcdf4')
ds_tke = xr.open_dataset(tke_path, engine='netcdf4')
ds_mke = xr.open_dataset(mke_path, engine='netcdf4')
ds_eke = xr.open_dataset(eke_path, engine='netcdf4')

print("Dataset sizes:")
print(f"  Continuous: {len(ds_cont.time)} timesteps")
print(f"  TKE: {len(ds_tke.time)} months")
print(f"  MKE: {len(ds_mke.time)} months")
print(f"  EKE: {len(ds_eke.time)} months")

print("\nTime coordinate alignment:")
print(f"  TKE[750]: {ds_tke.time.values[750]}")
print(f"  MKE[750]: {ds_mke.time.values[750]}")
print(f"  EKE[750]: {ds_eke.time.values[750]}")

# Check if they all point to same month
print("\nAre they the same month?")
print(f"  TKE == MKE: {ds_tke.time.values[750] == ds_mke.time.values[750]}")
print(f"  TKE == EKE: {ds_tke.time.values[750] == ds_eke.time.values[750]}")
print(f"  MKE == EKE: {ds_mke.time.values[750] == ds_eke.time.values[750]}")

# Manually verify month 750 calculation
print("\n=== MANUAL VERIFICATION FOR JULY 2020 ===")

# Get July 2020 data from continuous file
july_data = ds_cont.sel(time='2020-07')
print(f"July 2020 continuous data: {len(july_data.time)} timesteps")

# Calculate components manually
u = july_data['u']
v = july_data['v']

# TKE
tke_manual = (0.5 * (u**2 + v**2)).mean(dim='time').mean()
print(f"\nManual TKE: {float(tke_manual):.4f}")
print(f"File TKE:   {float(ds_tke['tke'].isel(time=750).mean()):.4f}")

# MKE  
u_mean = u.mean(dim='time')
v_mean = v.mean(dim='time')
mke_manual = (0.5 * (u_mean**2 + v_mean**2)).mean()
print(f"\nManual MKE: {float(mke_manual):.4f}")
print(f"File MKE:   {float(ds_mke['mke'].isel(time=750).mean()):.4f}")

# EKE
u_anom = u - u_mean
v_anom = v - v_mean
eke_manual = (0.5 * (u_anom**2 + v_anom**2)).mean(dim='time').mean()
print(f"\nManual EKE: {float(eke_manual):.4f}")
print(f"File EKE:   {float(ds_eke['eke'].isel(time=750).mean()):.4f}")

# Check balance
print(f"\n=== BUDGET CHECK ===")
tke_val = float(tke_manual)
mke_val = float(mke_manual)
eke_val = float(eke_manual)
print(f"TKE: {tke_val:.4f}")
print(f"MKE + EKE: {mke_val + eke_val:.4f}")
print(f"Difference: {abs(tke_val - (mke_val + eke_val)):.4f}")
print(f"Error: {abs(tke_val - (mke_val + eke_val))/tke_val * 100:.4f}%")

ds_cont.close()
ds_tke.close()
ds_mke.close()
ds_eke.close()
```

```{python june16-missing-ERA5_1958-2024_std_res_vars_continuous?}
import xarray as xr
import pandas as pd

continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
ds = xr.open_dataset(continuous_path, engine='netcdf4')

# Check for June 2016
print("Checking for June 2016...")
try:
    june_2016 = ds.sel(time='2016-06')
    print(f"June 2016 found: {len(june_2016.time)} timesteps")
except:
    print("June 2016 NOT FOUND in continuous file!")

# Check what months exist in 2016
year_2016 = ds.sel(time='2016')
months_2016 = pd.to_datetime(year_2016.time.values).month.unique()
print(f"\nMonths present in 2016: {sorted(months_2016)}")
print(f"Missing months: {set(range(1,13)) - set(months_2016)}")

ds.close()
```

```{python std-june16-inspection}
import xarray as xr
import os

# Check 1: Verify the exact path
std_dir = "E:/ERA5_data/processed/temp_monthly_std"  # Note: you have extra nested folder!
june_file = os.path.join(std_dir, "std_2016_06.nc")

print(f"Looking for file at: {june_file}")
print(f"File exists: {os.path.exists(june_file)}")

# Check 2: List all files in directory
print(f"\nFiles in {std_dir}:")
try:
    files = os.listdir(std_dir)
    june_files = [f for f in files if '2016_06' in f]
    print(f"  Files with '2016_06': {june_files}")
except Exception as e:
    print(f"  Error listing directory: {e}")

# Check 3: Try to open the June 2016 file directly
print("\nAttempting to open June 2016 file...")
try:
    ds_june = xr.open_dataset("E:/ERA5_data/processed/temp_monthly_std/std_2016_06.nc", engine='netcdf4')
    print(f"  SUCCESS! File opened.")
    print(f"  Dimensions: {dict(ds_june.dims)}")
    print(f"  Variables: {list(ds_june.data_vars)}")
    print(f"  Time coord exists: {'time' in ds_june.coords}")
    if 'time' in ds_june.coords:
        print(f"  Time length: {len(ds_june.time)}")
        if len(ds_june.time) > 0:
            print(f"  First time: {ds_june.time.values[0]}")
            print(f"  Last time: {ds_june.time.values[-1]}")
    ds_june.close()
except Exception as e:
    print(f"  FAILED to open: {e}")

# Check 4: Check continuous file for June 2016 more carefully
print("\nChecking continuous file...")
continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
ds_cont = xr.open_dataset(continuous_path, engine='netcdf4')

print(f"Total timesteps in continuous: {len(ds_cont.time)}")

# Try different ways to get June 2016
print("\nSearching for June 2016 in continuous file:")

# Method 1: Direct selection
try:
    june_2016 = ds_cont.sel(time='2016-06')
    print(f"  Method 1 (sel): {len(june_2016.time)} timesteps")
except Exception as e:
    print(f"  Method 1 failed: {e}")

# Method 2: Slice by date range
try:
    june_2016 = ds_cont.sel(time=slice('2016-06-01', '2016-06-30'))
    print(f"  Method 2 (slice): {len(june_2016.time)} timesteps")
    if len(june_2016.time) > 0:
        print(f"    First: {june_2016.time.values[0]}")
        print(f"    Last: {june_2016.time.values[-1]}")
except Exception as e:
    print(f"  Method 2 failed: {e}")

# Method 3: Check what's around June 2016
print("\n  Times around June 2016:")
may_june_july = ds_cont.sel(time=slice('2016-05-25', '2016-07-05'))
print(f"    May 25 - July 5: {len(may_june_july.time)} timesteps")
if len(may_june_july.time) > 0:
    for i, t in enumerate(may_june_july.time.values):
        if i < 5 or i >= len(may_june_july.time) - 5:
            print(f"      {t}")
        elif i == 5:
            print(f"      ... ({len(may_june_july.time) - 10} more) ...")

ds_cont.close()
```

```{python}
import xarray as xr
import pandas as pd

continuous_path = "E:/ERA5_data/processed/ERA5_1958-2024_std_res_vars_continuous.nc"
ds = xr.open_dataset(continuous_path)

# Convert time to pandas DatetimeIndex
times = pd.to_datetime(ds.time.values)

# Compute time differences between consecutive timesteps
gaps = times[1:] - times[:-1]

# Convert to Series for .mode()
gaps_series = pd.Series(gaps)

print("Expected 6-hour spacing:", gaps_series.mode()[0])
print("\nDetected time gaps > 6 hours:")
print(times[:-1][gaps > pd.Timedelta('6H')])

ds.close()
```

## Reynolds decomposition on raw data

```{python}
# ============================================================================
# ARCTIC STORMINESS CLIMATOLOGY (1958-2024)
# SENSITIVITY TEST: Bandpass Filter Only (No Prior Deseasonalization)
# ============================================================================
#
# PURPOSE: Test whether the 2-6 day Butterworth bandpass filter alone is 
#          sufficient to isolate synoptic-scale eddies, without first 
#          subtracting monthly climatological means.
#
# RATIONALE :
#   1. The bandpass filter removes ALL frequencies outside 2-6 days, including
#      the seasonal cycle (period = 365 days >> 6 days). So deseasonalization
#      may be redundant.
#   
#   2. The original approach of subtracting step-function monthly climatologies
#      creates discontinuities at month boundaries. These sudden jumps have
#      broadband Fourier spectra that could partially bleed into the 2-6 day
#      passband, potentially contaminating the storm signal.
#   
#   3. By applying the filter directly to raw winds, we avoid these artifacts
#      entirely and let the filter do all the work of defining "background flow"
#      (anything slower than 6 days).
#
# COMPARISON: Results should be compared with the original methodology to
#             quantify whether deseasonalization affects the final storminess
#             trends. If results are nearly identical, the simpler approach
#             (filter only) is preferred.
#
# ============================================================================

import xarray as xr
import os
import cupy as cp
import cupyx.scipy.signal as cusignal
from scipy import signal
import numpy as np
from datetime import datetime

# Define paths
PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
OUTPUT_DIR = os.path.join(PROCESSED_DATA_DIR, "raw_reynolds_decomposition")

# Create output directory if it doesn't exist
os.makedirs(OUTPUT_DIR, exist_ok=True)

def log_message(msg):
    """Log messages with timestamp"""
    timestamp = datetime.now().strftime("%Y-%m-%d %H:%M:%S")
    print(f"[{timestamp}] {msg}")

log_message("=" * 70)
log_message("SENSITIVITY TEST: BANDPASS FILTER WITHOUT DESEASONALIZATION")
log_message("Following Stefan Brönnimann's methodological suggestion")
log_message("=" * 70)

# ============================================================================
# NOTE: STEP 3 (MONTHLY CLIMATOLOGY) IS INTENTIONALLY SKIPPED
# ============================================================================
#
# In the original methodology, we calculated monthly climatological means and
# subtracted them before filtering. Here we skip this entirely because:
#
#   1. The 2-6 day bandpass filter will remove the seasonal cycle anyway
#      (seasonal period = 365 days, which is far outside the 6-day cutoff)
#
#   2. Subtracting step-function monthly means creates artificial discontinuities
#      at month boundaries (e.g., Jan 31 23:00 → Feb 1 00:00). These jumps
#      contain high-frequency energy that could contaminate the 2-6 day band.
#
#   3. This test will reveal whether the deseasonalization step affects results
#      or is simply redundant preprocessing.
#
# ============================================================================

# ============================================================================
# STEP 4 (MODIFIED): DIRECT BANDPASS FILTERING OF RAW WINDS
# ============================================================================
#
# WHAT: Apply 2-6 day bandpass filter directly to raw wind components (u, v)
#       WITHOUT any prior deseasonalization or anomaly calculation.
#
# WHY:  The Butterworth bandpass filter is designed to:
#       - REMOVE: All variability with periods > 6 days (including seasonal cycle,
#         interannual variability, planetary waves, blocking patterns)
#       - REMOVE: All variability with periods < 2 days (diurnal cycle, turbulence,
#         high-frequency noise)
#       - RETAIN: Only 2-6 day variability (synoptic-scale storms/cyclones)
#
#       The filter itself defines what constitutes "background flow" (anything
#       slower than 6 days) without needing explicit climatology subtraction.
#
# DIFFERENCE FROM ORIGINAL:
#       Original:  u_eddy = filter(u - u_climatology)
#       This test: u_eddy = filter(u)
#
#       If results are equivalent, the simpler approach is preferred.
#
# ============================================================================

log_message("\n" + "=" * 70)
log_message("STEP 4 (MODIFIED): DIRECT BANDPASS FILTERING")
log_message("=" * 70)

def vector_bandpass_filter_gpu(data_array, time_dim, fs, lowcut_days, highcut_days, order=5):
    """
    Apply 2-6 day bandpass filter using GPU acceleration.
    
    This implements the Butterworth bandpass filter following Yau & Chang (2020):
    - Filters out slow-moving features (>6 days) including seasonal cycle
    - Filters out high-frequency noise (<2 days) including diurnal cycle
    - Retains synoptic-scale eddies (storms) in the 2-6 day period band
    - Uses GPU (CuPy) for massive speedup on large datasets
    
    Parameters:
    -----------
    data_array : xarray.DataArray
        Input data to filter (raw u or v wind components)
    time_dim : str
        Name of time dimension (usually 'time')
    fs : float
        Sampling frequency (4 = four times per day for 6-hourly data)
    lowcut_days : float
        Lower period cutoff in days (6.0 days = removes slower features)
    highcut_days : float
        Upper period cutoff in days (2.0 days = removes faster features)
    order : int
        Butterworth filter order (5 is standard)
    
    Returns:
    --------
    xarray.DataArray : Filtered data (eddy components)
    
    NOTES ON WHAT THE FILTER REMOVES:
    ---------------------------------
    The 6-day high-pass component removes:
      - Seasonal cycle (365 days)
      - Monthly variations (~30 days)
      - Intraseasonal oscillations (30-60 days)
      - Planetary waves (typically 10-30 days)
      - Blocking patterns (5-15 days, partially)
    
    The 2-day low-pass component removes:
      - Diurnal cycle (1 day)
      - Semi-diurnal tides (0.5 days)
      - High-frequency turbulence (< 1 day)
      - Mesoscale phenomena (hours)
    """
    
    # Convert period (days) to frequency (cycles per timestep)
    # For 6-hourly data: fs = 4 samples/day
    lowcut_freq = 1 / lowcut_days   # Low frequency bound (long period cutoff)
    highcut_freq = 1 / highcut_days  # High frequency bound (short period cutoff)
    
    # Design Butterworth bandpass filter coefficients
    # Order 5 forward-backward = effective order 10
    b, a = signal.butter(order, [lowcut_freq, highcut_freq], btype='band', fs=fs)
    
    def filter_func_gpu(x):
        """
        Inner function that does the actual GPU filtering.
        
        Process:
        1. Transfer data to GPU (CuPy array)
        2. Detrend to remove any linear drift (important for numerical stability)
        3. Apply forward-backward filter (filtfilt) for zero phase shift
        4. Transfer result back to CPU (NumPy array)
        
        NOTE: The detrend here removes only the LINEAR trend within each chunk,
        not the seasonal cycle. This is a numerical stability measure, not
        a deseasonalization step.
        """
        x_gpu = cp.asarray(x.astype('float64'))
        filtered_gpu = cusignal.filtfilt(b, a, cusignal.detrend(x_gpu))
        return cp.asnumpy(filtered_gpu)
    
    # Apply filter along time dimension using xarray's apply_ufunc
    return xr.apply_ufunc(
        filter_func_gpu, 
        data_array,
        input_core_dims=[[time_dim]],
        output_core_dims=[[time_dim]],
        exclude_dims={time_dim},
        dask='parallelized',
        output_dtypes=[data_array.dtype],
        keep_attrs=True
    )


# Define file paths
continuous_path = os.path.join(PROCESSED_DATA_DIR, "ERA5_1958-2025_std_res_vars_continuous.nc")
eddy_path = os.path.join(OUTPUT_DIR, "ERA5_1958-2024_eddy_components_raw.nc")

# Check if already processed
if os.path.exists(eddy_path):
    log_message("Checking existing eddy components file...")
    try:
        ds_test = xr.open_dataset(eddy_path, engine='netcdf4')
        if 'u_eddy' in ds_test and 'v_eddy' in ds_test:
            log_message(">> Eddy components already exist and are complete - skipping Step 4")
            ds_test.close()
        else:
            log_message(">> Eddy components file is incomplete - will reprocess")
            ds_test.close()
            os.remove(eddy_path)
    except:
        log_message(">> Eddy components file is corrupted - will reprocess")
        if os.path.exists(eddy_path):
            os.remove(eddy_path)

if not os.path.exists(eddy_path):
    # ------------------------------------------------------------------------
    # 4.1: LOAD RAW DATA (NO CLIMATOLOGY NEEDED)
    # ------------------------------------------------------------------------
    log_message("Loading continuous dataset...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    log_message(f"  Loaded: {len(ds.time)} timesteps")
    log_message(f"  Period: {ds.time.values[0]} to {ds.time.values[-1]}")
    
    # NOTE: We skip climatology loading entirely!
    log_message("\n  >>> SKIPPING monthly climatology subtraction <<<")
    log_message("  >>> Bandpass filter will be applied directly to raw winds <<<")
    
    # ------------------------------------------------------------------------
    # 4.2: NO ANOMALY CALCULATION - USE RAW WINDS DIRECTLY
    # ------------------------------------------------------------------------
    log_message("\nPreparing raw wind components for filtering...")
    log_message("  NOTE: No deseasonalization applied")
    log_message("  The bandpass filter will remove the seasonal cycle directly")
    
    # Use raw winds directly (no subtraction of climatology)
    u_raw = ds['u']
    v_raw = ds['v']
    
    log_message(f"\nRaw wind statistics:")
    log_message(f"  u range: {float(u_raw.min()):.2f} to {float(u_raw.max()):.2f} m/s")
    log_message(f"  v range: {float(v_raw.min()):.2f} to {float(v_raw.max()):.2f} m/s")
    log_message(f"  Note: These contain the full signal including seasonal cycle")
    
    # ------------------------------------------------------------------------
    # 4.3: SETUP CHUNKED PROCESSING WITH OVERLAP
    # ------------------------------------------------------------------------
    log_message("\n=== BANDPASS FILTERING (2-6 DAYS) ===")
    log_message("Applying filter directly to raw winds (no prior deseasonalization)")
    log_message("METHOD: Overlap-save algorithm for continuous filtering")
    
    chunk_size = 10000  # ~2.5 years per chunk
    overlap = 500       # ~125 days overlap (>> 6-day filter period)
    
    total_timesteps = len(ds.time)
    n_chunks = int(np.ceil(total_timesteps / chunk_size))
    
    log_message(f"\nProcessing strategy:")
    log_message(f"  - {n_chunks} temporal chunks of {chunk_size} timesteps each")
    log_message(f"  - {overlap} timestep overlap between chunks (~125 days)")
    log_message(f"  - Overlap exceeds filter transition band by >10x for safety")
    
    u_eddy_list = []
    v_eddy_list = []
    
    # ------------------------------------------------------------------------
    # 4.4: PROCESS EACH CHUNK WITH GPU FILTERING
    # ------------------------------------------------------------------------
    for i in range(n_chunks):
        start_idx = max(0, i * chunk_size - overlap)
        end_idx = min(total_timesteps, (i + 1) * chunk_size + overlap)
        
        log_message(f"\n--- Chunk {i+1}/{n_chunks} ---")
        log_message(f"  Processing timesteps {start_idx} to {end_idx} ({end_idx - start_idx} total)")
        
        # Extract temporal chunk of RAW winds (not anomalies)
        u_chunk = u_raw.isel(time=slice(start_idx, end_idx))
        v_chunk = v_raw.isel(time=slice(start_idx, end_idx))
        
        # Apply bandpass filter on GPU
        log_message(f"  Applying 2-6 day bandpass filter on GPU...")
        log_message(f"    - Removing periods > 6 days (including 365-day seasonal cycle)")
        log_message(f"    - Removing periods < 2 days (diurnal cycle, turbulence)")
        u_eddy_chunk = vector_bandpass_filter_gpu(u_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        v_eddy_chunk = vector_bandpass_filter_gpu(v_chunk, 'time', fs=4, lowcut_days=6.0, highcut_days=2.0)
        
        # Trim overlap regions (discard edge effects)
        log_message(f"  Trimming edge effects...")
        if i == 0:
            # First chunk: keep start, trim end
            trim_end = -overlap if i < n_chunks - 1 else None
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(None, trim_end))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(None, trim_end))
            log_message(f"    First chunk: trim last {overlap} points")
        elif i == n_chunks - 1:
            # Last chunk: trim start, keep end
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, None))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, None))
            log_message(f"    Last chunk: trim first {overlap} points")
        else:
            # Middle chunks: trim both ends
            u_eddy_trimmed = u_eddy_chunk.isel(time=slice(overlap, -overlap))
            v_eddy_trimmed = v_eddy_chunk.isel(time=slice(overlap, -overlap))
            log_message(f"    Middle chunk: trim {overlap} points from each end")
        
        u_eddy_list.append(u_eddy_trimmed)
        v_eddy_list.append(v_eddy_trimmed)
        
        log_message(f"  >> Chunk filtered: {len(u_eddy_trimmed.time)} timesteps retained")
    
    # ------------------------------------------------------------------------
    # 4.5: COMBINE ALL CHUNKS
    # ------------------------------------------------------------------------
    log_message("\nCombining all filtered chunks into continuous time series...")
    u_eddy = xr.concat(u_eddy_list, dim='time')
    v_eddy = xr.concat(v_eddy_list, dim='time')
    
    log_message(f"Final check: {len(u_eddy.time)} timesteps (expected: {total_timesteps})")
    if len(u_eddy.time) != total_timesteps:
        log_message(f"WARNING: Timestep mismatch!")
    else:
        log_message(">> All timesteps accounted for")
    
    # Restore original time coordinates
    u_eddy['time'] = ds['time']
    v_eddy['time'] = ds['time']
    
    log_message(f"\nFinal filtered dataset:")
    log_message(f"  - Total timesteps: {len(u_eddy.time):,}")
    log_message(f"  - Time span: {str(u_eddy.time.values[0])[:10]} to {str(u_eddy.time.values[-1])[:10]}")
    log_message(f"  - u_eddy range: {float(u_eddy.min()):.2f} to {float(u_eddy.max()):.2f} m/s")
    log_message(f"  - v_eddy range: {float(v_eddy.min()):.2f} to {float(v_eddy.max()):.2f} m/s")
    
    log_message("\nVERIFICATION THAT SEASONAL CYCLE WAS REMOVED:")
    log_message("  The filtered u_eddy and v_eddy should oscillate around zero")
    log_message("  (If seasonal cycle remained, we'd see systematic offsets)")
    log_message(f"  u_eddy mean: {float(u_eddy.mean()):.6f} m/s (should be ~0)")
    log_message(f"  v_eddy mean: {float(v_eddy.mean()):.6f} m/s (should be ~0)")
    
    # Truncate to analysis period (1958-2024)
    log_message("\nTruncating filtered data to analysis period (1958-2024)...")
    u_eddy = u_eddy.sel(time=slice('1958-01-01', '2024-12-31'))
    v_eddy = v_eddy.sel(time=slice('1958-01-01', '2024-12-31'))

    log_message(f"Final analysis dataset:")
    log_message(f"  Timesteps: {len(u_eddy.time)}")
    log_message(f"  Period: {str(u_eddy.time.values[0])[:10]} to {str(u_eddy.time.values[-1])[:10]}")
    
    # ------------------------------------------------------------------------
    # 4.6: SAVE EDDY COMPONENTS
    # ------------------------------------------------------------------------
    log_message("\nSaving filtered eddy components to file...")
    
    eddy_ds = xr.Dataset(
        {'u_eddy': u_eddy, 'v_eddy': v_eddy},
        attrs={
            'description': 'Bandpass-filtered (2-6 day) wind components - NO prior deseasonalization',
            'methodology': 'Sensitivity test: filter applied directly to raw winds',
            'comparison_with': 'ERA5_1958-2024_eddy_components_monthly.nc (original with deseasonalization)'
        }
    )
    
    # Use compression
    encoding = {
        'u_eddy': {'zlib': True, 'complevel': 4},
        'v_eddy': {'zlib': True, 'complevel': 4}
    }
    
    # Force computation before saving
    log_message("  Computing data arrays (loading into RAM)...")
    eddy_ds = eddy_ds.compute()
    
    log_message("  Writing to disk...")
    eddy_ds.to_netcdf(eddy_path, encoding=encoding, format='NETCDF4')
    
    # Get file size
    file_size_gb = os.path.getsize(eddy_path) / (1024**3)
    log_message(f"  >> File saved: {file_size_gb:.2f} GB")
    
    # Clean up
    ds.close()
    del u_raw, v_raw, u_eddy, v_eddy, u_eddy_list, v_eddy_list, eddy_ds
    
    log_message(">> Step 4 complete: Eddy components saved successfully")


# ============================================================================
# STEP 5: CALCULATE KINETIC ENERGY BUDGET
# ============================================================================
#
# NOTE: Steps 5.1 (TKE) and 5.2 (MKE) use the raw continuous data and are
# identical to the original methodology. We only recalculate these if they
# don't already exist in the main processed folder.
#
# Steps 5.3 (EKE) and 5.4 (Storminess Index) are recalculated using the new
# eddy components from the filter-only approach, and saved to the new folder.
#
# ============================================================================

log_message("\n" + "=" * 70)
log_message("STEP 5: KINETIC ENERGY BUDGET")
log_message("=" * 70)

# Use existing TKE and MKE from main folder (they don't depend on deseasonalization)
tke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_TKE_monthly.nc')
mke_path = os.path.join(PROCESSED_DATA_DIR, 'ERA5_MKE_monthly.nc')

# New EKE and storminess for sensitivity test
eke_path = os.path.join(OUTPUT_DIR, 'ERA5_EKE_monthly_raw.nc')
storm_path = os.path.join(OUTPUT_DIR, 'ERA5_storminess_index_monthly_raw.nc')

# ------------------------------------------------------------------------
# 5.1: TKE (use existing or calculate)
# ------------------------------------------------------------------------
log_message("\n5.1: Total Kinetic Energy (TKE)...")
if not os.path.exists(tke_path):
    log_message("  Calculating TKE from raw winds...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    tke_instant = 0.5 * (ds['u']**2 + ds['v']**2)
    tke_monthly = tke_instant.resample(time='ME').mean(dim='time')
    tke_monthly.name = 'tke'
    tke_monthly.to_netcdf(tke_path)
    ds.close()
    log_message(">> TKE calculated and saved")
else:
    log_message(">> TKE already exists in main folder (reusing)")

# ------------------------------------------------------------------------
# 5.2: MKE (use existing or calculate)
# ------------------------------------------------------------------------
log_message("\n5.2: Mean Kinetic Energy (MKE)...")
if not os.path.exists(mke_path):
    log_message("  Calculating MKE from monthly-mean winds...")
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    u_monthly_mean = ds['u'].resample(time='ME').mean(dim='time')
    v_monthly_mean = ds['v'].resample(time='ME').mean(dim='time')
    mke_monthly = 0.5 * (u_monthly_mean**2 + v_monthly_mean**2)
    mke_monthly.name = 'mke'
    mke_monthly.to_netcdf(mke_path)
    ds.close()
    log_message(">> MKE calculated and saved")
else:
    log_message(">> MKE already exists in main folder (reusing)")

# ------------------------------------------------------------------------
# 5.3: EKE (recalculate using new eddy components)
# ------------------------------------------------------------------------
log_message("\n5.3: Eddy Kinetic Energy (EKE) - using filter-only eddy components...")
log_message("     Decomposition relative to EACH MONTH'S OWN MEAN for budget closure")

if not os.path.exists(eke_path):
    ds = xr.open_dataset(continuous_path, engine='netcdf4')
    
    log_message("  Processing month-by-month to ensure proper decomposition...")
    
    # Group data by month-end periods
    grouped = ds.resample(time='ME')
    
    eke_monthly_list = []
    month_coords = []
    
    for label, group in grouped:
        # For this specific month's data:
        # Calculate the mean wind for THIS month
        u_mean_this_month = group['u'].mean(dim='time')
        v_mean_this_month = group['v'].mean(dim='time')
        
        # Calculate anomalies relative to THIS month's mean
        u_anom = group['u'] - u_mean_this_month
        v_anom = group['v'] - v_mean_this_month
        
        # Calculate instantaneous EKE
        eke_instant = 0.5 * (u_anom**2 + v_anom**2)
        
        # Calculate mean EKE for this month
        eke_this_month = eke_instant.mean(dim='time')
        
        eke_monthly_list.append(eke_this_month)
        month_coords.append(label)
    
    log_message(f"  Processed {len(eke_monthly_list)} months")
    
    # Combine into single DataArray
    eke_monthly = xr.concat(eke_monthly_list, dim='time')
    eke_monthly = eke_monthly.assign_coords(time=month_coords)
    eke_monthly.name = 'eke'
    
    eke_monthly.to_netcdf(eke_path)
    ds.close()
    log_message(">> EKE calculated with correct decomposition")
else:
    log_message(">> EKE already exists")

# ------------------------------------------------------------------------
# 5.4: STORMINESS INDEX (using new eddy components)
# ------------------------------------------------------------------------
log_message("\n5.4: Calculating Storminess Index from filter-only eddy components...")

if not os.path.exists(storm_path):
    log_message("  Opening eddy components (filter-only version)...")
    ds_eddy = xr.open_dataset(eddy_path, engine='netcdf4')
    
    log_message(f"  File dimensions: {ds_eddy['u_eddy'].dims}")
    
    log_message("  Processing in monthly batches...")
    
    storminess_monthly_list = []
    time_labels = []
    
    month_count = 0
    for time_label, group in ds_eddy.resample(time='ME'):
        month_count += 1
        if month_count % 100 == 0:
            log_message(f"    Processed {month_count} months...")
        
        # Get data as numpy arrays
        u_eddy = group['u_eddy'].values
        v_eddy = group['v_eddy'].values
        
        # Calculate instantaneous EKE from filtered eddies
        eke_instant = 0.5 * (u_eddy**2 + v_eddy**2)
        
        # Storminess = variance of EKE over the month
        # axis=-1 means last dimension (time)
        storminess_month = np.var(eke_instant, axis=-1)
        
        # Create DataArray for this month
        storm_da = xr.DataArray(
            storminess_month,
            dims=['latitude', 'longitude'],
            coords={
                'latitude': group['latitude'].values,
                'longitude': group['longitude'].values
            }
        )
        
        storminess_monthly_list.append(storm_da)
        time_labels.append(time_label)
    
    log_message(f"  Combining {len(storminess_monthly_list)} months...")
    storminess_monthly = xr.concat(storminess_monthly_list, dim='time')
    storminess_monthly = storminess_monthly.assign_coords(time=time_labels)
    storminess_monthly.name = 'storminess_index'
    storminess_monthly.attrs = {
        'description': 'Storminess index (variance of bandpass-filtered EKE)',
        'methodology': 'Sensitivity test: no prior deseasonalization',
        'units': 'm^4/s^4'
    }
    
    log_message("  Saving Storminess Index...")
    storminess_monthly.to_netcdf(storm_path)
    
    ds_eddy.close()
    
    size_mb = os.path.getsize(storm_path) / (1024**2)
    log_message(f">> Storminess Index calculated: {size_mb:.1f} MB")
else:
    log_message(">> Storminess Index already exists")

# ------------------------------------------------------------------------
# 5.5: VERIFY ENERGY BUDGET CLOSURE
# ------------------------------------------------------------------------
log_message("\n" + "=" * 70)
log_message("VERIFYING ENERGY BUDGET CLOSURE")
log_message("=" * 70)

try:
    ds_tke = xr.open_dataset(tke_path, engine='netcdf4')
    ds_mke = xr.open_dataset(mke_path, engine='netcdf4')
    ds_eke = xr.open_dataset(eke_path, engine='netcdf4')
    
    sample_months = [50, 100, 200, 400, 600, 750, 800]
    all_errors = []
    
    log_message("\nTesting energy budget at sample months:")
    
    for month_idx in sample_months:
        if month_idx >= len(ds_tke.time):
            continue
            
        tke_val = float(ds_tke['tke'].isel(time=month_idx).mean())
        mke_val = float(ds_mke['mke'].isel(time=month_idx).mean())
        eke_val = float(ds_eke['eke'].isel(time=month_idx).mean())
        
        sum_val = mke_val + eke_val
        diff = abs(tke_val - sum_val)
        rel_error = (diff / tke_val) * 100
        all_errors.append(rel_error)
        
        month_date = str(ds_tke.time.values[month_idx])[:7]
        log_message(f"Month {month_idx} ({month_date}): error = {rel_error:.4f}%")
    
    mean_error = np.mean(all_errors)
    log_message(f"\nMean relative error: {mean_error:.4f}%")
    
    if mean_error < 0.01:
        log_message("SUCCESS - ENERGY BUDGET VERIFIED (< 0.01% error)")
    else:
        log_message("WARNING - Budget error > 0.01%")
    
    ds_tke.close()
    ds_mke.close()
    ds_eke.close()
    
except Exception as e:
    log_message(f"ERROR - Could not verify energy budget: {e}")
    import traceback
    traceback.print_exc()

# ============================================================================
# FINAL SUMMARY
# ============================================================================

log_message("\n" + "=" * 70)
log_message("*** SENSITIVITY TEST COMPLETE ***")
log_message("=" * 70)

log_message("\nOUTPUT FILES (in raw_reynolds_decomposition/):")
for path, name in [
    (eddy_path, "Eddy components (filter-only)"),
    (eke_path, "EKE (monthly)"),
    (storm_path, "Storminess Index")
]:
    if os.path.exists(path):
        size_mb = os.path.getsize(path) / (1024**2)
        log_message(f"  ✓ {name}: {size_mb:.1f} MB")
    else:
        log_message(f"  ✗ {name}: MISSING!")

log_message("\nNEXT STEPS:")
log_message("  1. Compare trend analysis results between methodologies")
log_message("  2. Check for month-boundary artifacts in original approach")
log_message("  3. If results are equivalent, adopt simpler filter-only method")
log_message("  4. Document sensitivity analysis in thesis methodology section")
```

```{python truncating-end-2024}
import xarray as xr
import numpy as np

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
OUTPUT_DIR = "E:/ERA5_data/processed/raw_reynolds_decomposition"

# ============================================================================
# 1. Truncate EKE to 1958-2024
# ============================================================================
eke_path = f"{OUTPUT_DIR}/ERA5_EKE_monthly_raw.nc"

print("Truncating EKE to 1958-2024...")
ds_eke = xr.open_dataset(eke_path, engine='netcdf4')
print(f"  Original time steps: {len(ds_eke.time)}")

ds_eke_truncated = ds_eke.sel(time=slice('1958-01-01', '2024-12-31'))
print(f"  After truncation: {len(ds_eke_truncated.time)}")

# Save truncated version (overwrite)
ds_eke_truncated.to_netcdf(eke_path + ".tmp")
ds_eke.close()
ds_eke_truncated.close()

import os
os.replace(eke_path + ".tmp", eke_path)
print("  >> EKE truncated and saved")
```

```{python}
import xarray as xr
import numpy as np

PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
OUTPUT_DIR = "E:/ERA5_data/processed/raw_reynolds_decomposition"

original_storm_path = f"{PROCESSED_DATA_DIR}/ERA5_storminess_index_monthly.nc"
storm_path = f"{OUTPUT_DIR}/ERA5_storminess_index_monthly_raw.nc"

print("Comparing storminess indices...")

ds_orig = xr.open_dataset(original_storm_path, engine='netcdf4')
ds_new = xr.open_dataset(storm_path, engine='netcdf4')

# Align time periods (use only overlapping period)
ds_orig = ds_orig.sel(time=slice('1958-01-01', '2024-12-31'))
ds_new = ds_new.sel(time=slice('1958-01-01', '2024-12-31'))

print(f"Original time steps: {len(ds_orig.time)}")
print(f"New time steps: {len(ds_new.time)}")

# Calculate domain-averaged time series
orig_ts = ds_orig['storminess_index'].mean(dim=['latitude', 'longitude']).values
new_ts = ds_new['storminess_index'].mean(dim=['latitude', 'longitude']).values

# Statistics
orig_mean = float(ds_orig['storminess_index'].mean())
new_mean = float(ds_new['storminess_index'].mean())
correlation = np.corrcoef(orig_ts, new_ts)[0, 1]
relative_diff = (new_mean - orig_mean) / orig_mean * 100

print(f"\nResults:")
print(f"  Original (with deseasonalization) mean: {orig_mean:.4f}")
print(f"  Filter-only (this test) mean:           {new_mean:.4f}")
print(f"  Relative difference:                    {relative_diff:.2f}%")
print(f"  Temporal correlation:                   {correlation:.6f}")

if correlation > 0.99:
    print("\n>> CONCLUSION: Deseasonalization appears REDUNDANT")
    print("   The two methods produce nearly identical results.")
elif correlation > 0.95:
    print("\n>> CONCLUSION: Methods show HIGH agreement")
else:
    print("\n>> CONCLUSION: Methods show NOTABLE differences")

ds_orig.close()
ds_new.close()

```

# Modified workflow

```{python}
# ============================================================================
# STEP 0: COMBINE RAW ERA5 FILES INTO YEARLY FILES (Skip final merge)
# ============================================================================
import xarray as xr
import os
from glob import glob
from datetime import datetime

DATA_DIRECTORY = "E:/ERA5_data/raw"
PROCESSED_DATA_DIR = "E:/ERA5_data/processed"
os.makedirs(PROCESSED_DATA_DIR, exist_ok=True)

def log(msg):
    print(f"[{datetime.now().strftime('%H:%M:%S')}] {msg}")

ENCODING = {
    'u': {'zlib': True, 'complevel': 4, 'dtype': 'float32'},
    'v': {'zlib': True, 'complevel': 4, 'dtype': 'float32'}
}

monthly_files = sorted(glob(os.path.join(DATA_DIRECTORY, "era5_850hpa_panarctic_*.nc")))
log(f"Found {len(monthly_files)} monthly files")

# Check latitude range
with xr.open_dataset(monthly_files[0]) as test:
    lat_max = float(test.latitude.max())
    log(f"Max latitude: {lat_max}°N")

log("Processing year by year...")

for year in range(1958, 2026):
    year_pattern = os.path.join(DATA_DIRECTORY, f"era5_850hpa_panarctic_{year}-*.nc")
    year_files = sorted(glob(year_pattern))
    
    if not year_files:
        continue
    
    yearly_output = os.path.join(PROCESSED_DATA_DIR, f"ERA5_panarctic_{year}.nc")
    
    if os.path.exists(yearly_output):
        log(f"Skipping {year} (already exists)")
        continue
    
    log(f"Processing {year} ({len(year_files)} files)...")
    
    ds = xr.open_mfdataset(year_files, combine='by_coords', parallel=False)
    ds = ds.rename({'valid_time': 'time'})
    
    if 'pressure_level' in ds.dims:
        ds = ds.squeeze('pressure_level', drop=True)
    
    # EXCLUDE 90°N to avoid ECMWF pole issue
    ds = ds.sel(latitude=ds.latitude[ds.latitude < 90])
    
    ds.to_netcdf(yearly_output, encoding=ENCODING)
    ds.close()

log("✓ Done! Yearly files ready.")
log("To use: xr.open_mfdataset('ERA5_panarctic_*.nc', combine='by_coords')")
```
