---
title: "Storminess_analysis_full_framework"
author: Francesco Lucano
format: html
engine: knitr
editor: visual
warning: false
error: false
toc: true
---

```{r setup-environment}
# ============================================================================
# SETUP: ENVIRONMENT, LIBRARIES, PATHS, AND GLOBAL PARAMETERS
# Arctic Storminess Climatology | University of Bern
# ============================================================================
#
# PURPOSE:
#   Single configuration block for all analytical phases.
#   Every path, variable name, colour, and statistical parameter used
#   downstream is defined here exactly once.
#
# ============================================================================

# ── 1. LIBRARIES ─────────────────────────────────────────────────────────────
# Core infrastructure
library(here)          # Portable file paths relative to project root
library(tidyverse)     # dplyr, ggplot2, tidyr, purrr, readr, stringr, forcats
library(glue)          # String interpolation for diagnostic messages
library(lubridate)     # Date/time manipulation

# NetCDF and spatial data
library(ncdf4)         # Read/write NetCDF files
library(ncmeta)        # NetCDF metadata inspection
library(stars)         # Spatio-temporal arrays (NetCDF → R)
library(terra)         # Raster/vector spatial operations
library(sf)            # Simple features (vector geometry)
library(rnaturalearth) # Natural Earth coastlines and boundaries
library(abind)         # Array binding across dimensions

# Time series and statistics
library(zoo)           # Rolling window functions, irregular time series
library(boot)          # tsboot() — Moving Blocks Bootstrap
library(zyp)           # Pre-Whitened Mann-Kendall + Sen's slope
library(blocklength)   # pwsd() — adaptive MBB block length selection
library(trend)         # Additional trend tests (Spearman, Pettitt)
library(modifiedmk)    # Modified Mann-Kendall variants (for reference)
library(nlme)          # Generalised Least Squares (GLS) with AR errors

# Visualisation
library(ggplot2)       # Core plotting (loaded via tidyverse, explicit for clarity)
library(patchwork)     # Multi-panel figure composition
library(viridis)       # Perceptually uniform colour scales
library(RColorBrewer)  # Categorical and sequential colour palettes
library(corrplot)      # Correlation matrix visualisation
library(ggcorrplot)    # ggplot2-based correlation matrices
library(fields)        # Colour image plots (spatial diagnostics)


# Parallelisation and progress
library(future)        # Parallel backend declaration
library(furrr)         # Parallel map functions (purrr + future)
library(foreach)       # For-each parallel loops
library(doParallel)    # Parallel backend for foreach
library(progressr)     # Progress bars compatible with future/furrr
library(pbapply)       # Progress-bar apply functions
library(tictoc)        # Execution time measurement

# Reporting
library(knitr)         # kable() tables in Quarto/RMarkdown output

# Python interoperability
library(reticulate)    # R–Python bridge


# ── 2. PYTHON ENVIRONMENT ─────────────────────────────────────────────────────
# The Python pipeline (Steps 1–3) was executed in the 'arctic-env' virtual
# environment. Reticulate is configured to use the same environment for any
# downstream Python calls (e.g., reading NetCDF metadata via xarray).
use_virtualenv("arctic-env", required = TRUE)


# ── 3. FILE PATHS ─────────────────────────────────────────────────────────────
# All paths are constructed relative to the project root via here::here().
# Update DATA_DIRECTORY if the processed data folder is located elsewhere.

# Root directories
DATA_DIRECTORY  <- here::here("Data", "processed")
OUTPUT_DATA_DIR <- here::here("Data", "processed", "analysis_ready")
FIGURES_DIR     <- here::here("Figures")

# ---- Input NetCDF files (from Python pipeline) ----
METRICS_FILE    <- file.path(DATA_DIRECTORY,
                             "ERA5_storminess_metrics_monthly.nc")
TKE_FILE        <- file.path(DATA_DIRECTORY, "ERA5_TKE_monthly.nc")
MKE_FILE        <- file.path(DATA_DIRECTORY, "ERA5_MKE_monthly.nc")
EKE_BUDGET_FILE <- file.path(DATA_DIRECTORY, "ERA5_EKE_budget_monthly.nc")

# ---- Prepared R data (from Phase 1) ----
PHASE1_DATA     <- file.path(OUTPUT_DATA_DIR, "Phase1_Prepared_Data.RData")

# ---- Phase-specific figure directories (created on demand) ----
FIG_PHASE1 <- file.path(FIGURES_DIR, "Phase1_Diagnostics")
FIG_PHASE2 <- file.path(FIGURES_DIR, "Phase2_Energy_Budget")
FIG_PHASE3 <- file.path(FIGURES_DIR, "Phase3_Trends")
FIG_PHASE4 <- file.path(FIGURES_DIR, "Phase4_Regional")
FIG_PHASE5 <- file.path(FIGURES_DIR, "Phase5_Synthesis")

for (d in c(OUTPUT_DATA_DIR,
            FIG_PHASE1, FIG_PHASE2, FIG_PHASE3,
            FIG_PHASE4, FIG_PHASE5)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}


# ── 4. VARIABLE DEFINITIONS ───────────────────────────────────────────────────
# These names must match the NetCDF variable names exactly (post-rename).
# eke_mean was renamed to eke_synoptic_mean to reflect its physical meaning:
# the monthly mean of the 1.2-6 day bandpass-filtered EKE (EKE_synoptic_bar,
# eq. 2.12 in the thesis), distinct from eke_budget (eq. 2.9).

METRIC_NAMES <- c(
  "eke_synoptic_mean",  # PRIMARY: Monthly mean synoptic-scale EKE (eq. 2.12)
  "eke_p90",            # 90th percentile of 6-hourly EKE within each month
  "eke_p95",            # 95th percentile
  "eke_p99",            # 99th percentile (extreme storm intensity)
  "exceed_p95_freq"     # Fraction of timesteps exceeding 1961-1990 P95 threshold
)

ENERGY_VARS <- c(
  "tke",                # Total Kinetic Energy (eq. 2.7)
  "mke",                # Mean Kinetic Energy  (eq. 2.8)
  "eke_budget"          # Budget-derived EKE, all sub-monthly variance (eq. 2.9)
)

ALL_VARS <- c(METRIC_NAMES, ENERGY_VARS)

# Labels for figures (native plotmath)
VAR_LABELS <- c(
  "eke_synoptic_mean" = "Mean~EKE[synoptic]~(m^2/s^2)",
  "eke_p90"           = "P90~EKE~(m^2/s^2)",
  "eke_p95"           = "P95~EKE~(m^2/s^2)",
  "eke_p99"           = "P99~EKE~(m^2/s^2)",
  "exceed_p95_freq"   = "P95~Exceedance~Frequency",
  "tke"               = "TKE~(m^2/s^2)",
  "mke"               = "MKE~(m^2/s^2)",
  "eke_budget"        = "EKE[budget]~(m^2/s^2)"
)

# Simple labels for table output (plain text, no Unicode superscripts)
VAR_LABELS_PLAIN <- c(
  "eke_synoptic_mean" = "Mean EKE_synoptic",
  "eke_p90"           = "P90 EKE",
  "eke_p95"           = "P95 EKE",
  "eke_p99"           = "P99 EKE",
  "exceed_p95_freq"   = "P95 Exceedance Freq.",
  "tke"               = "TKE",
  "mke"               = "MKE",
  "eke_budget"        = "EKE_budget"
)

# Short plotmath labels (no units) — for embedding variable names in compound
# plot titles, e.g. paste0('"STL Decomposition: "~', VAR_LABELS_SHORT[["eke_synoptic_mean"]])
# Strip everything from the first "~(" onward (removes the "(m^2/s^2)" suffix).
VAR_LABELS_SHORT <- sub("~\\(.*\\)$", "", VAR_LABELS)


# ── 5. ANALYSIS PARAMETERS ────────────────────────────────────────────────────

# Spatial domain
ARCTIC_LAT_THRESHOLD <- 60    # Core Arctic domain: >= 60°N

# Temporal boundaries
ANALYSIS_START_YEAR  <- 1957  # December 1957 is first monthly observation
ANALYSIS_END_YEAR    <- 2025
ANNUAL_START_YEAR    <- 1958  # Annual means start 1958 (full calendar years only)
                               # See Phase 1 §1.6.3 boundary note

# Reference period for climatological thresholds (exceedance frequency)
# Chosen to precede onset of Arctic Amplification (Overland et al. 2004;
# Chylek et al. 2022) and rapid sea-ice decline (Stroeve & Notz 2018).
REFERENCE_PERIOD <- c(1961, 1990)

# Seasons (DJF attribution: December of year N → DJF of year N+1)
SEASONS      <- c("DJF", "MAM", "JJA", "SON")
SEASON_NAMES <- c(
  "DJF" = "Winter (DJF)",
  "MAM" = "Spring (MAM)",
  "JJA" = "Summer (JJA)",
  "SON" = "Autumn (SON)"
)

# Statistical parameters (Phase 3)
ALPHA      <- 0.05    # Significance level for all hypothesis tests
N_BOOT     <- 2000    # MBB bootstrap replicates (2000 balances precision/speed)
DECADES    <- 10      # Slope conversion factor: per year → per decade
B_FALLBACK <- 3L      # MBB block length fallback when pwsd() returns b < 1
                      # b = 3 exceeds typical annual ACF decorrelation time
                      # while preserving ~23 non-overlapping blocks (N/b, N=68)


# ── 6. COLOUR PALETTES ────────────────────────────────────────────────────────
# Colour-blind safe palette (Wong 2011, Nature Methods).
# Storminess metrics: blue family. Energy components: warm family.

METRIC_COLORS <- c(
  "eke_synoptic_mean" = "#0072B2",  # Blue
  "eke_p90"           = "#56B4E9",  # Sky blue
  "eke_p95"           = "#009E73",  # Teal green
  "eke_p99"           = "#9467BD",  # Purple
  "exceed_p95_freq"   = "#D55E00"   # Vermillion
)

ENERGY_COLORS <- c(
  "tke"        = "#1B9E77",   # Green
  "mke"        = "#D95F02",   # Orange
  "eke_budget" = "#7570B3"    # Periwinkle
)

ALL_COLORS <- c(METRIC_COLORS, ENERGY_COLORS)

SEASON_COLORS <- c(
  "DJF" = "#4575B4",  # Blue   — winter
  "MAM" = "#91CF60",  # Green  — spring
  "JJA" = "#FC8D59",  # Orange — summer
  "SON" = "#D73027"   # Red    — autumn (hypothesis season)
)


# ── 7. REGIONAL DOMAIN DEFINITIONS ───────────────────────────────────────────
# Sub-regions for Phase 4 analysis. Defined as named lists for easy indexing.
# Coordinate convention: (lon_min, lon_max, lat_min, lat_max)

REGIONS <- list(
  PanArctic = list(
    label   = "Pan-Arctic (\u226560\u00b0N)",
    lat_min = 60,  lat_max = 90,
    lon_min = -180, lon_max = 180
  ),
  FramStrait = list(
    label   = "Fram Strait (76\u00b0N\u201381\u00b0N)",
    lat_min = 76,  lat_max = 81,
    lon_min = -10,  lon_max = 20
  ),
  WestSvalbard = list(
    label   = "West Svalbard (10\u00b0E\u201316\u00b0E, 76\u00b0N\u201381\u00b0N)",
    lat_min = 76,  lat_max = 81,
    lon_min = 10,   lon_max = 16
  ),
  EastSvalbard = list(
    label   = "East Svalbard (16\u00b0E\u201328\u00b0E, 76\u00b0N\u201381\u00b0N)",
    lat_min = 76,  lat_max = 81,
    lon_min = 16,   lon_max = 28
  )
)


# ── 8. VERIFICATION ───────────────────────────────────────────────────────────
# Confirm that all required input files exist before any phase is run.
# This prevents cryptic downstream errors from missing files.

required_files <- list(
  "Storminess metrics NetCDF" = METRICS_FILE,
  "TKE monthly NetCDF"        = TKE_FILE,
  "MKE monthly NetCDF"        = MKE_FILE,
  "EKE budget monthly NetCDF" = EKE_BUDGET_FILE,
  "Phase 1 prepared RData"    = PHASE1_DATA
)

cat("Checking required input files:\n")
all_ok <- TRUE
for (desc in names(required_files)) {
  path   <- required_files[[desc]]
  exists <- file.exists(path)
  cat(sprintf("  %s  %s\n",
              ifelse(exists, "\u2713", "\u2717"),
              desc))
  if (!exists) all_ok <- FALSE
}

if (!all_ok) {
  warning(paste(
    "\nOne or more required files are missing.",
    "Run the Python pipeline (Steps 1-3) and Phase 1 before proceeding.",
    sep = "\n"
  ))
} else {
  cat("\n\u2713 All input files present. Environment ready.\n")
}

cat(sprintf("\nProject root:     %s\n", here::here()))
cat(sprintf("Data directory:   %s\n", DATA_DIRECTORY))
cat(sprintf("Figures root:     %s\n", FIGURES_DIR))
cat(sprintf("Analysis period:  Dec %d – Dec %d (817 months)\n",
            ANALYSIS_START_YEAR, ANALYSIS_END_YEAR))
cat(sprintf("Annual trends:    %d – %d (N = %d years)\n",
            ANNUAL_START_YEAR, ANALYSIS_END_YEAR,
            ANALYSIS_END_YEAR - ANNUAL_START_YEAR + 1))
cat(sprintf("Reference period: %d – %d\n",
            REFERENCE_PERIOD[1], REFERENCE_PERIOD[2]))
cat(sprintf("Significance:     \u03b1 = %.2f\n", ALPHA))
```

-   PHASE 1: Data Preparation & Validation

    -   1.1 Data loading and standardization

    -   1.2 Domain refinement (\>=60°N)

    -   1.3 Area-weighted time series derivation

    -   1.4 Energy budget closure verification

    -   1.5 Diagnostic pipeline (applied uniformly to all variables):

        -   1.5.1 Deseasonalisation

        -   1.5.2 Autocorrelation assessment (ACF/PACF)

        -   1.5.3 Effective sample size (neff)

        -   1.5.4 Variance decomposition (STL, primary metric only)

-   PHASE 2: Energy budget characterisation

    -   2.1 Partitioning climatology (fEKE, MKE/EKE ratio)

    -   2.2 Seasonal cycle of energy components

    -   2.3 EKE definition comparison (budget vs synoptic, r = 0.885)

-   PHASE 3: Trend analysis

    -   3.1 Pan-Arctic trend analysis (all storminess metrics + energy components, simultaneously)

    -   3.2 Seasonal decomposition (all metrics by DJF/MAM/JJA/SON)

    -   3.3 Autumn intensification hypothesis test

    -   3.4 Spatial trend patterns + field significance

    -   3.5 Period comparison (pre- vs. post-regime shift, if warranted by 2.1–2.3)

-   PHASE 4: Regional Sub-Analysis (apply full Phase 3-4 workflow to each region)

    -   4.1 Fram Strait sector

    -   4.2 West Svalbard sector (10°E–16°E, 76°N–81°N)

    -   4.3 East Svalbard sector (16°E–28°E, 76°N–81°N)

-   PHASE 5: Synthesis and hypothesis testing

    -   5.1 Summary of all trend results

    -   5.2 Autumn feedback loop hypothesis evaluation

    -   5.3 Comparison with sea-ice loss timelines
    


# Phase 1: Foundational Analysis

This chapter details the foundational data processing and diagnostic procedures applied to the monthly Pan-Arctic storminess index for the period 1958–2025. The objective of this phase was to prepare a robust, spatially-aggregated time series for the core Arctic region and to formally assess its statistical properties. This assessment is a mandatory prerequisite for the subsequent trend and change-point analyses, as the validity of those advanced methods depends entirely on the rigorous characterization of the data's underlying structure.

## 1.1 Data Loading and Standardization

The analysis commenced with the `ERA5_storminess_metrics_monthly.nc` file, a NetCDF dataset containing five complementary storminess metrics derived from the bandpass-filtered synoptic-scale Eddy Kinetic Energy (EKE) calculated in Section 2.5 of the Methodology. Initial inspection revealed the data arrays were structured with dimensions of `[longitude, latitude, time]`, an unconventional ordering for climatological time series analysis.

The dataset contains the following variables:

```         
Variables:
  eke_synoptic_mean [m²/s²]              - Monthly mean synoptic-scale EKE (PRIMARY)
  eke_p90 [m²/s²]               - Monthly 90th percentile EKE
  eke_p95 [m²/s²]               - Monthly 95th percentile EKE
  eke_p99 [m²/s²]               - Monthly 99th percentile EKE
  exceed_p95_freq [fraction]    - Fraction exceeding climatological P95
  eke_p95_climatology [m²/s²]   - Reference P95 threshold (1961-1990)
```

To align with standard conventions and optimize computational efficiency for temporal operations, the array dimensions were permuted to a `[time, latitude, longitude]` structure. The time coordinate was decoded from its "days since origin" format and verified to correctly span the full analysis period from January 1958 through December 2025.

### 1.1.1 Rationale for Multi-Metric Approach

Following the methodological framework established by Yau and Chang (2020) and the IPCC AR6 WG1 Chapter 11 recommendations for extreme event analysis, a suite of complementary metrics was calculated to capture different aspects of storminess variability:

1.  **Mean EKE** ($\overline{EKE}$): The primary metric, validated by Yau and Chang (2020) as achieving the highest correlation with surface weather impacts among Eulerian storm track metrics. This represents the average synoptic-scale storm activity within each month.

2.  **Percentile Metrics** (P90, P95, P99): These capture within-month extreme storm intensity at different severity levels. A month characterized by several intense storms interspersed with calm periods will exhibit elevated percentile values even if mean EKE is moderate, providing sensitivity to episodic high-intensity events most relevant to coastal erosion (Irrgang et al. 2022).

3.  **Exceedance Frequency**: This metric detects changes in extreme storm frequency over time, independent of changes in mean storminess. By comparing each timestep against a fixed climatological threshold (P95 from 1961–1990), it reveals whether extreme events are becoming more or less common relative to the pre-Arctic Amplification baseline. This approach follows the rationale established by Zhang et al. (2011), fixing the exceedance probability during the base period to create a static reference that allows monitoring of physical shifts in extreme storm probability.

This multi-metric approach allows detection of changes in the storm intensity distribution that would be obscured by analyzing mean values alone.

```{r phase1}
# ============================================================================
# BOUNDARY NOTE: December 1957 and the Annual Aggregation Rule
# ============================================================================
#
# The analysis period begins on 1 December 1957. This extension was made
# deliberately to ensure that DJF 1958 (Dec 1957 + Jan 1958 + Feb 1958)
# constitutes a complete seasonal unit, yielding N = 68 for all four seasons.
#
# December 1957 is a fully valid monthly observation, its filtered EKE
# values are reliable, having been protected by the November 1957 filter
# warm-up buffer. It is therefore INCLUDED in:
#
#   storm_df            (817-row monthly data frame)
#   All monthly analyses: descriptive stats, ACF/PACF, STL, neff
#   Seasonal aggregation: Dec 1957 is attributed to DJF 1958
#     via storm_df$season_year and correctly counted in that season
#
# However, December 1957 is EXCLUDED from annual means because a "1957 annual mean" would be computed from a     #  single December value is not
#
# The filter year >= 1958 applied below enforces this rule. Annual means
# therefore span 1958–2025 (N = 68 complete years), consistent with all
# thesis captions and the stated study period.
#
# This distinction does NOT affect the seasonal analysis: storm_df retains
# all 817 months, and the season_year convention ensures DJF 1958 is
# complete with N = 3 months.
# ============================================================================

# ============================================================================
# PHASE 1: DATA LOADING AND EXPLORATORY ANALYSIS
# Arctic Storminess Climatology (1957-12 to 2025-12)
# ============================================================================
cat("\n", rep("=", 70), "\n")
cat("PHASE 1: FOUNDATIONAL ANALYSIS (Corrected Metrics)\n")
cat(rep("=", 70), "\n\n")

# --- Check if file exists ---
if (!file.exists(METRICS_FILE)) {
  stop(glue("
    ERROR: Metrics file not found!
    Expected: {METRICS_FILE}
    
    Please run the corrected Step 3 Python script first:
    Step3_Storminess_Metrics_Corrected.py
  "))
}

cat("Loading storminess metrics file:\n")
cat(glue("  {METRICS_FILE}\n\n"))

# --- Open NetCDF and inspect structure ---
nc <- nc_open(METRICS_FILE)

cat("NetCDF File Information:\n")
cat(rep("-", 40), "\n")

# Print dimensions
cat("\nDimensions:\n")
for (dim_name in names(nc$dim)) {
  cat(sprintf("  %s: %d\n", dim_name, nc$dim[[dim_name]]$len))
}

# Print variables
cat("\nVariables:\n")
for (var_name in names(nc$var)) {
  var_units <- ncatt_get(nc, var_name, "units")$value
  if (is.null(var_units) || var_units == "") var_units <- "dimensionless"
  cat(sprintf("  %s [%s]\n", var_name, var_units))
}

# --- Extract coordinate variables ---
time_raw  <- ncvar_get(nc, "time")
lat       <- ncvar_get(nc, "latitude")
lon       <- ncvar_get(nc, "longitude")

# Report the raw time units string for transparency
time_units <- ncatt_get(nc, "time", "units")$value
cat(sprintf("\nTime units (raw): %s\n", time_units))

# ── TIME COORDINATE RECONSTRUCTION ──────────────────────────────────────────
# The NetCDF time units string contains a floating-point encoding artefact
# produced by xarray's pandas-based timestamp encoding:
#   "days since 1957-12-31 23:59:59.999999999"
# R cannot parse sub-nanosecond origin strings reliably via ncdf4. The time
# coordinate is therefore reconstructed analytically from first principles:
# 817 monthly steps beginning December 1957.
#
# Verification: seq from 1957-12-01 by month for 817 steps ends 2025-12-01.
# ( 1957-12 → 2025-12 = 68 years × 12 months + 1 = 817 months ) 
# ----------------------------------------------------------------------------

N_MONTHS_EXPECTED <- nc$dim$time$len   # Read directly from file (=817)

time_dates <- seq.Date(
  from       = as.Date("1957-12-01"),
  by         = "month",
  length.out = N_MONTHS_EXPECTED
)

# Sanity check
cat(sprintf("\nTime coordinate reconstructed analytically (%d months):\n",
            length(time_dates)))
cat(sprintf("  First: %s\n", format(time_dates[1])))
cat(sprintf("  Last:  %s\n", format(time_dates[length(time_dates)])))

if (format(time_dates[length(time_dates)]) != "2025-12-01") {
  stop(sprintf(
    "ERROR: Final date is %s, expected 2025-12-01.\n%s",
    format(time_dates[length(time_dates)]),
    "Check N_MONTHS_EXPECTED and the analysis start date."
  ))
}
cat("Date sequence verified\n")

# --- Extract all metric arrays ---
cat("\nLoading metric arrays...\n")
metrics_list <- list()
for (metric in METRIC_NAMES) {
  if (metric %in% names(nc$var)) {
    metrics_list[[metric]] <- ncvar_get(nc, metric)
    cat(sprintf("%s: dim = [%s]\n",
                metric,
                paste(dim(metrics_list[[metric]]), collapse = " x ")))
  } else {
    cat(sprintf("%s: NOT FOUND in file\n", metric))
  }
}

# Also load climatological threshold if present
if ("eke_p95_climatology" %in% names(nc$var)) {
  eke_p95_clim <- ncvar_get(nc, "eke_p95_climatology")
  cat(sprintf("eke_p95_climatology: dim = [%s]\n",
              paste(dim(eke_p95_clim), collapse = " x ")))
}

nc_close(nc)
```

## 1.2 Domain Refinement to the Core Arctic

The original dataset extended south to 55°N latitude. To focus the analysis specifically on the high-Arctic and minimize the influence of mid-latitude dynamics, the spatial domain was refined to include only latitudes greater than or equal to 60°N. This resulted in refined metric arrays with dimensions of \[817, 120, 1440\] (time × latitude × longitude), covering a latitude range from 60.00°N to 89.75°N. The 90°N grid point was excluded during the initial data processing due to documented ECMWF issues with spherical harmonic interpolation at the pole singularity, which produces erroneous near-zero wind values in the native ERA5 grid. All subsequent analyses were performed on this core Arctic domain.

The 60°N threshold was selected to:

1.  Capture the locations of the most intense Arctic cyclones, primarily over the Norwegian, Greenland, and Barents Seas as well as the central Arctic Ocean (Vessey et al. 2022)

2.  Include all major Arctic coastlines relevant to erosion studies

3.  Focus on the region where Arctic-specific feedback mechanisms (e.g., sea-ice loss amplification) are expected to be most pronounced

```{r}
# ----------------------------------------------------------------------------
# SECTION 1.2: DIMENSION ORDERING AND TIME DECODING
# ----------------------------------------------------------------------------
# dim(storminess) = [360, 35, 804] so, [lon, lat, time] instead of the correct [lat, lon, time]
                  #  ↓    ↓    ↓
                  # lon  lat  time

cat("\n", rep("-", 40), "\n")
cat("Checking dimension ordering...\n")

# Determine current dimension order
# Expected from Python script: [time, latitude, longitude]
# But R/NetCDF may read as [lon, lat, time] depending on how it was written

sample_metric <- metrics_list[[1]]
n_dims <- length(dim(sample_metric))

if (n_dims != 3) {
 stop("ERROR: Expected 3D array [time, lat, lon]")
}

# Infer dimension order from sizes
dim_sizes <- dim(sample_metric)
n_time_expected <- length(time_raw)
n_lat_expected <- length(lat)
n_lon_expected <- length(lon)

cat(sprintf("  Array dimensions: [%s]\n", paste(dim_sizes, collapse = ", ")))
cat(sprintf("  Expected: time=%d, lat=%d, lon=%d\n", 
           n_time_expected, n_lat_expected, n_lon_expected))

# Check if transposition is needed
if (dim_sizes[1] == n_lon_expected && dim_sizes[3] == n_time_expected) {
 cat("  Detected order: [lon, lat, time] → Transposing to [time, lat, lon]\n")
 
 for (metric in names(metrics_list)) {
   metrics_list[[metric]] <- aperm(metrics_list[[metric]], c(3, 2, 1))
 }
 cat("All metrics transposed\n")
 
} else if (dim_sizes[1] == n_time_expected) {
 cat("Correct order: [time, lat, lon]\n")
 
} else {
 warning("Unexpected dimension ordering - please verify manually")
}

# Verify final dimensions
cat(sprintf("\nFinal array dimensions: [%s]\n",
           paste(dim(metrics_list[[1]]), collapse = " x ")))

# time_dates was reconstructed analytically in §1.1 (seq.Date from 1957-12-01).
# Parsing the raw time_units string is unreliable for this file — see §1.1 note.
cat(sprintf("  time_dates: %s → %s (%d months)\n",
            format(time_dates[1]), format(time_dates[length(time_dates)]),
            length(time_dates)))
```

```{r}
# ============================================================================
# SECTION 1.2: REFINING THE PAN-ARCTIC DOMAIN to >= 60°N
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat(sprintf("Refining domain to Core Arctic (>= %d°N)...\n", ARCTIC_LAT_THRESHOLD))

# Original domain
cat(sprintf("  Original latitude range: %.2f°N to %.2f°N\n", min(lat), max(lat)))

# Find indices for Arctic subset
arctic_lat_indices <- which(lat >= ARCTIC_LAT_THRESHOLD)
lat_arctic <- lat[arctic_lat_indices]

cat(sprintf("  New latitude range: %.2f°N to %.2f°N\n", 
           min(lat_arctic), max(lat_arctic)))
cat(sprintf("  Latitude points: %d → %d\n", 
           length(lat), length(arctic_lat_indices)))

# Subset all metrics to Arctic domain
metrics_arctic <- list()
for (metric in names(metrics_list)) {
 # Subset: keep all time, selected latitudes, all longitudes
 metrics_arctic[[metric]] <- metrics_list[[metric]][, arctic_lat_indices, ]
}

cat(sprintf("All metrics subsetted to Arctic domain\n"))
cat(sprintf("New array dimensions: [%s]\n", 
           paste(dim(metrics_arctic[[1]]), collapse = " x ")))
```

## 1.3 Derivation of the Area-Weighted Pan-Arctic Time Series

To analyze the overall storminess of the core Arctic, the spatial data were aggregated into single representative time series for each metric. A simple arithmetic mean across all grid cells is statistically invalid for geophysical data on a regular latitude-longitude grid, as it fails to account for the convergence of meridians toward the pole. This grants disproportionate weight to the physically smaller high-latitude grid cells, introducing a significant bias into the domain average (Peixoto and Oort 1992).

To correct for this distortion, an area-weighted average was calculated for each of the 816 monthly time steps. The weight for each grid cell, $w_j$, is proportional to the cosine of its latitude, $\phi_j$:

$$w_j = \cos(\phi_j)$$

The area-weighted metric value, $\overline{X_t}$, at each time step $t$ was then calculated as:

$$\overline{X_t} = \frac{\sum_{i=1}^{N_{lon}} \sum_{j=1}^{N_{lat}} X_{i,j,t} \cdot w_j}{\sum_{i=1}^{N_{lon}} \sum_{j=1}^{N_{lat}} w_j}$$

where $X_{i,j,t}$ is the metric value at longitude $i$, latitude $j$, and time $t$.

A diagnostic check confirmed the necessity of this procedure by comparing weighted and unweighted spatial means.

The consistent positive bias for EKE metrics indicates that unweighted means systematically underestimate the domain-average storminess. This occurs because cosine weighting assigns greater influence to lower Arctic latitudes (near 60°N), where the North Atlantic storm track produces higher EKE values. The bias magnitude increases with percentile level because extreme values exhibit greater spatial heterogeneity. The final output of this step is a set of monthly time series objects in R, which serve as the basis for all subsequent temporal analyses.

```{r}
# ============================================================================
# SECTION 1.3 Create Area-Weighted Time Series
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Creating area-weighted Pan-Arctic time series...\n")

# --- Calculate cosine-latitude weights ---
# Weight = cos(latitude in radians)
weights_arctic <- cos(lat_arctic * pi / 180)

# Create weight matrix matching spatial dimensions [lat, lon]
n_lat_arctic <- length(lat_arctic)
n_lon <- length(lon)

weights_matrix <- matrix(weights_arctic,
                        nrow = n_lat_arctic,
                        ncol = n_lon,
                        byrow = FALSE)

# --- Calculate area-weighted mean for each metric ---
ts_weighted <- list()
ts_unweighted <- list()  # For diagnostic comparison

for (metric in names(metrics_arctic)) {
 data_array <- metrics_arctic[[metric]]
 
 # Apply weighted mean across spatial dimensions for each timestep
 ts_weighted[[metric]] <- apply(data_array, 1, function(slice) {
   weighted.mean(slice, weights_matrix, na.rm = TRUE)
 })
 
 # Also calculate unweighted for comparison
 ts_unweighted[[metric]] <- apply(data_array, 1, mean, na.rm = TRUE)
 
 # Report bias
 bias <- mean(ts_weighted[[metric]]) - mean(ts_unweighted[[metric]])
 cat(sprintf("  %s: weighted mean = %.4f, bias = %.4f\n", 
             metric, mean(ts_weighted[[metric]]), bias))
}

cat("\n Area-weighting applied to all metrics\n")
cat("  (Non-zero bias confirms area-weighting is necessary)\n")
```

```{r}
# ============================================================================
# SECTION 1.3: CREATE TIME SERIES OBJECTS AND DATA FRAMES
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Creating time series objects...\n")

# --- Create R ts objects (for trend package functions) ---
ts_objects <- list()
for (metric in names(ts_weighted)) {
 ts_objects[[metric]] <- ts(ts_weighted[[metric]], 
                            start = c(year(time_dates[1]), month(time_dates[1])),
                            frequency = 12)
}

# --- Create master data frame for ggplot and analysis ---
storm_df <- data.frame(
 date = time_dates,
 year = year(time_dates),
 month = month(time_dates),
 season = case_when(
   month(time_dates) %in% c(12, 1, 2) ~ "DJF",
   month(time_dates) %in% c(3, 4, 5) ~ "MAM",
   month(time_dates) %in% c(6, 7, 8) ~ "JJA",
   month(time_dates) %in% c(9, 10, 11) ~ "SON"
 )
)

# Add all metrics as columns
for (metric in names(ts_weighted)) {
 storm_df[[metric]] <- ts_weighted[[metric]]
}

# Create season_year for proper DJF grouping
# (December of year N belongs to DJF of year N+1)
storm_df$season_year <- storm_df$year
storm_df$season_year[storm_df$month == 12] <- storm_df$season_year[storm_df$month == 12] + 1

cat(sprintf("  ✓ Created data frame with %d rows and %d columns\n", 
           nrow(storm_df), ncol(storm_df)))
```

## 1.4 Assessment of Temporal Autocorrelation

This diagnostic is a critical prerequisite for the entire statistical framework. Standard statistical tests, including Ordinary Least Squares (OLS) regression, operate on the strict assumption that data points (or model residuals) are independent. Climate time series violate this assumption due to persistence, or "memory," arising from the inertia of the climate system.

### 1.4.1 Deseasonalization Requirement

As demonstrated by the STL decomposition (Section 1.4.3), the seasonal cycle accounts for 70.9% of total variance in the Pan-Arctic storminess time series. To obtain valid estimates of atmospheric persistence, autocorrelation must be calculated on deseasonalized anomalies, monthly values minus the climatological mean for that specific calendar month (calculated over the 1961–1990 reference period), rather than raw data.

### 1.4.2 Autocorrelation and Partial Autocorrelation Functions

To quantify the degree of persistence in the deseasonalized Pan-Arctic storminess metrics, both the Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF) were computed for lags up to 36 months. The ACF measures the correlation between observations separated by $k$ time steps:

$$\rho_k = \frac{\text{Cov}(X_t, X_{t-k})}{\text{Var}(X_t)}$$

The PACF measures the correlation at lag $k$ after removing the linear dependence on intermediate lags, providing insight into the appropriate autoregressive order for error modelling.

### 1.4.3 Effective Sample Size

When autocorrelation is present, the effective number of independent observations is smaller than the nominal sample size. Following Zwiers and von Storch (1995), the effective sample size was calculated as:

$$n_{eff} = \frac{N}{1 + 2\sum_{k=1}^{K}\left(1 - \frac{k}{N}\right)\rho_k}$$

where $N$ is the original sample size and $\rho_k$ is the autocorrelation at lag $k$.

All lag-1 values exceed the threshold of $\rho_1 > 0.1$ that indicates violation of the independence assumption for monthly atmospheric data (Santer et al. 2000), confirming that autocorrelation-corrected methods are required. The modest lag-1 autocorrelation ($\rho_1 \approx 0.15$) reflects genuine atmospheric memory, likely arising from slower-varying boundary forcing such as sea surface temperatures and sea ice extent that impart persistence to monthly-mean atmospheric statistics.

The PACF exhibits a significant spike at lag 1 followed by rapid decay toward the confidence bounds, confirming that an AR(1) error structure is appropriate for GLS regression analyses.

### 1.4.4 Variance Decomposition

To further characterize the temporal structure, Seasonal-Trend decomposition using LOESS (STL) was applied to the primary metric (mean EKE).

```{r}
# ============================================================================
# Section 1.4: Assessment of Temporal Autocorrelation (CORRECTED)
# Calculate ACF on DESEASONALIZED ANOMALIES, not raw data
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Assessing temporal autocorrelation on DESEASONALIZED data...\n\n")

# ----------------------------------------------------------------------------
# Step 1: Create deseasonalized anomalies for each metric
# ----------------------------------------------------------------------------

# Function to deseasonalize by removing monthly climatology
deseasonalize <- function(df, metric_col) {
  df %>%
    group_by(month) %>%
    mutate(anomaly = .data[[metric_col]] - mean(.data[[metric_col]], na.rm = TRUE)) %>%
    ungroup() %>%
    pull(anomaly)
}

# Create anomalies for all metrics
storm_df_anom <- storm_df %>%
  mutate(
    eke_synoptic_mean_anom = deseasonalize(., "eke_synoptic_mean"),
    eke_p90_anom = deseasonalize(., "eke_p90"),
    eke_p95_anom = deseasonalize(., "eke_p95"),
    eke_p99_anom = deseasonalize(., "eke_p99"),
    exceed_p95_freq_anom = deseasonalize(., "exceed_p95_freq")
  )

# Create ts objects for anomalies
# Use time_dates[1] so start matches ts_objects (Dec 1957, not Jan 1958)
ts_start_anom <- c(year(time_dates[1]), month(time_dates[1]))
ts_objects_anom <- list(
  eke_synoptic_mean = ts(storm_df_anom$eke_synoptic_mean_anom, start = ts_start_anom, frequency = 12),
  eke_p90           = ts(storm_df_anom$eke_p90_anom,           start = ts_start_anom, frequency = 12),
  eke_p95           = ts(storm_df_anom$eke_p95_anom,           start = ts_start_anom, frequency = 12),
  eke_p99           = ts(storm_df_anom$eke_p99_anom,           start = ts_start_anom, frequency = 12),
  exceed_p95_freq   = ts(storm_df_anom$exceed_p95_freq_anom,   start = ts_start_anom, frequency = 12)
)

# ----------------------------------------------------------------------------
# Step 2: Calculate ACF on anomalies and compare with raw
# ----------------------------------------------------------------------------

acf_results <- list()
acf_comparison <- data.frame()

for (metric in names(ts_objects)) {
  ts_raw <- ts_objects[[metric]]
  ts_anom <- ts_objects_anom[[metric]]
  
  # ACF on raw data (for comparison)
  acf_raw <- acf(ts_raw, lag.max = 24, plot = FALSE)
  rho1_raw <- acf_raw$acf[2]
  
  # ACF on deseasonalized anomalies (CORRECT approach)
  acf_anom <- acf(ts_anom, lag.max = 24, plot = FALSE)
  rho1_anom <- acf_anom$acf[2]
  
  acf_results[[metric]] <- acf_anom
  
  # Calculate effective sample size using ANOMALY autocorrelation
  N <- length(ts_anom)
  rho_k <- acf_anom$acf[-1]  # All lags from 1 to 24
  k <- 1:length(rho_k)
  neff <- N / (1 + 2 * sum((1 - k/N) * pmax(rho_k, 0)))  # Only positive correlations
  
  # Store comparison
  acf_comparison <- rbind(acf_comparison, data.frame(
    metric = metric,
    rho1_raw = rho1_raw,
    rho1_anomaly = rho1_anom,
    inflation_factor = rho1_raw / rho1_anom,
    N = N,
    neff = neff,
    dof_reduction_pct = (1 - neff/N) * 100
  ))
  
  cat(sprintf("--- %s ---\n", metric))
  cat(sprintf("  ρ₁ (raw data):        %.3f  [INFLATED by seasonality]\n", rho1_raw))
  cat(sprintf("  ρ₁ (anomalies):       %.3f  [TRUE atmospheric memory]\n", rho1_anom))
  cat(sprintf("  Inflation factor:     %.2fx\n", rho1_raw / rho1_anom))
  cat(sprintf("  Effective sample size: %.1f (%.1f%% reduction)\n", neff, (1 - neff/N) * 100))
  cat("\n")
}

print(knitr::kable(acf_comparison, digits = 3, format = "simple"))
```

```{r}
# ============================================================================
# SECTION 1.4: DESCRIPTIVE STATISTICS
# ============================================================================

cat(rep("-", 40), "\n")
cat("Descriptive Statistics (Pan-Arctic >= 60°N):\n")
cat(rep("-", 40), "\n\n")

# Create summary table
summary_stats <- data.frame(
 Metric = character(),
 Units = character(),
 Mean = numeric(),
 SD = numeric(),
 Min = numeric(),
 Max = numeric(),
 Lag1_ACF = numeric(),
 stringsAsFactors = FALSE
)

for (metric in names(ts_weighted)) {
 ts_data <- ts_weighted[[metric]]
 rho1 <- acf_results[[metric]]$acf[2]
 
 # Determine units
 if (metric == "exceed_p95_freq") {
   units <- "fraction (0-1)"
 } else {
   units <- "m²/s²"
 }
 
 summary_stats <- rbind(summary_stats, data.frame(
   Metric = metric,
   Units = units,
   Mean = mean(ts_data, na.rm = TRUE),
   SD = sd(ts_data, na.rm = TRUE),
   Min = min(ts_data, na.rm = TRUE),
   Max = max(ts_data, na.rm = TRUE),
   Lag1_ACF = rho1
 ))
}

print(knitr::kable(summary_stats, digits = 4, format = "simple"))
```

## 1.5 Verification of Exceedance Frequency Calculation

As an internal consistency check, the exceedance frequency metric was validated against its theoretical expectation. By definition, the 95th percentile threshold was calculated from the 1961–1990 reference period; therefore, approximately 5% of timesteps during this period should exceed the threshold.

The verification yielded:

```         
Reference period: 1961-1990
Mean exceedance frequency: 0.0500 (5.00%)
Expected (by definition of P95): ~0.05 (5%)
✓ VERIFICATION PASSED
```

The perfect agreement confirms correct implementation of the percentile threshold calculation and validates the exceedance frequency metric for subsequent trend analysis. Deviations from 5% in other time periods will indicate genuine changes in extreme storm frequency relative to the climatological baseline.

```{r}
# ============================================================================
# SECTION 1.5: VERIFICATION - EXCEEDANCE FREQUENCY CHECK
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Verification: Exceedance Frequency During Reference Period\n")
cat(rep("-", 40), "\n\n")

if ("exceed_p95_freq" %in% names(storm_df)) {
 # Filter to reference period (1981-2010)
 ref_period_idx <- storm_df$year >= REFERENCE_PERIOD[1] & 
                   storm_df$year <= REFERENCE_PERIOD[2]
 
 ref_exceed_freq <- mean(storm_df$exceed_p95_freq[ref_period_idx], na.rm = TRUE)
 
 cat(sprintf("Reference period: %d-%d\n", REFERENCE_PERIOD[1], REFERENCE_PERIOD[2]))
 cat(sprintf("Mean exceedance frequency: %.4f (%.2f%%)\n", 
             ref_exceed_freq, ref_exceed_freq * 100))
 cat(sprintf("Expected (by definition of P95): ~0.05 (5%%)\n"))
 
 if (abs(ref_exceed_freq - 0.05) < 0.01) {
   cat("✓ VERIFICATION PASSED: Exceedance frequency is approximately 5%%\n")
 } else {
   cat("⚠ WARNING: Exceedance frequency deviates from expected 5%%\n")
   cat("  This may indicate an issue with the P95 climatology calculation\n")
 }
} else {
 cat("  (exceed_p95_freq not available for verification)\n")
}
```

## Intermediate Summary of Phase 1 Outputs

The foundational analysis produced the following validated datasets for subsequent phases:

| Output | Description | Dimensions |
|------------------------|------------------------|------------------------|
| `metrics_arctic` | Gridded monthly metrics (Arctic subset) | \[816 × 120 × 1440\] per metric |
| `ts_objects` | Area-weighted monthly time series | 816 values per metric |
| `storm_df` | Master data frame with date/season columns | 816 rows × 10 columns |
| `acf_results` | ACF diagnostics for each metric | Lags 0–24 |

: Summary of Phase 1 output objects {#tbl-outputs}

**Key Statistical Properties:**

| Metric          | Mean        | SD          | Lag-1 ACF |
|-----------------|-------------|-------------|-----------|
| eke_synoptic_mean        | 12.27 m²/s² | 2.80 m²/s²  | 0.153     |
| eke_p90         | 29.47 m²/s² | 6.72 m²/s²  | 0.149     |
| eke_p95         | 40.75 m²/s² | 9.18 m²/s²  | 0.152     |
| eke_p99         | 66.39 m²/s² | 14.82 m²/s² | 0.160     |
| exceed_p95_freq | 0.0504      | 0.0218      | 0.125     |

: Descriptive statistics for Pan-Arctic storminess metrics {#tbl-descriptive}

-   **Temporal coverage**: 817 months (Decemeber 1957 – December 2025)
-   **Spatial domain**: 60.00°N – 89.75°N (Core Arctic, excluding pole)
-   **Autocorrelation**: All metrics exhibit $\rho_1>0.1$, requiring autocorrelation-corrected statistical methods

The formal characterization of the data autocorrelation structure establishes the statistical foundation for all subsequent trend and hypothesis testing analyses. The multi-metric approach enables independent assessment of changes in average storminess (mean EKE), extreme storm intensity (percentiles), and extreme storm frequency (exceedance rate).

## 1.6 Energy Budget

```{r}
# ============================================================================
# SECTION 1.6: ENERGY BUDGET INTEGRATION
# ============================================================================
#
# PURPOSE:
#   Load the TKE, MKE, and EKE_budget monthly fields (computed in the
#   Python pipeline, Step 2), spatially aggregate them using the same
#   cosine-latitude weighting as the storminess metrics, verify budget
#   closure, and run the full diagnostic pipeline (deseasonalisation,
#   ACF/PACF, n_eff) on the energy components.
#
# INPUTS:
#   - ERA5_TKE_monthly.nc     
#   - ERA5_MKE_monthly.nc       
#   - ERA5_EKE_budget_monthly.nc 
#   - lat_arctic, lon            
#   - storm_df                   
#   - deseasonalize()            
#
# OUTPUTS:
#   - energy_df: monthly data frame with tke, mke, eke_budget, 
#                mke_eke_ratio, eke_fraction
#   - energy_annual: annual means of all energy components
#   - acf_results_energy: ACF diagnostics for energy components
#   - acf_comparison_energy: summary table (rho1, n_eff, DoF reduction)
#   - 3 diagnostic figures 
#
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("SECTION 1.5: ENERGY BUDGET INTEGRATION\n")
cat(rep("=", 70), "\n\n")


# --------------------------------------------------------------------------
# 1.6.1  Load and Spatially Aggregate Energy Components
# --------------------------------------------------------------------------
# Each NetCDF contains a single variable on the same lat/lon grid as the
# storminess metrics. We apply the identical cosine-latitude weighting
# (eq. 2.21–2.22 in the methodology) to produce Pan-Arctic time series.
# --------------------------------------------------------------------------

cat("--- 1.5.1 Loading Energy Budget Components ---\n\n")

# TKE_FILE, MKE_FILE, EKE_BUDGET_FILE are already defined in the setup-environment chunk.
# Check all three files exist before proceeding
energy_files <- list(TKE = TKE_FILE, MKE = MKE_FILE, EKE_budget = EKE_BUDGET_FILE)
for (name in names(energy_files)) {
  if (!file.exists(energy_files[[name]])) {
    stop(sprintf("Energy budget file not found: %s\n  Expected: %s", 
                 name, energy_files[[name]]))
  }
  cat(sprintf("Found: %s\n", basename(energy_files[[name]])))
}

# Function to load a single energy component NetCDF and return an
# area-weighted Pan-Arctic (≥60°N) monthly time series.
#
# This replicates the exact same aggregation procedure used for the
# storminess metrics in §1.3, ensuring that all variables entering the
# diagnostic pipeline have been processed identically.

load_energy_component <- function(filepath, varname, lat_threshold = 60) {
  
  nc <- nc_open(filepath)
  data_raw <- ncvar_get(nc, varname)
  lat_nc   <- ncvar_get(nc, "latitude")
  lon_nc   <- ncvar_get(nc, "longitude")
  nc_close(nc)
  
  # Handle dimension ordering: NetCDF may store as [lon, lat, time]
  # We need [time, lat, lon] for consistent indexing
  if (dim(data_raw)[1] == length(lon_nc) && dim(data_raw)[3] != length(lon_nc)) {
    data_array <- aperm(data_raw, c(3, 2, 1))  # [time, lat, lon]
  } else {
    data_array <- data_raw
  }
  
  # Subset to Arctic domain (≥ lat_threshold)
  arctic_idx   <- which(lat_nc >= lat_threshold)
  data_arctic  <- data_array[, arctic_idx, ]
  lat_subset   <- lat_nc[arctic_idx]
  
  # Cosine-latitude weights (same as eq. 2.21 in methodology)
  weights     <- cos(lat_subset * pi / 180)
  weights_mat <- matrix(weights, nrow = length(lat_subset), ncol = length(lon_nc))
  
  # Area-weighted spatial mean for each month (eq. 2.22)
  ts_weighted <- apply(data_arctic, 1, function(slice) {
    weighted.mean(slice, weights_mat, na.rm = TRUE)
  })
  
  return(ts_weighted)
}

# Load all three components
tke_ts        <- load_energy_component(TKE_FILE, "tke")
mke_ts        <- load_energy_component(MKE_FILE, "mke")
eke_budget_ts <- load_energy_component(EKE_BUDGET_FILE, "eke_budget")

cat(sprintf("\n  Loaded %d months of energy budget data\n", length(tke_ts)))

# Verify temporal alignment with storminess data
stopifnot(
  "Energy budget length must match storminess data" = 
    length(tke_ts) == nrow(storm_df)
)
cat("Temporal alignment verified (both have 817 months)\n")
```

```{r}
# --------------------------------------------------------------------------
# 1.6.2  Budget Closure Verification
# --------------------------------------------------------------------------
# The Reynolds decomposition guarantees TKE = MKE + EKE exactly in theory.
# In practice, numerical precision of the computation introduces a tiny
# residual. We verify that |TKE - (MKE + EKE)| / TKE < 0.01%.
# --------------------------------------------------------------------------

cat("\n--- 1.5.2 Verifying Energy Budget Closure ---\n\n")

budget_residual <- tke_ts - (mke_ts + eke_budget_ts)
relative_error  <- mean(abs(budget_residual)) / mean(tke_ts) * 100

cat(sprintf("  Mean TKE:         %.3f m²/s²\n", mean(tke_ts)))
cat(sprintf("  Mean MKE:         %.3f m²/s²\n", mean(mke_ts)))
cat(sprintf("  Mean EKE_budget:  %.3f m²/s²\n", mean(eke_budget_ts)))
cat(sprintf("  Relative closure error: %.6f%%\n", relative_error))

if (relative_error < 0.01) {
  cat("BUDGET CLOSURE VERIFIED (error < 0.01%)\n")
} else {
  cat("WARNING: Budget error exceeds tolerance\n")
}
```

```{r}
# --------------------------------------------------------------------------
# 1.6.3  Merge Energy Components into Master Data Frame
# --------------------------------------------------------------------------
# We add TKE, MKE, EKE_budget and two derived partitioning metrics to
# storm_df, so that all variables are in a single data frame for
# subsequent analysis.
#
# Derived metrics:
#   f_EKE       = EKE_budget / TKE     (eddy fraction, eq. 2.X in methodology)
#   mke_eke_ratio = MKE / EKE_budget   (mean-to-eddy ratio)
# --------------------------------------------------------------------------

cat("\n--- 1.5.3 Merging into Master Data Frame ---\n")

storm_df <- storm_df %>%
  mutate(
    tke           = tke_ts,
    mke           = mke_ts,
    eke_budget    = eke_budget_ts,
    eke_fraction  = eke_budget_ts / tke_ts,
    mke_eke_ratio = mke_ts / eke_budget_ts
  )

cat(sprintf("storm_df now has %d columns\n", ncol(storm_df)))
cat(sprintf("New columns: tke, mke, eke_budget, eke_fraction, mke_eke_ratio\n"))

# Also create annual means (needed for some diagnostics and later for Phase 2)
energy_annual <- storm_df %>%
  group_by(year) %>%
  summarise(
    tke           = mean(tke),
    mke           = mean(mke),
    eke_budget    = mean(eke_budget),
    eke_synoptic_mean      = mean(eke_synoptic_mean),
    mke_eke_ratio = mean(mke_eke_ratio),
    eke_fraction  = mean(eke_fraction),
    .groups       = "drop"
  ) %>%
  filter(year >= 1958)  # exclude Dec 1957 — see boundary note below

cat(sprintf("energy_annual: %d years\n", nrow(energy_annual)))
```

```{r}
# --------------------------------------------------------------------------
# 1.6.4  Diagnostic Pipeline: Energy Components
# --------------------------------------------------------------------------
# Apply the SAME diagnostic procedure used for storminess metrics in §1.4:
#   1. Deseasonalise (subtract monthly climatology)
#   2. Compute ACF/PACF on anomalies
#   3. Calculate effective sample size (n_eff)
#
# This ensures that when Phase 2 applies trend tests to these variables,
# the autocorrelation structure has been characterised and the tests can
# be appropriately configured.
#
# NOTE: The deseasonalize() function was defined in §1.4 and is already
# available in the environment.
# --------------------------------------------------------------------------

cat("\n--- 1.5.4 Diagnostic Pipeline: Energy Components ---\n\n")
cat("Assessing temporal autocorrelation on DESEASONALIZED energy data...\n\n")

# Energy metrics to diagnose (ENERGY_VARS is defined in the setup-environment chunk)
ENERGY_METRICS <- ENERGY_VARS

# Step 1: Create deseasonalised anomalies
storm_df <- storm_df %>%
  mutate(
    tke_anom        = deseasonalize(., "tke"),
    mke_anom        = deseasonalize(., "mke"),
    eke_budget_anom = deseasonalize(., "eke_budget")
  )

# Create ts objects for anomalies (needed for acf/pacf)
# Use time_dates[1] so start matches the storminess ts objects (Dec 1957)
ts_start_energy <- c(year(time_dates[1]), month(time_dates[1]))
ts_objects_energy_anom <- list(
  tke        = ts(storm_df$tke_anom,        start = ts_start_energy, frequency = 12),
  mke        = ts(storm_df$mke_anom,        start = ts_start_energy, frequency = 12),
  eke_budget = ts(storm_df$eke_budget_anom, start = ts_start_energy, frequency = 12)
)

# Also create raw ts objects (for inflation factor comparison)
ts_objects_energy_raw <- list(
  tke        = ts(storm_df$tke,        start = ts_start_energy, frequency = 12),
  mke        = ts(storm_df$mke,        start = ts_start_energy, frequency = 12),
  eke_budget = ts(storm_df$eke_budget, start = ts_start_energy, frequency = 12)
)

# Step 2: ACF/PACF + n_eff (identical procedure to §1.4)
acf_results_energy    <- list()
acf_comparison_energy <- data.frame()

for (metric in ENERGY_METRICS) {
  
  ts_raw  <- ts_objects_energy_raw[[metric]]
  ts_anom <- ts_objects_energy_anom[[metric]]
  
  # ACF on raw data (for comparison — will show seasonal inflation)
  acf_raw  <- acf(ts_raw, lag.max = 24, plot = FALSE)
  rho1_raw <- acf_raw$acf[2]
  
  # ACF on deseasonalised anomalies (CORRECT approach)
  acf_anom  <- acf(ts_anom, lag.max = 24, plot = FALSE)
  rho1_anom <- acf_anom$acf[2]
  
  acf_results_energy[[metric]] <- acf_anom
  
  # Effective sample size (eq. 2.25 in methodology)
  N     <- length(ts_anom)
  rho_k <- acf_anom$acf[-1]           # All lags from 1 to 24
  k     <- 1:length(rho_k)
  neff  <- N / (1 + 2 * sum((1 - k/N) * pmax(rho_k, 0)))
  
  # Store results
  acf_comparison_energy <- rbind(acf_comparison_energy, data.frame(
    metric            = metric,
    rho1_raw          = rho1_raw,
    rho1_anomaly      = rho1_anom,
    inflation_factor  = rho1_raw / rho1_anom,
    N                 = N,
    neff              = neff,
    dof_reduction_pct = (1 - neff / N) * 100
  ))
  
  cat(sprintf("--- %s ---\n", metric))
  cat(sprintf("  ρ₁ (raw data):    %.3f  [INFLATED by seasonality]\n", rho1_raw))
  cat(sprintf("  ρ₁ (anomalies):   %.3f  [TRUE atmospheric memory]\n", rho1_anom))
  cat(sprintf("  Inflation factor: %.2fx\n", rho1_raw / rho1_anom))
  cat(sprintf("  Effective sample size: %.1f (%.1f%% reduction)\n\n", 
              neff, (1 - neff / N) * 100))
}

# Print summary table
cat(rep("-", 40), "\n")
cat("Energy Component Autocorrelation Diagnostics:\n")
cat(rep("-", 40), "\n\n")
print(knitr::kable(acf_comparison_energy, digits = 3, format = "simple"))
```

```{r}
# ============================================================================
# Section 1.6.5: SAVE PREPARED DATA
# ============================================================================
cat("\n", rep("-", 40), "\n")
cat("Saving prepared data objects...\n")

# OUTPUT_DATA_DIR and PHASE1_DATA are defined in the setup-environment chunk.
# The output directory is also created there, so no dir.create() needed here.

save(
  # Coordinate data
  lat, lon, lat_arctic, time_dates,

  # Area-weighted time series (storminess metrics)
  ts_weighted, ts_objects,

  # Master data frame (includes storminess metrics + energy columns:
  # tke, mke, eke_budget, eke_fraction, mke_eke_ratio, and _anom columns)
  storm_df,

  # Annual means (energy + synoptic EKE)
  energy_annual,

  # ACF diagnostics — storminess metrics
  acf_results,
  acf_comparison,

  # ACF diagnostics — energy components
  acf_results_energy,
  acf_comparison_energy,

  # Summary statistics
  summary_stats,

  # Energy budget closure error (%)
  relative_error,

  # Climatological threshold (2D, small)
  eke_p95_clim,

  # Configuration
  ARCTIC_LAT_THRESHOLD, REFERENCE_PERIOD,

  file = PHASE1_DATA
)
cat(sprintf("Saved to: %s\n", PHASE1_DATA))

write_csv(storm_df, file.path(OUTPUT_DATA_DIR, "storm_metrics_panarctic_monthly.csv"))
cat(sprintf("Saved CSV: %s\n", file.path(OUTPUT_DATA_DIR, "storm_metrics_panarctic_monthly.csv")))
```

```{r}
# ============================================================================
# PHASE 1: DIAGNOSTIC PLOTS
# Arctic Storminess Climatology (1958-2025)
# 
# PURPOSE: Generate all foundational diagnostic figures for Phase 1
#
# PLOTS GENERATED:
#   1. Time series overview (all metrics)
#   2. ACF plots (all metrics)
#   3. Seasonal boxplots
#   4. Climatological seasonal cycle (12-month)
#   5. STL decomposition
#   6. Spatial climatology maps (polar projection)
#   7. Distribution plots (histograms + density)
#   8. Metric correlation matrix
#   9. Anomaly time series (deseasonalized)
#  10. Seasonal cycle by decade (stability check)
#
# INPUT: Phase1_Prepared_Data.RData
# ============================================================================

# ============================================================================
# SETUP
# ============================================================================

# Load Phase 1 data
# PHASE1_DATA is defined in the setup-environment chunk — do NOT reassign DATA_DIRECTORY here.
if (!file.exists(PHASE1_DATA)) {
  stop("Phase 1 data not found. Run Phase 1 script first.")
}
load(PHASE1_DATA)

cat("Loaded Phase 1 prepared data\n")

# Figure output directory (FIG_PHASE1 defined and created by the setup-environment chunk)
fig_dir <- FIG_PHASE1

# Colour aliases — plotting code uses lowercase names; values come from the setup chunk
# so palettes stay consistent across all phases.
metric_colors <- METRIC_COLORS
season_colors <- SEASON_COLORS

# Metrics to plot (METRIC_NAMES defined in the setup-environment chunk)
METRICS <- METRIC_NAMES

# ============================================================================
# PLOT 1: TIME SERIES OVERVIEW (ALL METRICS)
# ============================================================================

cat("\n--- Generating Plot 1: Time Series Overview ---\n")

# Prepare data
ts_long <- storm_df %>%
  select(date, all_of(METRICS)) %>%
  pivot_longer(-date, names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = METRICS))

p1 <- ggplot(ts_long, aes(x = date, y = value, color = metric)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1,
             labeller = as_labeller(VAR_LABELS, default = label_parsed)) +
  scale_color_manual(values = metric_colors, guide = "none") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Pan-Arctic (≥60°N) Storminess Metrics: Monthly Time Series",
    subtitle = "1958–2025 | Raw data (light) with LOESS smoother (bold)",
    x = NULL,
    y = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", hjust = 0),
    strip.background = element_rect(fill = "gray95"),
    panel.grid.minor = element_blank()
  )

ggsave(file.path(fig_dir, "Fig01_TimeSeries_AllMetrics.png"),
       p1, width = 12, height = 14, dpi = 300)

# ============================================================================
# PLOT 2: ACF AND PACF PLOTS (ON DESEASONALIZED ANOMALIES)
# ============================================================================

cat("\n--- Generating Plot 2: ACF and PACF Plots (Anomalies) ---\n")

png(file.path(fig_dir, "Fig02_ACF_PACF_Anomalies.png"),
    width = 12, height = 14, units = "in", res = 300)

par(mfrow = c(5, 2), 
    mar = c(4, 4, 3, 1),
    oma = c(0, 0, 3, 0))

for (metric in METRICS) {
  if (metric %in% names(ts_objects_anom)) {
    ts_anom <- ts_objects_anom[[metric]]
    
    # Get rho1 values for annotation
    rho1_anom <- acf(ts_anom, lag.max = 1, plot = FALSE)$acf[2]
    pacf1_anom <- pacf(ts_anom, lag.max = 36, plot = FALSE)$acf[1]
    
    # Build plotmath titles from the setup-chunk VAR_LABELS_SHORT lookup
    acf_title  <- parse(text = paste0('"ACF: "~',  VAR_LABELS_SHORT[[metric]], '~"(anomalies)"'))[[1]]
    pacf_title <- parse(text = paste0('"PACF: "~', VAR_LABELS_SHORT[[metric]], '~"(anomalies)"'))[[1]]

    # ACF plot
    acf(ts_anom, lag.max = 36,
        main = acf_title,
        xlab = "Lag (months)",
        ylab = "Autocorrelation",
        col = metric_colors[metric],
        lwd = 2)
    mtext(bquote(rho[1] == .(sprintf("%.3f", rho1_anom))),
          side = 3, line = -1.5, adj = 0.95, cex = 0.8)

    # PACF plot
    pacf(ts_anom, lag.max = 36,
         main = pacf_title,
         xlab = "Lag (months)",
         ylab = "Partial Autocorrelation",
         col = metric_colors[metric],
         lwd = 2)
    mtext(bquote(phi[1] == .(sprintf("%.3f", pacf1_anom))),
          side = 3, line = -1.5, adj = 0.95, cex = 0.8)
  }
}

mtext("Autocorrelation Functions: Deseasonalized Anomalies (1958–2025)",
      outer = TRUE, line = 1, cex = 1.2, font = 2)

dev.off()

# ============================================================================
# PLOT 3: SEASONAL BOXPLOTS
# ============================================================================

cat("\n--- Generating Plot 3: Seasonal Boxplots ---\n")

# Prepare data with proper season ordering
storm_df_season <- storm_df %>%
  mutate(season = factor(season, levels = c("DJF", "MAM", "JJA", "SON")))

# Create boxplots for primary metrics
p3a <- ggplot(storm_df_season, aes(x = season, y = eke_synoptic_mean, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = parse(text = VAR_LABELS[["eke_synoptic_mean"]])[[1]],
       y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3b <- ggplot(storm_df_season, aes(x = season, y = eke_p95, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = parse(text = VAR_LABELS[["eke_p95"]])[[1]],
       y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3c <- ggplot(storm_df_season, aes(x = season, y = eke_p99, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = parse(text = VAR_LABELS[["eke_p99"]])[[1]],
       y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3d <- ggplot(storm_df_season, aes(x = season, y = exceed_p95_freq, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = parse(text = VAR_LABELS[["exceed_p95_freq"]])[[1]],
       y = "Fraction", x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3_combined <- (p3a | p3b | p3c | p3d) +
  plot_annotation(
    title = "Seasonal Distribution of Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 | Red dashed line = climatological 5% threshold",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig03_Seasonal_Boxplots.png"),
       p3_combined, width = 14, height = 5, dpi = 300)

# ============================================================================
# PLOT 4: CLIMATOLOGICAL SEASONAL CYCLE (12-MONTH)
# ============================================================================

cat("\n--- Generating Plot 4: Climatological Seasonal Cycle ---\n")

# Calculate monthly climatology (mean for each calendar month)
monthly_clim <- storm_df %>%
  group_by(month) %>%
  summarise(
    eke_synoptic_mean_avg = mean(eke_synoptic_mean, na.rm = TRUE),
    eke_synoptic_mean_sd = sd(eke_synoptic_mean, na.rm = TRUE),
    eke_p95_avg = mean(eke_p95, na.rm = TRUE),
    eke_p95_sd = sd(eke_p95, na.rm = TRUE),
    exceed_freq_avg = mean(exceed_p95_freq, na.rm = TRUE),
    exceed_freq_sd = sd(exceed_p95_freq, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(month_name = month.abb[month])

# Plot for Mean EKE
p4a <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = eke_synoptic_mean_avg - eke_synoptic_mean_sd, 
                  ymax = eke_synoptic_mean_avg + eke_synoptic_mean_sd),
              fill = "#0072B2", alpha = 0.3) +
  geom_line(aes(y = eke_synoptic_mean_avg), color = "#0072B2", linewidth = 1.2) +
  geom_point(aes(y = eke_synoptic_mean_avg), color = "#0072B2", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = parse(text = VAR_LABELS[["eke_synoptic_mean"]])[[1]],
       y = expression("m"^2*"/s"^2),
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Plot for P95 EKE
p4b <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = eke_p95_avg - eke_p95_sd, 
                  ymax = eke_p95_avg + eke_p95_sd),
              fill = "#009E73", alpha = 0.3) +
  geom_line(aes(y = eke_p95_avg), color = "#009E73", linewidth = 1.2) +
  geom_point(aes(y = eke_p95_avg), color = "#009E73", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = parse(text = VAR_LABELS[["eke_p95"]])[[1]],
       y = expression("m"^2*"/s"^2),
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Plot for Exceedance Frequency
p4c <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = exceed_freq_avg - exceed_freq_sd,
                  ymax = exceed_freq_avg + exceed_freq_sd),
              fill = "#D55E00", alpha = 0.3) +
  geom_line(aes(y = exceed_freq_avg), color = "#D55E00", linewidth = 1.2) +
  geom_point(aes(y = exceed_freq_avg), color = "#D55E00", size = 3) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = parse(text = VAR_LABELS[["exceed_p95_freq"]])[[1]],
       y = "Fraction",
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p4_combined <- (p4a | p4b | p4c) +
  plot_annotation(
    title = "Climatological Seasonal Cycle of Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 mean ± 1 SD",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig04_Seasonal_Cycle_Climatology.png"),
       p4_combined, width = 14, height = 5, dpi = 300)

# ============================================================================
# PLOT 5: STL DECOMPOSITION
# ============================================================================

cat("\n--- Generating Plot 5: STL Decomposition ---\n")

# STL decomposition for primary metric (eke_synoptic_mean)
stl_result <- stl(ts_objects$eke_synoptic_mean, s.window = "periodic")

# Extract components
stl_df <- data.frame(
  date = time_dates,
  observed = as.vector(ts_objects$eke_synoptic_mean),
  seasonal = as.vector(stl_result$time.series[, "seasonal"]),
  trend = as.vector(stl_result$time.series[, "trend"]),
  remainder = as.vector(stl_result$time.series[, "remainder"])
)

# Calculate component contributions
var_total <- var(stl_df$observed, na.rm = TRUE)
var_seasonal <- var(stl_df$seasonal, na.rm = TRUE)
var_trend <- var(stl_df$trend, na.rm = TRUE)
var_remainder <- var(stl_df$remainder, na.rm = TRUE)

pct_seasonal <- (var_seasonal / var_total) * 100
pct_trend <- (var_trend / var_total) * 100
pct_remainder <- (var_remainder / var_total) * 100

cat(sprintf("  Variance decomposition:\n"))
cat(sprintf("    Seasonal: %.1f%%\n", pct_seasonal))
cat(sprintf("    Trend:    %.1f%%\n", pct_trend))
cat(sprintf("    Residual: %.1f%%\n", pct_remainder))

# Create STL plot
stl_long <- stl_df %>%
  pivot_longer(-date, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component, 
                       levels = c("observed", "seasonal", "trend", "remainder")),
    component_label = case_when(
      component == "observed" ~ sprintf("Observed Data"),
      component == "seasonal" ~ sprintf("Seasonal (%.1f%% of variance)", pct_seasonal),
      component == "trend" ~ sprintf("Trend (%.1f%% of variance)", pct_trend),
      component == "remainder" ~ sprintf("Remainder (%.1f%% of variance)", pct_remainder)
    ),
    component_label = factor(component_label, levels = unique(component_label))
  )

p5 <- ggplot(stl_long, aes(x = date, y = value)) +
  geom_line(color = "#0072B2", linewidth = 0.4) +
  facet_wrap(~ component_label, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  labs(
    title = parse(text = paste0('"STL Decomposition: "~', VAR_LABELS_SHORT[["eke_synoptic_mean"]], '~"(Primary Metric)"'))[[1]],
    subtitle = "Pan-Arctic (≥60°N) | Seasonal-Trend decomposition using LOESS",
    x = NULL,
    y = expression("m"^2*"/s"^2)
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", hjust = 0),
    strip.background = element_rect(fill = "gray95"),
    panel.grid.minor = element_blank()
  )

ggsave(file.path(fig_dir, "Fig05_STL_Decomposition.png"),
       p5, width = 12, height = 10, dpi = 300)

# Also save STL for P95 (extreme intensity)
stl_p95 <- stl(ts_objects$eke_p95, s.window = "periodic")

stl_p95_df <- data.frame(
  date = time_dates,
  observed = as.vector(ts_objects$eke_p95),
  seasonal = as.vector(stl_p95$time.series[, "seasonal"]),
  trend = as.vector(stl_p95$time.series[, "trend"]),
  remainder = as.vector(stl_p95$time.series[, "remainder"])
)

# Store for later use
stl_results <- list(
  eke_synoptic_mean = stl_result,
  eke_p95 = stl_p95
)

# ============================================================================
# PLOT 6: SPATIAL CLIMATOLOGY MAPS (POLAR PROJECTION)
# Using proven methodology from spatial trend analysis
# NOTE: For visualization, we use the FULL dataset (all available latitudes)
#       even though the statistical analysis focuses on ≥60°N
# ============================================================================

cat("\n--- Generating Plot 6: Spatial Climatology Maps (Polar) ---\n")

# Calculate temporal mean at each grid point for each metric
# Use FULL domain (metrics_list) not Arctic subset (metrics_arctic)
# Result: [lat, lon] matrices
climatology_maps_full <- list()
for (metric in c("eke_synoptic_mean", "eke_p95", "exceed_p95_freq")) {
  if (metric %in% names(metrics_list)) {
    # Apply mean across time dimension (dim 1) - using FULL dataset
    climatology_maps_full[[metric]] <- apply(metrics_list[[metric]], c(2, 3), mean, na.rm = TRUE)
  }
}

cat(sprintf("  Climatology maps calculated (full domain: %.1f°N to %.1f°N)\n", 
            min(lat), max(lat)))

# ============================================================================
# STEP 1: CREATE A SPATIALLY-AWARE `stars` OBJECT (WITH WRAPPING)
# ============================================================================

# 1. Prepare Longitude Vector with Wrapping
# ----------------------------------------------------------------------------
# We extend the longitude vector by one point to close the gap at 180 deg
lon_vec <- as.vector(lon)
lat_vec <- as.vector(lat) 

# Create a new longitude vector that adds one point (start + 360)
lon_wrapped <- c(lon_vec, lon_vec[1] + 360)

cat(sprintf("  Wrapping longitude: Original %d points -> Wrapped %d points\n", 
            length(lon_vec), length(lon_wrapped)))

# 2. Helper function to wrap matrices
# ----------------------------------------------------------------------------
# Input: [lat, lon] matrix
# Output: [lon_wrapped, lat] matrix
wrap_data <- function(mat) {
  # Transpose to [lon, lat]
  mat_t <- t(mat)
  # Append the first row (first longitude) to the end
  mat_wrapped <- rbind(mat_t, mat_t[1, ])
  return(mat_wrapped)
}

# 3. Create Wrapped Data List
# ----------------------------------------------------------------------------
data_list_plot <- list(
  eke_synoptic_mean    = wrap_data(climatology_maps_full$eke_synoptic_mean),
  eke_p95     = wrap_data(climatology_maps_full$eke_p95),
  exceed_freq = wrap_data(climatology_maps_full$exceed_p95_freq)
)

# 4. Create Dimensions and Stars Object
# ----------------------------------------------------------------------------
# IMPORTANT: Use 'lon_wrapped' here, not the original 'lon_vec'
dims <- st_dimensions(x = lon_wrapped, y = lat_vec)

# Create the stars object and set CRS
clim_stars <- st_as_stars(data_list_plot, dimensions = dims) %>%
  st_set_crs(4326)

cat("  Stars object created (gap repaired)\n")

# ============================================================================
# STEP 2: REPROJECT ALL SPATIAL DATA
# ============================================================================

# Define the target projection: Lambert Azimuthal Equal-Area, centered on North Pole
arctic_crs <- "+proj=laea +lat_0=90 +lon_0=0"

# Warp the raster data to polar projection
clim_polar <- st_warp(clim_stars, crs = arctic_crs)

# Load and transform coastlines
world <- ne_countries(scale = "medium", returnclass = "sf")
world_polar <- st_transform(world, crs = arctic_crs)

cat("  Data reprojected to polar CRS\n")

# ============================================================================
# STEP 3: CREATE CARTOGRAPHIC ELEMENTS
# ============================================================================

# Define two radii for "two-ring" masking approach
view_radius <- 3700000  # 3700 km - data boundary
plot_radius <- 4400000  # 4400 km - space for labels

# Create circular mask
circle_mask <- st_sfc(st_point(c(0, 0)), crs = arctic_crs) %>%
  st_buffer(dist = view_radius)

# Apply mask to data
clim_polar_masked <- clim_polar[circle_mask]

# Create inverse mask (square with circular hole)
square_box <- st_as_sfc(st_bbox(c(
  xmin = -plot_radius, xmax = plot_radius,
  ymin = -plot_radius, ymax = plot_radius
), crs = arctic_crs))
inverse_mask <- st_difference(square_box, circle_mask)

# Create graticule lines
graticules <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = c(60, 70, 80),
  crs = 4326
) %>%
  st_transform(crs = arctic_crs)

# Create latitude labels (positioned at 180° longitude for visibility)
lat_labels <- data.frame(
  lon = 180,
  lat = c(60, 70, 80)
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = arctic_crs)
lat_labels$label <- c("60°N", "70°N", "80°N")

# Create longitude labels (positioned in the blank space between radii)
lon_labels <- data.frame(
  lon = seq(0, 300, by = 60),
  lat = 54  # Falls in blank space between view_radius and plot_radius
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = arctic_crs)
lon_labels$label <- c("0°", "60°E", "120°E", "180°", "120°W", "60°W")

cat("  Cartographic elements created\n")

# ==========================================================================
# STEP 4: Compute derived fields
# ==========================================================================

# --- 0a. Shared colour limits for Mean EKE and P95 EKE ---
shared_max <- max(
  max(clim_polar_masked[["eke_p95"]],  na.rm = TRUE),
  max(clim_polar_masked[["eke_synoptic_mean"]], na.rm = TRUE)
)
shared_max_round <- ceiling(shared_max / 10) * 10
shared_breaks    <- seq(0, shared_max_round, length.out = 5)

cat(sprintf("  Shared EKE colour limits: [0, %d]\n", shared_max_round))
cat(sprintf("  Legend breaks: %s\n", paste(shared_breaks, collapse = ", ")))

# --- 0b. Compute P95/Mean ratio and rebuild stars object ---
climatology_maps_full[["p95_mean_ratio"]] <-
  climatology_maps_full$eke_p95 / climatology_maps_full$eke_synoptic_mean

# Add wrapped ratio to data list
data_list_plot[["p95_mean_ratio"]] <- wrap_data(climatology_maps_full$p95_mean_ratio)

# Rebuild stars object with the extra attribute
dims <- st_dimensions(x = lon_wrapped, y = lat_vec)

clim_stars <- st_as_stars(data_list_plot, dimensions = dims) %>%
  st_set_crs(4326)

clim_polar        <- st_warp(clim_stars, crs = arctic_crs)
clim_polar_masked <- clim_polar[circle_mask]

ratio_range  <- range(clim_polar_masked[["p95_mean_ratio"]], na.rm = TRUE)
median_ratio <- median(clim_polar_masked[["p95_mean_ratio"]], na.rm = TRUE)
cat(sprintf("  P95/Mean ratio — median: %.2f, range: [%.2f, %.2f]\n",
            median_ratio, ratio_range[1], ratio_range[2]))


# FIGURE 6a: MEAN EKE + P95 EKE  (shared colour axis)
# ==========================================================================
cat("\n--- Generating Figure 6a: Mean EKE + P95 EKE ---\n")

# --- Panel (a): Mean EKE ---
p6a_mean <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = eke_synoptic_mean), na.rm = TRUE) +
  scale_fill_viridis(
    name   = expression(atop("EKE", "[m"^2*"/s"^2*"]")),
    option = "C",
    na.value = "transparent",
    limits = c(0, shared_max_round),
    breaks = shared_breaks,
    oob    = scales::squish
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label),
               size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label),
               size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius),
           ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = parse(text = paste0('"(a) "~', VAR_LABELS_SHORT[["eke_synoptic_mean"]]))[[1]]) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width  = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- Panel (b): P95 EKE ---
p6a_p95 <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = eke_p95), na.rm = TRUE) +
  scale_fill_viridis(
    name   = expression(atop("EKE", "[m"^2*"/s"^2*"]")),
    option = "C",
    na.value = "transparent",
    limits = c(0, shared_max_round),
    breaks = shared_breaks,
    oob    = scales::squish
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label),
               size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label),
               size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius),
           ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = parse(text = paste0('"(b) "~', VAR_LABELS_SHORT[["eke_p95"]]))[[1]]) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width  = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- Combine with shared legend ---
# guides = "collect" merges the two identical legends into one.
# Without it, patchwork would show two separate (but identical) colourbars.
fig6a <- p6a_mean + p6a_p95 +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title    = "Spatial Climatology of EKE Intensity (1958–2025)",
    subtitle = "Pan-Arctic (≥55°N) | Lambert Azimuthal Equal-Area Projection | Shared colour scale",
    theme = theme(
      plot.title    = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5)
    )
  )

print(fig6a)

ggsave(file.path(fig_dir, "Fig06a_EKE_Intensity_Climatology.png"),
       fig6a, width = 14, height = 7, dpi = 300, bg = "white")


# FIGURE 6b: P95/MEAN RATIO + EXCEEDANCE FREQUENCY
# ==========================================================================
cat("\n--- Generating Figure 6b: P95/Mean Ratio + Exceedance Frequency ---\n")

# --- Panel (a): P95/Mean EKE Ratio ---
p6b_ratio <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = p95_mean_ratio), na.rm = TRUE) +
  scale_fill_distiller(
    name      = expression(atop("P95 / Mean", "EKE ratio")),
    palette   = "RdYlBu",
    direction = -1,
    na.value  = "transparent",
    breaks    = scales::pretty_breaks(n = 4)
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label),
               size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label),
               size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius),
           ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = parse(text = paste0('"(a) P95 / Mean "~', VAR_LABELS_SHORT[["eke_synoptic_mean"]], '~"Ratio"'))[[1]]) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width  = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- Panel (b): P95 Exceedance Frequency ---
p6b_exceed <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = exceed_freq), na.rm = TRUE) +
  scale_fill_viridis(
    name     = "Exceedance\nFrequency",
    option   = "D",
    na.value = "transparent"
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label),
               size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label),
               size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius),
           ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = parse(text = paste0('"(b) "~', VAR_LABELS_SHORT[["exceed_p95_freq"]]))[[1]]) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width  = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- Combine ---
# No guides = "collect" here because the two panels have DIFFERENT scales.
fig6b <- p6b_ratio + p6b_exceed +
  plot_layout(ncol = 2) +
  plot_annotation(
    title    = "Spatial Climatology of Storminess Distribution Properties (1958–2025)",
    subtitle = "Pan-Arctic (≥55°N) | Lambert Azimuthal Equal-Area Projection",
    theme = theme(
      plot.title    = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5)
    )
  )

print(fig6b)

ggsave(file.path(fig_dir, "Fig06b_EKE_Distribution_Climatology.png"),
       fig6b, width = 14, height = 7, dpi = 300, bg = "white")


# Also save individual high-res versions
# ==========================================================================
ggsave(file.path(fig_dir, "Fig06a1_MeanEKE_Polar.png"),
       p6a_mean, width = 8, height = 8, dpi = 300, bg = "white")
ggsave(file.path(fig_dir, "Fig06a2_P95EKE_Polar.png"),
       p6a_p95, width = 8, height = 8, dpi = 300, bg = "white")
ggsave(file.path(fig_dir, "Fig06b1_P95MeanRatio_Polar.png"),
       p6b_ratio, width = 8, height = 8, dpi = 300, bg = "white")
ggsave(file.path(fig_dir, "Fig06b2_ExceedFreq_Polar.png"),
       p6b_exceed, width = 8, height = 8, dpi = 300, bg = "white")

# ============================================================================
# PLOT 7: DISTRIBUTION PLOTS (HISTOGRAMS + DENSITY)
# ============================================================================

cat("\n--- Generating Plot 7: Distribution Plots ---\n")

# Prepare data
dist_data <- storm_df %>%
  select(all_of(METRICS)) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = METRICS))

p7 <- ggplot(dist_data, aes(x = value, fill = metric)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.7, color = "white") +
  geom_density(color = "black", linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free", ncol = 3,
             labeller = as_labeller(VAR_LABELS, default = label_parsed)) +
  scale_fill_manual(values = metric_colors, guide = "none") +
  labs(
    title = "Distribution of Monthly Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 | n = 816 months",
    x = "Value",
    y = "Density"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold")
  )

ggsave(file.path(fig_dir, "Fig07_Distributions.png"),
       p7, width = 12, height = 8, dpi = 300)

# ============================================================================
# PLOT 8: METRIC CORRELATION MATRIX
# ============================================================================

cat("\n--- Generating Plot 8: Correlation Matrix ---\n")

# Calculate correlation matrix — keep METRICS as row/col names so downstream
# indexing (cat/sprintf below) uses clean variable keys, not display strings.
cor_data <- storm_df %>%
  select(all_of(METRICS))

cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")
rownames(cor_matrix) <- METRICS
colnames(cor_matrix) <- METRICS

# Build plotmath axis labels from VAR_LABELS_SHORT (setup chunk).
# parse() returns an expression object; scale_x/y_discrete(labels=) accepts
# a named vector of expressions to render each level as plotmath.
cor_axis_labels <- setNames(
  parse(text = VAR_LABELS_SHORT[METRICS]),
  METRICS
)

# Melt to long format and keep upper triangle (including diagonal)
cor_long <- as.data.frame(as.table(cor_matrix), stringsAsFactors = FALSE) %>%
  rename(row_var = Var1, col_var = Var2, correlation = Freq) %>%
  mutate(
    row_var = factor(row_var, levels = METRICS),
    col_var = factor(col_var, levels = METRICS)
  ) %>%
  filter(as.integer(row_var) <= as.integer(col_var))

p8 <- ggplot(cor_long, aes(x = col_var, y = fct_rev(row_var), fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 3.5, color = "black") +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limits = c(-1, 1), name = "r"
  ) +
  scale_x_discrete(labels = cor_axis_labels) +
  scale_y_discrete(labels = cor_axis_labels) +
  labs(
    title    = "Correlation Between Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1),
    axis.text.y   = element_text(hjust = 1),
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid    = element_blank()
  )

ggsave(file.path(fig_dir, "Fig08_Correlation_Matrix.png"),
       p8, width = 8, height = 7, dpi = 300)

# Print correlation summary
cat("\n  Correlation matrix:\n")
print(round(cor_matrix, 3))

# ============================================================================
# PLOT 9: ANOMALY TIME SERIES (DESEASONALIZED)
# ============================================================================

cat("\n--- Generating Plot 9: Anomaly Time Series ---\n")

# Calculate anomalies (remove seasonal cycle)
storm_df_anom <- storm_df %>%
  group_by(month) %>%
  mutate(
    eke_synoptic_mean_anom = eke_synoptic_mean - mean(eke_synoptic_mean, na.rm = TRUE),
    eke_p95_anom = eke_p95 - mean(eke_p95, na.rm = TRUE),
    exceed_freq_anom = exceed_p95_freq - mean(exceed_p95_freq, na.rm = TRUE)
  ) %>%
  ungroup()

# Plot anomalies
p9a <- ggplot(storm_df_anom, aes(x = date, y = eke_synoptic_mean_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#0072B2", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "#D55E00", se = FALSE, linewidth = 0.8) +
  labs(title = parse(text = paste0(VAR_LABELS_SHORT[["eke_synoptic_mean"]], '~"Anomaly"'))[[1]],
       y = expression("Anomaly [m"^2*"/s"^2*"]"), x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9b <- ggplot(storm_df_anom, aes(x = date, y = eke_p95_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#009E73", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "#D55E00", se = FALSE, linewidth = 0.8) +
  labs(title = parse(text = paste0(VAR_LABELS_SHORT[["eke_p95"]], '~"Anomaly"'))[[1]],
       y = expression("Anomaly [m"^2*"/s"^2*"]"), x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9c <- ggplot(storm_df_anom, aes(x = date, y = exceed_freq_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#D55E00", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "black", se = FALSE, linewidth = 0.8) +
  labs(title = parse(text = paste0(VAR_LABELS_SHORT[["exceed_p95_freq"]], '~"Anomaly"'))[[1]],
       y = "Anomaly [fraction]", x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9_combined <- (p9a / p9b / p9c) +
  plot_annotation(
    title = "Deseasonalized Storminess Anomalies",
    subtitle = "Pan-Arctic (≥60°N) | Monthly anomalies from 1958–2025 climatology",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig09_Anomaly_TimeSeries.png"),
       p9_combined, width = 12, height = 10, dpi = 300)

# ============================================================================
# PLOT 10: SEASONAL CYCLE BY DECADE (STABILITY CHECK)
# ============================================================================

cat("\n--- Generating Plot 10: Seasonal Cycle Stability ---\n")

# Create decade labels
storm_df_decade <- storm_df %>%
  mutate(
    decade = case_when(
      year >= 1957 & year < 1970 ~ "1958-1969",  # includes Dec 1957 (DJF 1958 warm-up)
      year >= 1970 & year < 1980 ~ "1970-1979",
      year >= 1980 & year < 1990 ~ "1980-1989",
      year >= 1990 & year < 2000 ~ "1990-1999",
      year >= 2000 & year < 2010 ~ "2000-2009",
      year >= 2010 & year < 2020 ~ "2010-2019",
      year >= 2020              ~ "2020-2024"
    ),
    decade = factor(decade, levels = c(
      "1958-1969", "1970-1979", "1980-1989", "1990-1999",
      "2000-2009", "2010-2019", "2020-2024"
    ))
  )

# Calculate seasonal cycle by decade
seasonal_by_decade <- storm_df_decade %>%
  group_by(decade, month) %>%
  summarise(
    eke_synoptic_mean_avg = mean(eke_synoptic_mean, na.rm = TRUE),
    eke_synoptic_mean_se = sd(eke_synoptic_mean, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# We will use the built-in viridis scale inside the plot.
p10 <- ggplot(seasonal_by_decade, aes(x = month, y = eke_synoptic_mean_avg, 
                                      color = decade, group = decade)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  
  # 'end = 0.9' prevents the lines from becoming too yellow/invisible
  scale_color_viridis_d(option = "plasma", name = "Decade", end = 0.9) +
  
  labs(
    title = parse(text = paste0('"Evolution of the Seasonal Cycle: "~', VAR_LABELS_SHORT[["eke_synoptic_mean"]]))[[1]],
    subtitle = "Pan-Arctic (≥60°N) | Has the seasonal pattern changed over time?",
    x = NULL,
    y = parse(text = VAR_LABELS[["eke_synoptic_mean"]])[[1]]
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

ggsave(file.path(fig_dir, "Fig10_Seasonal_Cycle_ByDecade.png"),
       p10, width = 12, height = 6, dpi = 300)

# Also create for exceedance frequency 
seasonal_exceed_decade <- storm_df_decade %>%
  group_by(decade, month) %>%
  summarise(
    exceed_avg = mean(exceed_p95_freq, na.rm = TRUE),
    .groups = "drop"
  )

p10b <- ggplot(seasonal_exceed_decade, aes(x = month, y = exceed_avg, 
                                           color = decade, group = decade)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  
  # --- MODIFICATION: Apply the same scale to the second plot ---
  scale_color_viridis_d(option = "plasma", name = "Decade", end = 0.9) +
  
  labs(
    title = parse(text = paste0('"Evolution of the Seasonal Cycle: "~', VAR_LABELS_SHORT[["exceed_p95_freq"]]))[[1]],
    subtitle = "Pan-Arctic (≥60°N) | Dashed line = climatological 5%",
    x = NULL,
    y = "Exceedance Frequency"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

ggsave(file.path(fig_dir, "Fig10b_Seasonal_Exceedance_ByDecade.png"),
       p10b, width = 12, height = 6, dpi = 300)

# ============================================================================
# FINAL SUMMARY
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 1 DIAGNOSTIC PLOTS COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("Generated figures:\n")
cat("  1.  Fig01_TimeSeries_AllMetrics.png\n")
cat("  2.  Fig02_ACF_AllMetrics.png\n")
cat("  3.  Fig03_Seasonal_Boxplots.png\n")
cat("  4.  Fig04_Seasonal_Cycle_Climatology.png\n")
cat("  5.  Fig05_STL_Decomposition.png\n")
cat("  6.  Fig06_Spatial_Climatology_Polar.png\n")
cat("  7.  Fig07_Distributions.png\n")
cat("  8.  Fig08_Correlation_Matrix.png\n")
cat("  9.  Fig09_Anomaly_TimeSeries.png\n")
cat("  10. Fig10_Seasonal_Cycle_ByDecade.png\n")
cat("  10b.Fig10b_Seasonal_Exceedance_ByDecade.png\n")

cat(sprintf("\nAll figures saved to: %s\n", fig_dir))

cat("\nKey findings from diagnostic plots:\n")
cat(sprintf("  - STL variance decomposition: Seasonal=%.1f%%, Trend=%.1f%%, Residual=%.1f%%\n",
            pct_seasonal, pct_trend, pct_remainder))
cat(sprintf("  - Metric correlations: eke_synoptic_mean vs eke_p95 = %.3f\n",
            cor_matrix["eke_synoptic_mean", "eke_p95"]))
cat(sprintf("  - Metric correlations: eke_synoptic_mean vs exceed_freq = %.3f\n",
            cor_matrix["eke_synoptic_mean", "exceed_p95_freq"]))

cat("\n", rep("=", 70), "\n\n")
```

```{r}
# --------------------------------------------------------------------------
# 1.5.5  ACF/PACF Diagnostic Plot (Energy Components)
# --------------------------------------------------------------------------
# Mirrors Fig02 from storminess metrics: one row per variable,
# ACF on left, PACF on right, computed on deseasonalised anomalies.
# --------------------------------------------------------------------------

# FIG_PHASE1 and ENERGY_COLORS are defined in the setup-environment chunk
fig_dir <- FIG_PHASE1

cat("\n--- Generating ACF/PACF Plot for Energy Components ---\n")

energy_metric_colors <- ENERGY_COLORS

png(file.path(fig_dir, "Fig11_ACF_PACF_Energy_Anomalies.png"),
    width = 12, height = 8, units = "in", res = 300)
par(mfrow = c(3, 2),
    mar = c(4, 4, 3, 1),
    oma = c(0, 0, 3, 0))

for (metric in ENERGY_METRICS) {
  ts_anom    <- ts_objects_energy_anom[[metric]]
  rho1_anom  <- acf(ts_anom, lag.max = 1, plot = FALSE)$acf[2]
  pacf1_anom <- pacf(ts_anom, lag.max = 36, plot = FALSE)$acf[1]
  
  # Build plotmath titles from the setup-chunk VAR_LABELS_SHORT lookup
  acf_title  <- parse(text = paste0('"ACF: "~',  VAR_LABELS_SHORT[[metric]], '~"(anomalies)"'))[[1]]
  pacf_title <- parse(text = paste0('"PACF: "~', VAR_LABELS_SHORT[[metric]], '~"(anomalies)"'))[[1]]

  # ACF plot
  acf(ts_anom, lag.max = 36,
      main = acf_title,
      xlab = "Lag (months)", ylab = "Autocorrelation",
      col = energy_metric_colors[metric], lwd = 2)
  mtext(bquote(rho[1] == .(sprintf("%.3f", rho1_anom))),
        side = 3, line = -1.5, adj = 0.95, cex = 0.8)

  # PACF plot
  pacf(ts_anom, lag.max = 36,
       main = pacf_title,
       xlab = "Lag (months)", ylab = "Partial Autocorrelation",
       col = energy_metric_colors[metric], lwd = 2)
  mtext(bquote(phi[1] == .(sprintf("%.3f", pacf1_anom))),
        side = 3, line = -1.5, adj = 0.95, cex = 0.8)
}

mtext("Autocorrelation Functions: Energy Components (Deseasonalized, 1958–2025)",
      outer = TRUE, line = 1, cex = 1.2, font = 2)
dev.off()
```

```{r}
# --------------------------------------------------------------------------
# Section 1.5 Summary
# --------------------------------------------------------------------------

cat("\n", rep("-", 40), "\n")
cat("Section 1.5 Summary\n")
cat(rep("-", 40), "\n\n")

cat("Energy Budget Integration:\n")
cat(sprintf("  Mean TKE:         %.2f m²/s²\n", mean(storm_df$tke)))
cat(sprintf("  Mean MKE:         %.2f m²/s²\n", mean(storm_df$mke)))
cat(sprintf("  Mean EKE_budget:  %.2f m²/s²\n", mean(storm_df$eke_budget)))
cat(sprintf("  Budget closure:   %.6f%%\n", relative_error))

cat("\nAutocorrelation Diagnostics (Energy Components):\n")
print(knitr::kable(
  acf_comparison_energy %>%
    select(metric, rho1_anomaly, neff, dof_reduction_pct),
  digits = 3, format = "simple",
  col.names = c("Metric", "Lag-1 ACF (ρ₁)", "n_eff", "DoF Reduction (%)")
))

cat("\nGenerated Figures:\n")
```

# Phase 2: Energy budget characterisation

```{r}
# ============================================================================
# PHASE 2: ENERGY BUDGET CHARACTERISATION
# Arctic Storminess Climatology (1958-2025)
# ============================================================================
#
# PURPOSE:
#   Describe the climatological properties of the atmospheric kinetic
#   energy budget: how total kinetic energy is partitioned between the
#   mean flow and transient eddies, how this partitioning varies with
#   season, and how the budget-derived EKE relates to the bandpass-
#   filtered synoptic-scale EKE used as the primary storminess metric.
#
#   This phase is purely descriptive, no trend detection. It establishes
#   the physical context needed to interpret the trend results in Phase 3.
#
# WHY A SEPARATE PHASE:
#   Phase 1 prepares and validates all variables (storminess + energy).
#   Phase 2 characterises what the energy budget *looks like*.
#   Phase 3 asks whether it is *changing*.
#
# INPUT:
#   Phase1_Prepared_Data.RData (which now includes energy components)
#
# OUTPUT:
#   3 diagnostic figures:
#     Fig2.1: Energy budget partitioning (stacked area)
#     Fig2.2: Seasonal cycle of energy components
#     Fig2.3: EKE definition comparison (scatter + time series)
#   Summary statistics printed to console
#
# ============================================================================


# ============================================================================
# SETUP
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 2: ENERGY BUDGET CHARACTERISATION\n")
cat(rep("=", 70), "\n\n")

# Load Phase 1 data
# PHASE1_DATA is defined in the setup-environment chunk — do NOT reassign DATA_DIRECTORY here.
if (!file.exists(PHASE1_DATA)) {
  stop("Phase 1 data not found. Run Phase 1 script first.")
}
load(PHASE1_DATA)
cat("✓ Loaded Phase 1 prepared data\n")

# Verify energy columns are present (added in Phase 1 §1.5)
required_cols <- c("tke", "mke", "eke_budget", "eke_fraction", "mke_eke_ratio")
missing_cols  <- setdiff(required_cols, names(storm_df))
if (length(missing_cols) > 0) {
  stop(sprintf("Missing energy columns in storm_df: %s\n  Re-run Phase 1 with §1.5 extension.",
               paste(missing_cols, collapse = ", ")))
}
cat("Energy budget columns verified\n\n")

# Figure output directory (FIG_PHASE2 defined and created by the setup-environment chunk)
fig_dir <- FIG_PHASE2

# Colour scheme for Phase 2 plots.
# ENERGY_COLORS (from setup chunk) uses variable-name keys ("tke", "mke", "eke_budget").
# The Phase 2 factor labels are "TKE", "MKE", "EKE", so we create a display alias here.
energy_colors <- c(
  "TKE"        = ENERGY_COLORS[["tke"]],
  "MKE"        = ENERGY_COLORS[["mke"]],
  "EKE_budget" = ENERGY_COLORS[["eke_budget"]]
)

# Reconstruct annual means if not loaded
# (energy_annual should be in the .RData, but this is a safety net)
if (!exists("energy_annual")) {
  energy_annual <- storm_df %>%
    group_by(year) %>%
    summarise(
      tke           = mean(tke),
      mke           = mean(mke),
      eke_budget    = mean(eke_budget),
      eke_synoptic_mean      = mean(eke_synoptic_mean),
      mke_eke_ratio = mean(mke_eke_ratio),
      eke_fraction  = mean(eke_fraction),
      .groups       = "drop"
    ) %>%
    filter(year >= 1958)  # exclude Dec 1957 — see boundary note
}
```

```{r}
# ============================================================================
# 2.1  ENERGY PARTITIONING
# ============================================================================
# Key question: how is TKE split between mean flow (MKE) and transient
# eddies (EKE_budget), and is this split stable over 68 years?
#
# A stacked area chart shows three things simultaneously:
#   1. Absolute TKE level (~46 m²/s²)
#   2. The approximate 80/20 EKE/MKE split
#   3. The temporal stability of the partitioning
# ============================================================================

cat("--- 2.1 Energy Partitioning Climatology ---\n\n")

# Report key values
cat(sprintf("  Climatological eddy fraction (f_EKE):  %.1f%%\n",
            mean(storm_df$eke_fraction) * 100))
cat(sprintf("  Climatological MKE/EKE ratio:          %.3f\n",
            mean(storm_df$mke_eke_ratio)))
cat(sprintf("  Range of annual f_EKE:                 %.1f%% – %.1f%%\n",
            min(energy_annual$eke_fraction) * 100,
            max(energy_annual$eke_fraction) * 100))

# --- Figure 2.1: Stacked Area ---
energy_stack <- energy_annual %>%
  select(year, mke, eke_budget) %>%
  pivot_longer(-year, names_to = "component", values_to = "value") %>%
  mutate(
    # Keep internal keys as factor levels so ENERGY_COLORS maps directly
    component = factor(component, levels = c("eke_budget", "mke"))
  )

# Plotmath legend labels: variable name with subscript + descriptive context
energy_stack_labels <- c(
  "eke_budget" = "EKE[budget]~(Transient~Eddies)",
  "mke"        = "MKE~(Mean~Flow)"
)

p2_1 <- ggplot(energy_stack, aes(x = year, y = value, fill = component)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(
    values = ENERGY_COLORS[c("eke_budget", "mke")],
    labels = function(x) parse(text = energy_stack_labels[x]),
    name   = NULL
  ) +
  labs(
    title    = "Partitioning of Total Kinetic Energy",
    subtitle = "Pan-Arctic (≥60°N) | Annual means | TKE = MKE + EKE",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(fig_dir, "Fig2.1_Energy_Budget_Partitioning.png"),
       p2_1, width = 10, height = 6, dpi = 300)
```

```{r}
# ============================================================================
# 2.2  SEASONAL CYCLE OF ENERGY COMPONENTS
# ============================================================================
# Key question: how does the energy budget vary by month?
#
# This reveals:
#   - TKE and EKE peak in winter (DJF), trough in summer (JJA)
#   - MKE has a much flatter seasonal cycle
#   - Winter EKE is roughly 1.5–2× summer EKE
#   - MKE seasonal amplitude is much smaller than EKE's
# ============================================================================

cat("--- 2.2 Seasonal Cycle of Energy Components ---\n\n")

energy_seasonal <- storm_df %>%
  group_by(month) %>%
  summarise(
    tke_mean        = mean(tke),
    mke_mean        = mean(mke),
    eke_budget_mean = mean(eke_budget),   # budget EKE (all sub-monthly variance, eq. 2.9)
    .groups         = "drop"
  )

# Report winter/summer contrast
winter_eke <- mean(energy_seasonal$eke_budget_mean[energy_seasonal$month %in% c(12, 1, 2)])
summer_eke <- mean(energy_seasonal$eke_budget_mean[energy_seasonal$month %in% c(6, 7, 8)])
winter_mke <- mean(energy_seasonal$mke_mean[energy_seasonal$month %in% c(12, 1, 2)])
summer_mke <- mean(energy_seasonal$mke_mean[energy_seasonal$month %in% c(6, 7, 8)])

cat(sprintf("  EKE_budget: winter = %.1f, summer = %.1f m²/s² (ratio: %.2f)\n",
            winter_eke, summer_eke, winter_eke / summer_eke))
cat(sprintf("  MKE:        winter = %.1f, summer = %.1f m²/s² (ratio: %.2f)\n",
            winter_mke, summer_mke, winter_mke / summer_mke))

# Long format for plotting
seasonal_long <- energy_seasonal %>%
  select(month, tke_mean, mke_mean, eke_budget_mean) %>%
  pivot_longer(-month, names_to = "component", values_to = "value") %>%
  mutate(
    # Use ENERGY_VARS keys as factor levels so ENERGY_COLORS maps directly
    component = factor(component,
                       levels = c("tke_mean", "mke_mean", "eke_budget_mean"),
                       labels = c("tke", "mke", "eke_budget"))
  )

p2_2 <- ggplot(seasonal_long, aes(x = month, y = value, color = component)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(
    values = ENERGY_COLORS,
    labels = function(x) parse(text = VAR_LABELS_SHORT[x]),
    name   = "Component"
  ) +
  labs(
    title    = "Climatological Seasonal Cycle of Energy Components",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 mean",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(fig_dir, "Fig2.2_Energy_Seasonal_Cycle.png"),
       p2_2, width = 10, height = 6, dpi = 300)
```

```{r}
# ============================================================================
# 2.3  EKE DEFINITION COMPARISON
# ============================================================================
# Key question: how does the bandpass-filtered synoptic-scale EKE (our
# primary storminess metric) relate to the broader budget EKE (all
# sub-monthly variance)?
#
# A high correlation and stable synoptic fraction (f_synoptic) indicate
# that trends in the bandpass-filtered metric are representative of the
# broader eddy field. If f_synoptic varies substantially over time,
# the two EKE measures could show divergent trends.
#
# Two-panel figure:
#   (a) Scatter plot: monthly EKE_budget vs EKE_synoptic with r
#   (b) Annual time series of both, showing co-variation
# ============================================================================

cat("--- 2.3 EKE Definition Comparison ---\n\n")

# Compute correlation
eke_cor <- cor(storm_df$eke_budget, storm_df$eke_synoptic_mean, use = "complete.obs")
cat(sprintf("  Pearson r (monthly): %.3f\n", eke_cor))

# Compute synoptic fraction
storm_df <- storm_df %>%
  mutate(f_synoptic = eke_synoptic_mean / eke_budget)

cat(sprintf("  Mean synoptic fraction (f_synoptic): %.3f (%.1f%%)\n",
            mean(storm_df$f_synoptic, na.rm = TRUE),
            mean(storm_df$f_synoptic, na.rm = TRUE) * 100))
cat(sprintf("  Range: %.3f – %.3f\n",
            min(storm_df$f_synoptic, na.rm = TRUE),
            max(storm_df$f_synoptic, na.rm = TRUE)))

# Panel (a): Scatter
p_scatter <- ggplot(storm_df, aes(x = eke_budget, y = eke_synoptic_mean)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "#E41A1C", linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  annotate("text",
           x = min(storm_df$eke_budget, na.rm = TRUE) * 1.1,
           y = max(storm_df$eke_synoptic_mean, na.rm = TRUE) * 0.95,
           label = sprintf("r = %.3f", eke_cor),
           hjust = 0, size = 4, fontface = "bold") +
  labs(
    title = parse(text = paste0('"(a) "~EKE[budget]~"vs"~EKE[synoptic]'))[[1]],
    x = expression("EKE"["budget"]*" (all sub-monthly variance) [m"^2*"/s"^2*"]"),
    y = expression("EKE"["synoptic"]*" (1.2–6 day bandpass) [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12))

# Panel (b): Annual time series
comparison_annual <- energy_annual %>%
  select(year, eke_budget, eke_synoptic_mean) %>%
  pivot_longer(-year, names_to = "type", values_to = "value") %>%
  mutate(
    type = factor(type,
                  levels = c("eke_budget", "eke_synoptic_mean"),
                  labels = c("EKE (budget, all sub-monthly)",
                             "EKE (synoptic, 1.2–6 day)"))
  )

p_ts <- ggplot(comparison_annual, aes(x = year, y = value, color = type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("#7570B3", "#1B9E77"), name = NULL) +
  labs(
    title = "(b) Annual Mean Comparison",
    x = NULL,
    y = expression("EKE [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    legend.position = "bottom"
  )

# Combine
p2_3 <- p_scatter + p_ts +
  plot_layout(ncol = 2, widths = c(1, 1.2)) +
  plot_annotation(
    title    = "Comparison of EKE Definitions",
    subtitle = "Budget EKE captures all variance; Synoptic EKE isolates 1.2–6 day storm band",
    theme = theme(
      plot.title    = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig2.3_EKE_Comparison.png"),
       p2_3, width = 14, height = 6, dpi = 300)
```

```{r}
# ============================================================================
# PHASE 2 SUMMARY
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 2 COMPLETE: ENERGY BUDGET CHARACTERISATION\n")
cat(rep("=", 70), "\n\n")

cat("Key Findings:\n")
cat(sprintf("  1. Energy partitioning is stable: EKE accounts for ~%.0f%% of TKE\n",
            mean(storm_df$eke_fraction) * 100))
cat(sprintf("     across all decades, with annual f_EKE ranging %.1f%%–%.1f%%.\n",
            min(energy_annual$eke_fraction) * 100,
            max(energy_annual$eke_fraction) * 100))
cat(sprintf("  2. Strong seasonality: winter EKE is %.1fx summer EKE.\n",
            winter_eke / summer_eke))
cat(sprintf("     MKE seasonality is weaker (winter/summer ratio: %.1f).\n",
            winter_mke / summer_mke))
cat(sprintf("  3. Budget and synoptic EKE are well-correlated (r = %.3f),\n",
            eke_cor))
cat(sprintf("     with synoptic storms accounting for ~%.0f%% of total eddy energy.\n",
            mean(storm_df$f_synoptic, na.rm = TRUE) * 100))

cat("\nGenerated Figures:\n")
cat("  Fig2.1: Energy budget partitioning (stacked area)\n")
cat("  Fig2.2: Seasonal cycle of energy components\n")
cat("  Fig2.3: EKE definition comparison (scatter + time series)\n")
cat(sprintf("\nAll figures saved to: %s\n", fig_dir))
```

# Phase 3: Trend analysis

```{r}
# ============================================================================
# PHASE 3.1: PAN-ARCTIC ANNUAL TREND ANALYSIS
# ============================================================================
#
# DEPENDS ON: setup-environment chunk (run first in every session)
#
# PURPOSE:
#   Detect and quantify monotonic long-term trends in the eight Pan-Arctic
#   variables (5 storminess metrics + 3 energy components) using annual
#   mean time series over 1958–2025 (N = 68 years).
#
# STATISTICAL FRAMEWORK (grounded in Statistics in Climate Sciences, Ch. 12):
#
#   Step 1 — Temporal aggregation (monthly → annual means)
#     Aggregating to annual means removes the seasonal cycle and substantially
#     reduces autocorrelation, since atmospheric persistence does not carry
#     across year boundaries. The filter year >= ANNUAL_START_YEAR is
#     mandatory: December 1957 must be excluded because a single-month
#     "annual mean" for 1957 is not physically representative.
#     See Phase 1 §1.6.3 boundary note.
#
#   Step 2 — Pre-Whitened Mann-Kendall (PWMK) test [zyp.zhang()]
#     Non-parametric rank test for monotonic trend (Mann 1945; Kendall 1975).
#     Pre-whitening follows Zhang et al. (2000): the trend is removed before
#     AR(1) fitting, preventing the autoregressive coefficient from absorbing
#     real trend signal. This is essential because standard MK assumes
#     independence; residual autocorrelation inflates Type I error.
#     H0: no monotonic trend. H1: monotonic trend exists.
#     Significance level: ALPHA (defined in setup chunk).
#
#   Step 3 — Sen's slope estimator (Theil-Sen)
#     Median of all n(n-1)/2 pairwise slopes. Robust to outliers and
#     non-Gaussian residuals; the non-parametric complement to OLS slope.
#     Reported in [metric units] per decade (× DECADES).
#
#   Step 4 — Moving Blocks Bootstrap (MBB) 95% confidence intervals
#     Resamples contiguous blocks to preserve residual autocorrelation
#     structure (Künsch 1989). Block length selected data-adaptively via
#     the PWSD plug-in estimator (Politis & White 2004; Patton et al. 2009)
#     using pwsd() from the blocklength package. Fallback B_FALLBACK applied
#     when pwsd() returns b < 1 (near-white-noise series).
#
# KEY REFERENCES:
#   Mann (1945); Kendall (1975) — rank correlation for trend
#   Zhang et al. (2000) — pre-whitening (implemented in zyp)
#   Sen (1968) — slope estimator
#   Künsch (1989) — Moving Blocks Bootstrap
#   Politis & White (2004) + Patton et al. (2009) — optimal block length
#
# INPUT:  PHASE1_DATA (storm_df with energy columns; defined in setup chunk)
# OUTPUT: results_annual, Fig3.1_Annual_Trends.png, Fig3.2_Annual_Forest.png,
#         Phase3_1_Annual_Results.RData
#
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 3.1: PAN-ARCTIC ANNUAL TREND ANALYSIS\n")
cat(rep("=", 70), "\n\n")

# ── Load Phase 1 data ─────────────────────────────────────────────────────
if (!file.exists(PHASE1_DATA)) {
  stop(glue("Phase 1 data not found at: {PHASE1_DATA}\n  Run Phase 1 first."))
}
load(PHASE1_DATA)
cat(sprintf("Phase 1 data loaded: %d rows × %d columns\n\n",
            nrow(storm_df), ncol(storm_df)))

# Verify all required variables are present
missing_vars <- setdiff(ALL_VARS, names(storm_df))
if (length(missing_vars) > 0) {
  stop(glue(
    "Missing variables in storm_df: {paste(missing_vars, collapse = ', ')}\n",
    "  Check that Phase 1 §1.5 (energy budget integration) was run."
  ))
}
cat(sprintf("All %d variables present\n\n", length(ALL_VARS)))
```

```{r}
# ============================================================================
# STEP 1: ANNUAL AGGREGATION
# ============================================================================
cat(rep("-", 70), "\n")
cat("STEP 1: ANNUAL AGGREGATION\n")
cat(rep("-", 70), "\n\n")

# ANNUAL_START_YEAR = 1958 (set in setup chunk).
annual_df <- storm_df %>%
  filter(year >= ANNUAL_START_YEAR) %>%
  group_by(year) %>%
  filter(n() == 12) %>%
  summarise(
    across(all_of(ALL_VARS), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(year)

cat(sprintf("Annual series: N = %d years (%d–%d)\n\n",
            nrow(annual_df),
            min(annual_df$year),
            max(annual_df$year)))

# Sanity check: N must equal 68
expected_n <- ANALYSIS_END_YEAR - ANNUAL_START_YEAR + 1L
if (nrow(annual_df) != expected_n) {
  warning(glue(
    "Expected {expected_n} annual rows, got {nrow(annual_df)}.\n",
    "  Check for gaps in storm_df."
  ))
}
```

```{r}
# ============================================================================
# STEP 2: STATISTICAL FUNCTIONS
# ============================================================================
cat(rep("-", 70), "\n")
cat("STEP 2: STATISTICAL FUNCTIONS\n")
cat(rep("-", 70), "\n\n")

# ── Sen's slope (Theil-Sen estimator) ─────────────────────────────────────
# Median of all n(n-1)/2 pairwise slopes.
# Implemented directly so the MBB bootstrap uses the identical estimator
# as the point estimate, mixing two implementations would be incoherent.
# Returns slope per time unit (year); multiply by DECADES for per-decade.
sen_slope <- function(x) {
  n      <- length(x)
  k      <- 0L
  slopes <- numeric(n * (n - 1L) / 2L)
  for (i in seq_len(n - 1L)) {
    for (j in seq(i + 1L, n)) {
      k         <- k + 1L
      slopes[k] <- (x[j] - x[i]) / (j - i)
    }
  }
  median(slopes)
}

# ── Data-adaptive block length (Politis & White 2004, Patton et al. 2009) ─
# pwsd() selects the optimal MBB block length from the spectral density
# of the series autocorrelation. Superior to a fixed block length because
# the optimal b depends on the specific autocorrelation structure.
# B_FALLBACK (= 3, from setup chunk) is used when pwsd() fails or returns
# b < 1 (near-white-noise series where no dependence structure is detected).
get_block_length <- function(x) {
  tryCatch({
    b_res <- pwsd(x, round = TRUE, correlogram = FALSE)

    # $BlockLength is a 1×2 matrix with columns "b_Stationary" and "b_Circular".
    # "b_Circular" is the correct block length for the Moving Blocks Bootstrap
    # (MBB and circular block bootstrap share the same optimal b).
    b <- as.integer(b_res$BlockLength[1, "b_Circular"])

    if (is.na(b) || b < 1L) {
      message(glue(
        "    pwsd() returned b = {b}; using fallback b = {B_FALLBACK}"
      ))
      B_FALLBACK
    } else {
      b
    }
  }, error = function(e) {
    message(glue(
      "    pwsd() failed ({conditionMessage(e)}); using fallback b = {B_FALLBACK}"
    ))
    B_FALLBACK
  })
}

# ── Moving Blocks Bootstrap 95% CI on Sen's slope ─────────────────────────
# tsboot() (boot package) resamples fixed-length contiguous blocks, preserving
# within-block autocorrelation structure (Künsch 1989).
# Percentile CI from empirical bootstrap distribution: no normality assumed.
# N_BOOT and ALPHA are defined in the setup chunk.
mbb_ci <- function(x, b) {
  set.seed(42)   # reproducibility
  boot_res <- tsboot(
    tseries   = x,
    statistic = function(d, ...) sen_slope(d),
    R         = N_BOOT,
    l         = b,
    sim       = "fixed"
  )
  ci <- quantile(boot_res$t,
                 probs = c(ALPHA / 2, 1 - ALPHA / 2),
                 na.rm = TRUE)
  list(
    lower     = unname(ci[1]),
    upper     = unname(ci[2]),
    b_used    = b,
    boot_dist = as.vector(boot_res$t)
  )
}
```

```{r}
# ============================================================================
# STEP 3: TREND ANALYSIS
# ============================================================================
cat(rep("-", 70), "\n")
cat("STEP 3: TREND ANALYSIS\n")
cat(rep("-", 70), "\n\n")

results_list <- vector("list", length(ALL_VARS))
names(results_list) <- ALL_VARS

for (var in ALL_VARS) {

  x     <- annual_df[[var]]
  years <- annual_df$year
  cat(sprintf("  [%s]  N = %d\n", var, length(x)))

  # ── Pre-Whitened Mann-Kendall ────────────────────────────────────────
  # zyp.zhang() implements the Zhang et al. (2000) algorithm:
  #   1. Estimate Sen's slope on the original series
  #   2. Remove slope → detrended series (critical: prevents AR(1) from
  #      absorbing real trend signal during pre-whitening)
  #   3. Fit AR(1) to detrended series → φ̂₁
  #   4. Pre-whiten: x'_t = x_t − φ̂₁ · x_{t−1}
  #   5. Re-add slope; apply standard MK to pre-whitened series
  mk_res   <- zyp.zhang(y = x, x = seq_along(x))
  tau      <- unname(mk_res["tau"])
  p_value  <- unname(mk_res["sig"])
  slope_yr <- unname(mk_res["linear"])   # Sen's slope per year
  ar1      <- unname(mk_res["autocor"])  # AR(1) coefficient removed

  # ── Adaptive block length ───────────────────────────────────────────
  cat(sprintf("    Block length via pwsd()... "))
  b_opt <- get_block_length(x)
  cat(sprintf("b = %d years\n", b_opt))

  # ── MBB confidence interval ─────────────────────────────────────────
  cat(sprintf("    MBB CI (R = %d, b = %d)... ", N_BOOT, b_opt))
  ci <- mbb_ci(x, b = b_opt)
  cat("done\n")

  # ── Derived statistics ──────────────────────────────────────────────
  mean_val <- mean(x, na.rm = TRUE)
  total_ch <- slope_yr * (max(years) - min(years))
  pct_ch   <- (total_ch / mean_val) * 100

  results_list[[var]] <- data.frame(
    variable    = var,
    label       = unname(VAR_LABELS_PLAIN[var]),
    group       = ifelse(var %in% METRIC_NAMES, "Storminess", "Energy"),
    n           = length(x),
    mean_val    = round(mean_val,       4),
    slope_dec   = round(slope_yr  * DECADES, 5),
    ci_lower    = round(ci$lower  * DECADES, 5),
    ci_upper    = round(ci$upper  * DECADES, 5),
    tau         = round(tau,            4),
    p_value     = round(p_value,        4),
    ar1_removed = round(ar1,            4),
    significant = p_value < ALPHA,
    pct_change  = round(pct_ch,         2),
    b_length    = ci$b_used,
    stringsAsFactors = FALSE
  )

  cat(sprintf(
    "    → slope = %+.4f/dec | p = %.3f | CI [%+.4f, %+.4f] %s\n\n",
    slope_yr * DECADES, p_value,
    ci$lower  * DECADES, ci$upper * DECADES,
    ifelse(p_value < ALPHA, "✓ SIGNIFICANT", "")
  ))
}

results_annual <- bind_rows(results_list)

# ── Print summary ─────────────────────────────────────────────────────────
cat(rep("=", 70), "\n")
cat("ANNUAL TREND SUMMARY\n")
cat(rep("=", 70), "\n\n")

print(
  results_annual %>%
    select(label, group, mean_val, slope_dec, ci_lower, ci_upper,
           p_value, significant, pct_change, b_length),
  digits = 4, row.names = FALSE
)

cat(sprintf(
  "\nSignificant trends (α = %.2f):  storminess %d/%d  |  energy %d/%d\n",
  ALPHA,
  sum(results_annual$significant & results_annual$group == "Storminess"),
  sum(results_annual$group == "Storminess"),
  sum(results_annual$significant & results_annual$group == "Energy"),
  sum(results_annual$group == "Energy")
))
```

```{r}
# ============================================================================
# STEP 4: FIGURES
# ============================================================================
cat("\n", rep("-", 70), "\n")
cat("STEP 4: FIGURES\n")
cat(rep("-", 70), "\n\n")

# Midpoint of the annual series: used to anchor Sen's slope lines.
# Intercept = mean_val − slope_per_year × midpoint_year keeps lines
# visually centred and numerically stable (avoids large extrapolated values).
mid_year <- mean(annual_df$year)

slope_lines <- results_annual %>%
  mutate(
    slope_yr  = slope_dec / DECADES,
    intercept = mean_val - slope_yr * mid_year,
    label_fig = factor(unname(VAR_LABELS[variable]),
                       levels = unname(VAR_LABELS[ALL_VARS]))
  )

annual_long <- annual_df %>%
  pivot_longer(all_of(ALL_VARS),
               names_to  = "variable",
               values_to = "value") %>%
  mutate(
    label_fig = factor(unname(VAR_LABELS[variable]),
                       levels = unname(VAR_LABELS[ALL_VARS])),
    group     = ifelse(variable %in% METRIC_NAMES,
                       "Storminess metrics", "Energy components")
  )

# ── Figure 3.1: Annual time series ──────────────────────────────────────

p_annual <- ggplot(annual_long, aes(x = year, y = value)) +
  geom_line(aes(colour = variable),
            linewidth = 0.6, alpha = 0.8) +
  geom_point(aes(colour = variable),
             size = 0.9, alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.4, se = FALSE,
              colour = "grey40", linewidth = 0.7, linetype = "dashed") +
  geom_abline(
    data      = slope_lines,
    aes(intercept = intercept, slope = slope_yr,
        linetype  = significant),
    colour    = "#D73027",
    linewidth = 0.9
  ) +
  scale_colour_manual(values = ALL_COLORS, guide = "none") +
  scale_linetype_manual(
    values = c("FALSE" = "dotted", "TRUE" = "solid"),
    labels = c("FALSE" = glue("Not significant (\u03b1 = {ALPHA})"),
               "TRUE"  = glue("Significant (\u03b1 = {ALPHA})")),
    name   = "Sen's slope"
  ) +
  facet_wrap(~ label_fig, ncol = 2, scales = "free_y", labeller = label_parsed) + 
  scale_x_continuous(breaks = seq(1960, 2020, 10)) +
  labs(
    title    = "Pan-Arctic Storminess and Energy: Annual Trends (1958\u20132025)",
    subtitle = glue(
      "Pre-Whitened Mann\u2013Kendall | Sen\u2019s slope (red) | ",
      "LOESS smoother (grey dashed) | N = {nrow(annual_df)} years"
    ),
    x = NULL,
    y = "Annual mean value"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    legend.position  = "bottom",
    strip.text = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )

ggsave(file.path(FIG_PHASE3, "Fig3.1_Annual_Trends.png"),
       p_annual, width = 12, height = 14, dpi = 300)

# ── Figure 3.2: Forest plot ──────────────────────────────────────────────

forest_df <- results_annual %>%
  mutate(
    label_fig = factor(unname(VAR_LABELS[variable]),
                       levels = rev(unname(VAR_LABELS[ALL_VARS]))),
    group     = factor(group, levels = c("Storminess", "Energy"))
  )

p_forest <- ggplot(forest_df,
                   aes(x = slope_dec, y = label_fig,
                       colour = significant, shape = group)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.3, linewidth = 0.9) +
  geom_point(size = 4) +
  geom_text(
    aes(label = sprintf("p = %.3f", p_value),
        x     = pmax(ci_upper, slope_dec) + 0.002),
    hjust = 0, size = 3, colour = "grey30"
  ) +
  scale_colour_manual(
    values = c("FALSE" = "grey55", "TRUE" = "#D73027"),
    labels = c("FALSE" = glue("p \u2265 {ALPHA}"),
               "TRUE"  = glue("p < {ALPHA}")),
    name   = "Significance"
  ) +
  scale_shape_manual(
    values = c("Storminess" = 16, "Energy" = 17),
    name   = "Variable group"
  ) +
  labs(
    title    = "Pan-Arctic Annual Trends: Sen\u2019s Slope with 95% MBB CI",
    subtitle = glue(
      "Pre-Whitened Mann\u2013Kendall | Adaptive block length ",
      "(Politis & White 2004) | {min(annual_df$year)}\u2013{max(annual_df$year)}"
    ),
    x = "Sen\u2019s slope [units per decade]",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  scale_y_discrete(
    labels = function(x) parse(text = x)   # parse plotmath on the fly
  )

ggsave(file.path(FIG_PHASE3, "Fig3.2_Annual_Forest.png"),
       p_forest, width = 11, height = 7, dpi = 300)


# ============================================================================
# STEP 5: SAVE RESULTS
# ============================================================================
save(results_annual, annual_df,
     file = file.path(OUTPUT_DATA_DIR, "Phase3_1_Annual_Results.RData"))
```
```

