---
title: "Storminess_analysis_full_framework"
author: Francesco Lucano
format: html
engine: knitr
editor: visual
warning: false
error: false
toc: true
---

```{r setup-environment}
# Load R Libraries
library(reticulate)
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(glue)
library(future)
library(furrr)
library(progressr)
library(tictoc)
library(ncdf4)
library(glue)
library(lubridate)
library(trend)
library(ggplot2)
library(viridis)
library(patchwork)
library(modifiedmk)
library(fields)
library(RColorBrewer)
library(dplyr)
library(zoo)
library(foreach)
library(doParallel)
library(pbapply)
library(stars)        
library(modifiedmk) 
library(nlme)
library(ncmeta)
library(abind)
library(knitr)
library(boot)
library(here)
library(corrplot)
library(ggcorrplot)

# Configure Python Environment
use_virtualenv("arctic-env", required = TRUE)

# Define paths - UPDATE THESE FOR YOUR SYSTEM
DATA_DIRECTORY <- here::here("Data", "processed")

# Input file (from corrected Step 3 Python script)
METRICS_FILE <- file.path(DATA_DIRECTORY, "ERA5_storminess_metrics_monthly.nc")

# Define which metrics to analyze
METRIC_NAMES <- c(
 "eke_mean",         # PRIMARY: Monthly mean synoptic-scale EKE
 "eke_p90",          # 90th percentile of monthly EKE
 "eke_p95",          # 95th percentile of monthly EKE  
 "eke_p99",          # 99th percentile of monthly EKE
 "exceed_p95_freq"   # Fraction exceeding climatological P95
)

# Analysis parameters
ARCTIC_LAT_THRESHOLD <- 60  # Core Arctic domain (>= 60°N)
REFERENCE_PERIOD <- c(1961, 1990)  # WMO climatological baseline
```

-   PHASE 1: Data Preparation & Validation

    -   1.1 Data loading and standardization

    -   1.2 Domain refinement (\>=60°N)

    -   1.3 Area-weighted time series derivation

    -   1.4 Autocorrelation assessment and effective sample size calculation neff calculation

-   PHASE 2: Energy Budget Diagnostics

    -   2.1 TKE, MKE, EKE_budget calculation and verification

    -   2.2 Budget closure confirmation

    -   2.3 MKE/EKE ratio trends (mechanistic interpretation)

-   PHASE 3: Primary Storminess Analysis (Mean EKE)

    -   3.1 Pan-Arctic trend analysis (eke_mean)

    -   3.2 Spatial trend patterns + field significance

    -   3.3 Seasonal decomposition

    -   3.4 Period comparison (pre- vs post-regime shift)

    -   3.5 Autumn intensification hypothesis test

-   PHASE 4: Extreme Storm Analysis (Percentiles)

    -   4.1 Extreme intensity trends (eke_p95, eke_p99)

    -   4.2 Extreme frequency trends (exceed_p95_freq)

    -   4.3 Spatial patterns of extreme changes

-   PHASE 5: Regional Sub-Analysis (apply full Phase 3-4 workflow to each region)

    -   5.1 Fram Strait sector

    -   5.2 West Svalbard sector (10°E–16°E, 76°N–81°N)

    -   5.3 East Svalbard sector (16°E–28°E, 76°N–81°N)

-   PHASE 6: Synthesis and hypothesis testing

    -   6.1 Summary of all trend results

    -   6.2 Autumn feedback loop hypothesis evaluation

    -   6.3 Comparison with sea-ice loss timelines

# Phase 1: Foundational Analysis

This chapter details the foundational data processing and diagnostic procedures applied to the monthly Pan-Arctic storminess index for the period 1958–2025. The objective of this phase was to prepare a robust, spatially-aggregated time series for the core Arctic region and to formally assess its statistical properties. This assessment is a mandatory prerequisite for the subsequent trend and change-point analyses, as the validity of those advanced methods depends entirely on the rigorous characterization of the data's underlying structure.

## 1.1 Data Loading and Standardization

The analysis commenced with the `ERA5_storminess_metrics_monthly.nc` file, a NetCDF dataset containing five complementary storminess metrics derived from the bandpass-filtered synoptic-scale Eddy Kinetic Energy (EKE) calculated in Section 2.5 of the Methodology. Initial inspection revealed the data arrays were structured with dimensions of `[longitude, latitude, time]`, an unconventional ordering for climatological time series analysis.

The dataset contains the following variables:

```         
Variables:
  eke_mean [m²/s²]              - Monthly mean synoptic-scale EKE (PRIMARY)
  eke_p90 [m²/s²]               - Monthly 90th percentile EKE
  eke_p95 [m²/s²]               - Monthly 95th percentile EKE
  eke_p99 [m²/s²]               - Monthly 99th percentile EKE
  exceed_p95_freq [fraction]    - Fraction exceeding climatological P95
  eke_p95_climatology [m²/s²]   - Reference P95 threshold (1961-1990)
```

To align with standard conventions and optimize computational efficiency for temporal operations, the array dimensions were permuted to a `[time, latitude, longitude]` structure. The time coordinate was decoded from its "days since origin" format and verified to correctly span the full analysis period from January 1958 through December 2025.

### 1.1.1 Rationale for Multi-Metric Approach

Following the methodological framework established by Yau and Chang (2020) and the IPCC AR6 WG1 Chapter 11 recommendations for extreme event analysis, a suite of complementary metrics was calculated to capture different aspects of storminess variability:

1.  **Mean EKE** ($\overline{EKE}$): The primary metric, validated by Yau and Chang (2020) as achieving the highest correlation with surface weather impacts among Eulerian storm track metrics. This represents the average synoptic-scale storm activity within each month.

2.  **Percentile Metrics** (P90, P95, P99): These capture within-month extreme storm intensity at different severity levels. A month characterized by several intense storms interspersed with calm periods will exhibit elevated percentile values even if mean EKE is moderate, providing sensitivity to episodic high-intensity events most relevant to coastal erosion (Irrgang et al. 2022).

3.  **Exceedance Frequency**: This metric detects changes in extreme storm frequency over time, independent of changes in mean storminess. By comparing each timestep against a fixed climatological threshold (P95 from 1961–1990), it reveals whether extreme events are becoming more or less common relative to the pre-Arctic Amplification baseline. This approach follows the rationale established by Zhang et al. (2011), fixing the exceedance probability during the base period to create a static reference that allows monitoring of physical shifts in extreme storm probability.

This multi-metric approach allows detection of changes in the storm intensity distribution that would be obscured by analyzing mean values alone.

```{r}
# ============================================================================
# PHASE 1: DATA LOADING AND EXPLORATORY ANALYSIS
# Arctic Storminess Climatology (1958-2025)
# ============================================================================
cat("\n", rep("=", 70), "\n")
cat("PHASE 1: FOUNDATIONAL ANALYSIS (Corrected Metrics)\n")
cat(rep("=", 70), "\n\n")

# --- Check if file exists ---
if (!file.exists(METRICS_FILE)) {
 stop(glue("
   ERROR: Metrics file not found!
   Expected: {METRICS_FILE}
   
   Please run the corrected Step 3 Python script first:
   Step3_Storminess_Metrics_Corrected.py
 "))
}

cat("Loading storminess metrics file:\n")
cat(glue("  {METRICS_FILE}\n\n"))

# --- Open NetCDF and inspect structure ---
nc <- nc_open(METRICS_FILE)

cat("NetCDF File Information:\n")
cat(rep("-", 40), "\n")

# Print dimensions
cat("\nDimensions:\n")
for (dim_name in names(nc$dim)) {
 cat(sprintf("  %s: %d\n", dim_name, nc$dim[[dim_name]]$len))
}

# Print variables
cat("\nVariables:\n")
for (var_name in names(nc$var)) {
 var_units <- ncatt_get(nc, var_name, "units")$value
 if (is.null(var_units) || var_units == "") var_units <- "dimensionless"
 cat(sprintf("  %s [%s]\n", var_name, var_units))
}

# --- Extract coordinate variables ---
time_raw <- ncvar_get(nc, "time")
lat <- ncvar_get(nc, "latitude")
lon <- ncvar_get(nc, "longitude")

# Get time units for decoding
time_units <- ncatt_get(nc, "time", "units")$value
cat(sprintf("\nTime units: %s\n", time_units))

# --- Extract all metric arrays ---
cat("\nLoading metric arrays...\n")

metrics_list <- list()
for (metric in METRIC_NAMES) {
 if (metric %in% names(nc$var)) {
   metrics_list[[metric]] <- ncvar_get(nc, metric)
   cat(sprintf("  ✓ %s: dim = [%s]\n", 
               metric, 
               paste(dim(metrics_list[[metric]]), collapse = " x ")))
 } else {
   cat(sprintf("  ✗ %s: NOT FOUND in file\n", metric))
 }
}

# Also load climatological threshold if present
if ("eke_p95_climatology" %in% names(nc$var)) {
 eke_p95_clim <- ncvar_get(nc, "eke_p95_climatology")
 cat(sprintf("  ✓ eke_p95_climatology: dim = [%s]\n",
             paste(dim(eke_p95_clim), collapse = " x ")))
}

nc_close(nc)
```

## 1.2 Domain Refinement to the Core Arctic

The original dataset extended south to 55°N latitude. To focus the analysis specifically on the high-Arctic and minimize the influence of mid-latitude dynamics, the spatial domain was refined to include only latitudes greater than or equal to 60°N. This resulted in refined metric arrays with dimensions of \[816, 120, 1440\] (time × latitude × longitude), covering a latitude range from 60.00°N to 89.75°N. The 90°N grid point was excluded during the initial data processing due to documented ECMWF issues with spherical harmonic interpolation at the pole singularity, which produces erroneous near-zero wind values in the native ERA5 grid. All subsequent analyses were performed on this core Arctic domain.

The 60°N threshold was selected to:

1.  Capture the locations of the most intense Arctic cyclones, primarily over the Norwegian, Greenland, and Barents Seas as well as the central Arctic Ocean (Vessey et al. 2022)

2.  Include all major Arctic coastlines relevant to erosion studies

3.  Focus on the region where Arctic-specific feedback mechanisms (e.g., sea-ice loss amplification) are expected to be most pronounced

```{r}
# ----------------------------------------------------------------------------zz
# SECTION 1.2: DIMENSION ORDERING AND TIME DECODING
# ----------------------------------------------------------------------------
# dim(storminess) = [360, 35, 804] so, [lon, lat, time] instead of the correct [lat, lon, time]
                  #  ↓    ↓    ↓
                  # lon  lat  time

cat("\n", rep("-", 40), "\n")
cat("Checking dimension ordering...\n")

# Determine current dimension order
# Expected from Python script: [time, latitude, longitude]
# But R/NetCDF may read as [lon, lat, time] depending on how it was written

sample_metric <- metrics_list[[1]]
n_dims <- length(dim(sample_metric))

if (n_dims != 3) {
 stop("ERROR: Expected 3D array [time, lat, lon]")
}

# Infer dimension order from sizes
dim_sizes <- dim(sample_metric)
n_time_expected <- length(time_raw)
n_lat_expected <- length(lat)
n_lon_expected <- length(lon)

cat(sprintf("  Array dimensions: [%s]\n", paste(dim_sizes, collapse = ", ")))
cat(sprintf("  Expected: time=%d, lat=%d, lon=%d\n", 
           n_time_expected, n_lat_expected, n_lon_expected))

# Check if transposition is needed
if (dim_sizes[1] == n_lon_expected && dim_sizes[3] == n_time_expected) {
 cat("  Detected order: [lon, lat, time] → Transposing to [time, lat, lon]\n")
 
 for (metric in names(metrics_list)) {
   metrics_list[[metric]] <- aperm(metrics_list[[metric]], c(3, 2, 1))
 }
 cat("  ✓ All metrics transposed\n")
 
} else if (dim_sizes[1] == n_time_expected) {
 cat("  ✓ Correct order: [time, lat, lon]\n")
 
} else {
 warning("Unexpected dimension ordering - please verify manually")
}

# Verify final dimensions
cat(sprintf("\nFinal array dimensions: [%s]\n", 
           paste(dim(metrics_list[[1]]), collapse = " x ")))

# --- Decode time coordinate ---
cat("\nDecoding time coordinate...\n")

# Parse time units string (e.g., "days since 1958-01-01")
origin_str <- sub(".*since ", "", time_units)
origin_date <- as.Date(origin_str)

if (grepl("hours since", time_units)) {
 time_dates <- origin_date + hours(time_raw)
} else if (grepl("days since", time_units)) {
 time_dates <- origin_date + days(time_raw)
} else {
 # Assume days
 time_dates <- origin_date + days(time_raw)
}

time_dates <- as.Date(time_dates)

cat(sprintf("  First date: %s\n", format(time_dates[1], "%Y-%m-%d")))
cat(sprintf("  Last date:  %s\n", format(time_dates[length(time_dates)], "%Y-%m-%d")))
cat(sprintf("  Total months: %d\n", length(time_dates)))
```

```{r}
# ============================================================================
# SECTION 1.2: REFINING THE PAN-ARCTIC DOMAIN to >= 60°N
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat(sprintf("Refining domain to Core Arctic (>= %d°N)...\n", ARCTIC_LAT_THRESHOLD))

# Original domain
cat(sprintf("  Original latitude range: %.2f°N to %.2f°N\n", min(lat), max(lat)))

# Find indices for Arctic subset
arctic_lat_indices <- which(lat >= ARCTIC_LAT_THRESHOLD)
lat_arctic <- lat[arctic_lat_indices]

cat(sprintf("  New latitude range: %.2f°N to %.2f°N\n", 
           min(lat_arctic), max(lat_arctic)))
cat(sprintf("  Latitude points: %d → %d\n", 
           length(lat), length(arctic_lat_indices)))

# Subset all metrics to Arctic domain
metrics_arctic <- list()
for (metric in names(metrics_list)) {
 # Subset: keep all time, selected latitudes, all longitudes
 metrics_arctic[[metric]] <- metrics_list[[metric]][, arctic_lat_indices, ]
}

cat(sprintf("  ✓ All metrics subsetted to Arctic domain\n"))
cat(sprintf("  New array dimensions: [%s]\n", 
           paste(dim(metrics_arctic[[1]]), collapse = " x ")))
```

## 1.3 Derivation of the Area-Weighted Pan-Arctic Time Series

To analyze the overall storminess of the core Arctic, the spatial data were aggregated into single representative time series for each metric. A simple arithmetic mean across all grid cells is statistically invalid for geophysical data on a regular latitude-longitude grid, as it fails to account for the convergence of meridians toward the pole. This grants disproportionate weight to the physically smaller high-latitude grid cells, introducing a significant bias into the domain average (Peixoto and Oort 1992).

To correct for this distortion, an area-weighted average was calculated for each of the 816 monthly time steps. The weight for each grid cell, $w_j$, is proportional to the cosine of its latitude, $\phi_j$:

$$w_j = \cos(\phi_j)$$

The area-weighted metric value, $\overline{X_t}$, at each time step $t$ was then calculated as:

$$\overline{X_t} = \frac{\sum_{i=1}^{N_{lon}} \sum_{j=1}^{N_{lat}} X_{i,j,t} \cdot w_j}{\sum_{i=1}^{N_{lon}} \sum_{j=1}^{N_{lat}} w_j}$$

where $X_{i,j,t}$ is the metric value at longitude $i$, latitude $j$, and time $t$.

A diagnostic check confirmed the necessity of this procedure by comparing 
weighted and unweighted spatial means:

| Metric          | Weighted Mean | Bias (Weighted − Unweighted) |
|-----------------|---------------|------------------------------|
| eke_mean        | 12.27 m²/s²   | +0.625 m²/s²                 |
| eke_p90         | 29.47 m²/s²   | +1.538 m²/s²                 |
| eke_p95         | 40.74 m²/s²   | +2.087 m²/s²                 |
| eke_p99         | 66.38 m²/s²   | +3.328 m²/s²                 |
| exceed_p95_freq | 0.0504        | −0.0004                      |

: Comparison of area-weighted versus unweighted spatial means {#tbl-weighting}

The consistent positive bias for EKE metrics indicates that unweighted means systematically underestimate the domain-average storminess. This occurs because cosine weighting assigns greater influence to lower Arctic latitudes (near 60°N), where the North Atlantic storm track produces higher EKE values. The bias magnitude increases with percentile level because extreme values exhibit greater spatial heterogeneity. The final output of this step is a set of monthly time series objects in R, which serve as the basis for all subsequent temporal analyses.

```{r}
# ============================================================================
# SECTION 1.3 Create Area-Weighted Time Series
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Creating area-weighted Pan-Arctic time series...\n")

# --- Calculate cosine-latitude weights ---
# Weight = cos(latitude in radians)
weights_arctic <- cos(lat_arctic * pi / 180)

# Create weight matrix matching spatial dimensions [lat, lon]
n_lat_arctic <- length(lat_arctic)
n_lon <- length(lon)

weights_matrix <- matrix(weights_arctic,
                        nrow = n_lat_arctic,
                        ncol = n_lon,
                        byrow = FALSE)

# --- Calculate area-weighted mean for each metric ---
ts_weighted <- list()
ts_unweighted <- list()  # For diagnostic comparison

for (metric in names(metrics_arctic)) {
 data_array <- metrics_arctic[[metric]]
 
 # Apply weighted mean across spatial dimensions for each timestep
 ts_weighted[[metric]] <- apply(data_array, 1, function(slice) {
   weighted.mean(slice, weights_matrix, na.rm = TRUE)
 })
 
 # Also calculate unweighted for comparison
 ts_unweighted[[metric]] <- apply(data_array, 1, mean, na.rm = TRUE)
 
 # Report bias
 bias <- mean(ts_weighted[[metric]]) - mean(ts_unweighted[[metric]])
 cat(sprintf("  %s: weighted mean = %.4f, bias = %.4f\n", 
             metric, mean(ts_weighted[[metric]]), bias))
}

cat("\n  ✓ Area-weighting applied to all metrics\n")
cat("  (Non-zero bias confirms area-weighting is necessary)\n")
```

```{r}
# ============================================================================
# SECTION 1.3: CREATE TIME SERIES OBJECTS AND DATA FRAMES
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Creating time series objects...\n")

# --- Create R ts objects (for trend package functions) ---
ts_objects <- list()
for (metric in names(ts_weighted)) {
 ts_objects[[metric]] <- ts(ts_weighted[[metric]], 
                            start = c(year(time_dates[1]), month(time_dates[1])),
                            frequency = 12)
}

# --- Create master data frame for ggplot and analysis ---
storm_df <- data.frame(
 date = time_dates,
 year = year(time_dates),
 month = month(time_dates),
 season = case_when(
   month(time_dates) %in% c(12, 1, 2) ~ "DJF",
   month(time_dates) %in% c(3, 4, 5) ~ "MAM",
   month(time_dates) %in% c(6, 7, 8) ~ "JJA",
   month(time_dates) %in% c(9, 10, 11) ~ "SON"
 )
)

# Add all metrics as columns
for (metric in names(ts_weighted)) {
 storm_df[[metric]] <- ts_weighted[[metric]]
}

# Create season_year for proper DJF grouping
# (December of year N belongs to DJF of year N+1)
storm_df$season_year <- storm_df$year
storm_df$season_year[storm_df$month == 12] <- storm_df$season_year[storm_df$month == 12] + 1

cat(sprintf("  ✓ Created data frame with %d rows and %d columns\n", 
           nrow(storm_df), ncol(storm_df)))
```

## 1.4 Assessment of Temporal Autocorrelation

This diagnostic is a critical prerequisite for the entire statistical framework. Standard statistical tests, including Ordinary Least Squares (OLS) regression, operate on the strict assumption that data points (or model residuals) are independent. Climate time series violate this assumption due to persistence, or "memory," arising from the inertia of the climate system.

### 1.4.1 Deseasonalization Requirement

As demonstrated by the STL decomposition (Section 1.4.3), the seasonal cycle accounts for 70.9% of total variance in the Pan-Arctic storminess time series. To obtain valid estimates of atmospheric persistence, autocorrelation must be calculated on deseasonalized anomalies, monthly values minus the climatological mean for that specific calendar month (calculated over the 1961–1990 reference period), rather than raw data.

### 1.4.2 Autocorrelation and Partial Autocorrelation Functions

To quantify the degree of persistence in the deseasonalized Pan-Arctic storminess metrics, both the Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF) were computed for lags up to 36 months. The ACF measures the correlation between observations separated by $k$ time steps:

$$\rho_k = \frac{\text{Cov}(X_t, X_{t-k})}{\text{Var}(X_t)}$$

The PACF measures the correlation at lag $k$ after removing the linear dependence on intermediate lags, providing insight into the appropriate autoregressive order for error modelling.

### 1.4.3 Effective Sample Size

When autocorrelation is present, the effective number of independent observations is smaller than the nominal sample size. Following Zwiers and von Storch (1995), the effective sample size was calculated as:

$$n_{eff} = \frac{N}{1 + 2\sum_{k=1}^{K}\left(1 - \frac{k}{N}\right)\rho_k}$$

where $N$ is the original sample size and $\rho_k$ is the autocorrelation at lag $k$.

The analysis of deseasonalized anomalies yielded the following results:

| Metric          | Lag-1 ACF ($\rho_1$) | $n_{eff}$ | DoF Reduction |
|-----------------|----------------------|-----------|---------------|
| eke_mean        | 0.153                | 313       | 61.7%         |
| eke_p90         | 0.149                | 316       | 61.3%         |
| eke_p95         | 0.152                | 308       | 62.2%         |
| eke_p99         | 0.160                | 313       | 61.6%         |
| exceed_p95_freq | 0.125                | 328       | 59.7%         |

: Autocorrelation diagnostics for deseasonalized Pan-Arctic storminess anomalies {#tbl-acf}

All lag-1 values exceed the threshold of $\rho_1 > 0.1$ that indicates violation of the independence assumption for monthly atmospheric data (Santer et al. 2000), confirming that autocorrelation-corrected methods are required. The modest lag-1 autocorrelation ($\rho_1 \approx 0.15$) reflects genuine atmospheric memory, likely arising from slower-varying boundary forcing such as sea surface temperatures and sea ice extent that impart persistence to monthly-mean atmospheric statistics.

The PACF exhibits a significant spike at lag 1 followed by rapid decay toward the confidence bounds, confirming that an AR(1) error structure is appropriate for GLS regression analyses.

### 1.4.4 Variance Decomposition

To further characterize the temporal structure, Seasonal-Trend decomposition using LOESS (STL) was applied to the primary metric (mean EKE). The variance contribution of each component was:

| Component | Variance Explained |
|-----------|-------------------|
| Seasonal  | 70.9%             |
| Trend     | 3.9%              |
| Remainder | 24.1%             |

: STL variance decomposition for Pan-Arctic mean EKE {#tbl-stl}

The seasonal cycle dominates the variance structure, reflecting the pronounced winter-summer contrast in Arctic storm activity. Critically, the trend component explains only 3.9% of total variance, indicating that any long-term systematic change in Pan-Arctic storminess is small relative to seasonal and interannual variability. This finding provides important context for interpreting subsequent trend analyses.

```{r}
# ============================================================================
# Section 1.4: Assessment of Temporal Autocorrelation (CORRECTED)
# Calculate ACF on DESEASONALIZED ANOMALIES, not raw data
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Assessing temporal autocorrelation on DESEASONALIZED data...\n\n")

# ----------------------------------------------------------------------------
# Step 1: Create deseasonalized anomalies for each metric
# ----------------------------------------------------------------------------

# Function to deseasonalize by removing monthly climatology
deseasonalize <- function(df, metric_col) {
  df %>%
    group_by(month) %>%
    mutate(anomaly = .data[[metric_col]] - mean(.data[[metric_col]], na.rm = TRUE)) %>%
    ungroup() %>%
    pull(anomaly)
}

# Create anomalies for all metrics
storm_df_anom <- storm_df %>%
  mutate(
    eke_mean_anom = deseasonalize(., "eke_mean"),
    eke_p90_anom = deseasonalize(., "eke_p90"),
    eke_p95_anom = deseasonalize(., "eke_p95"),
    eke_p99_anom = deseasonalize(., "eke_p99"),
    exceed_p95_freq_anom = deseasonalize(., "exceed_p95_freq")
  )

# Create ts objects for anomalies
ts_objects_anom <- list(
  eke_mean = ts(storm_df_anom$eke_mean_anom, start = c(1958, 1), frequency = 12),
  eke_p90 = ts(storm_df_anom$eke_p90_anom, start = c(1958, 1), frequency = 12),
  eke_p95 = ts(storm_df_anom$eke_p95_anom, start = c(1958, 1), frequency = 12),
  eke_p99 = ts(storm_df_anom$eke_p99_anom, start = c(1958, 1), frequency = 12),
  exceed_p95_freq = ts(storm_df_anom$exceed_p95_freq_anom, start = c(1958, 1), frequency = 12)
)

# ----------------------------------------------------------------------------
# Step 2: Calculate ACF on anomalies and compare with raw
# ----------------------------------------------------------------------------

acf_results_anom <- list()
acf_comparison <- data.frame()

for (metric in names(ts_objects)) {
  ts_raw <- ts_objects[[metric]]
  ts_anom <- ts_objects_anom[[metric]]
  
  # ACF on raw data (for comparison)
  acf_raw <- acf(ts_raw, lag.max = 24, plot = FALSE)
  rho1_raw <- acf_raw$acf[2]
  
  # ACF on deseasonalized anomalies (CORRECT approach)
  acf_anom <- acf(ts_anom, lag.max = 24, plot = FALSE)
  rho1_anom <- acf_anom$acf[2]
  
  acf_results_anom[[metric]] <- acf_anom
  
  # Calculate effective sample size using ANOMALY autocorrelation
  N <- length(ts_anom)
  rho_k <- acf_anom$acf[-1]  # All lags from 1 to 24
  k <- 1:length(rho_k)
  neff <- N / (1 + 2 * sum((1 - k/N) * pmax(rho_k, 0)))  # Only positive correlations
  
  # Store comparison
  acf_comparison <- rbind(acf_comparison, data.frame(
    metric = metric,
    rho1_raw = rho1_raw,
    rho1_anomaly = rho1_anom,
    inflation_factor = rho1_raw / rho1_anom,
    N = N,
    neff = neff,
    dof_reduction_pct = (1 - neff/N) * 100
  ))
  
  cat(sprintf("--- %s ---\n", metric))
  cat(sprintf("  ρ₁ (raw data):        %.3f  [INFLATED by seasonality]\n", rho1_raw))
  cat(sprintf("  ρ₁ (anomalies):       %.3f  [TRUE atmospheric memory]\n", rho1_anom))
  cat(sprintf("  Inflation factor:     %.2fx\n", rho1_raw / rho1_anom))
  cat(sprintf("  Effective sample size: %.1f (%.1f%% reduction)\n", neff, (1 - neff/N) * 100))
  cat("\n")
}

print(knitr::kable(acf_comparison, digits = 3, format = "simple"))
```

```{r}
# ============================================================================
# SECTION 1.4: DESCRIPTIVE STATISTICS
# ============================================================================

cat(rep("-", 40), "\n")
cat("Descriptive Statistics (Pan-Arctic >= 60°N):\n")
cat(rep("-", 40), "\n\n")

# Create summary table
summary_stats <- data.frame(
 Metric = character(),
 Units = character(),
 Mean = numeric(),
 SD = numeric(),
 Min = numeric(),
 Max = numeric(),
 Lag1_ACF = numeric(),
 stringsAsFactors = FALSE
)

for (metric in names(ts_weighted)) {
 ts_data <- ts_weighted[[metric]]
 rho1 <- acf_results[[metric]]$acf[2]
 
 # Determine units
 if (metric == "exceed_p95_freq") {
   units <- "fraction (0-1)"
 } else {
   units <- "m²/s²"
 }
 
 summary_stats <- rbind(summary_stats, data.frame(
   Metric = metric,
   Units = units,
   Mean = mean(ts_data, na.rm = TRUE),
   SD = sd(ts_data, na.rm = TRUE),
   Min = min(ts_data, na.rm = TRUE),
   Max = max(ts_data, na.rm = TRUE),
   Lag1_ACF = rho1
 ))
}

print(knitr::kable(summary_stats, digits = 4, format = "simple"))
```

## 1.5 Verification of Exceedance Frequency Calculation

As an internal consistency check, the exceedance frequency metric was validated against its theoretical expectation. By definition, the 95th percentile threshold was calculated from the 1961–1990 reference period; therefore, approximately 5% of timesteps during this period should exceed the threshold.

The verification yielded:

```         
Reference period: 1961-1990
Mean exceedance frequency: 0.0500 (5.00%)
Expected (by definition of P95): ~0.05 (5%)
✓ VERIFICATION PASSED
```

The perfect agreement confirms correct implementation of the percentile threshold calculation and validates the exceedance frequency metric for subsequent trend analysis. Deviations from 5% in other time periods will indicate genuine changes in extreme storm frequency relative to the climatological baseline.

```{r}
# ============================================================================
# SECTION 1.5: VERIFICATION - EXCEEDANCE FREQUENCY CHECK
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Verification: Exceedance Frequency During Reference Period\n")
cat(rep("-", 40), "\n\n")

if ("exceed_p95_freq" %in% names(storm_df)) {
 # Filter to reference period (1981-2010)
 ref_period_idx <- storm_df$year >= REFERENCE_PERIOD[1] & 
                   storm_df$year <= REFERENCE_PERIOD[2]
 
 ref_exceed_freq <- mean(storm_df$exceed_p95_freq[ref_period_idx], na.rm = TRUE)
 
 cat(sprintf("Reference period: %d-%d\n", REFERENCE_PERIOD[1], REFERENCE_PERIOD[2]))
 cat(sprintf("Mean exceedance frequency: %.4f (%.2f%%)\n", 
             ref_exceed_freq, ref_exceed_freq * 100))
 cat(sprintf("Expected (by definition of P95): ~0.05 (5%%)\n"))
 
 if (abs(ref_exceed_freq - 0.05) < 0.01) {
   cat("✓ VERIFICATION PASSED: Exceedance frequency is approximately 5%%\n")
 } else {
   cat("⚠ WARNING: Exceedance frequency deviates from expected 5%%\n")
   cat("  This may indicate an issue with the P95 climatology calculation\n")
 }
} else {
 cat("  (exceed_p95_freq not available for verification)\n")
}
```

## 1.6 Summary of Phase 1 Outputs


The foundational analysis produced the following validated datasets for subsequent phases:

| Output           | Description                              | Dimensions              |
|------------------|------------------------------------------|-------------------------|
| `metrics_arctic` | Gridded monthly metrics (Arctic subset)  | [816 × 120 × 1440] per metric |
| `ts_objects`     | Area-weighted monthly time series        | 816 values per metric   |
| `storm_df`       | Master data frame with date/season columns | 816 rows × 10 columns |
| `acf_results`    | ACF diagnostics for each metric          | Lags 0–24               |

: Summary of Phase 1 output objects {#tbl-outputs}

**Key Statistical Properties:**

| Metric          | Mean          | SD            | Lag-1 ACF |
|-----------------|---------------|---------------|-----------|
| eke_mean        | 12.27 m²/s²   | 2.80 m²/s²    | 0.648     |
| eke_p90         | 29.47 m²/s²   | 6.72 m²/s²    | 0.649     |
| eke_p95         | 40.74 m²/s²   | 9.18 m²/s²    | 0.658     |
| eke_p99         | 66.38 m²/s²   | 14.82 m²/s²   | 0.664     |
| exceed_p95_freq | 0.0504        | 0.0218        | 0.583     |

: Descriptive statistics for Pan-Arctic storminess metrics {#tbl-descriptive}

- **Temporal coverage**: 816 months (January 1958 – December 2025)
- **Spatial domain**: 60.00°N – 89.75°N (Core Arctic, excluding pole)
- **Autocorrelation**: All metrics exhibit $\rho_1 \approx 0.6$, requiring 
  autocorrelation-corrected statistical methods

The formal characterization of the data's autocorrelation structure establishes the statistical foundation for all subsequent trend and hypothesis testing analyses. The multi-metric approach enables independent assessment of changes in average storminess (mean EKE), extreme storm intensity (percentiles), and extreme storm frequency (exceedance rate).

```{r}
# ============================================================================
# Section 1.6: SAVE PREPARED DATA
# ============================================================================

cat("\n", rep("-", 40), "\n")
cat("Saving prepared data objects...\n")

# Create output directory if needed
output_dir <- file.path(DATA_DIRECTORY, "analysis_ready")
if (!dir.exists(output_dir)) {
 dir.create(output_dir, recursive = TRUE)
}

# Save R objects for use in subsequent phases
save(
 # Coordinate data
 lat, lon, lat_arctic, time_dates,
 
 # Gridded data (Arctic subset)
 metrics_arctic,
 
 # Area-weighted time series
 ts_weighted, ts_objects,
 
 # Master data frame
 storm_df,
 
 # ACF results
 acf_results,
 
 # Summary statistics
 summary_stats,
 
 # Climatological threshold (if available)
 eke_p95_clim,
 
 # Configuration
 ARCTIC_LAT_THRESHOLD, REFERENCE_PERIOD,
 
 file = file.path(output_dir, "Phase1_Prepared_Data.RData")
)

cat(sprintf("  ✓ Saved to: %s\n", file.path(output_dir, "Phase1_Prepared_Data.RData")))

# Also save storm_df as CSV for inspection
write_csv(storm_df, file.path(output_dir, "storm_metrics_panarctic_monthly.csv"))
cat(sprintf("  ✓ Saved CSV: %s\n", file.path(output_dir, "storm_metrics_panarctic_monthly.csv")))
```

```{r}
# ============================================================================
# PHASE 1: DIAGNOSTIC PLOTS
# Arctic Storminess Climatology (1958-2025)
# 
# PURPOSE: Generate all foundational diagnostic figures for Phase 1
#
# PLOTS GENERATED:
#   1. Time series overview (all metrics)
#   2. ACF plots (all metrics)
#   3. Seasonal boxplots
#   4. Climatological seasonal cycle (12-month)
#   5. STL decomposition
#   6. Spatial climatology maps (polar projection)
#   7. Distribution plots (histograms + density)
#   8. Metric correlation matrix
#   9. Anomaly time series (deseasonalized)
#  10. Seasonal cycle by decade (stability check)
#
# INPUT: Phase1_Prepared_Data.RData
# ============================================================================

# ============================================================================
# SETUP
# ============================================================================

# Load Phase 1 data
data_file <- here::here("Data", "processed", "analysis_ready", "Phase1_Prepared_Data.RData")
if (!file.exists(data_file)) {
  stop("Phase 1 data not found. Run Phase 1 script first.")
}
load(data_file)

cat("✓ Loaded Phase 1 prepared data\n")

# Create output directory
fig_dir <- here::here("Figures", "Phase1_Diagnostics")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

# Define color schemes
metric_colors <- c(
  "eke_mean" = "#0072B2",
  "eke_p90" = "#56B4E9", 
  "eke_p95" = "#009E73",
  "eke_p99" = "#9467BD",
  "exceed_p95_freq" = "#D55E00"
)

season_colors <- c(
  "DJF" = "#4575B4",  # Winter - blue
  "MAM" = "#91CF60",  # Spring - green
  "JJA" = "#FC8D59",  # Summer - orange
  "SON" = "#D73027"   # Autumn - red
)

# Metrics to plot
METRICS <- c("eke_mean", "eke_p90", "eke_p95", "eke_p99", "exceed_p95_freq")

# ============================================================================
# PLOT 1: TIME SERIES OVERVIEW (ALL METRICS)
# ============================================================================

cat("\n--- Generating Plot 1: Time Series Overview ---\n")

# Prepare data
ts_long <- storm_df %>%
  select(date, all_of(METRICS)) %>%
  pivot_longer(-date, names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric, levels = METRICS),
    metric_label = case_when(
      metric == "eke_mean" ~ "Mean EKE [m²/s²]",
      metric == "eke_p90" ~ "P90 EKE [m²/s²]",
      metric == "eke_p95" ~ "P95 EKE [m²/s²]",
      metric == "eke_p99" ~ "P99 EKE [m²/s²]",
      metric == "exceed_p95_freq" ~ "P95 Exceedance Frequency"
    )
  )

p1 <- ggplot(ts_long, aes(x = date, y = value, color = metric)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, linewidth = 0.8) +
  facet_wrap(~ metric_label, scales = "free_y", ncol = 1) +
  scale_color_manual(values = metric_colors, guide = "none") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Pan-Arctic (≥60°N) Storminess Metrics: Monthly Time Series",
    subtitle = "1958–2025 | Raw data (light) with LOESS smoother (bold)",
    x = NULL,
    y = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", hjust = 0),
    strip.background = element_rect(fill = "gray95"),
    panel.grid.minor = element_blank()
  )

ggsave(file.path(fig_dir, "Fig01_TimeSeries_AllMetrics.png"),
       p1, width = 12, height = 14, dpi = 300)
cat("  ✓ Saved: Fig01_TimeSeries_AllMetrics.png\n")

# ============================================================================
# PLOT 2: ACF AND PACF PLOTS (ON DESEASONALIZED ANOMALIES)
# ============================================================================

cat("\n--- Generating Plot 2: ACF and PACF Plots (Anomalies) ---\n")

png(file.path(fig_dir, "Fig02_ACF_PACF_Anomalies.png"),
    width = 12, height = 14, units = "in", res = 300)

par(mfrow = c(5, 2), 
    mar = c(4, 4, 3, 1),
    oma = c(0, 0, 3, 0))

for (metric in METRICS) {
  if (metric %in% names(ts_objects_anom)) {
    ts_anom <- ts_objects_anom[[metric]]
    
    # Get rho1 values for annotation
    rho1_anom <- acf(ts_anom, lag.max = 1, plot = FALSE)$acf[2]
    pacf1_anom <- pacf(ts_anom, lag.max = 36, plot = FALSE)$acf[1]
    
    # ACF plot
    acf(ts_anom, lag.max = 36,
        main = paste0("ACF: ", metric, " (anomalies)"),
        xlab = "Lag (months)",
        ylab = "Autocorrelation",
        col = metric_colors[metric],
        lwd = 2)
    mtext(bquote(rho[1] == .(sprintf("%.3f", rho1_anom))), 
          side = 3, line = -1.5, adj = 0.95, cex = 0.8)
    
    # PACF plot
    pacf(ts_anom, lag.max = 36,
         main = paste0("PACF: ", metric, " (anomalies)"),
         xlab = "Lag (months)",
         ylab = "Partial Autocorrelation",
         col = metric_colors[metric],
         lwd = 2)
    mtext(bquote(phi[1] == .(sprintf("%.3f", pacf1_anom))), 
          side = 3, line = -1.5, adj = 0.95, cex = 0.8)
  }
}

mtext("Autocorrelation Functions: Deseasonalized Anomalies (1958–2025)",
      outer = TRUE, line = 1, cex = 1.2, font = 2)

dev.off()
cat("  ✓ Saved: Fig02_ACF_PACF_Anomalies.png\n")

# ============================================================================
# PLOT 3: SEASONAL BOXPLOTS
# ============================================================================

cat("\n--- Generating Plot 3: Seasonal Boxplots ---\n")

# Prepare data with proper season ordering
storm_df_season <- storm_df %>%
  mutate(season = factor(season, levels = c("DJF", "MAM", "JJA", "SON")))

# Create boxplots for primary metrics
p3a <- ggplot(storm_df_season, aes(x = season, y = eke_mean, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = "Mean EKE", y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3b <- ggplot(storm_df_season, aes(x = season, y = eke_p95, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = "P95 EKE", y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3c <- ggplot(storm_df_season, aes(x = season, y = eke_p99, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = "P99 EKE", y = expression("m"^2*"/s"^2), x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3d <- ggplot(storm_df_season, aes(x = season, y = exceed_p95_freq, fill = season)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8, outlier.alpha = 0.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(values = season_colors, guide = "none") +
  labs(title = "P95 Exceedance Freq.", y = "Fraction", x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p3_combined <- (p3a | p3b | p3c | p3d) +
  plot_annotation(
    title = "Seasonal Distribution of Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 | Red dashed line = climatological 5% threshold",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig03_Seasonal_Boxplots.png"),
       p3_combined, width = 14, height = 5, dpi = 300)
cat("  ✓ Saved: Fig03_Seasonal_Boxplots.png\n")

# ============================================================================
# PLOT 4: CLIMATOLOGICAL SEASONAL CYCLE (12-MONTH)
# ============================================================================

cat("\n--- Generating Plot 4: Climatological Seasonal Cycle ---\n")

# Calculate monthly climatology (mean for each calendar month)
monthly_clim <- storm_df %>%
  group_by(month) %>%
  summarise(
    eke_mean_avg = mean(eke_mean, na.rm = TRUE),
    eke_mean_sd = sd(eke_mean, na.rm = TRUE),
    eke_p95_avg = mean(eke_p95, na.rm = TRUE),
    eke_p95_sd = sd(eke_p95, na.rm = TRUE),
    exceed_freq_avg = mean(exceed_p95_freq, na.rm = TRUE),
    exceed_freq_sd = sd(exceed_p95_freq, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(month_name = month.abb[month])

# Plot for Mean EKE
p4a <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = eke_mean_avg - eke_mean_sd, 
                  ymax = eke_mean_avg + eke_mean_sd),
              fill = "#0072B2", alpha = 0.3) +
  geom_line(aes(y = eke_mean_avg), color = "#0072B2", linewidth = 1.2) +
  geom_point(aes(y = eke_mean_avg), color = "#0072B2", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Mean EKE", 
       y = expression("m"^2*"/s"^2),
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Plot for P95 EKE
p4b <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = eke_p95_avg - eke_p95_sd, 
                  ymax = eke_p95_avg + eke_p95_sd),
              fill = "#009E73", alpha = 0.3) +
  geom_line(aes(y = eke_p95_avg), color = "#009E73", linewidth = 1.2) +
  geom_point(aes(y = eke_p95_avg), color = "#009E73", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "P95 EKE",
       y = expression("m"^2*"/s"^2),
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# Plot for Exceedance Frequency
p4c <- ggplot(monthly_clim, aes(x = month)) +
  geom_ribbon(aes(ymin = exceed_freq_avg - exceed_freq_sd, 
                  ymax = exceed_freq_avg + exceed_freq_sd),
              fill = "#D55E00", alpha = 0.3) +
  geom_line(aes(y = exceed_freq_avg), color = "#D55E00", linewidth = 1.2) +
  geom_point(aes(y = exceed_freq_avg), color = "#D55E00", size = 3) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "P95 Exceedance Frequency",
       y = "Fraction",
       x = NULL) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p4_combined <- (p4a | p4b | p4c) +
  plot_annotation(
    title = "Climatological Seasonal Cycle of Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 mean ± 1 SD",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig04_Seasonal_Cycle_Climatology.png"),
       p4_combined, width = 14, height = 5, dpi = 300)
cat("  ✓ Saved: Fig04_Seasonal_Cycle_Climatology.png\n")

# ============================================================================
# PLOT 5: STL DECOMPOSITION
# ============================================================================

cat("\n--- Generating Plot 5: STL Decomposition ---\n")

# STL decomposition for primary metric (eke_mean)
stl_result <- stl(ts_objects$eke_mean, s.window = "periodic")

# Extract components
stl_df <- data.frame(
  date = time_dates,
  observed = as.vector(ts_objects$eke_mean),
  seasonal = as.vector(stl_result$time.series[, "seasonal"]),
  trend = as.vector(stl_result$time.series[, "trend"]),
  remainder = as.vector(stl_result$time.series[, "remainder"])
)

# Calculate component contributions
var_total <- var(stl_df$observed, na.rm = TRUE)
var_seasonal <- var(stl_df$seasonal, na.rm = TRUE)
var_trend <- var(stl_df$trend, na.rm = TRUE)
var_remainder <- var(stl_df$remainder, na.rm = TRUE)

pct_seasonal <- (var_seasonal / var_total) * 100
pct_trend <- (var_trend / var_total) * 100
pct_remainder <- (var_remainder / var_total) * 100

cat(sprintf("  Variance decomposition:\n"))
cat(sprintf("    Seasonal: %.1f%%\n", pct_seasonal))
cat(sprintf("    Trend:    %.1f%%\n", pct_trend))
cat(sprintf("    Residual: %.1f%%\n", pct_remainder))

# Create STL plot
stl_long <- stl_df %>%
  pivot_longer(-date, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component, 
                       levels = c("observed", "seasonal", "trend", "remainder")),
    component_label = case_when(
      component == "observed" ~ sprintf("Observed Data"),
      component == "seasonal" ~ sprintf("Seasonal (%.1f%% of variance)", pct_seasonal),
      component == "trend" ~ sprintf("Trend (%.1f%% of variance)", pct_trend),
      component == "remainder" ~ sprintf("Remainder (%.1f%% of variance)", pct_remainder)
    ),
    component_label = factor(component_label, levels = unique(component_label))
  )

p5 <- ggplot(stl_long, aes(x = date, y = value)) +
  geom_line(color = "#0072B2", linewidth = 0.4) +
  facet_wrap(~ component_label, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  labs(
    title = "STL Decomposition: Mean EKE (Primary Metric)",
    subtitle = "Pan-Arctic (≥60°N) | Seasonal-Trend decomposition using LOESS",
    x = NULL,
    y = expression("m"^2*"/s"^2)
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", hjust = 0),
    strip.background = element_rect(fill = "gray95"),
    panel.grid.minor = element_blank()
  )

ggsave(file.path(fig_dir, "Fig05_STL_Decomposition.png"),
       p5, width = 12, height = 10, dpi = 300)
cat("  ✓ Saved: Fig05_STL_Decomposition.png\n")

# Also save STL for P95 (extreme intensity)
stl_p95 <- stl(ts_objects$eke_p95, s.window = "periodic")

stl_p95_df <- data.frame(
  date = time_dates,
  observed = as.vector(ts_objects$eke_p95),
  seasonal = as.vector(stl_p95$time.series[, "seasonal"]),
  trend = as.vector(stl_p95$time.series[, "trend"]),
  remainder = as.vector(stl_p95$time.series[, "remainder"])
)

# Store for later use
stl_results <- list(
  eke_mean = stl_result,
  eke_p95 = stl_p95
)

# ============================================================================
# PLOT 6: SPATIAL CLIMATOLOGY MAPS (POLAR PROJECTION)
# Using proven methodology from spatial trend analysis
# NOTE: For visualization, we use the FULL dataset (all available latitudes)
#       even though the statistical analysis focuses on ≥60°N
# ============================================================================

cat("\n--- Generating Plot 6: Spatial Climatology Maps (Polar) ---\n")

# Calculate temporal mean at each grid point for each metric
# Use FULL domain (metrics_list) not Arctic subset (metrics_arctic)
# Result: [lat, lon] matrices
climatology_maps_full <- list()
for (metric in c("eke_mean", "eke_p95", "exceed_p95_freq")) {
  if (metric %in% names(metrics_list)) {
    # Apply mean across time dimension (dim 1) - using FULL dataset
    climatology_maps_full[[metric]] <- apply(metrics_list[[metric]], c(2, 3), mean, na.rm = TRUE)
  }
}

cat(sprintf("  Climatology maps calculated (full domain: %.1f°N to %.1f°N)\n", 
            min(lat), max(lat)))

# ============================================================================
# STEP 1: CREATE A SPATIALLY-AWARE `stars` OBJECT (WITH WRAPPING)
# ============================================================================

# 1. Prepare Longitude Vector with Wrapping
# ----------------------------------------------------------------------------
# We extend the longitude vector by one point to close the gap at 180 deg
lon_vec <- as.vector(lon)
lat_vec <- as.vector(lat) 

# Create a new longitude vector that adds one point (start + 360)
lon_wrapped <- c(lon_vec, lon_vec[1] + 360)

cat(sprintf("  Wrapping longitude: Original %d points -> Wrapped %d points\n", 
            length(lon_vec), length(lon_wrapped)))

# 2. Helper function to wrap matrices
# ----------------------------------------------------------------------------
# Input: [lat, lon] matrix
# Output: [lon_wrapped, lat] matrix
wrap_data <- function(mat) {
  # Transpose to [lon, lat]
  mat_t <- t(mat)
  # Append the first row (first longitude) to the end
  mat_wrapped <- rbind(mat_t, mat_t[1, ])
  return(mat_wrapped)
}

# 3. Create Wrapped Data List
# ----------------------------------------------------------------------------
data_list_plot <- list(
  eke_mean    = wrap_data(climatology_maps_full$eke_mean),
  eke_p95     = wrap_data(climatology_maps_full$eke_p95),
  exceed_freq = wrap_data(climatology_maps_full$exceed_p95_freq)
)

# 4. Create Dimensions and Stars Object
# ----------------------------------------------------------------------------
# IMPORTANT: Use 'lon_wrapped' here, not the original 'lon_vec'
dims <- st_dimensions(x = lon_wrapped, y = lat_vec)

# Create the stars object and set CRS
clim_stars <- st_as_stars(data_list_plot, dimensions = dims) %>%
  st_set_crs(4326)

cat("  Stars object created (gap repaired)\n")

# ============================================================================
# STEP 2: REPROJECT ALL SPATIAL DATA
# ============================================================================

# Define the target projection: Lambert Azimuthal Equal-Area, centered on North Pole
arctic_crs <- "+proj=laea +lat_0=90 +lon_0=0"

# Warp the raster data to polar projection
clim_polar <- st_warp(clim_stars, crs = arctic_crs)

# Load and transform coastlines
world <- ne_countries(scale = "medium", returnclass = "sf")
world_polar <- st_transform(world, crs = arctic_crs)

cat("  Data reprojected to polar CRS\n")

# ============================================================================
# STEP 3: CREATE CARTOGRAPHIC ELEMENTS
# ============================================================================

# Define two radii for "two-ring" masking approach
view_radius <- 3700000  # 3700 km - data boundary
plot_radius <- 4400000  # 4400 km - space for labels

# Create circular mask
circle_mask <- st_sfc(st_point(c(0, 0)), crs = arctic_crs) %>%
  st_buffer(dist = view_radius)

# Apply mask to data
clim_polar_masked <- clim_polar[circle_mask]

# Create inverse mask (square with circular hole)
square_box <- st_as_sfc(st_bbox(c(
  xmin = -plot_radius, xmax = plot_radius,
  ymin = -plot_radius, ymax = plot_radius
), crs = arctic_crs))
inverse_mask <- st_difference(square_box, circle_mask)

# Create graticule lines
graticules <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = c(60, 70, 80),
  crs = 4326
) %>%
  st_transform(crs = arctic_crs)

# Create latitude labels (positioned at 180° longitude for visibility)
lat_labels <- data.frame(
  lon = 180,
  lat = c(60, 70, 80)
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = arctic_crs)
lat_labels$label <- c("60°N", "70°N", "80°N")

# Create longitude labels (positioned in the blank space between radii)
lon_labels <- data.frame(
  lon = seq(0, 300, by = 60),
  lat = 54  # Falls in blank space between view_radius and plot_radius
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = arctic_crs)
lon_labels$label <- c("0°", "60°E", "120°E", "180°", "120°W", "60°W")

cat("  Cartographic elements created\n")

# ============================================================================
# STEP 4: ASSEMBLE THE PLOTS
# ============================================================================

# --- MAP 1: Mean EKE Climatology ---
p6a <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = eke_mean), na.rm = TRUE) +
  scale_fill_viridis(
    name = expression(atop("Mean EKE", "[m"^2*"/s"^2*"]")),
    option = "C",
    na.value = "transparent"
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label), size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label), size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius), ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = "(a) Mean EKE Climatology") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- MAP 2: P95 EKE Climatology ---
p6b <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = eke_p95), na.rm = TRUE) +
  scale_fill_viridis(
    name = expression(atop("P95 EKE", "[m"^2*"/s"^2*"]")),
    option = "C",
    na.value = "transparent"
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label), size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label), size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius), ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = "(b) P95 EKE Climatology") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# --- MAP 3: Exceedance Frequency Climatology ---
p6c <- ggplot() +
  geom_stars(data = clim_polar_masked, aes(fill = exceed_freq), na.rm = TRUE) +
  scale_fill_viridis(
    name = "Exceedance\nFrequency",
    option = "D",
    na.value = "transparent"
  ) +
  geom_sf(data = world_polar, fill = NA, color = "black", linewidth = 0.2) +
  geom_sf(data = graticules, color = "gray50", linetype = "dashed", linewidth = 0.25) +
  geom_sf(data = inverse_mask, fill = "white", color = NA) +
  geom_sf_text(data = lat_labels, aes(label = label), size = 2.5, color = "black", hjust = 1.2) +
  geom_sf_text(data = lon_labels, aes(label = label), size = 2.5, color = "black") +
  coord_sf(xlim = c(-plot_radius, plot_radius), ylim = c(-plot_radius, plot_radius), expand = FALSE) +
  labs(title = "(c) P95 Exceedance Frequency") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    legend.position = "right",
    legend.key.width = unit(0.5, "cm"),
    legend.key.height = unit(1.5, "cm"),
    panel.background = element_rect(fill = "gray95", color = NA),
    plot.margin = margin(5, 5, 5, 5)
  )

# Combine panels
p6_combined <- p6a + p6b + p6c +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Spatial Climatology of Storminess Metrics (1958-2025)",
    subtitle = "Full domain visualization (≥55°N) | Lambert Azimuthal Equal-Area Projection",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5)
    )
  )

# Print to viewer
print(p6_combined)

# Save combined figure
ggsave(file.path(fig_dir, "Fig06_Spatial_Climatology_Polar.png"),
       p6_combined, width = 20, height = 8, dpi = 300, bg = "white")
cat("  ✓ Saved: Fig06_Spatial_Climatology_Polar.png\n")

# Also save individual high-res versions
ggsave(file.path(fig_dir, "Fig06a_MeanEKE_Climatology_Polar.png"),
       p6a, width = 8, height = 8, dpi = 300, bg = "white")
ggsave(file.path(fig_dir, "Fig06b_P95EKE_Climatology_Polar.png"),
       p6b, width = 8, height = 8, dpi = 300, bg = "white")
ggsave(file.path(fig_dir, "Fig06c_ExceedFreq_Climatology_Polar.png"),
       p6c, width = 8, height = 8, dpi = 300, bg = "white")
cat("  ✓ Saved individual polar maps\n")
cat("  ✓ Saved individual polar maps\n")

# ============================================================================
# PLOT 7: DISTRIBUTION PLOTS (HISTOGRAMS + DENSITY)
# ============================================================================

cat("\n--- Generating Plot 7: Distribution Plots ---\n")

# Prepare data
dist_data <- storm_df %>%
  select(all_of(METRICS)) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = METRICS))

p7 <- ggplot(dist_data, aes(x = value, fill = metric)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.7, color = "white") +
  geom_density(color = "black", linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free", ncol = 3) +
  scale_fill_manual(values = metric_colors, guide = "none") +
  labs(
    title = "Distribution of Monthly Storminess Metrics",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 | n = 816 months",
    x = "Value",
    y = "Density"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold")
  )

ggsave(file.path(fig_dir, "Fig07_Distributions.png"),
       p7, width = 12, height = 8, dpi = 300)
cat("  ✓ Saved: Fig07_Distributions.png\n")

# ============================================================================
# PLOT 8: METRIC CORRELATION MATRIX
# ============================================================================

cat("\n--- Generating Plot 8: Correlation Matrix ---\n")

# Calculate correlation matrix
cor_data <- storm_df %>%
  select(all_of(METRICS))

cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

# Rename for display
rownames(cor_matrix) <- c("Mean EKE", "P90 EKE", "P95 EKE", "P99 EKE", "Exceed. Freq.")
colnames(cor_matrix) <- rownames(cor_matrix)

# Create correlation plot
png(file.path(fig_dir, "Fig08_Correlation_Matrix.png"),
    width = 8, height = 7, units = "in", res = 300)

corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         order = "original",
         addCoef.col = "black",
         number.cex = 0.9,
         tl.col = "black",
         tl.srt = 45,
         col = colorRampPalette(c("#B2182B", "white", "#2166AC"))(200),
         title = "Correlation Between Storminess Metrics\n(Pan-Arctic, 1958–2025)",
         mar = c(0, 0, 2, 0))

dev.off()
cat("  ✓ Saved: Fig08_Correlation_Matrix.png\n")

# Print correlation summary
cat("\n  Correlation matrix:\n")
print(round(cor_matrix, 3))

# ============================================================================
# PLOT 9: ANOMALY TIME SERIES (DESEASONALIZED)
# ============================================================================

cat("\n--- Generating Plot 9: Anomaly Time Series ---\n")

# Calculate anomalies (remove seasonal cycle)
storm_df_anom <- storm_df %>%
  group_by(month) %>%
  mutate(
    eke_mean_anom = eke_mean - mean(eke_mean, na.rm = TRUE),
    eke_p95_anom = eke_p95 - mean(eke_p95, na.rm = TRUE),
    exceed_freq_anom = exceed_p95_freq - mean(exceed_p95_freq, na.rm = TRUE)
  ) %>%
  ungroup()

# Plot anomalies
p9a <- ggplot(storm_df_anom, aes(x = date, y = eke_mean_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#0072B2", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "#D55E00", se = FALSE, linewidth = 0.8) +
  labs(title = "Mean EKE Anomaly", y = expression("Anomaly [m"^2*"/s"^2*"]"), x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9b <- ggplot(storm_df_anom, aes(x = date, y = eke_p95_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#009E73", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "#D55E00", se = FALSE, linewidth = 0.8) +
  labs(title = "P95 EKE Anomaly", y = expression("Anomaly [m"^2*"/s"^2*"]"), x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9c <- ggplot(storm_df_anom, aes(x = date, y = exceed_freq_anom)) +
  geom_hline(yintercept = 0, color = "gray50") +
  geom_line(color = "#D55E00", alpha = 0.5, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, color = "black", se = FALSE, linewidth = 0.8) +
  labs(title = "Exceedance Frequency Anomaly", y = "Anomaly [fraction]", x = NULL) +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p9_combined <- (p9a / p9b / p9c) +
  plot_annotation(
    title = "Deseasonalized Storminess Anomalies",
    subtitle = "Pan-Arctic (≥60°N) | Monthly anomalies from 1958–2025 climatology",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(fig_dir, "Fig09_Anomaly_TimeSeries.png"),
       p9_combined, width = 12, height = 10, dpi = 300)
cat("  ✓ Saved: Fig09_Anomaly_TimeSeries.png\n")

# ============================================================================
# PLOT 10: SEASONAL CYCLE BY DECADE (STABILITY CHECK)
# ============================================================================

cat("\n--- Generating Plot 10: Seasonal Cycle Stability ---\n")

# Create decade labels
storm_df_decade <- storm_df %>%
  mutate(
    decade = case_when(
      year >= 1958 & year < 1970 ~ "1958-1969",
      year >= 1970 & year < 1980 ~ "1970-1979",
      year >= 1980 & year < 1990 ~ "1980-1989",
      year >= 1990 & year < 2000 ~ "1990-1999",
      year >= 2000 & year < 2010 ~ "2000-2009",
      year >= 2010 & year < 2020 ~ "2010-2019",
      year >= 2020 ~ "2020-2024"
    ),
    decade = factor(decade)
  )

# Calculate seasonal cycle by decade
seasonal_by_decade <- storm_df_decade %>%
  group_by(decade, month) %>%
  summarise(
    eke_mean_avg = mean(eke_mean, na.rm = TRUE),
    eke_mean_se = sd(eke_mean, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Define decade colors
decade_colors <- colorRampPalette(c("#2166AC", "#D1E5F0", "#F7F7F7", 
                                    "#FDDBC7", "#B2182B"))(7)
names(decade_colors) <- levels(seasonal_by_decade$decade)

p10 <- ggplot(seasonal_by_decade, aes(x = month, y = eke_mean_avg, 
                                       color = decade, group = decade)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(values = decade_colors, name = "Decade") +
  labs(
    title = "Evolution of the Seasonal Cycle: Mean EKE",
    subtitle = "Pan-Arctic (≥60°N) | Has the seasonal pattern changed over time?",
    x = NULL,
    y = expression("Mean EKE [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

ggsave(file.path(fig_dir, "Fig10_Seasonal_Cycle_ByDecade.png"),
       p10, width = 12, height = 6, dpi = 300)
cat("  ✓ Saved: Fig10_Seasonal_Cycle_ByDecade.png\n")

# Also create for exceedance frequency (important for autumn hypothesis)
seasonal_exceed_decade <- storm_df_decade %>%
  group_by(decade, month) %>%
  summarise(
    exceed_avg = mean(exceed_p95_freq, na.rm = TRUE),
    .groups = "drop"
  )

p10b <- ggplot(seasonal_exceed_decade, aes(x = month, y = exceed_avg, 
                                            color = decade, group = decade)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(values = decade_colors, name = "Decade") +
  labs(
    title = "Evolution of the Seasonal Cycle: P95 Exceedance Frequency",
    subtitle = "Pan-Arctic (≥60°N) | Dashed line = climatological 5%",
    x = NULL,
    y = "Exceedance Frequency"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

ggsave(file.path(fig_dir, "Fig10b_Seasonal_Exceedance_ByDecade.png"),
       p10b, width = 12, height = 6, dpi = 300)
cat("  ✓ Saved: Fig10b_Seasonal_Exceedance_ByDecade.png\n")

# ============================================================================
# SAVE ADDITIONAL OUTPUTS
# ============================================================================

cat("\n--- Saving additional outputs ---\n")

# Save STL results
save(stl_results, stl_df, stl_p95_df,
     file = file.path(fig_dir, "STL_decomposition_results.RData"))
cat("  ✓ Saved: STL_decomposition_results.RData\n")

# Save climatology maps (full domain for visualization)
save(climatology_maps_full,
     file = file.path(fig_dir, "climatology_maps_full_domain.RData"))
cat("  ✓ Saved: climatology_maps_full_domain.RData\n")

# Save correlation matrix
write.csv(cor_matrix, file.path(fig_dir, "correlation_matrix.csv"))
cat("  ✓ Saved: correlation_matrix.csv\n")

# ============================================================================
# FINAL SUMMARY
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 1 DIAGNOSTIC PLOTS COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("Generated figures:\n")
cat("  1.  Fig01_TimeSeries_AllMetrics.png\n")
cat("  2.  Fig02_ACF_AllMetrics.png\n")
cat("  3.  Fig03_Seasonal_Boxplots.png\n")
cat("  4.  Fig04_Seasonal_Cycle_Climatology.png\n")
cat("  5.  Fig05_STL_Decomposition.png\n")
cat("  6.  Fig06_Spatial_Climatology_Polar.png\n")
cat("  7.  Fig07_Distributions.png\n")
cat("  8.  Fig08_Correlation_Matrix.png\n")
cat("  9.  Fig09_Anomaly_TimeSeries.png\n")
cat("  10. Fig10_Seasonal_Cycle_ByDecade.png\n")
cat("  10b.Fig10b_Seasonal_Exceedance_ByDecade.png\n")

cat(sprintf("\nAll figures saved to: %s\n", fig_dir))

cat("\nKey findings from diagnostic plots:\n")
cat(sprintf("  - STL variance decomposition: Seasonal=%.1f%%, Trend=%.1f%%, Residual=%.1f%%\n",
            pct_seasonal, pct_trend, pct_remainder))
cat(sprintf("  - Metric correlations: eke_mean vs eke_p95 = %.3f\n", 
            cor_matrix["Mean EKE", "P95 EKE"]))
cat(sprintf("  - Metric correlations: eke_mean vs exceed_freq = %.3f\n",
            cor_matrix["Mean EKE", "Exceed. Freq."]))

cat("\n", rep("=", 70), "\n\n")
```

# Phase 2: Energy Budget Diagnostics

This phase examines the atmospheric kinetic energy budget to provide mechanistic context for interpreting storminess trends. Following the Reynolds decomposition framework (Lorenz 1955; Peixoto and Oort 1992), the total kinetic energy of the atmosphere can be partitioned into mean and eddy components:

$$\overline{TKE} = MKE + EKE$$

where:
- **TKE** (Total Kinetic Energy): Complete kinetic energy of the atmospheric flow
- **MKE** (Mean Kinetic Energy): Energy in the time-mean circulation (jet streams, stationary waves)
- **EKE** (Eddy Kinetic Energy): Energy in transient disturbances (cyclones, anticyclones)

This budget framework enables mechanistic interpretation of any observed storminess changes. An increase in EKE could result from:

1. **Overall atmospheric energization**: TKE increases, with energy distributed 
   to both MKE and EKE
2. **Energy redistribution**: TKE remains stable but the mean flow becomes more 
   unstable, converting MKE to EKE
3. **Decreased mean flow with stable eddies**: MKE decreases while EKE remains 
   unchanged

By analyzing all three components together, we can distinguish between these physically distinct mechanisms.

**Note**: The EKE calculated for budget closure (EKE_budget) captures all sub-monthly variance, which differs from the bandpass-filtered synoptic-scale EKE (1.2–6 day periods) used as the primary storminess metric. Both quantities are physically meaningful but serve different analytical purposes.

```{r}
# ============================================================================
# PHASE 2: ENERGY BUDGET DIAGNOSTICS
# Arctic Storminess Climatology (1958-2025)
# ============================================================================

# ============================================================================
# SETUP
# ============================================================================

# Define paths
DATA_DIRECTORY <- here::here("Data", "processed")
RESULTS_DIR <- here::here("Data", "processed", "analysis_ready")
FIG_DIR <- here::here("Figures", "Phase2_EnergyBudget")

# Create directories if needed
if (!dir.exists(RESULTS_DIR)) dir.create(RESULTS_DIR, recursive = TRUE)
if (!dir.exists(FIG_DIR)) dir.create(FIG_DIR, recursive = TRUE)

# Output file for Phase 2 results
PHASE2_OUTPUT <- file.path(RESULTS_DIR, "Phase2_EnergyBudget_Results.RData")

# ----------------------------------------------------------------------------
# Load Phase 1 Prepared Data
# ----------------------------------------------------------------------------

phase1_file <- file.path(RESULTS_DIR, "Phase1_Prepared_Data.RData")

if (!file.exists(phase1_file)) {
  stop(glue("
    ERROR: Phase 1 data not found!
    Expected: {phase1_file}
    Please run Phase 1 script first.
  "))
}

load(phase1_file)
cat("✓ Loaded Phase 1 prepared data\n")
cat(sprintf("  - Time series: %d months (%d-%d)\n", 
            length(time_dates), 
            year(min(time_dates)), 
            year(max(time_dates))))

# ----------------------------------------------------------------------------
# Check if Phase 2 already completed (skip if results exist)
# ----------------------------------------------------------------------------

if (file.exists(PHASE2_OUTPUT)) {
  cat("\n✓ Phase 2 results already exist. Loading saved results...\n")
  load(PHASE2_OUTPUT)
  cat("  To re-run analysis, delete:", PHASE2_OUTPUT, "\n")
} else {
  
  cat("\n", rep("=", 70), "\n")
  cat("PHASE 2: ENERGY BUDGET DIAGNOSTICS\n")
  cat(rep("=", 70), "\n\n")

  # ==========================================================================
  # 2.1 Load Energy Budget Files
  # ==========================================================================
  
  cat("--- 2.1 Loading Energy Budget Components ---\n\n")
  
  TKE_FILE <- file.path(DATA_DIRECTORY, "ERA5_TKE_monthly.nc")
  MKE_FILE <- file.path(DATA_DIRECTORY, "ERA5_MKE_monthly.nc")
  EKE_BUDGET_FILE <- file.path(DATA_DIRECTORY, "ERA5_EKE_budget_monthly.nc")
  
  # Check files exist
  energy_files <- list(TKE = TKE_FILE, MKE = MKE_FILE, EKE_budget = EKE_BUDGET_FILE)
  for (name in names(energy_files)) {
    if (!file.exists(energy_files[[name]])) {
      stop(sprintf("File not found: %s", energy_files[[name]]))
    }
    cat(sprintf("  ✓ Found: %s\n", basename(energy_files[[name]])))
  }
  
  # Function to load and process energy component
  load_energy_component <- function(filepath, varname, lat_arctic, lon) {
    nc <- nc_open(filepath)
    data_raw <- ncvar_get(nc, varname)
    lat_nc <- ncvar_get(nc, "latitude")
    lon_nc <- ncvar_get(nc, "longitude")
    nc_close(nc)
    
    # Permute to [time, lat, lon] if needed
    if (dim(data_raw)[1] == length(lon_nc) && dim(data_raw)[3] != length(lon_nc)) {
      data_array <- aperm(data_raw, c(3, 2, 1))
    } else {
      data_array <- data_raw
    }
    
    # Subset to Arctic (>= 60°N)
    arctic_idx <- which(lat_nc >= 60)
    data_arctic <- data_array[, arctic_idx, ]
    lat_subset <- lat_nc[arctic_idx]
    
    # Area-weighted mean
    weights <- cos(lat_subset * pi / 180)
    weights_mat <- matrix(weights, nrow = length(lat_subset), ncol = length(lon_nc))
    
    ts_weighted <- apply(data_arctic, 1, function(slice) {
      weighted.mean(slice, weights_mat, na.rm = TRUE)
    })
    
    return(list(
      ts = ts_weighted,
      array = data_arctic,
      lat = lat_subset
    ))
  }
  
  # Load all components
  tke_data <- load_energy_component(TKE_FILE, "tke", lat_arctic, lon)
  mke_data <- load_energy_component(MKE_FILE, "mke", lat_arctic, lon)
  eke_budget_data <- load_energy_component(EKE_BUDGET_FILE, "eke_budget", lat_arctic, lon)
  
  cat(sprintf("\n  Loaded %d months of energy budget data\n", length(tke_data$ts)))
  
  # ==========================================================================
  # 2.2 Budget Closure Verification
  # ==========================================================================
  
  cat("\n--- 2.2 Verifying Energy Budget Closure ---\n\n")
  
  budget_residual <- tke_data$ts - (mke_data$ts + eke_budget_data$ts)
  budget_closure <- list(
    mean_tke = mean(tke_data$ts),
    mean_mke = mean(mke_data$ts),
    mean_eke_budget = mean(eke_budget_data$ts),
    mean_residual = mean(abs(budget_residual)),
    relative_error_pct = mean(abs(budget_residual)) / mean(tke_data$ts) * 100,
    verified = mean(abs(budget_residual)) / mean(tke_data$ts) * 100 < 0.01
  )
  
  cat(sprintf("  Mean TKE: %.3f m²/s²\n", budget_closure$mean_tke))
  cat(sprintf("  Mean MKE: %.3f m²/s²\n", budget_closure$mean_mke))
  cat(sprintf("  Mean EKE_budget: %.3f m²/s²\n", budget_closure$mean_eke_budget))
  cat(sprintf("  Relative closure error: %.6f%%\n", budget_closure$relative_error_pct))
  
  if (budget_closure$verified) {
    cat("  ✓ BUDGET CLOSURE VERIFIED\n")
  } else {
    cat("  ⚠ WARNING: Budget error exceeds tolerance\n")
  }
  
  # ==========================================================================
  # 2.3 Create Energy Budget Data Frame
  # ==========================================================================
  
  cat("\n--- 2.3 Creating Energy Budget Data Frame ---\n")
  
  energy_df <- data.frame(
    date = time_dates,
    year = year(time_dates),
    month = month(time_dates),
    tke = tke_data$ts,
    mke = mke_data$ts,
    eke_budget = eke_budget_data$ts
  ) %>%
    mutate(
      mke_eke_ratio = mke / eke_budget,
      eke_fraction = eke_budget / tke,
      mke_fraction = mke / tke,
      season = case_when(
        month %in% c(12, 1, 2) ~ "DJF",
        month %in% c(3, 4, 5) ~ "MAM",
        month %in% c(6, 7, 8) ~ "JJA",
        month %in% c(9, 10, 11) ~ "SON"
      )
    )
  
  cat(sprintf("  ✓ Energy data frame: %d rows\n", nrow(energy_df)))
  
  # Summary statistics
  energy_summary <- data.frame(
    Component = c("TKE", "MKE", "EKE_budget", "MKE/EKE ratio"),
    Mean = c(mean(energy_df$tke), mean(energy_df$mke), 
             mean(energy_df$eke_budget), mean(energy_df$mke_eke_ratio)),
    SD = c(sd(energy_df$tke), sd(energy_df$mke), 
           sd(energy_df$eke_budget), sd(energy_df$mke_eke_ratio)),
    Min = c(min(energy_df$tke), min(energy_df$mke), 
            min(energy_df$eke_budget), min(energy_df$mke_eke_ratio)),
    Max = c(max(energy_df$tke), max(energy_df$mke), 
            max(energy_df$eke_budget), max(energy_df$mke_eke_ratio))
  )
  
  cat("\n")
  print(kable(energy_summary, digits = 3, format = "simple",
              caption = "Energy Budget Components (Pan-Arctic ≥60°N)"))
  
  cat(sprintf("\n  Mean EKE fraction of TKE: %.1f%%\n", mean(energy_df$eke_fraction) * 100))
  cat(sprintf("  Mean MKE fraction of TKE: %.1f%%\n", mean(energy_df$mke_fraction) * 100))
  
  # ==========================================================================
  # 2.4 Annual Aggregation for Trend Analysis
  # ==========================================================================
  
  cat("\n--- 2.4 Annual Aggregation ---\n")
  
  energy_annual <- energy_df %>%
    group_by(year) %>%
    summarise(
      tke = mean(tke),
      mke = mean(mke),
      eke_budget = mean(eke_budget),
      mke_eke_ratio = mean(mke_eke_ratio),
      eke_fraction = mean(eke_fraction),
      .groups = "drop"
    )
  
  cat(sprintf("  ✓ Annual means: %d years\n", nrow(energy_annual)))
  
  # ==========================================================================
  # 2.5 Trend Analysis of Energy Components
  # ==========================================================================
  
  cat("\n--- 2.5 Energy Component Trends ---\n\n")
  
# Function to perform trend analysis and return results
  analyze_trend <- function(data_vector, varname) {
    # Ensure it's a numeric vector (not ts object)
    x <- as.numeric(data_vector)
    n <- length(x)
    
    # Pre-whitened Mann-Kendall test
    # If this succeeds, it often returns a VECTOR, not a list
    mk_result <- tryCatch({
      mkttest(x)
    }, error = function(e) {
      # Fallback to standard MK if pre-whitened fails
      # This returns a LIST (class "htest")
      mk.test(x)
    })
    
    # Sen's slope
    sens_result <- sens.slope(x)
    
    # Calculate percent change over record
    years <- length(x)
    total_change <- sens_result$estimates * years
    pct_change <- (total_change / mean(x)) * 100
    
    # --- FIX START: Extract Statistics Correctly ---
    if (is.list(mk_result)) {
      # It is a list (likely from mk.test fallback)
      stat_val <- as.numeric(mk_result$statistic)
      p_val <- as.numeric(mk_result$p.value)
    } else {
      # It is a vector (likely from modifiedmk::mkttest)
      # Usually: [1] Z-value, [2] Sen's, [3] S, [4] Var, [5] P-value
      # We attempt to extract by name, or fall back to indices
      
      # Extract Z (Statistic)
      if ("Z-Value" %in% names(mk_result)) {
        stat_val <- as.numeric(mk_result["Z-Value"])
      } else if ("Corrected Z" %in% names(mk_result)) {
        stat_val <- as.numeric(mk_result["Corrected Z"])
      } else {
        stat_val <- as.numeric(mk_result[1]) # Assume 1st element is Z
      }
      
      # Extract P-value
      if ("P-value" %in% names(mk_result)) {
        p_val <- as.numeric(mk_result["P-value"])
      } else {
        # Often the last element or specific index 5 in modifiedmk
        p_val <- as.numeric(mk_result[length(mk_result)]) 
      }
    }
    
    # Return Results List
    list(
      variable = varname,
      n_years = n,
      sens_slope = as.numeric(sens_result$estimates),
      sens_slope_pct_per_decade = (as.numeric(sens_result$estimates) / mean(x)) * 100 * 10,
      total_change = total_change,
      pct_change = pct_change,
      mk_statistic = stat_val,  
      p_value = p_val,           
      significant = p_val < 0.05
    )
  }
  
  # Analyze each component
  trend_vars <- c("tke", "mke", "eke_budget", "mke_eke_ratio")
  energy_trends <- list()
  
  for (varname in trend_vars) {
    energy_trends[[varname]] <- analyze_trend(energy_annual[[varname]], varname)
    
    result <- energy_trends[[varname]]
    cat(sprintf("%s:\n", toupper(varname)))
    cat(sprintf("  Sen's slope: %.5f per year (%.3f%% per decade)\n", 
                result$sens_slope, result$sens_slope_pct_per_decade))
    cat(sprintf("  Total change (1958-2025): %.3f (%.1f%%)\n", 
                result$total_change, result$pct_change))
    cat(sprintf("  p-value: %.4f\n", result$p_value))
    cat(sprintf("  Trend: %s\n\n", 
                ifelse(result$significant, "SIGNIFICANT", "not significant")))
  }
  
  # Convert to data frame for easy table generation
  energy_trends_df <- bind_rows(lapply(energy_trends, as.data.frame))
  
  # ==========================================================================
  # 2.6 Save Phase 2 Results
  # ==========================================================================
  
  cat("\n--- 2.6 Saving Phase 2 Results ---\n")
  
  # Compile all results
  phase2_results <- list(
    # Data frames
    energy_df = energy_df,
    energy_annual = energy_annual,
    energy_summary = energy_summary,
    
    # Trend analysis
    energy_trends = energy_trends,
    energy_trends_df = energy_trends_df,
    
    # Budget verification
    budget_closure = budget_closure,
    
    # Metadata
    metadata = list(
      created = Sys.time(),
      n_months = nrow(energy_df),
      n_years = nrow(energy_annual),
      period = c(min(energy_df$year), max(energy_df$year))
    )
  )
  
  save(phase2_results, energy_df, energy_annual, energy_trends_df,
       file = PHASE2_OUTPUT)
  
  cat(sprintf("  ✓ Saved: %s\n", PHASE2_OUTPUT))
  
  cat("\n", rep("=", 70), "\n")
  cat("PHASE 2 COMPLETE\n")
  cat(rep("=", 70), "\n")

}  # End of if-else for checking existing results

# ============================================================================
# SUMMARY TABLE (always runs, using loaded or computed results)
# ============================================================================

cat("\n--- Phase 2 Summary ---\n\n")

cat("Energy Budget Trends (Pan-Arctic ≥60°N, 1958-2025):\n\n")
print(kable(energy_trends_df %>%
              select(variable, sens_slope, sens_slope_pct_per_decade, 
                     p_value, significant) %>%
              mutate(
                sens_slope = sprintf("%.5f", sens_slope),
                sens_slope_pct_per_decade = sprintf("%.3f%%", sens_slope_pct_per_decade),
                p_value = sprintf("%.4f", p_value),
                significant = ifelse(significant, "Yes", "No")
              ),
            col.names = c("Variable", "Sen's Slope [/yr]", "Change [%/decade]", 
                          "p-value", "Significant"),
            format = "simple"))
```

```{r}
# ============================================================================
# PHASE 2: DIAGNOSTIC PLOTS
# Energy Budget Diagnostics
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 2: DIAGNOSTIC PLOTS\n")
cat(rep("=", 70), "\n\n")

# Load Phase 2 results if not already in environment
if (!exists("energy_df") || !exists("energy_annual")) {
  load(PHASE2_OUTPUT)
  cat("✓ Loaded Phase 2 results\n\n")
}

# Define color scheme for energy components
energy_colors <- c(
  "TKE" = "#1B9E77",
  "MKE" = "#D95F02", 
  "EKE" = "#7570B3"
)

# ============================================================================
# PLOT 2.1: TIME SERIES OF ENERGY COMPONENTS (MONTHLY)
# ============================================================================

cat("--- Generating Plot 2.1: Energy Components Time Series ---\n")

# Prepare long-format data
energy_long <- energy_df %>%
  select(date, tke, mke, eke_budget) %>%
  pivot_longer(-date, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component, 
                       levels = c("tke", "mke", "eke_budget"),
                       labels = c("TKE", "MKE", "EKE (budget)"))
  )

p2_1 <- ggplot(energy_long, aes(x = date, y = value, color = component)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, linewidth = 0.8) +
  scale_color_manual(values = c("TKE" = "#1B9E77", "MKE" = "#D95F02", 
                                "EKE (budget)" = "#7570B3"),
                     name = "Component") +
  scale_x_date(date_breaks = "10 years", date_labels = "%Y") +
  labs(
    title = "Atmospheric Kinetic Energy Budget Components",
    subtitle = "Pan-Arctic (≥60°N) | Monthly values with LOESS smoother | 1958–2025",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(FIG_DIR, "Fig2.1_Energy_Components_TimeSeries.png"),
       p2_1, width = 12, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.1_Energy_Components_TimeSeries.png\n")

# ============================================================================
# PLOT 2.2: ANNUAL ENERGY COMPONENTS WITH TRENDS
# ============================================================================

cat("\n--- Generating Plot 2.2: Annual Energy with Trends ---\n")

# Prepare annual long-format data
energy_annual_long <- energy_annual %>%
  select(year, tke, mke, eke_budget) %>%
  pivot_longer(-year, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component, 
                       levels = c("tke", "mke", "eke_budget"),
                       labels = c("TKE", "MKE", "EKE (budget)"))
  )

p2_2 <- ggplot(energy_annual_long, aes(x = year, y = value, color = component)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.2) +
  scale_color_manual(values = c("TKE" = "#1B9E77", "MKE" = "#D95F02", 
                                "EKE (budget)" = "#7570B3"),
                     name = "Component") +
  labs(
    title = "Annual Mean Kinetic Energy Components with Linear Trends",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 | Shading = 95% CI",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(FIG_DIR, "Fig2.2_Energy_Annual_Trends.png"),
       p2_2, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.2_Energy_Annual_Trends.png\n")

# ============================================================================
# PLOT 2.3: ENERGY BUDGET COMPOSITION (STACKED AREA)
# ============================================================================

cat("\n--- Generating Plot 2.3: Energy Budget Composition ---\n")

# Annual data for cleaner visualization
energy_stack <- energy_annual %>%
  select(year, mke, eke_budget) %>%
  pivot_longer(-year, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component,
                       levels = c("eke_budget", "mke"),
                       labels = c("EKE (Transient Eddies)", "MKE (Mean Flow)"))
  )

p2_3 <- ggplot(energy_stack, aes(x = year, y = value, fill = component)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = c("EKE (Transient Eddies)" = "#7570B3", 
                               "MKE (Mean Flow)" = "#D95F02"),
                    name = NULL) +
  labs(
    title = "Partitioning of Total Kinetic Energy",
    subtitle = "Pan-Arctic (≥60°N) | Annual means | TKE = MKE + EKE",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(FIG_DIR, "Fig2.3_Energy_Budget_Composition.png"),
       p2_3, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.3_Energy_Budget_Composition.png\n")

# ============================================================================
# PLOT 2.4: MKE/EKE RATIO TIME SERIES
# ============================================================================

cat("\n--- Generating Plot 2.4: MKE/EKE Ratio ---\n")

# Get trend info for annotation
ratio_trend <- energy_trends_df %>% filter(variable == "mke_eke_ratio")

p2_4 <- ggplot(energy_annual, aes(x = year, y = mke_eke_ratio)) +
  geom_line(color = "gray50", alpha = 0.5) +
  geom_point(color = "#E7298A", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "#E7298A", 
              fill = "#E7298A", alpha = 0.2, linewidth = 1) +
  geom_hline(yintercept = mean(energy_annual$mke_eke_ratio), 
             linetype = "dashed", color = "gray30") +
  annotate("text", x = 1965, y = max(energy_annual$mke_eke_ratio) * 0.98,
           label = sprintf("Sen's slope: %.4f/yr\np = %.4f (%s)",
                           ratio_trend$sens_slope,
                           ratio_trend$p_value,
                           ifelse(ratio_trend$significant, "significant", "n.s.")),
           hjust = 0, size = 3.5, fontface = "italic") +
  labs(
    title = "Mean-to-Eddy Kinetic Energy Ratio",
    subtitle = "Pan-Arctic (≥60°N) | Higher values = stronger mean flow relative to eddies",
    x = NULL,
    y = "MKE / EKE Ratio"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(file.path(FIG_DIR, "Fig2.4_MKE_EKE_Ratio.png"),
       p2_4, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.4_MKE_EKE_Ratio.png\n")

# ============================================================================
# PLOT 2.5: EKE FRACTION OF TKE
# ============================================================================

cat("\n--- Generating Plot 2.5: EKE Fraction of TKE ---\n")

p2_5 <- ggplot(energy_annual, aes(x = year, y = eke_fraction * 100)) +
  geom_line(color = "gray50", alpha = 0.5) +
  geom_point(color = "#7570B3", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "#7570B3",
              fill = "#7570B3", alpha = 0.2, linewidth = 1) +
  geom_hline(yintercept = mean(energy_annual$eke_fraction) * 100,
             linetype = "dashed", color = "gray30") +
  labs(
    title = "Eddy Kinetic Energy as Fraction of Total",
    subtitle = "Pan-Arctic (≥60°N) | Higher values = more energy in transient eddies",
    x = NULL,
    y = "EKE / TKE (%)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(file.path(FIG_DIR, "Fig2.5_EKE_Fraction.png"),
       p2_5, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.5_EKE_Fraction.png\n")

# ============================================================================
# PLOT 2.6: SEASONAL CYCLE OF ENERGY COMPONENTS
# ============================================================================

cat("\n--- Generating Plot 2.6: Seasonal Cycle ---\n")

energy_seasonal <- energy_df %>%
  group_by(month) %>%
  summarise(
    tke_mean = mean(tke),
    tke_sd = sd(tke),
    mke_mean = mean(mke),
    mke_sd = sd(mke),
    eke_mean = mean(eke_budget),
    eke_sd = sd(eke_budget),
    .groups = "drop"
  )

# Long format for plotting
seasonal_long <- energy_seasonal %>%
  select(month, tke_mean, mke_mean, eke_mean) %>%
  pivot_longer(-month, names_to = "component", values_to = "value") %>%
  mutate(
    component = factor(component,
                       levels = c("tke_mean", "mke_mean", "eke_mean"),
                       labels = c("TKE", "MKE", "EKE"))
  )

p2_6 <- ggplot(seasonal_long, aes(x = month, y = value, color = component)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(values = c("TKE" = "#1B9E77", "MKE" = "#D95F02", 
                                "EKE" = "#7570B3"),
                     name = "Component") +
  labs(
    title = "Climatological Seasonal Cycle of Energy Components",
    subtitle = "Pan-Arctic (≥60°N) | 1958–2025 mean",
    x = NULL,
    y = expression("Kinetic Energy [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

ggsave(file.path(FIG_DIR, "Fig2.6_Energy_Seasonal_Cycle.png"),
       p2_6, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.6_Energy_Seasonal_Cycle.png\n")

# ============================================================================
# PLOT 2.7: SEASONAL MKE/EKE RATIO
# ============================================================================

cat("\n--- Generating Plot 2.7: Seasonal MKE/EKE Ratio ---\n")

ratio_seasonal <- energy_df %>%
  group_by(month) %>%
  summarise(
    ratio_mean = mean(mke_eke_ratio),
    ratio_sd = sd(mke_eke_ratio),
    .groups = "drop"
  )

p2_7 <- ggplot(ratio_seasonal, aes(x = month, y = ratio_mean)) +
  geom_ribbon(aes(ymin = ratio_mean - ratio_sd, ymax = ratio_mean + ratio_sd),
              fill = "#E7298A", alpha = 0.3) +
  geom_line(color = "#E7298A", linewidth = 1.2) +
  geom_point(color = "#E7298A", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Seasonal Cycle of MKE/EKE Ratio",
    subtitle = "Pan-Arctic (≥60°N) | Mean ± 1 SD | Higher in summer = weaker relative eddy activity",
    x = NULL,
    y = "MKE / EKE Ratio"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(file.path(FIG_DIR, "Fig2.7_MKE_EKE_Ratio_Seasonal.png"),
       p2_7, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.7_MKE_EKE_Ratio_Seasonal.png\n")

# ============================================================================
# PLOT 2.8: COMPARISON - EKE_BUDGET vs SYNOPTIC-SCALE EKE
# ============================================================================

cat("\n--- Generating Plot 2.8: EKE Budget vs Synoptic EKE ---\n")

# Merge with storminess metrics from Phase 1
comparison_df <- energy_df %>%
  select(date, year, month, eke_budget) %>%
  left_join(
    storm_df %>% select(date, eke_mean),
    by = "date"
  ) %>%
  rename(
    eke_budget = eke_budget,
    eke_synoptic = eke_mean
  )

# Correlation
eke_cor <- cor(comparison_df$eke_budget, comparison_df$eke_synoptic, use = "complete.obs")

# Scatter plot
p2_8a <- ggplot(comparison_df, aes(x = eke_budget, y = eke_synoptic)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "#E41A1C", linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  annotate("text", 
           x = min(comparison_df$eke_budget, na.rm = TRUE) * 1.1,
           y = max(comparison_df$eke_synoptic, na.rm = TRUE) * 0.95,
           label = sprintf("r = %.3f", eke_cor),
           hjust = 0, size = 4, fontface = "bold") +
  labs(
    title = "(a) EKE Budget vs Synoptic-Scale EKE",
    x = expression("EKE"["budget"]*" (all sub-monthly variance) [m"^2*"/s"^2*"]"),
    y = expression("EKE"["synoptic"]*" (1.2–6 day bandpass) [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12))

# Time series comparison (annual)
comparison_annual <- comparison_df %>%
  group_by(year) %>%
  summarise(
    eke_budget = mean(eke_budget),
    eke_synoptic = mean(eke_synoptic),
    .groups = "drop"
  ) %>%
  pivot_longer(-year, names_to = "type", values_to = "value") %>%
  mutate(
    type = factor(type,
                  levels = c("eke_budget", "eke_synoptic"),
                  labels = c("EKE (budget, all sub-monthly)", 
                             "EKE (synoptic, 1.2–6 day)"))
  )

p2_8b <- ggplot(comparison_annual, aes(x = year, y = value, color = type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_manual(values = c("#7570B3", "#1B9E77"), name = NULL) +
  labs(
    title = "(b) Annual Mean Comparison",
    x = NULL,
    y = expression("EKE [m"^2*"/s"^2*"]")
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "bottom"
  )

p2_8_combined <- p2_8a + p2_8b +
  plot_layout(ncol = 2, widths = c(1, 1.2)) +
  plot_annotation(
    title = "Comparison of EKE Definitions",
    subtitle = "Budget EKE captures all variance; Synoptic EKE isolates 1.2–6 day storm band",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )
  )

ggsave(file.path(FIG_DIR, "Fig2.8_EKE_Comparison.png"),
       p2_8_combined, width = 14, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.8_EKE_Comparison.png\n")

# ============================================================================
# PLOT 2.9: DECADAL CHANGES IN ENERGY PARTITIONING
# ============================================================================

cat("\n--- Generating Plot 2.9: Decadal Energy Partitioning ---\n")

energy_decadal <- energy_df %>%
  mutate(
    decade = case_when(
      year >= 1958 & year < 1970 ~ "1958-1969",
      year >= 1970 & year < 1980 ~ "1970-1979",
      year >= 1980 & year < 1990 ~ "1980-1989",
      year >= 1990 & year < 2000 ~ "1990-1999",
      year >= 2000 & year < 2010 ~ "2000-2009",
      year >= 2010 & year < 2020 ~ "2010-2019",
      year >= 2020 ~ "2020-2025"
    ),
    decade = factor(decade)
  ) %>%
  group_by(decade) %>%
  summarise(
    mke_fraction = mean(mke_fraction) * 100,
    eke_fraction = mean(eke_fraction) * 100,
    mke_eke_ratio = mean(mke_eke_ratio),
    .groups = "drop"
  )

p2_9 <- ggplot(energy_decadal, aes(x = decade)) +
  geom_col(aes(y = mke_fraction, fill = "MKE"), width = 0.7) +
  geom_col(aes(y = -eke_fraction, fill = "EKE"), width = 0.7) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_fill_manual(values = c("MKE" = "#D95F02", "EKE" = "#7570B3"),
                    name = "Component") +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%"),
    breaks = seq(-60, 60, by = 20)
  ) +
  labs(
    title = "Decadal Changes in Energy Partitioning",
    subtitle = "Pan-Arctic (≥60°N) | Fraction of TKE in each component",
    x = NULL,
    y = "Fraction of TKE"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(file.path(FIG_DIR, "Fig2.9_Decadal_Partitioning.png"),
       p2_9, width = 10, height = 6, dpi = 300)
cat("  ✓ Saved: Fig2.9_Decadal_Partitioning.png\n")

# ============================================================================
# SUMMARY
# ============================================================================

cat("\n", rep("=", 70), "\n")
cat("PHASE 2 DIAGNOSTIC PLOTS COMPLETE\n")
cat(rep("=", 70), "\n\n")

cat("Generated figures:\n")
cat("  2.1  Energy Components Time Series (monthly)\n")
cat("  2.2  Annual Energy with Linear Trends\n")
cat("  2.3  Energy Budget Composition (stacked)\n")
cat("  2.4  MKE/EKE Ratio Time Series\n")
cat("  2.5  EKE Fraction of TKE\n")
cat("  2.6  Seasonal Cycle of Energy Components\n")
cat("  2.7  Seasonal MKE/EKE Ratio\n")
cat("  2.8  EKE Budget vs Synoptic EKE Comparison\n")
cat("  2.9  Decadal Changes in Energy Partitioning\n")

cat(sprintf("\nAll figures saved to: %s\n", FIG_DIR))

cat("\nKey findings:\n")
cat(sprintf("  - EKE fraction of TKE: %.1f%% (mean)\n", 
            mean(energy_df$eke_fraction) * 100))
cat(sprintf("  - MKE fraction of TKE: %.1f%% (mean)\n", 
            mean(energy_df$mke_fraction) * 100))
cat(sprintf("  - Correlation EKE_budget vs EKE_synoptic: r = %.3f\n", eke_cor))
cat(sprintf("  - MKE/EKE ratio trend: %s (p = %.4f)\n",
            ifelse(ratio_trend$significant, "SIGNIFICANT", "not significant"),
            ratio_trend$p_value))

cat("\n", rep("=", 70), "\n\n")
```






# Phase 3

```{r}
# Load prepared data from Phase 1
data_file <- here::here("Data", "processed", "analysis_ready", "Phase1_Prepared_Data.RData")
# This loads:
#   - storm_df          : Master data frame with all metrics
#   - ts_objects        : List of ts objects for each metric
#   - metrics_arctic    : List of 3D arrays [time, lat, lon] for spatial analysis
#   - acf_results       : ACF results for autocorrelation reference
#   - lat_arctic, lon   : Coordinate vectors
#   - time_dates        : Date vector
#   - summary_stats     : Descriptive statistics table

if (!file.exists(data_file)) {
  stop(glue("
    ERROR: Phase 1 data not found!
    Expected: {data_file}
    
    Please run Phase 1 script first.
  "))
}

load(data_file)
cat("✓ Loaded Phase 1 prepared data\n\n")

# Define output directories
results_dir <- here::here("Data", "processed", "trend_results")
fig_dir <- here::here("Figures", "Phase2_Trends")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

# ============================================================================
# CONFIGURATION
# ============================================================================

# Metrics to analyze (in order of priority)
METRICS_TO_ANALYZE <- c(
  "eke_mean",         # PRIMARY
  "eke_p95",          # Extreme intensity
  "eke_p99",          # Extreme intensity (more extreme)
  "exceed_p95_freq"   # Extreme frequency
)

# Bootstrap parameters
N_BOOTSTRAP <- 5000
BLOCK_LENGTH_MONTHLY <- 12  # For monthly data
BLOCK_LENGTH_ANNUAL <- 3    # For annual data

# Significance level
ALPHA <- 0.05
```

